### 2017-03-27 submission/manuscript.Rmd
first manuscript commit

---
title: "**Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
csl: plos.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---
\vspace{35mm}

Running title: Unique taxa differentiate proximal and distal human colon microbiota

\vspace{35mm}

Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3^, and Patrick D. Schloss^1$\dagger$^

\vspace{35mm}

$\dagger$ Corresponding author: pschloss@umich.edu

1\. Department of Microbiology and Immunology, University of Michigan Medical School, Ann Arbor, Michigan 48109

2\. Pennslyvania State University, Hershey, Pennslyvania ??

3\. Department of Internal Medicine, Division of Gastroenterology, University of Michigan Medical School, Ann Arbor, Michigan

\newpage
\linenumbers

### Abstract

### 2017-03-28 submission/manuscript.Rmd
manuscript updates

@@
 ### Abstract
+
+Colorectal cancer (CRC) remains a leading cause of death worldwide.
+Tumors of the proximal (right) and distal (left) colon are morpologically and genetically distinct.
+Previous work from our group found that microbial dysbiosis is associated with the development of colorectal cancer tumors in studies of both mice and humans.
+Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease.
+In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon.
+We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure.
+This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
+16S rRNA gene sequencing was performed on proximal and distal mucosal biopsies, luminal and exit stool for 20 healthy individuals.
+Diversity analysis of each location revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
+Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites.
+Random Forest classification models were built using taxa abundance and sample location revealed distinct populations that were found in each location.
+Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas and Actinomyces were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925).
+The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755).
+The left mucosa was found to have high populations of Finegoldia, Murdochiella and Porphyromonas.
+Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end.
+Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon.
+Taken together, our results have identified distinct bacterial populations distinct of the proximal and distal colon.
+Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.
+
+### Introduction
+
+- colon cancer 
+- right and left tumors are morphologically and genetically distinct
+- the microbiome is of healthy people and patients with CRC is distinguishable at the OTU level
+- specific keystone species are associated with CRC tumors
+- do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
+- nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
+- bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
+- first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? 
+- how does the microbiome vary spatially along the colon?
+
+### Results
+
+### Discussion
+
+### Acknowledgments
+
+### Methods
+
+### Figure Legends

### 2017-03-29 submission/manuscript.Rmd
manuscript updates

@@
 - right and left tumors are morphologically and genetically distinct
-- the microbiome is of healthy people and patients with CRC is distinguishable at the OTU level
+- the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
 - specific keystone species are associated with CRC tumors
@@
 - first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? 
-- how does the microbiome vary spatially along the colon?
+- mucosal vs lumenal populations studies have also been complicated by bowel prep 
+- how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations 
+- hypothesis: there are distinct bacteria on the left and right side that can be distinguished by 16S sequencing of stool
 

### 2017-03-29 submission/manuscript.Rmd
manuscript outline

@@
 - colon cancer 
-- right and left tumors are morphologically and genetically distinct
+- right and left tumors are morphologically and genetically distinct (why Q is important)
 - the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
 - specific keystone species are associated with CRC tumors
-- do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
+- do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? (central Q)
 - nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
 - bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
-- first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? 
-- mucosal vs lumenal populations studies have also been complicated by bowel prep 
+- first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? (data needed to answer Q)
+- mucosal vs lumenal populations studies have also been complicated by bowel prep (what methods are used to get data? what analysis must be applied?- ML)
 - how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations 
@@
 
+- what data were obtained? what were the results of the analyses?
+
 ### Discussion
+
+-how did the analyses answer the central question?
+
+-what does this answer tell us about the broader field?
 

### 2017-04-17 submission/manuscript.Rmd
updated manuscript outline

--- 2017-03-28 submission/manuscript.Rmd
@@
 ---
-title: "**Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota**"
-bibliography: references.bib
-output:
-  pdf_document:
-    includes:
-      in_header: header.tex
-    keep_tex: yes
-csl: plos.csl #Get themes at https://github.com/citation-style-language/styles
+title: '**Machine learning identifies unique taxa differentiating proximal and distal
+  human colonic microbiota**'
 fontsize: 11pt
 geometry: margin=1.0in
+output:
+  html_document: default
+  pdf_document:
+    includes: null
+    keep_tex: yes
+csl: plos.csl
 ---
@@
 - colon cancer 
-- right and left tumors are morphologically and genetically distinct
-- the microbiome is of healthy people and patients with CRC is distinguishable at the OTU level
+- right and left tumors are morphologically and genetically distinct (why Q is important)
+- the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
 - specific keystone species are associated with CRC tumors
-- do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
+- Central Question: do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
 - nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
 - bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
-- first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? 
-- how does the microbiome vary spatially along the colon?
+- how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations
+
+- What data is needed to answer this question?
+first, need to characterize healthy species that are specific to each location.
+then we can ask, if/how disruption affects tumorigenesis?
+so what is needed is paired samples from healthy people that have not undergone bowel prep, from the right/left and lumen/mucosa
+
+- what methods are used to get the data? unprepped colonoscopy
+
+- What analysis must be applied? - diversity analysis and machine learning to pull out small variations/differences among the noisy data set.
+
+- hypothesis: there are distinct bacteria on the left and right side that can be distinguished by 16S sequencing of stool
 
@@
 
+- what data were obtained? 16S sequencing results for samples taken from healthy human colons that had not undergone bowel prep
+Figure 1: sample collection schematic
+
+- what were the results of the analyses? the results of the analyses are as follows:
+
+#### Diversity analyses
+- there are similar levels (within error margins) of each phyla for each location, but there is noise due to interpatient variability.
+Overall, there is also a similar level of diversity in each location.
+However, each patient has a different pattern of diversity over the locations, some are high in R, some are high in L, etc.
+Figure 2:
+* 2A: relative abundance by phyla 
+* 2B: simpson diversity 
+* 2C: NMDS/ordination of some kind
+
+#### similarity analyses
+- we can compare sides, samples or patients by looking at distances between samples.
+here we used the thetayc calculator to make comparisons.
+The right mucosa and lumen are most similar to each other.
+As you move closer to the exit, the samples diverge.
+With the exception of the left lumen- this sample best resembles the stool collected from exit as it is...
+exiting.
+
+Figure 3:
+* 3A: the right mucosa and lumen are most similar to each other 
+* 3B: the left lumen is most similar than exit stool
+
+#### Random Forest classification models 
+- we can use machine learning models to build classification models that identify specific features in noisy data sets that allow for separation.
+- we used a Random Forest model to separate samples from each side or site
+
+Figure 4:
+* 4A: Classifying lumen vs mucosal samples on each side (L vs L, R vs R)
+* 4B: Classifying side of lumen or side of mucosa (L vs R lumen, L vs R mucosa)
+
+#### Important OTUs 
+- the Random Forest model lets us pick out features that dictate the classification model separation values
+
+Figure 5:
+* 5A: There are differentially abundant OTUs in lumen vs mucosa
+* 5B: There are differentially abundant OTUs in right vs left lumen, right vs left mucosa
+
+- There is some commonality between the OTUs - drivers of left sidedness are consistent in both lumen and mucosa (more of this in discussion)
+
 ### Discussion
+
+Recap of results, interesting discussion points about diversity and previous studies of diversity along the gut axis of healthy people
+
+Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)
+
+-how did the analyses answer the central question?
+The analyses identified OTUs that are relatively unique to each location, among a wide variation differences between study participants
+
+-what does this answer tell us about the broader field?
+This answer told us that there are specific differences of bacteria in each side and that we can use this information to form hypotheses about how specific microbial/ecological variations from the host can affect oncogenesis.
+It tells us too that sample collection methods are important, and what we have found in unprepped bowels do/do not match published findings.
+
+-what's next?
+cancer associated OTUs on one side or the other, compare to previous studies.
+But really to understand what's going on in the various regions of the gut we need more information than just 16S
 
@@
 ### Methods
+Study participants/IRB
+
+Sample collection
+
+Sample processing/DNA extraction
+
+16S rRNA sequencing
+
+Statistical analysis/Random Forest
 
 ### Figure Legends
+
+Figure 1
+
+Figure 2
+
+Figure 3
+
+Figure 4
+
+Figure 5
+
+### References

### 2017-04-17 submission/manuscript.tex
updated manuscript outline

--- 2017-03-29 submission/manuscript.Rmd
@@
----
-title: "**Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota**"
-bibliography: references.bib
-output:
-  pdf_document:
-    includes:
-      in_header: header.tex
-    keep_tex: yes
-csl: plos.csl #Get themes at https://github.com/citation-style-language/styles
-fontsize: 11pt
-geometry: margin=1.0in
----
-\vspace{35mm}
+\documentclass[11pt,]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage[margin=1.0in]{geometry}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
 
-Running title: Unique taxa differentiate proximal and distal human colon microbiota
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
+
+%%% Change title format to be more compact
+\usepackage{titling}
+
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
+
+\setlength{\droptitle}{-2em}
+  \title{\textbf{Machine learning identifies unique taxa differentiating proximal
+and distal human colonic microbiota}}
+  \pretitle{\vspace{\droptitle}\centering\huge}
+  \posttitle{\par}
+  \author{}
+  \preauthor{}\postauthor{}
+  \date{}
+  \predate{}\postdate{}
+
+\begin{document}
+
+\maketitle
 
@@
 
-Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3^, and Patrick D. Schloss^1$\dagger$^
+Running title: Unique taxa differentiate proximal and distal human colon
+microbiota
 
@@
 
-$\dagger$ Corresponding author: pschloss@umich.edu
+Kaitlin J. Flynn\textsuperscript{1}, Charles C.
+Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2},
+Danielle Kimberly Turgeon\textsuperscript{3}, and Patrick D.
+Schloss\textsuperscript{1\(\dagger\)}
 
-1\. Department of Microbiology and Immunology, University of Michigan Medical School, Ann Arbor, Michigan 48109
+\vspace{35mm}
 
-2\. Pennslyvania State University, Hershey, Pennslyvania ??
+\(\dagger\) Corresponding author:
+\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
 
-3\. Department of Internal Medicine, Division of Gastroenterology, University of Michigan Medical School, Ann Arbor, Michigan
+1. Department of Microbiology and Immunology, University of Michigan
+Medical School, Ann Arbor, Michigan 48109
+
+2. Pennslyvania State University, Hershey, Pennslyvania ??
+
+3. Department of Internal Medicine, Division of Gastroenterology,
+University of Michigan Medical School, Ann Arbor, Michigan
 
@@
 
-### Abstract
+\subsubsection{Abstract}\label{abstract}
 
 Colorectal cancer (CRC) remains a leading cause of death worldwide.
-Tumors of the proximal (right) and distal (left) colon are morpologically and genetically distinct.
-Previous work from our group found that microbial dysbiosis is associated with the development of colorectal cancer tumors in studies of both mice and humans.
-Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease.
-In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon.
-We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure.
-This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
-16S rRNA gene sequencing was performed on proximal and distal mucosal biopsies, luminal and exit stool for 20 healthy individuals.
-Diversity analysis of each location revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
-Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites.
-Random Forest classification models were built using taxa abundance and sample location revealed distinct populations that were found in each location.
-Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas and Actinomyces were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925).
-The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755).
-The left mucosa was found to have high populations of Finegoldia, Murdochiella and Porphyromonas.
-Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end.
-Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon.
-Taken together, our results have identified distinct bacterial populations distinct of the proximal and distal colon.
-Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.
+Tumors of the proximal (right) and distal (left) colon are
+morpologically and genetically distinct. Previous work from our group
+found that microbial dysbiosis is associated with the development of
+colorectal cancer tumors in studies of both mice and humans. Analysis of
+the fecal microbiota from healthy and CRC patients further revealed
+different microbial signatures associated with disease. In this study,
+we extended our observations of the fecal microbiome to analysis of the
+proximal and distal human colon. We used a two-colonoscope approach on
+subjects that had not undergone standard bowel preparation procedure.
+This technique allowed us to characterize the native proximal and distal
+luminal and mucosal microbiome without prior chemical disruption. 16S
+rRNA gene sequencing was performed on proximal and distal mucosal
+biopsies, luminal and exit stool for 20 healthy individuals. Diversity
+analysis of each location revealed that each site contained a diverse
+community, and that a patient's samples were more similar to each other
+than to that of other individuals. Since we could not differentiate
+sites along the colon based on community structure or community
+membership alone, we employed the machine-learning algorithm Random
+Forest to identify key species that distinguish biogeographical sites.
+Random Forest classification models were built using taxa abundance and
+sample location revealed distinct populations that were found in each
+location. Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas
+and Actinomyces were most likely to be found in mucosal samples versus
+luminal samples (AUC = 0.925). The classification model performed well
+(AUC = 0.912) when classifying mucosal samples into proximal or distal
+sides, but separating luminal samples from each side proved more
+challenging (AUC = 0.755). The left mucosa was found to have high
+populations of Finegoldia, Murdochiella and Porphyromonas. Proximal and
+distal luminal samples were comprised of many of the same taxa, likely
+reflecting the fact that stool moves along the colon from the proximal
+to distal end. Finally, comparison of all samples to fecal samples taken
+at exit uncovered that the feces were most similar to samples taken from
+the left lumen, again reflecting the anatomical structure of the colon.
+Taken together, our results have identified distinct bacterial
+populations distinct of the proximal and distal colon. Further
+investigation of these bacteria may elucidate if and how these groups
+contribute to differential oncogenesis processes on the respective sides
+of the colon.
 
-### Introduction
+\subsubsection{Introduction}\label{introduction}
 
-- colon cancer 
-- right and left tumors are morphologically and genetically distinct
-- the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
-- specific keystone species are associated with CRC tumors
-- do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
-- nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
-- bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
-- first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? 
-- mucosal vs lumenal populations studies have also been complicated by bowel prep 
-- how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations 
-- hypothesis: there are distinct bacteria on the left and right side that can be distinguished by 16S sequencing of stool
+\begin{itemize}
+\item
+  colon cancer
+\item
+  right and left tumors are morphologically and genetically distinct
+  (why Q is important)
+\item
+  the microbiome of healthy people and patients with CRC is
+  distinguishable at the OTU level
+\item
+  specific keystone species are associated with CRC tumors
+\item
+  Central Question: do the populations of bacteria in the left and right
+  colon contribute to the development of specific subtypes of CRC?
+\item
+  nearly all studies of the gut microbiome have been from shed stool or
+  colonoscopy with bowel prep
+\item
+  bowel prep changes the mucosal/fecal microbiome and what can be
+  analyzed from it.
+\item
+  how does the microbiome vary spatially along the colon? we know that
+  the environment varies dramatically but we dont know how those
+  environmental factors specifically affect bacterial populations
+\item
+  What data is needed to answer this question? first, need to
+  characterize healthy species that are specific to each location. then
+  we can ask, if/how disruption affects tumorigenesis? so what is needed
+  is paired samples from healthy people that have not undergone bowel
+  prep, from the right/left and lumen/mucosa
+\item
+  what methods are used to get the data? unprepped colonoscopy
+\item
+  What analysis must be applied? - diversity analysis and machine
+  learning to pull out small variations/differences among the noisy data
+  set.
+\item
+  hypothesis: there are distinct bacteria on the left and right side
+  that can be distinguished by 16S sequencing of stool
+\end{itemize}
 
-### Results
+\subsubsection{Results}\label{results}
 
-### Discussion
+\begin{itemize}
+\item
+  what data were obtained? 16S sequencing results for samples taken from
+  healthy human colons that had not undergone bowel prep Figure 1:
+  sample collection schematic
+\item
+  what were the results of the analyses? the results of the analyses are
+  as follows:
+\end{itemize}
 
-### Acknowledgments
+\paragraph{Descriptive analysis}\label{descriptive-analysis}
 
-### Methods
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  there are similar levels (within error margins) of each phyla for each
+  location, but there is noise due to interpatient variability. Overall,
+  there is also a similar level of diversity in each location. However,
+  each patient has a different pattern of diversity over the locations,
+  some are high in R, some are high in L, etc. Figure 2:
+\item
+  2A: relative abundance by phyla
+\item
+  2B: simpson diversity
+\item
+  2C: NMDS/ordination of some kind
+\end{itemize}
 
-### Figure Legends
+\subsubsection{Discussion}\label{discussion}
+
+-how did the analyses answer the central question?
+
+-what does this answer tell us about the broader field?
+
+\subsubsection{Acknowledgments}\label{acknowledgments}
+
+\subsubsection{Methods}\label{methods}
+
+Study participants/IRB
+
+Sample collection
+
+Sample processing/DNA extraction
+
+16S rRNA sequencing
+
+Statistical analysis/Random Forest
+
+\subsubsection{Figure Legends}\label{figure-legends}
+
+\end{document}

### 2017-04-18 code/otu_barplots.R
manuscript and plot updates

@@ -31,16 +31,34 @@ rel_abund <- 100*test/unique(apply(test, 1, sum)) #unique is bad
 #Create vector of OTUs with median abundances >1%
 OTUs_1 <- apply(rel_abund, 2, max) > 1
 OTU_list <- colnames(rel_abund)[OTUs_1]
+
 #get df of just top OTUs
 rel_abund_top <- rel_abund[, OTUs_1]
+
 rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
 
 source('code/Sum_OTU_by_Tax.R')
 source('code/sum_shared.R')
 
+#this code makes the df for family level RA sorted by % abundance thanks to nick 
+#add column of sites to rel_abund_top
+locat <- meta_file$location
+rel_abund_top <- cbind(site = locat, rel_abund_top)
+
+#subset file to just include location 
+
+locat <- subset
+
+#for above, just subset df from rel_meta instead to get the right location/OTU table 
+rel_fam <- sum_OTU_by_tax_level(2, rel_abund_top, tax_file)
+
 #use this code to assign phyla to each OTU in the shared file 
 shared_phyla <- get_tax_level_shared(subsample, tax, 2)
 phyla_met <- merge(meta_file, shared_phyla, by.x='group', by.y='row.names')
+
+shared_fam <- get_tax_level_shared(subsample, tax, 5)
+fam_met <- merge(meta_file, shared_fam, by.x='group', by.y='row.names')
+
 
 #try to get the df organized to work as a boxplot - no median calculation 
 phyla_test <- phyla_met[, c("location","Firmicutes","Bacteroidetes","Proteobacteria","Verrucomicrobia","Actinobacteria","Fusobacteria")]
@@ -49,6 +67,11 @@ phyla_test <- phyla_met[, c("location","Firmicutes","Bacteroidetes","Proteobacte
 phyla_loc <- aggregate(phyla_met[, 7:ncol(phyla_met)], list(phyla_met$location), median)
 phyla_upper <- aggregate(phyla_met[, 7:ncol(phyla_met)], list(phyla_met$location), FUN= quantile, probs =0.75)
 phyla_lower <- aggregate(phyla_met[, 7:ncol(phyla_met)], list(phyla_met$location), FUN= quantile, probs =0.25)
+
+fam_loc <- aggregate(fam_met[,7:ncol(fam_met)], list(fam_met$location), median)
+
+#??
+
 
 #could do all three of these lines in one line with summarize
 #could do dplyr way, start with melted DF 

### 2017-04-19 code/otu_barplots.R
family plot attempted

@@ -45,12 +45,19 @@ source('code/sum_shared.R')
 locat <- meta_file$location
 rel_abund_top <- cbind(site = locat, rel_abund_top)
 
-#subset file to just include location 
 
-locat <- subset
-
-#for above, just subset df from rel_meta instead to get the right location/OTU table 
+#this is what i want! 
 rel_fam <- sum_OTU_by_tax_level(2, rel_abund_top, tax_file)
+
+#add a column of locations to the df
+loc_names <- meta_file$location
+rel_fam[(ncol(rel_fam)+1)] <- loc_names
+colnames(rel_fam)[ncol(rel_fam)] <- "location"
+
+fam_med <- aggregate(rel_fam[, 1:(ncol(rel_fam)-1)], list(rel_fam$location), median)
+
+
+barplot(t(fam_med))
 
 #use this code to assign phyla to each OTU in the shared file 
 shared_phyla <- get_tax_level_shared(subsample, tax, 2)
@@ -70,7 +77,16 @@ phyla_lower <- aggregate(phyla_met[, 7:ncol(phyla_met)], list(phyla_met$location
 
 fam_loc <- aggregate(fam_met[,7:ncol(fam_met)], list(fam_met$location), median)
 
-#??
+#reorganize to make next calculations easier 
+rownames(fam_loc) <- fam_loc$Group.1
+fam_loc <- fam_loc[,-1]
+fam_abund <- 100*fam_loc/4231
+
+#need to sort fam_abund by 
+
+#do this to remove 0s
+
+fam_abund <- fam_abund[,colSums(fam_abund^2) !=0]
 
 
 #could do all three of these lines in one line with summarize
@@ -114,6 +130,14 @@ phylamelt_up <- melt(phyla_abund_up[, phylanames], id.vars=1)
 phyla_abund_low <- cbind(group=rownames(phyla_lower), phyla_lower)
 rownames(phyla_abund_low) <- c()
 phylamelt_low <- melt(phyla_abund_low[, phylanames], id.vars=1)
+
+fam_abund <- cbind(group=rownames(fam_abund), fam_abund)
+rownames(fam_abund) <- c()
+famnames <- colnames(fam_abund)
+fam_melt <- melt(fam_abund[, famnames ], id.vars=1)
+
+ggplot(fam_melt, aes(x=group, y=value, fill=variable)) +geom_bar(stat = 'identity')
+
 
 #merge them riiite
 names(phylamelt_up)[3] <- "upper"

### 2017-04-25 code/otu_barplots.R
top ten barplots

@@ -131,12 +131,55 @@ phyla_abund_low <- cbind(group=rownames(phyla_lower), phyla_lower)
 rownames(phyla_abund_low) <- c()
 phylamelt_low <- melt(phyla_abund_low[, phylanames], id.vars=1)
 
-fam_abund <- cbind(group=rownames(fam_abund), fam_abund)
-rownames(fam_abund) <- c()
-famnames <- colnames(fam_abund)
-fam_melt <- melt(fam_abund[, famnames ], id.vars=1)
+fam_abund2 <- cbind(group=rownames(fam_abund), fam_abund)
+rownames(fam_abund2) <- c()
+famnames <- colnames(fam_abund2)
+fam_melt <- melt(fam_abund2[, famnames ], id.vars=1)
 
-ggplot(fam_melt, aes(x=group, y=value, fill=variable)) +geom_bar(stat = 'identity')
+
+#this is a stacked bar chart, because there are like 33 groups. need to rank first then plot top 10 or so
+
+ggplot(fam_melt, aes(x=variable, y=value, fill=variable)) +geom_bar(stat = 'identity') +facet_wrap(~group) +theme_bw()
+
+#ok still need to sort these by top 10 OTUs 
+
+fam_ordered <- fam_abund[order(fam_abund[,1:ncol(fam_abund)]),]
+
+rownames(fam_abund2) <- fam_abund2$group
+fam_abund2 <- fam_abund2[,-1]
+
+decrease_sort <- function(x){
+  sort(x, decreasing = TRUE)
+}
+
+fam_ordered <- t(apply(fam_abund2, 1, FUN = function(x) decrease_sort(x)))
+
+#so the above sort (ha) of works. but we lose colnames because the distribution isnt the same for each location
+#duh 
+#maybe it would be better to order and them one at a time like in a loop
+
+LB_fam <- subset(fam_abund2, rownames(fam_abund2) == 'LB')
+rownames(LB_fam) <- LB_fam$group
+LB_fam <- LB_fam[,-1]
+LB_fam <- t(apply(LB_fam, 1, FUN = function(x) decrease_sort(x)))
+
+LB_melted <- melt(LB_fam)
+#select first ten rows
+LB_ten <- LB_melted[1:10,]
+
+ggplot(LB_ten, aes(x=Var2, y=value)) + geom_bar(stat='identity') +theme_bw()
+ 
[…]

### 2017-04-26 submission/manuscript.Rmd
top ten barplots
manuscript with results written

@@
 
-Running title: Unique taxa differentiate proximal and distal human colon microbiota
+Running title: Specific taxa differentiate proximal and distal human colon microbiota
 
@@
 
-- what data were obtained? 16S sequencing results for samples taken from healthy human colons that had not undergone bowel prep
-Figure 1: sample collection schematic
+#### Microbial membership and diversity of the proximal and distal colon 
+Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1).
+In addition, individuals collected stool at home one week prior to the procedure.
+To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
+The relative abundance of each phyla for each site is shown in Figure 2A.
+This data combines the samples from all participants in the study.
+Each site is primarily dominated by Firmicutes and Bacteriodetes, consistent with known variability in human microbiome research.
+((Is anything statistically significant here?
+use Doherty's code)).
+Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
+That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on.
+Thus there is no clear pattern of changes in microbial diversity along the gut axis.
 
-- what were the results of the analyses? the results of the analyses are as follows:
+* 2C: NMDS/ordination of some kind ??? maybe if i calculate inter/intrapersonal variability
 
-#### Diversity analyses
-- there are similar levels (within error margins) of each phyla for each location, but there is noise due to interpatient variability.
-Overall, there is also a similar level of diversity in each location.
-However, each patient has a different pattern of diversity over the locations, some are high in R, some are high in L, etc.
-Figure 2:
-* 2A: relative abundance by phyla 
-* 2B: simpson diversity 
-* 2C: NMDS/ordination of some kind
-
-#### similarity analyses
-- we can compare sides, samples or patients by looking at distances between samples.
-here we used the thetayc calculator to make comparisons.
-The right mucosa and lumen are most similar to each other.
-As you move closer to the exit, the samples diverge.
-With the exception of the left lumen- this sample best resembles the stool collected from exit as it is...
-exiting.
-
-Figure 3:
-* 3A: the right mucosa and lumen are most similar to each other 
-* 3B: the left lumen is most similar than exit stool
+To compare similarity between sides or sites, we calculated theta YC distances from OTU abundances and compared distances between sies for all patients.
+Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
+However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
+Next, we calculated thetaYC distances to examine how each sample compares to the stool at exit.
+Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (P < 0.05, Wilcoxon, BH adjustment).
+Further, there was an even more divergent difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
+This indicates that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
+((mention the other 2 comparisons?
+"among the four comparisons made..."))
 
 #### Random Forest classification models 
-- we can use machine learning models to build classification models that identify specific features in noisy data sets that allow for separation.
-- we used a Random Forest model to separate samples from each side or site
+To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances.
+The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A).
+The constructed model used ((Xopt?
+need to follow up with this too)) features for the left and ((Xopt)) for the right.
+The models performed very well when classifying these samples (AUC X and Y).
 
@@
 * 4A: Classifying lumen vs mucosal samples on each side (L vs L, R vs R)
-* 4B: Classifying side of lumen or side of mucosa (L vs R lumen, L vs R mucosa)
+* 4B: There are differentially abundant OTUs in lumen vs mucosa
 
@@
 Figure 5:
-* 5A: There are differentially abundant OTUs in lumen vs mucosa
+* 5A: Classifying side of lumen or side of mucosa (L vs R lumen, L vs R mucosa)
 * 5B: There are differentially abundant OTUs in right vs left lumen, right vs left mucosa
@@
 - There is some commonality between the OTUs - drivers of left sidedness are consistent in both lumen and mucosa (more of this in discussion)
+
+### exit comparison models? other findings- fuso etc
 

### 2017-04-26 code/otu_barplots.R

@@ -116,6 +116,7 @@ phyla_RA <- data.frame(phyla_test[1], apply(phyla_test[2:ncol(phyla_test)],2, RA
 phylaRAnames <- colnames(phyla_RA[,1:7])
 phyla_RAmelt <- melt(phyla_RA[, phylaRAnames], id.vars=1)
 
+
 #gather AND PLOT OMG :D
 #put rownames back in their own column - you can get around this w dplyr 
 phyla_abund <- cbind(group=rownames(phyla_abund), phyla_abund)
@@ -154,9 +155,8 @@ decrease_sort <- function(x){
 
 fam_ordered <- t(apply(fam_abund2, 1, FUN = function(x) decrease_sort(x)))
 
-#so the above sort (ha) of works. but we lose colnames because the distribution isnt the same for each location
-#duh 
-#maybe it would be better to order and them one at a time like in a loop
+#a ridiculously long way to get the top ten families and select them. 
+#like a really ridiculously long way to go about that
 
 LB_fam <- subset(fam_abund2, rownames(fam_abund2) == 'LB')
 rownames(LB_fam) <- LB_fam$group
@@ -164,20 +164,37 @@ LB_fam <- LB_fam[,-1]
 LB_fam <- t(apply(LB_fam, 1, FUN = function(x) decrease_sort(x)))
 
 LB_melted <- melt(LB_fam)
-#select first ten rows
-LB_ten <- LB_melted[1:10,]
+LB_ten <- LB_melted[1:9,]
+LB_six <- LB_melted[1:6,]
 
-ggplot(LB_ten, aes(x=Var2, y=value)) + geom_bar(stat='identity') +theme_bw()
- 
 #use the top ten column to select on overall dataset then facet!
-
 topten <- as.character(LB_ten$Var2)
+topsix <- as.character(LB_six$Var2)
 
 fam_ten <- fam_abund2[, topten]
 fam_ten <- cbind(group=rownames(fam_ten), fam_ten)
 rownames(fam_ten) <- c()
 famtennames <- colnames(fam_ten)
 fam_tenmelt <- melt(fam_ten[, famtennames ], id.vars=1)
+
+#to get RA without median for boxplot 
+
+fam_RA <- fam_met[, c("location",topsix)]
+family_RA <- data.frame(fam_RA[1], apply(fam_RA[2:ncol(fam_RA)],2, RA))
+famRAnames <- colnames(family_RA[,1:6])
+fam_RAmelt <- melt(family_RA[, famRAnames], id.vars=1)
+
[…]

### 2017-04-26 submission/manuscript.Rmd
results section updated

@@
 
-#### Random Forest classification models 
+#### Random Forest classification models identify important OTUs on each side
 To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances.
@@
 The models performed very well when classifying these samples (AUC X and Y).
-
-Figure 4:
-* 4A: Classifying lumen vs mucosal samples on each side (L vs L, R vs R)
-* 4B: There are differentially abundant OTUs in lumen vs mucosa
-
-#### Important OTUs 
-- the Random Forest model lets us pick out features that dictate the classification model separation values
+The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
+For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B).
+The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Anaerococcus_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4B).
+These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.
 

### 2017-05-01 submission/manuscript.Rmd
results tweaked on manuscript

@@
 
-Figure 5:
-* 5A: Classifying side of lumen or side of mucosa (L vs R lumen, L vs R mucosa)
-* 5B: There are differentially abundant OTUs in right vs left lumen, right vs left mucosa
+Next, we built a model to differentiate the left and right microbiome.
+The model performed best when distinguishing the left versus right mucosa (Figure 5A) compared to the left versus right lumen.
+The model was able to explain ((X%)) of the variance.
+OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B).
+The model was less-effective at classifying the right and left lumen ((AUC)).
+However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).
 

### 2017-05-01 submission/manuscript.Rmd
manuscript and oncogenic otu code added

@@
 Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1).
-In addition, individuals collected stool at home one week prior to the procedure.
+Participants also collected stool at home one week prior to the procedure.
 To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
@@
 That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on.
-Thus there is no clear pattern of changes in microbial diversity along the gut axis.
+Thus there was no clear pattern of changes in microbial diversity along the gut axis.
 
@@
 
-To compare similarity between sides or sites, we calculated theta YC distances from OTU abundances and compared distances between sies for all patients.
+To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals.
 Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
 However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
-Next, we calculated thetaYC distances to examine how each sample compares to the stool at exit.
-Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (P < 0.05, Wilcoxon, BH adjustment).
-Further, there was an even more divergent difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
-This indicates that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
+Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool.
+Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
+Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
+This finding suggests that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
 ((mention the other 2 comparisons?
@@
 For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B).
-The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Anaerococcus_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4B).
+The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4B).
 These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.
@@
 However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).
-
-- There is some commonality between the OTUs - drivers of left sidedness are consistent in both lumen and mucosa (more of this in discussion)
+This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also drive sidedness- such as _Anaerococcus_ and _Finegoldia_.
 
@@
 ### Discussion
+Here we identified bacterial taxa that are specific to the lumen and mucosa of the right and left human colon from samples collected during unprepared colonoscopy.
+We found that all locations contained a range of phyla and a range of diversity, but that there was a wide variability between subjects.
+Pairwise comparisons of each of the sites revealed that the right mucosa and lumen were most similar to each other.
+Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the left lumen is most similar to stool at exit.
+Random Forest algorithms built on OTU abundances from each sample identified microbes that are particular to each location of the colon.
 
@@
 
--how did the analyses answer the central question?
-The analyses identified OTUs that are relatively unique to each location, among a wide variation differences between study participants
+Put this stuff into context with Sears PNAS paper
+
+Fuso found on the right side?
 

### 2017-05-01 code/oncogenic_otus.R

@@ -1,0 +1,33 @@
+#looking for cancer OTUs in KWS samples
+
+#first gotta find the OTU numbers for the cancer bugs
+
+#fuso OTU numbers: Otu00179, Otu00184, Otu00314, Otu00472, Otu01440
+
+meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+subsampled_file <- read.table(file='data/mothur/kws_final.an.0.03.subsample.shared', sep = '\t', header=T, row.names=2)
+
+shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+subsampled_meta <- merge(meta_file, subsampled_file, by.x='group', by.y='row.names')
+
+
+only_fuso <- shared_meta[, colnames(shared_meta) %in% c('group', 'location', 'Otu00179', 'Otu00184', 'Otu00314', 'Otu00472', 'Otu01440')]
+
+only_subsampled_fuso <- subsampled_meta[, colnames(subsampled_meta) %in% c('group', 'location', 'Otu00179', 'Otu00184', 'Otu00314', 'Otu00472', 'Otu01440')]
+
+only_fuso[,8] <- only_fuso[,3] + only_fuso[,4] + only_fuso[,5] + only_fuso[,6] + only_fuso[,7]
+
+names(only_fuso)[8] <- "total_fuso"
+
+ggplot(only_fuso, aes(x=location, y=total_fuso)) + geom_point() +geom_jitter()
+
+ggplot(only_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter()
+
+ggplot(only_subsampled_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter()
+
+
+#ok probably now should get seq and blast each otu to figure out which is nucleatum 
+
+#do this with subsampled file maybe too? 

### 2017-05-09 code/oncogenic_otus.R
oncogenic otus updated with results

@@ -28,6 +28,26 @@ ggplot(only_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter()
 ggplot(only_subsampled_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter()
 
 
-#ok probably now should get seq and blast each otu to figure out which is nucleatum 
+fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]
 
-#do this with subsampled file maybe too? 
+fuso179[,5] <- (fuso179[,4]/4321)*100
+names(fuso179)[5] <- "Otu00179_relAbund"
+
+ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter() +theme_bw()
+
+#other oncogenic otus - Otu00152 is P. asaccharolytica. Otu00248 is p micra
+
+p_asach <- shared_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00152")]
+
+p_asach[,4] <- (p_asach[,3]/4321) *100
+names(p_asach)[4] <- "Otu00152_abund"
+
+ggplot(p_asach, aes(x=location, y=Otu00152_abund)) +geom_jitter() +theme_bw()
+
+p_micra <- shared_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00248")]
+
+p_micra[,4] <- (p_micra[,3]/4321) *100
+names(p_micra)[4] <- "Otu00248_abund"
+
+ggplot(p_micra, aes(x=location, y=Otu00248_abund)) +geom_jitter() +theme_bw()
+

### 2017-05-10 submission/manuscript.Rmd
oncogenic otus updated with results
discussion updates

@@
 
-### exit comparison models? other findings- fuso etc
+### exit comparison models?
+
+##draft of oncogenic OTU paragraph, this may go in results or discussion
+
+Some of the healthy patients have Fusobacterium OTUs.
+The most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_.
+It is the only OTU to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
+((cite joe and niels work somewhere in here, maybe in the discussion?)).
+The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample.
+((Table or something?)).
+OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors.
+OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides.
+Samples that were positive for _P.
+asacharolytica_ ranged in abundances from 0.01% - 16%.
+Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248.
+Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa.
 
@@
 Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)
+some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20)
 
@@
 
-Fuso found on the right side?
+We detected _F.
+nucleatum_ and _P.
+asacharolytica_ in 7 and 5 of our patients, respectively.
+These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel).
+Interestingly, while _F.
+nucleatum_ was found on both sides of the colon, _P.
+asacharolytica_ was only detected in the left mucosa.
+Not much is known about the distribution of _P.
+asacharolytica_ but given its documented anaerobic characteristics, it might not be surprising that it resides in the less-oygen-rich left bowel (cite something).
+In studies examining bacteria on colorectal cancer tumors, _F.
+nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F.
+nucleatum_ decreases along the colon to rectum (cite Mima 2016).
+Of the (8) individuals positive for _F.
+nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumena and left mucosa.
+That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F.
+nucleatum_ at all (Mima 2016).
+
+Fuso found on the right side? discussion about cancer-specific OTUS
+cant pull out ETBF or pks+ ecoli with 16S
+found fuso and porpyhromonas, but these are healthy people
+what is rate of colonization with these bugs in healthy in literature?
+of course presence does not precede cancer or IBD necessarily
 
@@
 
--what's next?
-cancer associated OTUs on one side or the other, compare to previous studies.
-But really to understand what's going on in the various regions of the gut we need more information than just 16S
+-what's next?  But really to understand what's going on in the various regions of the gut we need more information than just 16S
 

### 2017-05-15 submission/manuscript.Rmd
manuscript discussion updates

@@
 
-* 2C: NMDS/ordination of some kind ??? maybe if i calculate inter/intrapersonal variability
+Similarly, when examining community structure between sites and sides, there is no clear pattern among individuals.
+Microbial communities do not separate on an ordination by the left or right sides in either the mucosal or lumenal samples (Figure 2C).
 
@@
 For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B).
-The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4B).
+The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C).
 These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.
@@
 
-### exit comparison models?
-
-##draft of oncogenic OTU paragraph, this may go in results or discussion
-
+### Bacterial OTUs associated with cancer are found in healthy individuals 
 Some of the healthy patients have Fusobacterium OTUs.
@@
 Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa.
+Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.
 
@@
 
-Recap of results, interesting discussion points about diversity and previous studies of diversity along the gut axis of healthy people
+discussion of diversity results- different patterns in each person.
+ref previous studies of diversity along gut axis of healthy people.
+in this same paper ( I think) can also reference difference in NMDS...
+if I end up including that.
+
+discussion of similarity results- may combine with previous paragraph here, prior studies, again underscore extreme variability of humans
+
+Discussion of RF models, how they compare to FIT, Joe's check joes paper, how well does model do in L vs R for FIT?
+when we look at models for CRC, cant predict R vs L, so create models
+
+discuss host response papers? what does that mean.
 
 Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)
-some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20)
+some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20) - not necessarily anything interesting there.
 
-Put this stuff into context with Sears PNAS paper
+Put this stuff into context with Sears PNAS paper on sidedness
 
@@
 asacharolytica_ in 7 and 5 of our patients, respectively.
-These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel).
+These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite).
 Interestingly, while _F.
@@
 nucleatum_ decreases along the colon to rectum (cite Mima 2016).
-Of the (8) individuals positive for _F.
-nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumena and left mucosa.
+Of the (8) (40%) individuals positive for _F.
+nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa.
 That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F.
 nucleatum_ at all (Mima 2016).
+In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F.
+nucleatum_ in mucosal biopsies (Strauss 2011).
+These studies have the added caveat of being sampled from prepared or cleaned colons.
+Thus, the presence of _F.
+nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers.
+Because of the spatial distribution of the _F.
+nucleatum_ in our sample set, we cannot develop a model for the role of _F.
+nucleatum_ in the healthy colon.
+Indeed, rather than the _F.
+nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016).
 
-Fuso found on the right side? discussion about cancer-specific OTUS
-cant pull out ETBF or pks+ ecoli with 16S
-found fuso and porpyhromonas, but these are healthy people
 what is rate of colonization with these bugs in healthy in literature?
-of course presence does not precede cancer or IBD necessarily
+
+discuss host response papers
 

### 2017-05-15 submission/manuscript.Rmd
manuscript

@@
 
-Similarly, when examining community structure between sites and sides, there is no clear pattern among individuals.
-Microbial communities do not separate on an ordination by the left or right sides in either the mucosal or lumenal samples (Figure 2C).
+To determine if the structure of the community is consistent between locations we constructed an NMDS ordination to examine the distances between each sample.
+We found that microbial communities do not cluster on an ordination by the left or right sides in either the mucosal or lumenal samples (Figure 2C).
+Previous human microbiome studies that have sampled multiple sites have found that individuals are typically more similar to themselves than others in the study (cite).
+To see if this effect was driving the organization of the ordination, we connected communities from the same individual using gray lines on Figure 2C.
+While some individuals did have samples that clustered near each other (shorter lines), there were just as many individuals who's samples diverged from each other (longer lines).
+Similar to the diversity metrics, when examining community structure between sites and sides, there is no clear pattern between or within individuals.
 

### 2017-05-23 submission/manuscript.Rmd
manuscript updates

@@
 This data combines the samples from all participants in the study.
-Each site is primarily dominated by Firmicutes and Bacteriodetes, consistent with known variability in human microbiome research.
+Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research.
 ((Is anything statistically significant here?
@@
 Random Forest algorithms built on OTU abundances from each sample identified microbes that are particular to each location of the colon.
+Finally, we were able to detect some bacterial OTUs associated with colorectal cancer in our healthy patient cohort.
+Using unprepped colonoscopy and machine learning, we have identified bacterial phyla specific to the healthy proximal and distal human colon.
+
+We identified a range of Simpson diversity values between patients and between each anatomical site.
+As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon.
+There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal, some patients had the opposite.
+An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005).
+When ordinating the subject communities to compare the right and left sides, we found that more than half of the samples from the same patient clustered near each other (Figure 2C).
+However, there were a few patients whose samples were very different from each other and did not ordinate in that manner.
+One explanation for this ____ - human diversity, sampling locations arent perfectly matched (but so what?)
 
@@
 
-discuss host response papers
-
 -what does this answer tell us about the broader field?

### 2017-05-23 code/community.R
new interpersonal differences plot

@@ -1,6 +1,6 @@
 # OTU bar plots to look at community membership
 
-pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson', 'vegan')
 for (dep in pack_used){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
@@ -180,6 +180,39 @@ ggplot(simpmeta, aes(x=site, y=invsimpson)) + geom_jitter(col=simpmeta$patient) 
 
 tyc <- read.table("data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
 
+#can i quantify the distances between samples from the same patient and samples from each side? 
+
+alltyc <- read.table("data/process/allshared.summary", sep = '\t', header = T, row.names=NULL)
+alltyc <- separate(alltyc, label, into= c('pt1', 'samp1'), sep="-", remove=F)
+alltyc <- separate(alltyc, comparison, into= c('pt2', 'samp2'), sep="-", remove=F)
+alltyc <- alltyc[-1]
+alltyc <- alltyc[-7]
+
+#ultimately want a plot of all points where pt1 == pt2 in one bar and all of the others in another column 
+#unite and make column of 0/1 for matches? then can plot 1 and 0s 
+#should i separate out lumen and mucosa ? sure or no not for now
+
+alltyc["same_pt"] <- NA
+
+for (i in 1:nrow(alltyc)){
+  if (alltyc$pt1[i] == alltyc$pt2[i]){
+    alltyc$same_pt[i] <- 1
+  }
+  else alltyc$same_pt[i] <- 0
+}
+
+alltyc[10] <- as.factor(alltyc[10])
+
+ggplot(alltyc, aes(x=as.factor(same_pt), y=thetayc)) + geom_jitter(width=0.1) + theme_bw()+
+  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 16), axis.title= element_text(size=18)) +
[…]

### 2017-05-23 code/diversity_analysis.R

@@ -10,7 +10,7 @@ simps <- read.table(file='data/mothur/kws_final.an.groups.summary', header = T)
 shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
 
 #load niel's script for properly , reading in thetayc distances
-#source(file = 'read.dist.R')
+source(file = 'code/read.dist.R')
 #tyc <-read.dist(file='kws.an.thetayc.0.03.lt.dist', input = "lt", make.square=F, diag=NA)
 
 #nmds analysis
@@ -68,10 +68,17 @@ stoolshared <- add_rownames(stoolshared, var = "group")
 stoolshared <- stoolshared[-2]
 stoolshared <- cbind(label=0.03, stoolshared[1:ncol(stoolshared)])
 
+all <- meta_file$group[meta_file$side=='left' | meta_file$side=='right']
+allshared <- shared_file[row.names(shared_file) %in% all,]
+allshared <- add_rownames(allshared, var = "group")
+allshared <- allshared[-2]
+allshared <- cbind(label=0.03, allshared[1:ncol(allshared)])
+
 write.table(leftshared, file='data/process/leftshared.shared', sep='\t', quote=F, row.names=F)
 write.table(rightshared, file='data/process/rightshared.shared', sep='\t', quote=F, row.names=F)
 write.table(mucosashared, file='data/process/mucosashared.shared', sep='\t', quote=F, row.names=F)
 write.table(stoolshared, file='data/process/stoolshared.shared', sep='\t', quote=F, row.names=F)
+write.table(allshared, file='data/process/allshared.shared', sep='\t', quote=F, row.names=F)
 
 #run all in mothur, dist.shared and then nmds. import back in. plot. dance! 
 
@@ -90,6 +97,27 @@ ggplot(mucosanmds, aes(x=axis1, y=axis2)) + geom_point(aes(color=as.factor(patie
 stoolnmds <- read.table(file='data/process/stoolshared.thetayc.0.03.lt.ave.nmds.axes', sep='\t', header=T)
 stoolnmds <- merge(stoolnmds, meta_file)
 ggplot(stoolnmds, aes(x=axis1, y=axis2)) + geom_point(aes(color=as.factor(patient), shape=side, size = 1.5)) +theme_bw() + ggtitle("Stool samples comparison")
+
+allnmds <- read.table(file='data/process/allshared.thetayc.0.03.lt.nmds.axes', sep = '\t', header = T)
[…]

### 2017-05-24 code/community.R
discussion tweaks

@@ -203,11 +203,11 @@ for (i in 1:nrow(alltyc)){
 
 alltyc[10] <- as.factor(alltyc[10])
 
-ggplot(alltyc, aes(x=as.factor(same_pt), y=thetayc)) + geom_jitter(width=0.1) + theme_bw()+
+ggplot(alltyc, aes(x=as.factor(same_pt), y=thetayc)) + geom_jitter(width=0.15, shape=21, size=3, fill='grey', col='black') + theme_bw()+
   theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 16), axis.title= element_text(size=18)) +
-  scale_x_discrete(labels=c("distance between patients", "distance within a patient")) +
+  scale_x_discrete(labels=c("interpersonal", "intrapersonal")) +
   theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")+ 
-  stat_summary(aes(x=as.factor(same_pt), y=thetayc), data = alltyc, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.3, col = "red")
+  stat_summary(aes(x=as.factor(same_pt), y=thetayc), data = alltyc, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.4, col = "black")
 
 
 wilcox.test(thetayc ~ same_pt, data = alltyc)

### 2017-05-25 code/oncogenic_otus.R
oncogenic OTUs

@@ -25,7 +25,7 @@ ggplot(only_fuso, aes(x=location, y=total_fuso)) + geom_point() +geom_jitter()
 
 ggplot(only_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter()
 
-ggplot(only_subsampled_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter()
+ggplot(only_subsampled_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter() +theme_bw()
 
 
 #ok probably now should get seq and blast each otu to figure out which is nucleatum 

### 2017-05-25 code/community.R
oncogenic

@@ -1,6 +1,6 @@
 # OTU bar plots to look at community membership
 
-pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson', 'vegan')
 for (dep in pack_used){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
@@ -180,6 +180,39 @@ ggplot(simpmeta, aes(x=site, y=invsimpson)) + geom_jitter(col=simpmeta$patient) 
 
 tyc <- read.table("data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
 
+#can i quantify the distances between samples from the same patient and samples from each side? 
+
+alltyc <- read.table("data/process/allshared.summary", sep = '\t', header = T, row.names=NULL)
+alltyc <- separate(alltyc, label, into= c('pt1', 'samp1'), sep="-", remove=F)
+alltyc <- separate(alltyc, comparison, into= c('pt2', 'samp2'), sep="-", remove=F)
+alltyc <- alltyc[-1]
+alltyc <- alltyc[-7]
+
+#ultimately want a plot of all points where pt1 == pt2 in one bar and all of the others in another column 
+#unite and make column of 0/1 for matches? then can plot 1 and 0s 
+#should i separate out lumen and mucosa ? sure or no not for now
+
+alltyc["same_pt"] <- NA
+
+for (i in 1:nrow(alltyc)){
+  if (alltyc$pt1[i] == alltyc$pt2[i]){
+    alltyc$same_pt[i] <- 1
+  }
+  else alltyc$same_pt[i] <- 0
+}
+
+alltyc[10] <- as.factor(alltyc[10])
+
+ggplot(alltyc, aes(x=as.factor(same_pt), y=thetayc)) + geom_jitter(width=0.15, shape=21, size=3, fill='grey', col='black') + theme_bw()+
[…]

### 2017-05-25 code/oncogenic_otus.R

@@ -28,6 +28,26 @@ ggplot(only_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter()
 ggplot(only_subsampled_fuso, aes(x=location, y=Otu00179)) +geom_point() + geom_jitter() +theme_bw()
 
 
-#ok probably now should get seq and blast each otu to figure out which is nucleatum 
+fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]
 
-#do this with subsampled file maybe too? 
+fuso179[,5] <- (fuso179[,4]/4321)*100
+names(fuso179)[5] <- "Otu00179_relAbund"
+
+ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter() +theme_bw()
+
+#other oncogenic otus - Otu00152 is P. asaccharolytica. Otu00248 is p micra
+
+p_asach <- shared_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00152")]
+
+p_asach[,4] <- (p_asach[,3]/4321) *100
+names(p_asach)[4] <- "Otu00152_abund"
+
+ggplot(p_asach, aes(x=location, y=Otu00152_abund)) +geom_jitter() +theme_bw()
+
+p_micra <- shared_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00248")]
+
+p_micra[,4] <- (p_micra[,3]/4321) *100
+names(p_micra)[4] <- "Otu00248_abund"
+
+ggplot(p_micra, aes(x=location, y=Otu00248_abund)) +geom_jitter() +theme_bw()
+

### 2017-05-25 submission/manuscript.Rmd
discussion tweaks
results

@@
 so what is needed is paired samples from healthy people that have not undergone bowel prep, from the right/left and lumen/mucosa
+
+this is a really good paper, good for framing intro and discussion : Numerical ecology validates a biogeographical distribution and gender-based effect on mucosa-associated bacteria along the human colon
 
@@
 An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005).
-When ordinating the subject communities to compare the right and left sides, we found that more than half of the samples from the same patient clustered near each other (Figure 2C).
-However, there were a few patients whose samples were very different from each other and did not ordinate in that manner.
-One explanation for this ____ - human diversity, sampling locations arent perfectly matched (but so what?)
+We next used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other.
+We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants.
+This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (Eckberg 2005, Aguierre de Carcer, 2015).
 
-discussion of diversity results- different patterns in each person.
-ref previous studies of diversity along gut axis of healthy people.
-in this same paper ( I think) can also reference difference in NMDS...
-if I end up including that.
+To get around the noisiness created by a diverse set of samples, we employed machine learning algorithms to identify microbes specific to each side.
 
-discussion of similarity results- may combine with previous paragraph here, prior studies, again underscore extreme variability of humans
-
-Discussion of RF models, how they compare to FIT, Joe's check joes paper, how well does model do in L vs R for FIT?
-when we look at models for CRC, cant predict R vs L, so create models
-
-discuss host response papers? what does that mean.
+Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors.
+Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone.
+- Could try to predict L vs R tumors from exit stool? long term. maybe. probably not. value of getting data about exit stool comparisons. real issue is diagnosing right though. 
+- note that we have FIT for all of these people but they're all below limit of detection (except for 2). when we look at models for CRC, cant predict R vs L, so create models for healthy
 
@@
 -what's next?  But really to understand what's going on in the various regions of the gut we need more information than just 16S 
+discuss host response papers? what does that mean. link bugs to host response on left and right? future directions or intro
 

### 2017-05-25 submission/manuscript.Rmd
discussion

@@
 To get around the noisiness created by a diverse set of samples, we employed machine learning algorithms to identify microbes specific to each side.
+For each comparison we identified top X OTUs that were strongly predictive of one site or another.
+Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (cite nature review).
+For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon (cite).
+The left mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_.
+The model was less effective at classifying the right and left lumenal contents, probably because the samples are arguably composed of the same material.
 
-Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors.
-Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone.
-- Could try to predict L vs R tumors from exit stool? long term. maybe. probably not. value of getting data about exit stool comparisons. real issue is diagnosing right though. 
-- note that we have FIT for all of these people but they're all below limit of detection (except for 2). when we look at models for CRC, cant predict R vs L, so create models for healthy
-
-Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)
+other factors besides oxygen? pH? 
+offer another model/schematic?
 some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20) - not necessarily anything interesting there.
-
-Put this stuff into context with Sears PNAS paper on sidedness
 
@@
 
-what is rate of colonization with these bugs in healthy in literature?
+what is rate of colonization with these bugs in healthy in literature? how did these bugs factor into our model?
+
+Put this stuff into context with Sears PNAS paper on sidedness
+
+how does bowel prep fit into all of this
+
+Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors.
+Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone.
+- Could try to predict L vs R tumors from exit stool? long term. maybe. probably not. value of getting data about exit stool comparisons. real issue is diagnosing right though. 
+- note that we have FIT for all of these people but they're all below limit of detection (except for 2).
+when we look at models for CRC, cant predict R vs L, so create models for healthy.
+can also cite niels stuff showing that FIT is representative of stool community.
+could it also be representative of left?
+- LINK TO EXIT STOOL COMPARISONS.
 

### 2017-05-25 submission/manuscript.Rmd
discussion

@@
 
+Discussion of the phyla present and at what levels?
+Good refs in the gut biogeography paper - mucosa vs lumen enrichment stuff "significant variability between sample locations less than 1 cm apart, which suggests that mucosal microbial populations occur in patches" there is also patchiness in fecal samples themselves: spatial niche partitioning in faeces.
+.
+Patchiness/sampling an issue as well as differences in human age, sex, genotype, adn definitely diet, pH, oxygen, etc.
+(does this go in the intro?)
+
 We identified a range of Simpson diversity values between patients and between each anatomical site.
 As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon.
-There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal, some patients had the opposite.
+There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite.
 An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005).
@@
 We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants.
-This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (Eckberg 2005, Aguierre de Carcer, 2015).
+This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (Eckberg 2005, Aguierre de Carcer, 2015, Zhang et al 2013).
 
@@
 
-other factors besides oxygen? pH? 
 offer another model/schematic?

### 2017-06-06 submission/manuscript.Rmd
working on the intro

@@
 Colorectal cancer (CRC) remains a leading cause of death worldwide.
-Tumors of the proximal (right) and distal (left) colon are morpologically and genetically distinct.
+Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct.
 Previous work from our group found that microbial dysbiosis is associated with the development of colorectal cancer tumors in studies of both mice and humans.
@@
 
-- colon cancer 
-- right and left tumors are morphologically and genetically distinct (why Q is important)
+Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States (cite).
+CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon.
+Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen.
+These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (cite).
+
+The human colon contains a number of microhabitats that enrich for different microbial membership at each site.
+Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract.
+
 - the microbiome of healthy people and patients with CRC is distinguishable at the OTU level

### 2017-06-06 submission/manuscript.Rmd
working on the intro

@@
 These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (cite).
+The proximal and distal human colon are comprised of different microhabitats that likely enrich for different microbial membership at each site.
+Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract in accordance with oxygen, nutrient and pH gradients (cite).
+Further, our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (cite).
 
-The human colon contains a number of microhabitats that enrich for different microbial membership at each site.
-Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract.
+These studies have been limited though by the sample collection methods.
+Nearly all human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures (cite).
+While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel (cite).
+Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but has been proven to disrupt the normal gut microbiome
 

### 2017-06-06 submission/manuscript.Rmd
intro done for now

@@
 Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract in accordance with oxygen, nutrient and pH gradients (cite).
-Further, our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (cite).
+Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
 
-These studies have been limited though by the sample collection methods.
-Nearly all human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures (cite).
+Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (cite).
+Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (cite).
+Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (cite).
+These species include the oral pathogens _Fusobacterium nucleatum_, _Porphyromonas asacharolytica_ and _Parvimonas micra_.
+Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (cite Niel).
+However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon.
+Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
+
+Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods.
+The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures (cite).
 While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel (cite).
-Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but has been proven to disrupt the normal gut microbiome
+Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (Jalanka 2015, Harrell 2012).
+As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
 
-- the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
-- specific keystone species are associated with CRC tumors
-- Central Question: do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
-- nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
-- bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
-- how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations
-
-- What data is needed to answer this question?
-first, need to characterize healthy species that are specific to each location.
-then we can ask, if/how disruption affects tumorigenesis?
-so what is needed is paired samples from healthy people that have not undergone bowel prep, from the right/left and lumen/mucosa
-
-this is a really good paper, good for framing intro and discussion : Numerical ecology validates a biogeographical distribution and gender-based effect on mucosa-associated bacteria along the human colon
-
-- what methods are used to get the data? unprepped colonoscopy
-
-- What analysis must be applied? - diversity analysis and machine learning to pull out small variations/differences among the noisy data set.
-
-- hypothesis: there are distinct bacteria on the left and right side that can be distinguished by 16S sequencing of stool
+In this study we sought to combine these approaches to characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
+Our design uses unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native flora in 20 healthy volunteers.
+To get around the noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
+We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes.
+By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.
 

### 2017-06-07 submission/manuscript.Rmd
progress with discussion

@@
 Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
+
+add something about sidedness of the gut - embryonic origins, different lymph and blood supplies - more background about male/female differences probably good too
 
@@
 
+#### Exit stool comparisons model - specifics here or above?
+
 ### Discussion
@@
 
-Discussion of the phyla present and at what levels?
-Good refs in the gut biogeography paper - mucosa vs lumen enrichment stuff "significant variability between sample locations less than 1 cm apart, which suggests that mucosal microbial populations occur in patches" there is also patchiness in fecal samples themselves: spatial niche partitioning in faeces.
-.
-Patchiness/sampling an issue as well as differences in human age, sex, genotype, adn definitely diet, pH, oxygen, etc.
-(does this go in the intro?)
+When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes.
+This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome.
+Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (cite).
+Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (cite).
+That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (cite).
+Thus, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
 
@@
 An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005).
-We next used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other.
+We used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other.
 We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants.
@@
 
-To get around the noisiness created by a diverse set of samples, we employed machine learning algorithms to identify microbes specific to each side.
+To get around the noisiness created by a diverse set of samples, we built a Random Forest model to identify microbes specific to each side.
 For each comparison we identified top X OTUs that were strongly predictive of one site or another.
@@
 
-offer another model/schematic?
-some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20) - not necessarily anything interesting there.
-
 We detected _F.
 nucleatum_ and _P.
-asacharolytica_ in 7 and 5 of our patients, respectively.
+asacharolytica_ in 7 and 5 of our subjects, respectively.
 These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite).
@@
 nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016).
+Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis.
 
-what is rate of colonization with these bugs in healthy in literature? how did these bugs factor into our model?
+how does bowel prep fit into all of this- do these match publisehd findings? sort of hard to differentiate bc so many humans?
 
-Put this stuff into context with Sears PNAS paper on sidedness
-
-how does bowel prep fit into all of this
-
+Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer.
 Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors.
 Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone.
-- Could try to predict L vs R tumors from exit stool? long term. maybe. probably not. value of getting data about exit stool comparisons. real issue is diagnosing right though. 
-- note that we have FIT for all of these people but they're all below limit of detection (except for 2).
-when we look at models for CRC, cant predict R vs L, so create models for healthy.
-can also cite niels stuff showing that FIT is representative of stool community.
-could it also be representative of left?
-- LINK TO EXIT STOOL COMPARISONS.
+In this study, all of our subjects were tested using FIT and all but two were found to be below the limit of detection (data not shown).
+Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (cite Niel fit paper).
+Our work already showed that the stool most accurately matched the samples from the left lumen.
+So we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
+But further studies would need to be done.
+Hard to get a complete set of matched samples from CRC patients though due to discomfort and insurance issues.
+One caveat of the previous study is that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.
+
+offer another model/schematic?
 
@@
 
--what's next?  But really to understand what's going on in the various regions of the gut we need more information than just 16S 
-discuss host response papers? what does that mean. link bugs to host response on left and right? future directions or intro
+By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect oncogenesis of proximal and distal CRC tumors.
+To this point, 16S community profiling studies are a good start but not enough information to probe these questions.
+Future studies examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.
 

### 2017-06-07 submission/manuscript.Rmd
discussion updated

@@
 
-how does bowel prep fit into all of this- do these match publisehd findings? sort of hard to differentiate bc so many humans?
+Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies.
+A rare report of a matched-colonoscopy study that sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (cite).
+This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (Shobar 2016).
+Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites.
+Further, the differences were not great enough to overcome interpersonal differences between subjects.
+So, bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel.
+Thus our findings of specific bacteria at each site of the colon are strengthened by the lack of bowel preparation.
 
@@
 To this point, 16S community profiling studies are a good start but not enough information to probe these questions.
-Future studies examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.
+Our sample set of matched proximal, distal, lumenal and mucosal samples from colons that have not undergone bowel preparation presents a unique opportunity to explore further Future studies examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.
 

### 2017-06-07 submission/manuscript.Rmd
methods

@@
 
-### Bacterial OTUs associated with cancer are found in healthy individuals 
+#### Bacterial OTUs associated with cancer are found in healthy individuals 
 Some of the healthy patients have Fusobacterium OTUs.
@@
 Not much is known about the distribution of _P.
-asacharolytica_ but given its documented anaerobic characteristics, it might not be surprising that it resides in the less-oygen-rich left bowel (cite something).
+asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (cite something, Albenberg, 2014).
 In studies examining bacteria on colorectal cancer tumors, _F.
@@
 Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone.
-In this study, all of our subjects were tested using FIT and all but two were found to be below the limit of detection (data not shown).
 Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (cite Niel fit paper).
@@
 offer another model/schematic?
-
--what does this answer tell us about the broader field?
-This answer told us that there are specific differences of bacteria in each side and that we can use this information to form hypotheses about how specific microbial/ecological variations from the host can affect oncogenesis.
-It tells us too that sample collection methods are important, and what we have found in unprepped bowels do/do not match published findings.
+what does this tell us about the broader field?
+do i need to tie in with the host any more here?
+(for instance, how microbes might interact with mutated genes/pathways).
+model, specific microbes on each side, change in mutation or metabolism affects change in community or precedes oncogenesis in a specific way?
 
 By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect oncogenesis of proximal and distal CRC tumors.
-To this point, 16S community profiling studies are a good start but not enough information to probe these questions.
-Our sample set of matched proximal, distal, lumenal and mucosal samples from colons that have not undergone bowel preparation presents a unique opportunity to explore further Future studies examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.
+To this point, 16S community profiling studies do not provide enough information to probe these questions.
+Our sample set of matched proximal, distal, lumenal and mucosal samples from colons that have not undergone bowel preparation presents a unique opportunity to explore further questions about the microbiome along the gut axis.
+Specifically, examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.
 
@@
 ### Methods
-Study participants/IRB
+#### Human subjects
+The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number XXXX.
+Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study.
+Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl or Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive.
+20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure.
+(Table of participant characteristics?
+gender, race split?)
 
-Sample collection
+#### Sample collection
+At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit supplies).
+At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco) and ship the samples to the University.
+Notably, subjects did not undergo any bowel preparation method prior to sampling.
+On procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System.
+Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate.
+A flexible sigmoidoscope was first inserted about 25 cm into the colon and endoscopy brush used to collect lumenal/stool contents.
+Two lumenal samples were collected and the contents immediately deposited into RNAlater (source) and flash-frozen in liquid nitrogen.
+The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
+Three (double check this) mucosal biopsies were collected and flash-frozen in RNAlater.
+These samples comprised the distal or left colon samples.
+The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the ascending colon.
+Samples were then collected as in the distal colon and the colonoscope withdrawn.
+All samples were stored at -80 C until study completion.
 
-Sample processing/DNA extraction
+#### Sample processing and sequencing
 
-16S rRNA sequencing
-
-Statistical analysis/Random Forest
+#### Statistical analysis/Random Forest
 

### 2017-06-08 submission/manuscript.Rmd
methods added

@@
 At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit supplies).
-At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco) and ship the samples to the University.
+At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.) and ship the samples to the University.
 Notably, subjects did not undergo any bowel preparation method prior to sampling.
@@
 #### Sample processing and sequencing
+DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories).
+For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples.
+The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich 2013).
+Sequences were curated and clustered into operational taxonomic units (OTUs) using mothur software as described previously (Schloss 2009).
+The resultant OTUs assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (wang 2007).
 

### 2017-06-09 submission/manuscript.Rmd
methods done for now

@@
 Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate.
-A flexible sigmoidoscope was first inserted about 25 cm into the colon and endoscopy brush used to collect lumenal/stool contents.
+A flexible sigmoidoscope was first inserted about 25cm into the colon and endoscopy brush used to collect lumenal/stool contents.
 Two lumenal samples were collected and the contents immediately deposited into RNAlater (source) and flash-frozen in liquid nitrogen.
@@
 
-#### Sample processing and sequencing
+#### Sample processing, sequencing and analysis
 DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories).
@@
 The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich 2013).
-Sequences were curated and clustered into operational taxonomic units (OTUs) using mothur software as described previously (Schloss 2009).
-The resultant OTUs assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (wang 2007).
+Sequences were curated using the mothur software as described previously (Schloss 2009).
+The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (wang 2007) and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
+Sequencing and analysis of a mock community revealed the error rate to be X%.
+Samples were rarefied to 4231 sequences per sample in order to reduce sampling bias.
 
-#### Statistical analysis/Random Forest
+Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur (cite).
+ThetaYC distances were calculated to determine the dissimilarity between two samples.
+Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location (cite Liaw 2002).
+To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.
+
+#### Statistical analysis
+Differences in community membership at the phyla level were tested using the analysis of molecular variance (AMOVA) metric in mothur.
+Differences in thetaYC distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
 
@@
 
-Figure 1
+####Figure 1
+Sampling strategy.
+A flexible sigmoidoscope was used to sample the distal colonic luminal contents and mucosa.
+The scope was inserted ~ 25cm into the patient and endoscopy brushes were used to sample the luminal contents (green star).
+A separate set of biopsy forceps was used to sample the distal mucosa (blue star).
+The sigmoidoscope was removed.
+A pediatric colonoscope was inserted and used to access the proximal colon.
+Biopsies were taken of the proximal luminal contents and mucosa as described.
+One week prior to the procedure stool was collected at home and sent into the laboratory.
 

### 2017-06-09 submission/manuscript.Rmd
full rough draft done

@@
 ### Acknowledgments 
+We thank all the individuals who volunteered for the study.
+This work was supported by the Rose and Lawrence C.
+Page Foundation (DKT).
+We would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman for their roles in study coordination, subject recruitment, sample collection and sample processing.
 
@@
 20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure.
-(Table of participant characteristics?
-gender, race split?)
+There were 13 female and 7 male subjects ranging in age from 25 to 64.
 
@@
 
-### Figure Legends
+### Figure legends
 
@@
 A flexible sigmoidoscope was used to sample the distal colonic luminal contents and mucosa.
-The scope was inserted ~ 25cm into the patient and endoscopy brushes were used to sample the luminal contents (green star).
+The scope was inserted ~ 25cm into the subject and endoscopy brushes were used to sample the luminal contents (green star).
 A separate set of biopsy forceps was used to sample the distal mucosa (blue star).
@@
 One week prior to the procedure stool was collected at home and sent into the laboratory.
+Representative images from one individual are shown.
 
-Figure 2
+####Figure 2
+Microbial membership and diversity of the proximal and distal human colon.
+A) Relative abundance of the top five bacterial phyla in each sampling site.
+Each box represents the median and confidence intervals.
+B) Simpson diversity of the microbial communities at each location.
+The lines represent the median values.
 
-Figure 3
+####Figure 3
+Similarity of microbial community structure between sites of the gut.
+ThetaYC distances are shown for interpersonal similarities between two sites – each point represents one individual.
+In (A), comparisons of the right and left mucosal and lumen are shown.
+In (B), comparisons of each site to the exit stool are shown.
 
-Figure 4
+####Figure 4
+Random Forest classifies the mucosa and lumen of each side of the colon.
+A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the left and right sides of the colon.
+(B) Top five OTUs that are most important for the classification model for the left mucosa and lumen (B) and the right mucosa and lumen (C).
 
-Figure 5
+####Figure 5 
+Random Forest classifies the left and right sides of the colon.
+A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying left lumen versus right lumen (orange) and left mucosa vs right mucosa (green).
+(B) Top five OTUs that are most important for the classification model for the left and right mucosa (B) and the left and right lumen (C).
+
+####Figure 6
+Location and abundance of cancer-associated OTUs.
+Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum_ (B) _Porphyromonas asacharolytica_ and (C) _Parvimonas micra_.
+
+###Supplemental material
+
+####Supplemental Figure 1
+Random Forest classifier comparing samples to exit stool. ROC curves for the RF model classifying exit stool versus the right or left lumen, pooled mucosa or pooled lumen are shown.
 

### 2017-06-12 submission/manuscript.Rmd
now with references

@@
 csl: plos.csl
+bibliography: references.bib 
 ---
@@
 
-Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States (cite).
+Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States.
 CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon.
 Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen.
-These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (cite).
-The proximal and distal human colon are comprised of different microhabitats that likely enrich for different microbial membership at each site.
-Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract in accordance with oxygen, nutrient and pH gradients (cite).
+These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (@Yamauchi2012).
+CRC has recently been described to fall into four molecular subtypes (@Guinney2015).
+Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms.
+In contrast, right-sided colon tumors are bulky and project into the lumen, often causing no symptoms until they have developed into carcinomas (@Benedix2010).
+Because of the absence of symptoms, right-sided tumors, which comprise up to 47% of CRCs, have a significantly higher mortality rate (@Benedix2010).
+The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
 Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
+Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.
 
-add something about sidedness of the gut - embryonic origins, different lymph and blood supplies - more background about male/female differences probably good too
-
-Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (cite).
-Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (cite).
-Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (cite).
+Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012).
+Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014).
+Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013).
+Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014).
 These species include the oral pathogens _Fusobacterium nucleatum_, _Porphyromonas asacharolytica_ and _Parvimonas micra_.
-Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (cite Niel).
+Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).
 However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon.
 Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
+Disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.
+However, the contribution of the gut microbiota to these subtypes is largely undefined.
 
 Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods.
-The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures (cite).
-While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel (cite).
-Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (Jalanka 2015, Harrell 2012).
+The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures.
+While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
+Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (@Jalanka2014, @Harrell2012).
 As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

### 2017-06-12 submission/manuscript.Rmd
more references

@@
 Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research.
-((Is anything statistically significant here?
-use Doherty's code)).
 Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
@@
 Thus there was no clear pattern of changes in microbial diversity along the gut axis.
-
-To determine if the structure of the community is consistent between locations we constructed an NMDS ordination to examine the distances between each sample.
-We found that microbial communities do not cluster on an ordination by the left or right sides in either the mucosal or lumenal samples (Figure 2C).
-Previous human microbiome studies that have sampled multiple sites have found that individuals are typically more similar to themselves than others in the study (cite).
-To see if this effect was driving the organization of the ordination, we connected communities from the same individual using gray lines on Figure 2C.
-While some individuals did have samples that clustered near each other (shorter lines), there were just as many individuals who's samples diverged from each other (longer lines).
-Similar to the diversity metrics, when examining community structure between sites and sides, there is no clear pattern between or within individuals.
 
@@
 Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
-This finding suggests that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
-((mention the other 2 comparisons?
-"among the four comparisons made..."))
+To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
+We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005).
+Thus interpersonal variation between subjects drives the differences between samples more than sample site or location.
+Overall, the results of the similarity analysis suggest that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
 
@@
 The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A).
-The constructed model used ((Xopt?
-need to follow up with this too)) features for the left and ((Xopt)) for the right.
-The models performed very well when classifying these samples (AUC X and Y).
+The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left.
+The models performed very well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively).
 The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
@@
 Next, we built a model to differentiate the left and right microbiome.
-The model performed best when distinguishing the left versus right mucosa (Figure 5A) compared to the left versus right lumen.
+The model performed best when distinguishing the right versus left mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen (AUC = 0.755).
 The model was able to explain ((X%)) of the variance.
 OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B).
-The model was less-effective at classifying the right and left lumen ((AUC)).
+The model was less-effective at classifying the right and left lumen (AUC = 0.755).
 However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).
@@
 #### Bacterial OTUs associated with cancer are found in healthy individuals 
-Some of the healthy patients have Fusobacterium OTUs.
-The most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_.
-It is the only OTU to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
-((cite joe and niels work somewhere in here, maybe in the discussion?)).
-The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample.
-((Table or something?)).
+Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs.
+Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_.
+It was the only sequence to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
+(@Castellarin2011).
+The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A).
 OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors.
-OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides.
+OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B).
 Samples that were positive for _P.
@@
 Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248.
-Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa.
+Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa (Figure 6C).
 Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.
-
-#### Exit stool comparisons model - specifics here or above?
 

### 2017-06-12 submission/manuscript.Rmd
refs almost done

@@
 Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014).
-These species include the oral pathogens _Fusobacterium nucleatum_, _Porphyromonas asacharolytica_ and _Parvimonas micra_.
+These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
 Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).
@@
 This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome.
-Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (cite).
-Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (cite).
-That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (cite).
+Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (@Hong2011).
+Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (@Stearns2011).
+That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014).
 Thus, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
@@
 There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite.
-An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005).
+An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (@Eckberg2005).
 We used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other.
 We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants.
-This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (Eckberg 2005, Aguierre de Carcer, 2015, Zhang et al 2013).
+This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (@Eckberg2005, @deCarcer2010, @Zhang2013).
 
@@
 For each comparison we identified top X OTUs that were strongly predictive of one site or another.
-Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (cite nature review).
-For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon (cite).
+Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (@Donaldson2015).
+For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon.
 The left mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_.
@@
 asacharolytica_ in 7 and 5 of our subjects, respectively.
-These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite).
+These bacteria have been shown to be predictive of colorectal cancer in humans (@Baxter2016) and have oncogenic properties in cell culture and in mice (@Sears2014).
 Interestingly, while _F.
@@
 Not much is known about the distribution of _P.
-asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (cite something, Albenberg, 2014).
+asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (@Albenberg2014).
 In studies examining bacteria on colorectal cancer tumors, _F.
 nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F.
-nucleatum_ decreases along the colon to rectum (cite Mima 2016).
+nucleatum_ decreases along the colon to rectum (@Mima2016).
 Of the (8) (40%) individuals positive for _F.
@@
 That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F.
-nucleatum_ at all (Mima 2016).
+nucleatum_ at all (@Mima2016).
 In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F.
-nucleatum_ in mucosal biopsies (Strauss 2011).
+nucleatum_ in mucosal biopsies (@Strauss2011).
 These studies have the added caveat of being sampled from prepared or cleaned colons.
@@
 Indeed, rather than the _F.
-nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016).
+nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (@Flynn2016).
 Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis.
+Finally, _Fusobacteria_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016).
+Thus, further analysis of the distribution and role of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.
 

### 2017-06-12 submission/manuscript.Rmd
refs done

@@
     keep_tex: yes
-csl: plos.csl
+csl: mbio.csl
 bibliography: references.bib 
@@
 Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies.
-A rare report of a matched-colonoscopy study that sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (cite).
-This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (Shobar 2016).
+A rare report of a matched-colonoscopy study that sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016).
+This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016).
 Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites.
@@
 Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors.
-Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone.
-Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (cite Niel fit paper).
-Our work already showed that the stool most accurately matched the samples from the left lumen.
-So we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
-But further studies would need to be done.
-Hard to get a complete set of matched samples from CRC patients though due to discomfort and insurance issues.
-One caveat of the previous study is that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.
-
-offer another model/schematic?
-what does this tell us about the broader field?
-do i need to tie in with the host any more here?
-(for instance, how microbes might interact with mutated genes/pathways).
-model, specific microbes on each side, change in mutation or metabolism affects change in community or precedes oncogenesis in a specific way?
+Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit and it performed much better than FIT alone.
+Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (@Baxter2016).
+One caveat of the FIT study was that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.
+Given that our results showed that the stool most accurately reflects the community of the left lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
 
@@
 The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
-Three (double check this) mucosal biopsies were collected and flash-frozen in RNAlater.
+Three mucosal biopsies were collected and flash-frozen in RNAlater.
 These samples comprised the distal or left colon samples.
@@
 For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples.
-The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich 2013).
-Sequences were curated using the mothur software as described previously (Schloss 2009).
-The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (wang 2007) and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
+The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich2013).
+Sequences were curated using the mothur software as described previously (@Schloss2009).
+The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (@Wang2007) and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
 Sequencing and analysis of a mock community revealed the error rate to be X%.
@@
 
-Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur (cite).
+Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur (@Schloss2009).
 ThetaYC distances were calculated to determine the dissimilarity between two samples.
-Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location (cite Liaw 2002).
+Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location (@Liaw2002).
 To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.
@@
 Location and abundance of cancer-associated OTUs.
-Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum_ (B) _Porphyromonas asacharolytica_ and (C) _Parvimonas micra_.
-
-###Supplemental material
-
-####Supplemental Figure 1
-Random Forest classifier comparing samples to exit stool. ROC curves for the RF model classifying exit stool versus the right or left lumen, pooled mucosa or pooled lumen are shown.
+Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum_ and (B) _Porphyromonas asacharolytica_
 

### 2017-06-12 submission/manuscript.Rmd
first draft for editing done

@@
   human colonic microbiota**'
+csl: mbio.csl
 fontsize: 11pt
-geometry: margin=1.0in
 output:
-  html_document: default
+  word_document: default
   pdf_document:
@@
     keep_tex: yes
-csl: mbio.csl
+  html_document: default
+geometry: margin=1.0in
 bibliography: references.bib
@@
 Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct.
-Previous work from our group found that microbial dysbiosis is associated with the development of colorectal cancer tumors in studies of both mice and humans.
+Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans.
 Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease.
@@
 This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
-16S rRNA gene sequencing was performed on proximal and distal mucosal biopsies, luminal and exit stool for 20 healthy individuals.
-Diversity analysis of each location revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
+16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and exit stool for 20 healthy individuals.
+Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
 Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites.
-Random Forest classification models were built using taxa abundance and sample location revealed distinct populations that were found in each location.
-Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas and Actinomyces were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925).
+Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location.
+_Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas_ and _Actinomyces_ were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925).
 The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755).
-The left mucosa was found to have high populations of Finegoldia, Murdochiella and Porphyromonas.
+The left mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_.
 Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end.
 Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon.
-Taken together, our results have identified distinct bacterial populations distinct of the proximal and distal colon.
+Taken together, our results have identified distinct bacterial populations of the proximal and distal colon.
 Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.
@@
 CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon.
-Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen.
-These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (@Yamauchi2012).
 CRC has recently been described to fall into four molecular subtypes (@Guinney2015).
@@
 #### Microbial membership and diversity of the proximal and distal colon 
-Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1).
+Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation(Figure 1).
 Participants also collected stool at home one week prior to the procedure.
@@
 That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on.
-Thus there was no clear pattern of changes in microbial diversity along the gut axis.
+Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.
 
@@
 #### Random Forest classification models identify important OTUs on each side
-To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances.
+To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances.
 The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A).
 The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left.
-The models performed very well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively).
+The models performed well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively).
 The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
@@
 Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs.
-Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_.
-It was the only sequence to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
+Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subsp animalis_.
+It was the only sequence to align with a _F.
+nucleatum_ strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
 (@Castellarin2011).
@@
 asacharolytica_ ranged in abundances from 0.01% - 16%.
-Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248.
-Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa (Figure 6C).
-Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.
+Thus, cancer-associated OTUs could be found in our sample set of 20 healthy individuals.
 
@@
 That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014).
-Thus, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
+Hence, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
 

### 2017-06-12 submission/manuscript.tex
first draft for editing done

@@
     \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
@@
 \usepackage[margin=1.0in]{geometry}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
+\usepackage{hyperref}
+\hypersetup{unicode=true,
             pdftitle={Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
+            pdfborder={0 0 0},
+            breaklinks=true}
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
 \setlength{\parindent}{0pt}
 \setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 \begin{document}
-
 \maketitle
@@
 
-Running title: Unique taxa differentiate proximal and distal human colon
-microbiota
+Running title: Specific taxa differentiate proximal and distal human
+colon microbiota
 
@@
 \newpage
+
 \linenumbers
@@
 Tumors of the proximal (right) and distal (left) colon are
-morpologically and genetically distinct. Previous work from our group
-found that microbial dysbiosis is associated with the development of
-colorectal cancer tumors in studies of both mice and humans. Analysis of
-the fecal microbiota from healthy and CRC patients further revealed
-different microbial signatures associated with disease. In this study,
-we extended our observations of the fecal microbiome to analysis of the
-proximal and distal human colon. We used a two-colonoscope approach on
-subjects that had not undergone standard bowel preparation procedure.
-This technique allowed us to characterize the native proximal and distal
-luminal and mucosal microbiome without prior chemical disruption. 16S
-rRNA gene sequencing was performed on proximal and distal mucosal
-biopsies, luminal and exit stool for 20 healthy individuals. Diversity
-analysis of each location revealed that each site contained a diverse
-community, and that a patient's samples were more similar to each other
-than to that of other individuals. Since we could not differentiate
-sites along the colon based on community structure or community
-membership alone, we employed the machine-learning algorithm Random
-Forest to identify key species that distinguish biogeographical sites.
-Random Forest classification models were built using taxa abundance and
-sample location revealed distinct populations that were found in each
-location. Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas
-and Actinomyces were most likely to be found in mucosal samples versus
-luminal samples (AUC = 0.925). The classification model performed well
-(AUC = 0.912) when classifying mucosal samples into proximal or distal
-sides, but separating luminal samples from each side proved more
+morphologically and genetically distinct. Previous work from our group
+found that microbial dysbiosis is associated with the development of CRC
+tumors in studies of both mice and humans. Analysis of the fecal
+microbiota from healthy and CRC patients further revealed different
+microbial signatures associated with disease. In this study, we extended
+our observations of the fecal microbiome to analysis of the proximal and
+distal human colon. We used a two-colonoscope approach on subjects that
+had not undergone standard bowel preparation procedure. This technique
+allowed us to characterize the native proximal and distal luminal and
+mucosal microbiome without prior chemical disruption. 16S rRNA gene
+sequencing was performed on proximal and distal mucosal and luminal
+biopsies and exit stool for 20 healthy individuals. Diversity analysis
+revealed that each site contained a diverse community, and that a
+patient's samples were more similar to each other than to that of other
+individuals. Since we could not differentiate sites along the colon
+based on community structure or community membership alone, we employed
+the machine-learning algorithm Random Forest to identify key species
+that distinguish biogeographical sites. Random Forest classification
+models were built using taxa abundance and sample location and revealed
+distinct populations that were found in each location.
+\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
+\emph{Actinomyces} were most likely to be found in mucosal samples
+versus luminal samples (AUC = 0.925). The classification model performed
+well (AUC = 0.912) when classifying mucosal samples into proximal or
+distal sides, but separating luminal samples from each side proved more
 challenging (AUC = 0.755). The left mucosa was found to have high
-populations of Finegoldia, Murdochiella and Porphyromonas. Proximal and
-distal luminal samples were comprised of many of the same taxa, likely
-reflecting the fact that stool moves along the colon from the proximal
-to distal end. Finally, comparison of all samples to fecal samples taken
-at exit uncovered that the feces were most similar to samples taken from
-the left lumen, again reflecting the anatomical structure of the colon.
-Taken together, our results have identified distinct bacterial
-populations distinct of the proximal and distal colon. Further
+populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
+Proximal and distal luminal samples were comprised of many of the same
+taxa, likely reflecting the fact that stool moves along the colon from
+the proximal to distal end. Finally, comparison of all samples to fecal
+samples taken at exit uncovered that the feces were most similar to
+samples taken from the left lumen, again reflecting the anatomical
+structure of the colon. Taken together, our results have identified
+distinct bacterial populations of the proximal and distal colon. Further
 investigation of these bacteria may elucidate if and how these groups
@@
 
-\begin{itemize}
-\item
-  colon cancer
-\item
-  right and left tumors are morphologically and genetically distinct
-  (why Q is important)
-\item
-  the microbiome of healthy people and patients with CRC is
-  distinguishable at the OTU level
-\item
-  specific keystone species are associated with CRC tumors
-\item
-  Central Question: do the populations of bacteria in the left and right
-  colon contribute to the development of specific subtypes of CRC?
-\item
-  nearly all studies of the gut microbiome have been from shed stool or
-  colonoscopy with bowel prep
-\item
-  bowel prep changes the mucosal/fecal microbiome and what can be
-  analyzed from it.
-\item
-  how does the microbiome vary spatially along the colon? we know that
-  the environment varies dramatically but we dont know how those
-  environmental factors specifically affect bacterial populations
-\item
-  What data is needed to answer this question? first, need to
-  characterize healthy species that are specific to each location. then
-  we can ask, if/how disruption affects tumorigenesis? so what is needed
-  is paired samples from healthy people that have not undergone bowel
-  prep, from the right/left and lumen/mucosa
-\item
-  what methods are used to get the data? unprepped colonoscopy
-\item
-  What analysis must be applied? - diversity analysis and machine
-  learning to pull out small variations/differences among the noisy data
-  set.
-\item
-  hypothesis: there are distinct bacteria on the left and right side
-  that can be distinguished by 16S sequencing of stool
-\end{itemize}
+Colorectal cancer (CRC) is the second-leading cause of cancer-related
+deaths in the United States. CRC tumors vary in structure, size,
+morbidity and symptomology depending upon their geographical location in
+the colon. CRC has recently been described to fall into four molecular
+subtypes ((1)). Of these subtypes, tumors that arise on the left side of
+the colon are infiltrating lesions that present with painful symptoms.
+In contrast, right-sided colon tumors are bulky and project into the
+lumen, often causing no symptoms until they have developed into
+carcinomas ((2)). Because of the absence of symptoms, right-sided
+tumors, which comprise up to 47\% of CRCs, have a significantly higher
+mortality rate ((2)). The left and right sides of the colon differ in
+the amount of inflammation present and the genomic instability of
+precancerous cells, respectively, as well as oxygen, pH and the presence
+of antimicrobial peptides ((3), (4), (5)). Microenvironments differ not
+only longitutinally along the colon, but latitudinally from the
+epithelium to mucosa to intestinal lumen, offering several sites for
+different microbial communities to flourish. Given this varied
+physiology of the proximal-distal axis of the colon, symbiotic microbes
+and their metabolites likely vary between sites.
+
+Several recent findings have shown that development and progression of
+CRC can be attributed to specific molecular events as a result of
+interactions between the gut microbiota and human host ((3)). Our group
+and others have found that the stool microbiome of patients with CRC is
+distinct from that of healthy people ((6)). Further studies manipulating
+the gut microbiome using antibiotics or other chemoagents in mice has
+shown that dysbiosis preceded and accelerated the development of CRC
+tumors ((7)). Comparison of the specific bacteria present on CRC tumors
+with those found on nearby healthy tissue has also identified specific
+bacterial species that are tumor-associated based on FISH staining and
+16S rRNA sequencing ((8)). These species include the oral pathogens
+\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
+Interestingly, the periodontal pathogens are also highly predictive of
+if a patient has CRC tumors or not in our classification studies ((9)).
+However, these studies have been performed comparing shed stool or in
+mice, which does not allow for analysis of paried samples from the
+proximal and distal sides of the colon. Similarly, comparisons of on- or
+off-tumor bacteria rarely have matched tissue from the other side of the
+colon from the same patient, limiting what conclusions can be drawn
+about the colonic microbiome overall, let alone at that specific site.
+Disruption of the healthy community could promote the initiation or
+proliferation of the distinct left and right CRC tumors. However, the
+contribution of the gut microbiota to these subtypes is largely
+undefined.
+
+Furthermore, profiles of the microbial biogeography of the gut have been
+limited by sample collection methods. The majority of human gut
+microbiome studies have been performed on whole shed stool or on samples
+collected during colonoscopy procedures. While the latter method allows
+investigators to acquire samples from inside the human colon, typically
+this procedure is preceded by the use of bowel preparation methods such
+as the consumption of laxatives to cleanse the bowel. Bowel preparation
+is essential for detecting cancerous or precancerous lesions in the
+colon, but complicates microbiome profiling as the chemicals strip the
+bowel of contents and disrupt the mucosal layer ((10), (11)). As such,
+what little information we do have about the biogeographical
+distribution of the microbes in the proximal and distal colon is
+confounded by the bowel preparation procedure.
+
+In this study we sought to combine these approaches to characterize the
+microbiome in the lumen and mucosa of the proximal and distal healthy
+human colon. Our design uses unprepared colonoscopy techniques to sample
+each location of the gut without prior disruption of the native flora in
+20 healthy volunteers. To get around the noise created by a diverse set
+of human microbiomes, we used a machine-learning classification
+algorithm trained on curated 16S rRNA sequencing reads to identify
+microbes specific to each location. We found that our classification
+models were able to separate mucosal and lumenal samples as well as
+differentiate between sides of the colon based on populations of
+specific microbes. By identifying the specific microbes we are poised to
+ask if and how the presence or disruption of the microbes at each site
+contribute to the development of the specific tumor subtypes of CRC in
+the proximal and distal human colon.
 
@@
 
-\begin{itemize}
-\item
-  what data were obtained? 16S sequencing results for samples taken from
-  healthy human colons that had not undergone bowel prep Figure 1:
-  sample collection schematic
-\item
-  what were the results of the analyses? the results of the analyses are
-  as follows:
-\end{itemize}
+\paragraph{Microbial membership and diversity of the proximal and distal
+colon}\label{microbial-membership-and-diversity-of-the-proximal-and-distal-colon}
 
-\paragraph{Descriptive analysis}\label{descriptive-analysis}
+Proximal and distal lumenal and mucosal samples were collected from 20
+healthy humans that had not undergone bowel preparation(Figure 1).
+Participants also collected stool at home one week prior to the
+procedure. To characterize the bacterial communities present at these
+sites, 16S rRNA gene sequencing was performed on extracted DNA from each
+sample. The relative abundance of each phyla for each site is shown in
+Figure 2A. This data combines the samples from all participants in the
+study. Each site is primarily dominated by \emph{Firmicutes} and
+\emph{Bacteriodetes}, consistent with known variability in human
+microbiome research. Likewise, samples had varying levels of diversity
+at each site, irrespective of the individual (Figure 2B). That is, while
+some individuals had a more diverse right mucosa, some had a more
+diverse left mucosa, and so on. Therefore we could not identify a clear
+pattern of changes in microbial diversity along the gut axis.
 
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  there are similar levels (within error margins) of each phyla for each
-  location, but there is noise due to interpatient variability. Overall,
-  there is also a similar level of diversity in each location. However,
-  each patient has a different pattern of diversity over the locations,
-  some are high in R, some are high in L, etc. Figure 2:
-\item
-  2A: relative abundance by phyla
-\item
-  2B: simpson diversity
-\item
-  2C: NMDS/ordination of some kind
-\end{itemize}
+To compare similarity between sides (left or right) or sites (lumen or
+mucosa), we calculated theta YC distances from OTU abundances and
+compared distances for all individuals. Again, across all patient
+samples we saw a range of theta YC distances when comparing sample
+locations (Figure 3A) and again those ranges did not follow a clear
+pattern on an individual basis. However, when comparing median distances
+between the right lumen and mucosa, the left versus right lumen, the
+left versus right mucosa and the left lumen and mucosa, we found that
+the right lumen and mucosa were most similar to each other than the
+other samples (P \textless{} 0.005, Wilcoxon, BH adjustment). Next, we
+calculated thetaYC distances to examine how each sample compared to the
+home-collected exit stool. Amidst variability between patients, we did
+identify significantly smaller thetaYc distance between the left lumenal
+sample and the exit stool (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
+adjustment). Further, there was an even larger difference in the
+comparisons of the left mucosa to the exit stool, indicating that the
+mucosa is quite different from the stool as compared to lumen (P
+\textless{} 0.0005, Wilcoxon, BH adjustment). To determine what factors
+may be driving the differences seen among the samples, we compared
+thetaYC distances between samples from all subjects (interpersonal)
+versus samples from within one subject (intrapersonal). We found that
+samples from one individual were overall much more similar to each other
+than to other study subjects (Figure 3C), consistent with previous human
+microbiome studies that have sampled multiple sites of the human colon
+(\textbf{???}). Thus interpersonal variation between subjects drives the
+differences between samples more than sample site or location. Overall,
+the results of the similarity analysis suggest that the contents of the
+left lumen are most representative of stool at exit, and the microbes
+remaining on the mucosa are much different.
+
+\paragraph{Random Forest classification models identify important OTUs
+on each
+side}\label{random-forest-classification-models-identify-important-otus-on-each-side}
+
+To identify OTUs that were distinct at each biogeographical site, we
+constructed several Random Forest machine-learning models built on OTU
+abundances. The first model we built to classify the lumen versus
+mucosal samples for the right and left sides, independently (Figure 4A).
+The constructed model used ((Xopt)) features for the right and ((Xopt))
+for the left. The models performed well when classifying these samples
+(AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most
+predictive of each site are identified by their greatest mean decrease
+in accuracy when removed from the model. For distinguishing the right
+lumen and mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
+\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
+were differentially abundant (Figure 4B). The model classifying the left
+lumen and mucosa identified OTUs from \emph{Turicibacter},
+\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
+\emph{Anaerococcus} genera that could distinguish lumen from mucosal
+samples (Figure 4C). These results indicate that there are fine
+differences between the different sites of the colon, and that these can
+be traced down to specific OTUs on each side.
+
+Next, we built a model to differentiate the left and right microbiome.
+The model performed best when distinguishing the right versus left
+mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen
+(AUC = 0.755). The model was able to explain ((X\%)) of the variance.
+OTUs that were differentially abundant between the left and right mucosa
+included members of the \emph{Porphyromonas}, \emph{Murdochiella},
+\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} genera
+(Figure 5B). The model was less-effective at classifying the right and
+left lumen (AUC = 0.755). However it did identify differentially
+abundant OTUs such as three members of the \emph{Bacteroides} genus, a
+\emph{Clostridium IV} OTU and an \emph{Oscillibacter} OTU (Figure 5C).
+This analysis found that some of the same OTUs that are distinct between
+the mucosa and lumen also drive sidedness- such as \emph{Anaerococcus}
+and \emph{Finegoldia}.
+
+\paragraph{Bacterial OTUs associated with cancer are found in healthy
+individuals}\label{bacterial-otus-associated-with-cancer-are-found-in-healthy-individuals}
+
+Given that specific bacterial species have been associated with
+colorectal cancer we probed our sample set for these OTUs. Among our 100
+samples, the most frequent sequence associated wtih the
+\emph{Fusobacterium} genus was OTU179, which BLASTs to
+\emph{Fusobacterium nucleatum subsp animalis}. It was the only sequence
+to align with a \emph{F. nucleatum} strain- the only species of
+\emph{Fusobacterium} known to have oncogenic properties and be found on
+the surfaces of colorectal cancer tumors. ((12)). The
+\emph{Fusobacterium} positive samples were located on the right and left
+mucosa and ranged up to 1\% of the sample (Figure 6A). OTU00152 is
+\emph{Porphyromonas} and the most frequent sequence in that OTU aligns
+to \emph{Porphyromonas asacharolytica}, another bacterium commonly
+detected and isolated from colorectal tumors. OTU152 was only detected
+on the left mucosa, and in fact was one of the OTUs the classification
+model identified as separating left and right sides (Figure 6B). Samples
+that were positive for \emph{P. asacharolytica} ranged in abundances
+from 0.01\% - 16\%. Thus, cancer-associated OTUs could be found in our
+sample set of 20 healthy individuals.
 
@@
 
--how did the analyses answer the central question?
+Here we identified bacterial taxa that are specific to the lumen and
+mucosa of the right and left human colon from samples collected during
+unprepared colonoscopy. We found that all locations contained a range of
+phyla and a range of diversity, but that there was a wide variability
+between subjects. Pairwise comparisons of each of the sites revealed
+that the right mucosa and lumen were most similar to each other.
+Further, comparison of colonoscopy-collected samples with samples
+collected from stool at home showed that the left lumen is most similar
+to stool at exit. Random Forest algorithms built on OTU abundances from
+each sample identified microbes that are particular to each location of
+the colon. Finally, we were able to detect some bacterial OTUs
+associated with colorectal cancer in our healthy patient cohort. Using
+unprepped colonoscopy and machine learning, we have identified bacterial
+phyla specific to the healthy proximal and distal human colon.
 
--what does this answer tell us about the broader field?
+When examining the relative abundance of the different phyla at each
+site, there is a wide amount of variation for each phyla with
+communities primarily dominated by the Bacteriodes and Firmicutes. This
+likely reflects not only the variability between human subjects, casued
+by differences in age, gender, diet, but also reports of microbial
+``patchiness'' in the gut microbiome. Several previous studies have
+noted that the bacteria recoverable from the same mucosal sample
+location can be vastly different when the samples are taken just 1 cm
+away from each other ((13)). Similar patchiness is also observed in
+lumenal contents and fecal samples themselves; there is observed
+separation of different interacting microbes along the length of a stool
+sample, for instance ((14)). That said, across our samples the mucosal
+samples harbor more Proteobacteria, consistent with previous studies
+comparing mucosal swabs to lumenal content in humans ((4)). Hence, the
+conclusions we can draw from phyla analysis are likely confounded by
+differences in sampling and patchiness between subjects.
+
+We identified a range of Simpson diversity values between patients and
+between each anatomical site. As this measurement takes both number of
+species and abundance of species into account, it seems that there is a
+varied level of each component along the human colon. There was no clear
+pattern of diversity when comparing patients- some patients had a higher
+level of diversity in the proximal than the distal colon, some patients
+had the opposite. An early study of microbial diversity and membership
+of three human subjects found a varied level of diversity in each
+subject, but that subjects were overall more similar to themselves than
+to each other (\textbf{???}). We used the thetaYC dissimilarity metric
+to compare the communities present at each anatomical location to each
+other. We found that while there was a wide range of similarity of each
+site within a person, the study participants were ultimately more
+similar to themselves than all of the other participants. This finding
+is consistent with numerous studies examining the biogeography of the
+colon in healthy humans (\textbf{???}, (15), (16)).
+
+To get around the noisiness created by a diverse set of samples, we
+built a Random Forest model to identify microbes specific to each side.
+For each comparison we identified top X OTUs that were strongly
+predictive of one site or another. Generally, OTUs identified in each
+location were consistent with known physiological gradients along the
+gut axis ((5)). For instance, the right mucosa harbored
+mucosa-associated facultative anaerobes such as \emph{Actinomyces} and
+\emph{Enterobacteraceae} and the aerobic \emph{Psuedomonas} consistent
+with the highest oxygen regions of the colon. The left mucosa was far
+more likely to host strictly anaerobic species such as
+\emph{Porphyromonas}, \emph{Anaerococcus}, \emph{Finegoldia} and
+\emph{Peptoniphilus}. The model was less effective at classifying the
+right and left lumenal contents, probably because the samples are
+arguably composed of the same material.
+
+We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 7 and 5
+of our subjects, respectively. These bacteria have been shown to be
+predictive of colorectal cancer in humans ((9)) and have oncogenic
+properties in cell culture and in mice ((17)). Interestingly, while
+\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
+asacharolytica} was only detected in the left mucosa. Not much is known
+about the distribution of \emph{P. asacharolytica} but given its
+documented anaerobic characteristics and asacharolytic metabolism, it
+might not be surprising that it resides in the less-oygen-rich and
+proteinaceous left mucosa ((4)). In studies examining bacteria on
+colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
+on right-sided tumors, and distribution of \emph{F. nucleatum} decreases
+along the colon to rectum ((18)). Of the (8) (40\%) individuals positive
+for \emph{F. nucleatum} in this study, the bacteria was spread across
+the right mucosa, left lumen and left mucosa. That said, in a previous
+study of 1,102 colorectal cancer cases, only 138 (11\%) of them
+contained detectable \emph{F. nucleatum} at all ((18)). In a separate
+study of patients with inflammatory bowel disease (IBD), 50\% of 22 IBD
+patients and 17.6\% of 34 people without disease contained \emph{F.
+nucleatum} in mucosal biopsies ((19)). These studies have the added
+caveat of being sampled from prepared or cleaned colons. Thus, the
+presence of \emph{F. nucleatum} in a healthy individual is not
+necessarily linked to the development of future colorectal cancers.
+Because of the spatial distribution of the \emph{F. nucleatum} in our
+sample set, we cannot develop a model for the role of \emph{F.
+nucleatum} in the healthy colon. Indeed, rather than the \emph{F.
+nucleatum} initiating tumorigenesis, these bacteria may instead arrive
+``late'' to the tumor microenvironment and feed on tumors that have
+alread begun to develop ((20)). Data examining bacterial biofilms on CRC
+tumors suggests that \emph{Fusobacteria} species are more commonly found
+both on proximal tumors and in biofilms, indicating that it is not only
+the presence of the bacteria but the organization of the tumor community
+that contributes to \emph{Fusobacterium's} role in tumorigenesis.
+Finally, \emph{Fusobacteria} and \emph{Porphyromonas} species have been
+known to not only co-occur on CRC tumors but also to synergistically
+promote tumorigenesis in an oral cancer model ((21), (20)). Thus,
+further analysis of the distribution and role of these pathogens may
+elucidate a mechanism for development of CRC tumors in the proximal or
+distal colon.
+
+Specific comparisons of our findings to previously published gut
+biogeography studies are additionally confounded by the use of bowel
+preparation methods in most other studies. A rare report of a
+matched-colonoscopy study that sampled 18 patient's colonic mucosa and
+lumenal contents prior to and after bowel cleansing ((22)). This group
+found that mucosa and lumenal samples were distinguishable prior to
+bowel cleansing, but that bowel preparation resulted in an increase in
+shared OTUs between each site ((22)). Bowel cleansing not only made the
+samples harder to distinguish, it resulted in decreases in diversity
+across sites. Further, the differences were not great enough to overcome
+interpersonal differences between subjects. So, bowel preparation
+clearly induces bias into the microbes recovered from sampling the lumen
+or mucosa of a prepared bowel. Thus our findings of specific bacteria at
+each site of the colon are strengthened by the lack of bowel
+preparation.
+
+Microbiome-based diagnostics are increasingly being explored as
+non-invasive tools to survey for the development of colon cancer. Random
+Forest models have been used by our group and others to increase the
+detection sensitivity of CRC tumors. Indeed, our group found that a
+classification model that used microbiome data in combination with Fecal
+Immunohistochemical Test (FIT) results could correctly identify both
+carcinoma and adenoma lesions from communities of stool at exit and it
+performed much better than FIT alone. Further work from our lab has
+shown that microbiome profiling of the FIT cartridge contents
+sufficiently represented the stool community ((9)). One caveat of the
+FIT study was that there was not sufficient information to test if a
+classification model could differentiate between proximal and distal CRC
+tumors based on exit stool sample alone, but we would hypothesize that
+would not be effective. Given that our results showed that the stool
+most accurately reflects the community of the left lumen, we likely
+cannot use Random Forest of stool samples to diagnose any changes in the
+proximal or mucosal communities.
+
+By revealing specific differences in microbial populations at each
+location in the gut via sampling an unprepared bowel, we can begin to
+form hypothesies about how specific host-microbe interactions can affect
+oncogenesis of proximal and distal CRC tumors. To this point, 16S
+community profiling studies do not provide enough information to probe
+these questions. Our sample set of matched proximal, distal, lumenal and
+mucosal samples from colons that have not undergone bowel preparation
+presents a unique opportunity to explore further questions about the
+microbiome along the gut axis. Specifically, examining metagenomic,
+metabolomic and host interactions at each site will be useful in further
+characterizing the host-microbe interactions in the development of
+proximal and distal colorectal cancer.
 
@@
 
+We thank all the individuals who volunteered for the study. This work
+was supported by the Rose and Lawrence C. Page Foundation (DKT). We
+would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman
+for their roles in study coordination, subject recruitment, sample
+collection and sample processing.
+
 \subsubsection{Methods}\label{methods}
 
-Study participants/IRB
+\paragraph{Human subjects}\label{human-subjects}
 
-Sample collection
+The procedures in this study and consent were approved by the
+Institutional Review Board at the University of Michigan Health System
+with protocol number XXXX. Subjects were recruited using the online
+recruitment platform and were pre-screened prior to enrollment in the
+study. Exclusion criteria included: use of asprin or NSAIDs within 7
+days, use of antibiotics within 3 months, current use of anticoagulants,
+known allergies to Fentanyl or Benadryl, prior history of colon disease,
+diabetes, abdominal surgery, respiratory, liver, kidney or brain
+impairments, undergoing current chemotherapy or radiation treatment and
+subjects that were pregnant or trying to conceive. 20 subjects that met
+the criteria were selected and provided signed informed consent prior to
+the procedure. There were 13 female and 7 male subjects ranging in age
+from 25 to 64.
 
-Sample processing/DNA extraction
+\paragraph{Sample collection}\label{sample-collection}
 
-16S rRNA sequencing
+At a baseline visit, subjects were consented and given a home collection
+stool kit (Source of kit supplies). At least one week prior to the
+scheduled colonoscopy, subjects were to collect whole stool at home and
+two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.)
+and ship the samples to the University. Notably, subjects did not
+undergo any bowel preparation method prior to sampling. On procedure
+day, subjects reported to the Michigan Clinical Research Unit at the
+University of Michigan Health System. Patients were consciously sedated
+using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
+sigmoidoscope was first inserted about 25cm into the colon and endoscopy
+brush used to collect lumenal/stool contents. Two lumenal samples were
+collected and the contents immediately deposited into RNAlater (source)
+and flash-frozen in liquid nitrogen. The brushes were withdrawn and
+biopsy forceps were used to collect mucosal biopsies on sections of the
+colon that were pink and free of stool matter. Three mucosal biopsies
+were collected and flash-frozen in RNAlater. These samples comprised the
+distal or left colon samples. The sigmoidoscope was then withdrawn and a
+pediatric colonoscope was inserted to reach the ascending colon. Samples
+were then collected as in the distal colon and the colonoscope
+withdrawn. All samples were stored at -80 C until study completion.
 
-Statistical analysis/Random Forest
+\paragraph{Sample processing, sequencing and
+analysis}\label{sample-processing-sequencing-and-analysis}
 
-\subsubsection{Figure Legends}\label{figure-legends}
+DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
+Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
+solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
+added to the bead beating step to enhance DNA recovery from mucosal
+samples. The resulting DNA was used as template for amplification of the
+V4 region of the 16S rRNA gene and fragments were sequenced on an
+Illumina MiSeq as previously described ((23)). Sequences were curated
+using the mothur software as described previously ((24)). The sequences
+were assigned taxonomic classification using a naive Bayesian classifier
+trained using a 16S rRNA gene training set from the Ribosomal Database
+Project (RDP) ((25)) and clustered into operational taxonomic units
+(OTUs) based on a 97\% similarity cutoff. Sequencing and analysis of a
+mock community revealed the error rate to be X\%. Samples were rarefied
+to 4231 sequences per sample in order to reduce sampling bias.
+
+Diversity analysis was performed using the Simpson diversity calculator
+and theta YC calculator metrics in mothur ((24)). ThetaYC distances were
+calculated to determine the dissimilarity between two samples. Random
+Forest classification models were built using the randomForest R package
+and resultant models were used to identify the OTUs that were most
+important for classifying each location ((26)). To get species-level
+information about sequences of interest, sequences were aligned using
+blastn and the species name was only used if the identity score was
+\textgreater{}= 99\%.
+
+\paragraph{Statistical analysis}\label{statistical-analysis}
+
+Differences in community membership at the phyla level were tested using
+the analysis of molecular variance (AMOVA) metric in mothur. Differences
+in thetaYC distances by location were tested using the Wilcoxon rank-sum
+test adjusted for multiple comparisons using the Benjamini-Hochberg
+procedure.
+
+\subsubsection{Figure legends}\label{figure-legends}
+
+\paragraph{Figure 1}\label{figure-1}
+
+Sampling strategy. A flexible sigmoidoscope was used to sample the
+distal colonic luminal contents and mucosa. The scope was inserted
+\textasciitilde{} 25cm into the subject and endoscopy brushes were used
+to sample the luminal contents (green star). A separate set of biopsy
+forceps was used to sample the distal mucosa (blue star). The
+sigmoidoscope was removed. A pediatric colonoscope was inserted and used
+to access the proximal colon. Biopsies were taken of the proximal
+luminal contents and mucosa as described. One week prior to the
+procedure stool was collected at home and sent into the laboratory.
+Representative images from one individual are shown.
+
+\paragraph{Figure 2}\label{figure-2}
+
+Microbial membership and diversity of the proximal and distal human
+colon. A) Relative abundance of the top five bacterial phyla in each
+sampling site. Each box represents the median and confidence intervals.
+B) Simpson diversity of the microbial communities at each location. The
+lines represent the median values.
+
+\paragraph{Figure 3}\label{figure-3}
+
+Similarity of microbial community structure between sites of the gut.
+ThetaYC distances are shown for interpersonal similarities between two
+sites -- each point represents one individual. In (A), comparisons of
+the right and left mucosal and lumen are shown. In (B), comparisons of
+each site to the exit stool are shown.
+
+\paragraph{Figure 4}\label{figure-4}
+
+Random Forest classifies the mucosa and lumen of each side of the colon.
+A) Receiver Operator Characteristic curves are shown for the 10-fold
+cross validation of the Random Forest model classifying lumen and
+mucosal samples for the left and right sides of the colon. (B) Top five
+OTUs that are most important for the classification model for the left
+mucosa and lumen (B) and the right mucosa and lumen (C).
+
+\paragraph{Figure 5}\label{figure-5}
+
+Random Forest classifies the left and right sides of the colon. A)
+Receiver Operator Characteristic curves are shown for the 10-fold cross
+validation of the Random Forest model classifying left lumen versus
+right lumen (orange) and left mucosa vs right mucosa (green). (B) Top
+five OTUs that are most important for the classification model for the
+left and right mucosa (B) and the left and right lumen (C).
+
+\paragraph{Figure 6}\label{figure-6}
+
+Location and abundance of cancer-associated OTUs. Relative abundance was
+calculated and plotted by sample site for each OTU of interest: (A)
+\emph{Fusobacterium nucleatum} and (B) \emph{Porphyromonas
+asacharolytica}
+
+\subsubsection*{References}\label{references}
+\addcontentsline{toc}{subsubsection}{References}
+
+\hypertarget{refs}{}
+\hypertarget{ref-Guinney2015}{}
+1. \textbf{Guinney J}, \textbf{Dienstmann R}, \textbf{Wang X},
+\textbf{Reyniès A de}, \textbf{Schlicker A}, \textbf{Soneson C},
+\textbf{Marisa L}, \textbf{Roepman P}, \textbf{Nyamundanda G},
+\textbf{Angelino P}, \textbf{Bot BM}, \textbf{Morris JS}, \textbf{Simon
+IM}, \textbf{Gerster S}, \textbf{Fessler E}, \textbf{Melo FDSE},
+\textbf{Missiaglia E}, \textbf{Ramay H}, \textbf{Barras D},
+\textbf{Homicsko K}, \textbf{Maru D}, \textbf{Manyam GC}, \textbf{Broom
+B}, \textbf{Boige V}, \textbf{Perez-Villamil B}, \textbf{Laderas T},
+\textbf{Salazar R}, \textbf{Gray JW}, \textbf{Hanahan D},
+\textbf{Tabernero J}, \textbf{Bernards R}, \textbf{Friend SH},
+\textbf{Laurent-Puig P}, \textbf{Medema JP}, \textbf{Sadanandam A},
+\textbf{Wessels L}, \textbf{Delorenzi M}, \textbf{Kopetz S},
+\textbf{Vermeulen L}, \textbf{Tejpar S}. 2015. The consensus molecular
+subtypes of colorectal cancer. Nature Medicine \textbf{21}:1350--1356.
+doi:\href{https://doi.org/10.1038/nm.3967}{10.1038/nm.3967}.
+
+\hypertarget{ref-Benedix2010}{}
+2. \textbf{Benedix F}, \textbf{Kube R}, \textbf{Meyer F},
+\textbf{Schmidt U}, \textbf{Gastinger I}, \textbf{Lippert H}. 2010.
+Comparison of 17, 641 patients with right- and left-sided colon cancer:
+Differences in epidemiology, perioperative course, histology, and
+survival. Diseases of the Colon \& Rectum \textbf{53}:57--64.
+doi:\href{https://doi.org/10.1007/dcr.0b013e3181c703a4}{10.1007/dcr.0b013e3181c703a4}.
+
+\hypertarget{ref-Yamauchi2012}{}
+3. \textbf{Yamauchi M}, \textbf{Lochhead P}, \textbf{Morikawa T},
+\textbf{Huttenhower C}, \textbf{Chan AT}, \textbf{Giovannucci E},
+\textbf{Fuchs C}, \textbf{Ogino S}. 2012. Colorectal cancer: A tale of
+two sides or a continuum?: Figure 1. Gut \textbf{61}:794--797.
+doi:\href{https://doi.org/10.1136/gutjnl-2012-302014}{10.1136/gutjnl-2012-302014}.
+
+\hypertarget{ref-Albenberg2014}{}
+4. \textbf{Albenberg L}, \textbf{Esipova TV}, \textbf{Judge CP},
+\textbf{Bittinger K}, \textbf{Chen J}, \textbf{Laughlin A},
+\textbf{Grunberg S}, \textbf{Baldassano RN}, \textbf{Lewis JD},
+\textbf{Li H}, \textbf{Thom SR}, \textbf{Bushman FD}, \textbf{Vinogradov
+SA}, \textbf{Wu GD}. 2014. Correlation between intraluminal oxygen
+gradient and radial partitioning of intestinal microbiota.
+Gastroenterology \textbf{147}:1055--1063.e8.
+doi:\href{https://doi.org/10.1053/j.gastro.2014.07.020}{10.1053/j.gastro.2014.07.020}.
+
+\hypertarget{ref-Donaldson2015}{}
+5. \textbf{Donaldson GP}, \textbf{Lee SM}, \textbf{Mazmanian SK}. 2015.
+Gut biogeography of the bacterial microbiota. Nature Reviews
+Microbiology \textbf{14}:20--32.
+doi:\href{https://doi.org/10.1038/nrmicro3552}{10.1038/nrmicro3552}.
+
+\hypertarget{ref-Zackular2014}{}
+6. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prevention Research \textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.capr-14-0129}{10.1158/1940-6207.capr-14-0129}.
+
+\hypertarget{ref-Zackular2013}{}
+7. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
+\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
+tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
+doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.
+
+\hypertarget{ref-Dejea2014}{}
+8. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
+\textbf{White JR}, \textbf{Welch JLM}, \textbf{Rossetti BJ},
+\textbf{Peterson SN}, \textbf{Snesrud EC}, \textbf{Borisy GG},
+\textbf{Lazarev M}, \textbf{Stein E}, \textbf{Vadivelu J},
+\textbf{Roslani AC}, \textbf{Malik AA}, \textbf{Wanyiri JW}, \textbf{Goh
+KL}, \textbf{Thevambiga I}, \textbf{Fu K}, \textbf{Wan F}, \textbf{Llosa
+N}, \textbf{Housseau F}, \textbf{Romans K}, \textbf{Wu X},
+\textbf{McAllister FM}, \textbf{Wu S}, \textbf{Vogelstein B},
+\textbf{Kinzler KW}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2014.
+Microbiota organization is a distinct feature of proximal colorectal
+cancers. Proceedings of the National Academy of Sciences
+\textbf{111}:18321--18326.
+doi:\href{https://doi.org/10.1073/pnas.1406199111}{10.1073/pnas.1406199111}.
+
+\hypertarget{ref-Baxter2016}{}
+9. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\hypertarget{ref-Jalanka2014}{}
+10. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
+\textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
+\textbf{Gowland P}, \textbf{Hoad C}, \textbf{Garsed K}, \textbf{Lam C},
+\textbf{Palva A}, \textbf{Spiller RC}, \textbf{Vos WM de}. 2014. Effects
+of bowel cleansing on the intestinal microbiota. Gut
+\textbf{64}:1562--1568.
+doi:\href{https://doi.org/10.1136/gutjnl-2014-307240}{10.1136/gutjnl-2014-307240}.
+
+\hypertarget{ref-Harrell2012}{}
+11. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
+\textbf{Young V}, \textbf{Lichtenstein L}, \textbf{Huang Y},
+\textbf{Hanauer S}, \textbf{Chang E}. 2012. Standard colonic lavage
+alters the natural state of mucosal-associated microbiota in the human
+colon. PLoS ONE \textbf{7}:e32545.
+doi:\href{https://doi.org/10.1371/journal.pone.0032545}{10.1371/journal.pone.0032545}.
+
+\hypertarget{ref-Castellarin2011}{}
+12. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
+\textbf{Dreolini L}, \textbf{Krzywinski M}, \textbf{Strauss J},
+\textbf{Barnes R}, \textbf{Watson P}, \textbf{Allen-Vercoe E},
+\textbf{Moore RA}, \textbf{Holt RA}. 2011. Fusobacterium nucleatum
+infection is prevalent in human colorectal carcinoma. Genome Research
+\textbf{22}:299--306.
+doi:\href{https://doi.org/10.1101/gr.126516.111}{10.1101/gr.126516.111}.
+
+\hypertarget{ref-Hong2011}{}
+13. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
+\textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
+analysis of the mucosal microbiota in healthy individuals reveals
+ubiquitous bacterial groups and micro-heterogeneity. PLoS ONE
+\textbf{6}:e25042.
+doi:\href{https://doi.org/10.1371/journal.pone.0025042}{10.1371/journal.pone.0025042}.
+
+\hypertarget{ref-Stearns2011}{}
+14. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
+\textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
+\textbf{Croitoru K}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}.
+2011. Bacterial biogeography of the human digestive tract. Scientific
+Reports \textbf{1}.
+doi:\href{https://doi.org/10.1038/srep00170}{10.1038/srep00170}.
+
+\hypertarget{ref-deCarcer2010}{}
+15. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
+\textbf{Kang S}, \textbf{Worthley D}, \textbf{Whitehall V},
+\textbf{Gordon I}, \textbf{McSweeney C}, \textbf{Leggett B},
+\textbf{Morrison M}. 2010. Numerical ecology validates a biogeographical
+distribution and gender-based effect on mucosa-associated bacteria along
+the human colon. The ISME Journal \textbf{5}:801--809.
+doi:\href{https://doi.org/10.1038/ismej.2010.177}{10.1038/ismej.2010.177}.
+
+\hypertarget{ref-Zhang2013}{}
+16. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
+\textbf{Xu J}, \textbf{Wen X}, \textbf{Ma Z (Sam)}, \textbf{Shi P}.
+2013. Spatial heterogeneity and co-occurrence patterns of human
+mucosal-associated intestinal microbiota. The ISME Journal
+\textbf{8}:881--893.
+doi:\href{https://doi.org/10.1038/ismej.2013.185}{10.1038/ismej.2013.185}.
+
+\hypertarget{ref-Sears2014}{}
+17. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
+and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
+doi:\href{https://doi.org/10.1016/j.chom.2014.02.007}{10.1016/j.chom.2014.02.007}.
+
+\hypertarget{ref-Mima2016}{}
+18. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
+\textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
+\textbf{Silva A da}, \textbf{Gu M}, \textbf{Li W}, \textbf{Hamada T},
+\textbf{Kosumi K}, \textbf{Hanyuda A}, \textbf{Liu L}, \textbf{Kostic
+AD}, \textbf{Giannakis M}, \textbf{Bullman S}, \textbf{Brennan CA},
+\textbf{Milner DA}, \textbf{Baba H}, \textbf{Garraway LA},
+\textbf{Meyerhardt JA}, \textbf{Garrett WS}, \textbf{Huttenhower C},
+\textbf{Meyerson M}, \textbf{Giovannucci EL}, \textbf{Fuchs CS},
+\textbf{Nishihara R}, \textbf{Ogino S}. 2016. Fusobacterium nucleatum in
+colorectal carcinoma tissue according to tumor location. Clinical and
+Translational Gastroenterology \textbf{7}:e200.
+doi:\href{https://doi.org/10.1038/ctg.2016.53}{10.1038/ctg.2016.53}.
+
+\hypertarget{ref-Strauss2011}{}
+19. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
+\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
+\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
+gut mucosa-derived fusobacterium nucleatum positively correlates with
+IBD status of the host. Inflammatory Bowel Diseases
+\textbf{17}:1971--1978.
+doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.
+
+\hypertarget{ref-Flynn2016}{}
+20. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
+Metabolic and community synergy of oral bacteria in colorectal cancer.
+mSphere \textbf{1}:e00102--16.
+doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
+
+\hypertarget{ref-Whitmore2014}{}
+21. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
+cancer. PLoS Pathogens \textbf{10}:e1003933.
+doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.
+
+\hypertarget{ref-Shobar2016}{}
+22. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
+\textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
+\textbf{Losurdo J}, \textbf{Engen PA}, \textbf{Sun Y}, \textbf{Koenig
+L}, \textbf{Mutlu EA}. 2016. The effects of bowel preparation on
+microbiota-related metrics differ in health and in inflammatory bowel
+disease and for the mucosal and luminal microbiota compartments.
+Clinical and Translational Gastroenterology \textbf{7}:e143.
+doi:\href{https://doi.org/10.1038/ctg.2015.54}{10.1038/ctg.2015.54}.
+
+\hypertarget{ref-Kozich2013}{}
+23. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-Schloss2009}{}
+24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
+\hypertarget{ref-Wang2007}{}
+25. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
+of rRNA sequences into the new bacterial taxonomy. Applied and
+Environmental Microbiology \textbf{73}:5261--5267.
+doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
+
+\hypertarget{ref-Liaw2002}{}
+26. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+regression by randomForest. R News: The Newsletter of the R Project
+\textbf{2}:18--22.
 

### 2017-06-13 submission/manuscript.Rmd
added chunks to manuscript

@@
 title: '**Machine learning identifies unique taxa differentiating proximal and distal
-  human colonic microbiota**'
+  human colonic microbiota*'
 csl: mbio.csl
@@
 \linenumbers
+```{r dependencies}
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')
+for (dep in pack_used){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
+                     quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+```
+
+```{r datasets}
+meta <- 'data/raw/kws_metadata.tsv'
+shared <- 'data/mothur/kws_final.an.shared'
+tax <- 'data/mothur/kws_final.an.cons.taxonomy'
+subsample <- 'data/mothur/kws_final.an.0.03.subsample.shared'
+
+meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+
+simps <- read.table(file='data/mothur/kws_final.an.groups.summary', header = T)
+metasimp <- merge(metadata, simps)
+
+```
+
+```{r otu_barplot}
+source('code/Sum_OTU_by_Tax.R')
+source('code/sum_shared.R')
+
+#use this code to assign phyla to each OTU in the shared file 
+shared_phyla <- get_tax_level_shared(subsample, tax, 2)
+phyla_met <- merge(meta_file, shared_phyla, by.x='group', by.y='row.names')
+
+#try to get the df organized to work as a boxplot - no median calculation 
+phyla_test <- phyla_met[, c("location","Firmicutes","Bacteroidetes","Proteobacteria","Verrucomicrobia","Actinobacteria","Fusobacteria")]
+subsampled_to <- 4231
+RA <- function(x) 100*x/subsampled_to
+
+phyla_RA <- data.frame(phyla_test[1], apply(phyla_test[2:ncol(phyla_test)],2, RA))
+phylaRAnames <- colnames(phyla_RA[,1:7])
+phyla_RAmelt <- melt(phyla_RA[, phylaRAnames], id.vars=1)
+
+```
+
+```{r simpson}
+
+```
 

### 2017-06-13 code/build_figure2.R

@@ -1,0 +1,66 @@
+##########################################
+# Builds Figure 2: OTUs and Simpson diversity by location 
+#
+#
+##########################################
+
+meta <- 'data/raw/kws_metadata.tsv'
+shared <- 'data/mothur/kws_final.an.shared'
+tax <- 'data/mothur/kws_final.an.cons.taxonomy'
+subsample <- 'data/mothur/kws_final.an.0.03.subsample.shared'
+
+meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+
+source('code/Sum_OTU_by_Tax.R')
+source('code/sum_shared.R')
+
+############################################
+# OTU boxplot
+
+#use this code to assign phyla to each OTU in the shared file 
+shared_phyla <- get_tax_level_shared(subsample, tax, 2)
+phyla_met <- merge(meta_file, shared_phyla, by.x='group', by.y='row.names')
+
+#try to get the df organized to work as a boxplot - no median calculation 
+phyla_test <- phyla_met[, c("location","Firmicutes","Bacteroidetes","Proteobacteria","Verrucomicrobia","Actinobacteria","Fusobacteria")]
+
+subsampled_to <- 4231
+RA <- function(x) 100*x/subsampled_to
+
+phyla_RA <- data.frame(phyla_test[1], apply(phyla_test[2:ncol(phyla_test)],2, RA))
+phylaRAnames <- colnames(phyla_RA[,1:7])
+phyla_RAmelt <- melt(phyla_RA[, phylaRAnames], id.vars=1)
+
+
+#for boxplot version
+positions <- c("RB", "RS", "LB", "LS", "SS")
+ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(color=variable)) + 
+  scale_color_discrete(guide=FALSE)+
+  geom_boxplot(aes(fill=variable), outlier.shape=21, outlier.size=2.5) + theme_bw() + 
+  theme(axis.text = element_text(size= 16), axis.title= element_text(size=18), legend.text=element_text(size=14), legend.title=element_text(size=16)) +
[…]

### 2017-06-13 code/diversity_analysis.R

@@ -9,7 +9,7 @@ simps <- read.table(file='data/mothur/kws_final.an.groups.summary', header = T)
 #fullshan <- merge(shannon, metadata)
 shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
 
-#load niel's script for properly , reading in thetayc distances
+#load niel's script for properly reading in thetayc distances
 source(file = 'code/read.dist.R')
 #tyc <-read.dist(file='kws.an.thetayc.0.03.lt.dist', input = "lt", make.square=F, diag=NA)
 
@@ -266,8 +266,6 @@ axis(1, at=1:5, labels=c("left biopsy", " ", "right biopsy", " ", "spon. stool")
 #somewhere simpmeta gets defined and idk where
 simpmeta <- merge(meta_file, simps)
 
-install.packages("wesanderson")
-library("wesanderson")
 simpmeta$location <- factor(simpmeta$location, c("LB","RB", "LS", "RS", "SS"))
 ggplot(simpmeta, aes(x=location, y=invsimpson, group =1)) +geom_point() +geom_jitter(width=0.2) +
   theme_bw() + ylab("Inverse Simpson Diversity") +

### 2017-06-13 code/otu_barplots.R

@@ -20,13 +20,9 @@ tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t'
 
 shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
 
-#make OTU abundance file
-#remove numOTUs column 
 #Create df with relative abundances
-
 test <- subset(shared_file, select = -c(numOtus, label))
-rel_abund <- 100*test/unique(apply(test, 1, sum)) #unique is bad
-
+rel_abund <- 100*test/unique(apply(test, 1, sum)) 
 
 #Create vector of OTUs with median abundances >1%
 OTUs_1 <- apply(rel_abund, 2, max) > 1
@@ -34,17 +30,15 @@ OTU_list <- colnames(rel_abund)[OTUs_1]
 
 #get df of just top OTUs
 rel_abund_top <- rel_abund[, OTUs_1]
-
 rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
 
 source('code/Sum_OTU_by_Tax.R')
 source('code/sum_shared.R')
 
-#this code makes the df for family level RA sorted by % abundance thanks to nick 
+#this code makes the df for family level RA sorted by % abundance 
 #add column of sites to rel_abund_top
 locat <- meta_file$location
 rel_abund_top <- cbind(site = locat, rel_abund_top)
-
 
 #this is what i want! 
 rel_fam <- sum_OTU_by_tax_level(2, rel_abund_top, tax_file)
@@ -82,8 +76,6 @@ rownames(fam_loc) <- fam_loc$Group.1
 fam_loc <- fam_loc[,-1]
 fam_abund <- 100*fam_loc/4231
 
-#need to sort fam_abund by 
-
 #do this to remove 0s
 
 fam_abund <- fam_abund[,colSums(fam_abund^2) !=0]
@@ -98,19 +90,20 @@ phyla_upper <- phyla_upper[, c("Group.1","Firmicutes","Bacteroidetes","Proteobac
 phyla_lower <- phyla_lower[, c("Group.1","Firmicutes","Bacteroidetes","Proteobacteria","Verrucomicrobia","Actinobacteria","Fusobacteria")]
 
 #get rel abundance - subsampled to 4231- define as variable (or calculate )
+subsampled_to <- 4231
 rownames(phyla_loc) <- phyla_loc$Group.1
[…]

### 2017-06-13 submission/manuscript.Rmd

--- 2017-06-12 submission/manuscript.Rmd
@@
 
+{>> SUMMARY Overall looks really good.
+I do agree that the discussion is long and could be shorter.
+I marked a paragraph that could probably be cut.
+Otherwise there are so many details that I think are unnecessary.
+Making a model outline figure would probably be a good move, and help with the organization and direction of the overall discussion.
+I also marked a section in the results that could probably be split into two sections.
+That might make the points clearer and more compelling.
+<<}
+
+{>> I would drop the method stuff in the title. The machine learning stuff. Maybe add words that convey why that is biologically important? <<}
+
 Running title: Specific taxa differentiate proximal and distal human colon microbiota
+
+{>> Continued... This running title is better. Maybe add "...and inform disease state" or something like that? Just for more of a hook.  <<}
 
@@
 Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3^, and Patrick D. Schloss^1$\dagger$^
+
+{>> Let me know if you want stuff like this automated. Such as the author affilitations. It is not terribly hard to do, just need to edit the latex template. <<}
 
@@
 
-Colorectal cancer (CRC) remains a leading cause of death worldwide.
+Colorectal cancer (CRC) remains a leading cause of death worldwide {>> "despite decades of research and improved preventative and therapeutic measures." or something like that.
+<<}.
 Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct.
@@
 Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease.
-In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon.
+We extended these observations of the fecal microbiome to include analysis of the proximal and distal human colon {>> I made this sentence shorter <<}.
 We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure.
 This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
-16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and exit stool for 20 healthy individuals.
+16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and {>> "exit stool" this is a funny way of saying it ;) <<} for 20 healthy individuals.
 Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
-Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites.
+Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm {>> switched this around <<} to identify key species that distinguish biogeographical sites.
 Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location.
@@
 Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon.
-Taken together, our results have identified distinct bacterial populations of the proximal and distal colon.
+Taken together, our results have identified distinct bacterial populations of the proximal and distal colon {>> Again though, state why this is important.
+Kind of setting up the next sentence, if that makes sense.
+<<}.
 Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.
@@
 CRC has recently been described to fall into four molecular subtypes (@Guinney2015).
-Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms.
-In contrast, right-sided colon tumors are bulky and project into the lumen, often causing no symptoms until they have developed into carcinomas (@Benedix2010).
-Because of the absence of symptoms, right-sided tumors, which comprise up to 47% of CRCs, have a significantly higher mortality rate (@Benedix2010).
+Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions present with painful symptoms and account for 53% of CRC cases {>> I moved the percent stuff up here from below.
+<<}.
+In contrast, right-sided colon tumors are bulky and project into the lumen, often remaining asymptomatic {>> I changed some wording here <<} until advanced carcinogenesis {>> Something like this because "they have developed into carcinomas" kind of requires you to have described the advancing stages.
+<<} (@Benedix2010).
+Because they remain asympomatic, right-sided tumors have resulted in a significantly higher mortality rate (@Benedix2010).
 The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
@@
 
-Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012).
+Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012) {>> You say several and cite one, unless this is a review?
+<<}.
 Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014).
 Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013).
-Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014).
+Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated {>> "based on FISH staining and 16S rRNA sequencing" not sure if you need to mention this.
+<<} (@Dejea2014).
 These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
-Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).
-However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon.
+These periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our classification studies (@Baxter2016) {>> I changed this sentence around a bit.
+<<}.
+Because these studies compared shed human stool or mice samples, they were unable to analyze the paried samples from the proximal and distal sides of the colon {>> I also fixed this up to be more active.
+<<}.
 Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
-Disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.
-However, the contribution of the gut microbiota to these subtypes is largely undefined.
+Due to these limitations, the contribution of the gut microbiota to these subtypes is largely undefined.
+Characterizing these changes could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.
+{>> Or something like that.
+I just moved stuff around at the end of the paragraph.
+<<}
 
-Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods.
+In addition to the lack of geographical location in prevous studies, the few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods {>> Changed this around <<}.
 The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures.
@@
 
-In this study we sought to combine these approaches to characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
-Our design uses unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native flora in 20 healthy volunteers.
-To get around the noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
+We combined {>> more confident language!
+:) <<} these approaches to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
+Our design used {>> past tense <<} unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native flora in 20 healthy volunteers.
+To address {>> too casual so I changed it <<} noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
 We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes.
 By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.
+
+{>> Good, I like how this section ends! <<}
+
+{>> I tried to get rid of the "x, y" structure of starting sentences. Stuff like "Interestingly, we found..." or "Furthermore, this showed...". <<}
 
@@
 This data combines the samples from all participants in the study.
-Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research.
+Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research {>> citations?
+<<}.
 Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
-That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on.
+That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, {>> "and so on" I wouldn't say that phrase, be more quantitative.
+<<}.
 Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.
 
+{>> Figure 1 could be vetorized to prevent pixelation. A minor thing, but helpful if you can do it. <<}
+
 To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals.
-Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
+Again, across all patient samples we observed {>> more scientific word <<} a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
 However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
-Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool.
+
+{>> Could do a new paragraph if you wanted. <<}
+
+Next we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool.
 Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
-Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
-To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
+Furthermore {>> "furthermore" <<}, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is {>> delete "quite" <<} different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
+To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal) {>> Switch this sentence around to make it more active <<}.
 We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005).
-Thus interpersonal variation between subjects drives the differences between samples more than sample site or location.
+Thus interpersonal variation between subjects drives the differences between samples more than sample site or location {>> I think that this is your most relevant finding.
+You might try to make this the main point of the section?
+Maybe have one result section for interpersonal variation and another for left lumen representing the exit stool.
+That might make your points clearer and more compelling.
+Just an idea though.
+<<}.
 Overall, the results of the similarity analysis suggest that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
@@
 #### Random Forest classification models identify important OTUs on each side
-To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances.
-The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A).
+To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest {>> drop the phrase "machine-learning" here <<} models built on OTU abundances.
+We built the first model to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A) {>> moved things around <<}.
 The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left.
@@
 These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.
+
+{>> That model is suspiciously good.
+Could it be confounded by samples being from the same patient?
+Also I am guessing this was done with caret?
+The ROC curve is lumpy, which suggests you have two real models in there.
+That is something we could talk about, but the AUC may be artificially large?
+<<}
 
@@
 
+{>> I think your reports on the AUC models seem fine. <<}
+
 #### Bacterial OTUs associated with cancer are found in healthy individuals 
-Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs.
-Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subsp animalis_.
+Given that specific bacterial species have been associated with colorectal cancer, {>> added comma <<} we probed our sample set for these OTUs.
+Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subsp animalis_ {>> not sure I like BLAST as a verb or whatever that is.
+I do it too, but try to avoid it.
+<<}.
 It was the only sequence to align with a _F.
@@
 
-When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes.
+When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes {>> I would go with past tense here too.
+<<}.
 This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome.
@@
 Hence, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
+
+{>> This is interesting but may be too tangential. Might be a candidate for cutting. <<}
 
@@
 
+{>> So all this info is great and you definitely know your stuff, but I don't think you need to go in depth that much.
+I see your note about including a model figure, and I think that would be a great idea.
+Reference that and exclude so much of the text.
+And modeling the discussion around that model will probably help clear up the direction and purpose.
+<<}
+
 Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer.
@@
 Given that our results showed that the stool most accurately reflects the community of the left lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
+
+{>> Again might be a bit too much information, but I do like the point you are making here. Pretty interesting. <<}
 

### 2017-06-13 submission/manuscript.Rmd
rmd updates

--- 2017-06-13 submission/manuscript.Rmd
@@
 \linenumbers
+
 ```{r dependencies}
+#setwd('~/Documents/Flynn_LRColon_XXXX_2017')
 pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')
@@
 ```{r datasets}
-meta <- 'data/raw/kws_metadata.tsv'
-shared <- 'data/mothur/kws_final.an.shared'
-tax <- 'data/mothur/kws_final.an.cons.taxonomy'
-subsample <- 'data/mothur/kws_final.an.0.03.subsample.shared'
+#setwd("~/Users/kaitlinflynn/Documents/Flynn_LRColon_XXXX_2017")
+meta <- '../data/raw/kws_metadata.tsv'
+shared <- '../data/mothur/kws_final.an.shared'
+tax <- '../data/mothur/kws_final.an.cons.taxonomy'
+subsample <- '../data/mothur/kws_final.an.0.03.subsample.shared'
 
-meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
-shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
-tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+meta_file <- read.table(file='../data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='../data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='../data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
 shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
 
-simps <- read.table(file='data/mothur/kws_final.an.groups.summary', header = T)
-metasimp <- merge(metadata, simps)
+simps <- read.table(file='../data/mothur/kws_final.an.groups.summary', header = T)
+simpmeta <- merge(meta_file, simps)
+
+tyc <- read.table("../data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
 
@@
 ```{r otu_barplot}
-source('code/Sum_OTU_by_Tax.R')
-source('code/sum_shared.R')
+source('../code/Sum_OTU_by_Tax.R')
+source('../code/sum_shared.R')
 
@@
 phyla_RAmelt <- melt(phyla_RA[, phylaRAnames], id.vars=1)
-
-```
-
-```{r simpson}
 

### 2017-06-13 code/build_figure2.R

@@ -37,7 +37,7 @@ phyla_RAmelt <- melt(phyla_RA[, phylaRAnames], id.vars=1)
 
 #for boxplot version
 positions <- c("RB", "RS", "LB", "LS", "SS")
-ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(color=variable)) + 
+phy_plot <- ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(color=variable)) + 
   scale_color_discrete(guide=FALSE)+
   geom_boxplot(aes(fill=variable), outlier.shape=21, outlier.size=2.5) + theme_bw() + 
   theme(axis.text = element_text(size= 16), axis.title= element_text(size=18), legend.text=element_text(size=14), legend.title=element_text(size=16)) +
@@ -54,13 +54,23 @@ simps <- read.table(file='data/mothur/kws_final.an.groups.summary', header = T)
 simpmeta <- merge(meta_file, simps)
 
 simpmeta$location <- factor(simpmeta$location, c("LB","RB", "LS", "RS", "SS"))
-ggplot(simpmeta, aes(x=location, y=invsimpson, group =1)) +geom_point() +geom_jitter(width=0.2) +
-  theme_bw() + ylab("Inverse Simpson Diversity") +
-  scale_x_discrete(labels=c("L Mucosa", "R Mucosa", "L Lumen", "R Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank()) +
-  stat_summary(aes(x=location, y=invsimpson), data = simpmeta, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.4) +
-  theme(axis.text = element_text(size= 16), axis.title= element_text(size=18))
-
-
+positions <- c("RB", "RS", "LB", "LS", "SS")
+simp_plot <- ggplot(simpmeta, aes(x=location, y=invsimpson, group =1)) +geom_point() +geom_jitter(width=0.2) +theme_bw() + ylab("Inverse Simpson Diversity") +
+  scale_x_discrete(limits = positions, breaks=positions, 
+                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 16), axis.title= element_text(size=18)) +
[…]

### 2017-06-13 code/build_figure3.R

@@ -1,0 +1,141 @@
+###########################################################
+# Builds Figure 3
+
+###########################################################
+
+pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson')
+for (dep in pack_used){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
+                     quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+tyc <- read.table("data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
+
+#thetayc distances
+#separate column values for comparisons
+tyc <- separate(tyc, label, into= c('pt1', 'samp1'), sep="-", remove=F)
+tyc <- separate(tyc, comparison, into= c('pt2', 'samp2'), sep="-", remove=F)
+tyc <- subset(tyc, select = -c(row.names, X))
+
+#subset data to only include patients comparisons to each other
+tyc <- subset(tyc, pt1==pt2)
+tyc <- unite_(tyc, "match", from=c('samp1', 'samp2'), sep="_", remove = F)
+
+#now plot- this is a plot of thetayc distances in all pairings for all patients 
+ggplot(tyc, aes(x=match, y=thetayc)) + geom_point(aes(col=match)) +facet_wrap(~pt1) +theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggplot(tyc, aes(x=match, y=thetayc)) + geom_boxplot() +theme_bw()
+
+lefttyc <- subset(tyc, samp1== 'LB' | samp1== 'LS')
[…]

### 2017-06-14 submission/manuscript.Rmd
working code into rmd

@@
 ```{r dependencies}
-#setwd('~/Documents/Flynn_LRColon_XXXX_2017')
 pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')
@@
 ```{r datasets}
-#setwd("~/Users/kaitlinflynn/Documents/Flynn_LRColon_XXXX_2017")
 meta <- '../data/raw/kws_metadata.tsv'
@@
 
+```
+
+```{r thetayc}
+tyc <- read.table("../data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
+
+#separate column values for comparisons
+tyc <- separate(tyc, label, into= c('pt1', 'samp1'), sep="-", remove=F)
+tyc <- separate(tyc, comparison, into= c('pt2', 'samp2'), sep="-", remove=F)
+tyc <- subset(tyc, select = -c(row.names, X))
+
+tyc <- subset(tyc, pt1==pt2)
+tyc <- unite_(tyc, "match", from=c('samp1', 'samp2'), sep="_", remove = F)
+
+stooltyc <- subset(tyc, match=='LB_RB' | match== 'LS_RS')
+leftandrighttyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
+lvsr <- rbind(stooltyc, leftandrighttyc)
+
+exittyc <- subset(tyc, samp2 == 'SS')
+
+pvalues <- c()
+
+Atyc <- subset(tyc, match=='RB_RS' | match=='LS_RS')
+pvalues <- c(pvalues, wilcox.test(thetayc~match, data=Atyc, paired=T)$p.value)
+
+btyc <- subset(tyc, match=='RB_RS'| match=='LB_RB')
+btyc <- btyc[-25,]
+pvalues <- c(pvalues, wilcox.test(thetayc~match, data=btyc, paired=T)$p.value)
+
+ctyc <- subset(tyc, match=='RB_RS'| match=='LB_LS')
+ctyc <- ctyc[-25,]
+pvalues <- c(pvalues, wilcox.test(thetayc~match, data=ctyc, paired=T)$p.value)
+
+dtyc <- subset(tyc, match == 'LS_RS' | match == 'LB_RB')
+dtyc <- dtyc[-25,]
+pvalues <- c(pvalues, wilcox.test(thetayc~match, data=dtyc, paired=T)$p.value)
+
+etyc <- subset(tyc, match == 'LS_RS' | match == 'LB_LS')
+etyc <- etyc[-25,]
+pvalues <- c(pvalues, wilcox.test(thetayc~match, data=etyc, paired=T)$p.value)
+
+ftyc <- subset(tyc, match == 'LB_RB' | match == 'LB_LS')
+pvalues <- c(pvalues, wilcox.test(thetayc~match, data=ftyc, paired=T)$p.value)
+
+pvalues <- p.adjust(pvalues, method = "BH")
+
+# now for exit comparisons
+
+stoolpvalues <- c()
+
+htyc <- subset(tyc, match=='RB_SS' | match=='RS_SS')
+htyc <- htyc[-25,]
+stoolpvalues <- c(stoolpvalues, wilcox.test(thetayc~match, data=htyc, paired=T)$p.value)
+
+ityc <- subset(tyc, match=='RB_SS' | match=='LB_SS')
+stoolpvalues <- c(stoolpvalues, wilcox.test(thetayc~match, data=ityc, paired=T)$p.value)
+
+jtyc <- subset(tyc, match=='RB_SS' | match=='LS_SS')
+stoolpvalues <- c(stoolpvalues, wilcox.test(thetayc~match, data=jtyc, paired=T)$p.value)
+
+ktyc <- subset(tyc, match=='RS_SS' | match=='LB_SS')
+ktyc <- ktyc[-25,]
+stoolpvalues <- c(stoolpvalues, wilcox.test(thetayc~match, data=ktyc, paired=T)$p.value)
+
+ltyc <- subset(tyc, match=='RS_SS' | match=='LS_SS')
+ltyc <- ltyc[-25,]
+stoolpvalues <- c(stoolpvalues, wilcox.test(thetayc~match, data=ltyc, paired=T)$p.value)
+
+mtyc <- subset(tyc, match=='LB_SS' | match=='LS_SS')
+stoolpvalues <- c(stoolpvalues, wilcox.test(thetayc~match, data=mtyc, paired=T)$p.value)
+
+stoolpvalues <- p.adjust(stoolpvalues, method = "BH")
+
+alltyc <- read.table("../data/process/allshared.summary", sep = '\t', header = T, row.names=NULL)
+alltyc <- separate(alltyc, label, into= c('pt1', 'samp1'), sep="-", remove=F)
+alltyc <- separate(alltyc, comparison, into= c('pt2', 'samp2'), sep="-", remove=F)
+alltyc <- alltyc[-1]
+alltyc <- alltyc[-7]
+
+#ultimately want a plot of all points where pt1 == pt2 in one bar and all of the others in another column 
+#unite and make column of 0/1 for matches? then can plot 1 and 0s 
+#should i separate out lumen and mucosa ? sure or no not for now
+
+alltyc["same_pt"] <- NA
+
+for (i in 1:nrow(alltyc)){
+  if (alltyc$pt1[i] == alltyc$pt2[i]){
+    alltyc$same_pt[i] <- 1
+  }
+  else alltyc$same_pt[i] <- 0
+}
+
+alltyc[10] <- as.factor(alltyc[10])
+
+inter_medians <- aggregate(thetayc ~ same_pt, alltyc, median)
 ```

### 2017-06-14 code/build_figre3new.R

@@ -27,60 +27,27 @@ tyc <- subset(tyc, select = -c(row.names, X))
 tyc <- subset(tyc, pt1==pt2)
 tyc <- unite_(tyc, "match", from=c('samp1', 'samp2'), sep="_", remove = F)
 
-#now plot- this is a plot of thetayc distances in all pairings for all patients 
-ggplot(tyc, aes(x=match, y=thetayc)) + geom_point(aes(col=match)) +facet_wrap(~pt1) +theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-ggplot(tyc, aes(x=match, y=thetayc)) + geom_boxplot() +theme_bw()
-
-lefttyc <- subset(tyc, samp1== 'LB' | samp1== 'LS')
-lefttyc <- subset(lefttyc, samp2== 'LB' | samp2== 'LS' | samp2=='SS')
-
-righttyc <- subset(tyc, samp1=='RB' | samp1 == 'RS')
-righttyc <- subset(righttyc, samp2=='RB' | samp2 == 'RS' | samp2 == 'SS')
-
-mucosatyc <- subset(tyc, samp1=='RB' | samp1 == 'LB')
-mucosatyc <- subset(mucosatyc, samp2=='RB' | samp2 == 'LB')
-
 stooltyc <- subset(tyc, match=='LB_RB' | match== 'LS_RS')
-
-exittyc <- subset(tyc, samp2 == 'SS')
-exitLtyc <- subset(exittyc, samp1 == 'LS' | samp1 == 'LB')
-exitRtyc <- subset(exittyc, samp1 == 'LS' | samp1 == 'RS')
-
-
 leftandrighttyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
-
 lvsr <- rbind(stooltyc, leftandrighttyc)
 
-ggplot(lefttyc, aes(x=match, y=thetayc)) + geom_boxplot() +theme_bw() +ggtitle("Left")
+exittyc <- subset(tyc, samp2 == 'SS')
 
-ggplot(righttyc, aes(x=match, y=thetayc)) + geom_boxplot() +theme_bw() +ggtitle("Right")
+#plots
+tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
+ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
[…]

### 2017-06-14 submission/manuscript.Rmd
more code in manuscript

@@
 
-alltyc[10] <- as.factor(alltyc[10])
+inter_medians <- aggregate(thetayc ~ same_pt, alltyc, median)
+```
 
-inter_medians <- aggregate(thetayc ~ same_pt, alltyc, median)
+```{r randomforest}
+#create dataframe with relative abundances
+rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+
+#Create vector of OTUs with median abundances >1%
+OTUs_1 <- apply(rel_abund, 2, max) > 1
+OTU_list <- colnames(rel_abund)[OTUs_1]
+#get df of just top OTUs
+rel_abund_top <- rel_abund[, OTUs_1]
+rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+seed <- 1
+n_trees <- 2001
+
+source('../code/random_functions.R')
+
+#####RandomForest###########################################################################################
+#build randomForest model for each location comparison using randomize_loc function 
+rf_left <- randomize_loc(rel_meta, "LB", "LS") #OOB 10.26%
+rf_right <- randomize_loc(rel_meta, "RB", "RS") #OOB 53%
+rf_bowel <- randomize_loc(rel_meta, "LB", "RB") #OOB 25.64%
+rf_lumen <- randomize_loc(rel_meta, "LS", "RS") #OOB 69.23%
+
+#and for each site
+rf_all <- randomize_site(rel_meta, "mucosa", "stool")
+
+#####AUCRF#####################################################################################
+
+# create RF model with AUCRF outputs top OTUs
+aucrf_data_left_bs <- auc_loc(rel_meta, "LB", "LS")
+aucrf_data_LRbowel <- auc_loc(rel_meta, "LB", "RB")
+aucrf_data_right_bs <- auc_loc(rel_meta, "RB", "RS")
+aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
+aucrf_data_allum <- auc_site(rel_meta, "mucosa", "stool")
+
+########Cross-Validation#################################################################
+#10 fold cross validation for all lumen vs mucosa 
+iters <- 100
+cv10f_aucs <- c()
+cv10f_all_resp <- c()
+cv10f_all_pred <- c()
+for(j in 1:iters){
+  set.seed(j)
+  sampling <- sample(1:nrow(aucrf_data_allum),nrow(aucrf_data_allum),replace=F)
+  cv10f_probs <- rep(NA,78)
+  for(i in seq(1,77,7)){
+    train <- aucrf_data_allum[sampling[-(i:(i+6))],]
+    test <- aucrf_data_allum[sampling[i:(i+6)],]
+    set.seed(seed)
+    temp_model <- AUCRF(site~., data=train, pdel=0.99, ntree=500)
+    cv10f_probs[sampling[i:(i+6)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  }
+  cv10f_roc <- roc(aucrf_data_allum$site~cv10f_probs)
+  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
+  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data_allum$site)
+  cv10f_aucs[j] <- cv10f_roc$auc #stores aucs for all iterations, can use to calc IQR
+}
+cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
+
+#10fold CV for L lumen vs L mucosa
+iters <- 100
+cv10f_aucs <- c()
+cv10f_all_resp_left_bs <- c()
+cv10f_all_pred_left_bs <- c()
+for(j in 1:iters){
+  set.seed(j)
+  sampling <- sample(1:nrow(aucrf_data_left_bs),nrow(aucrf_data_left_bs),replace=F)
+  cv10f_probs <- rep(NA,39)
+  for(i in seq(1,36,4)){
+    train_left_bs <- aucrf_data_left_bs[sampling[-(i:(i+3))],]
+    test_left_bs <- aucrf_data_left_bs[sampling[i:(i+3)],]
+    set.seed(seed)
+    temp_model_left_bs <- AUCRF(location~., data=train_left_bs, pdel=0.99, ntree=500)
+    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_left_bs$RFopt, test_left_bs, type='prob')[,2]
+  }
+  cv10f_roc_left_bs <- roc(aucrf_data_left_bs$location~cv10f_probs)
+  cv10f_all_pred_left_bs <- c(cv10f_all_pred_left_bs, cv10f_probs)
+  cv10f_all_resp_left_bs <- c(cv10f_all_resp_left_bs, aucrf_data_left_bs$location)
+  cv10f_aucs[j] <- cv10f_roc_left_bs$auc #stores aucs for all iterations, can use to calc IQR
+}
+cv10f_roc_left_bs <- roc(cv10f_all_resp_left_bs~cv10f_all_pred_left_bs)
+
+#10fold CV for R lumen vs R mucosa
+iters <- 100
+cv10f_aucs <- c()
+cv10f_all_resp_right_bs <- c()
+cv10f_all_pred_right_bs <- c()
+for(j in 1:iters){
+  set.seed(j)
+  sampling <- sample(1:nrow(aucrf_data_right_bs),nrow(aucrf_data_right_bs),replace=F)
+  cv10f_probs <- rep(NA,39)
+  for(i in seq(1,36,4)){
+    train_right_bs <- aucrf_data_right_bs[sampling[-(i:(i+3))],]
+    test_right_bs <- aucrf_data_right_bs[sampling[i:(i+3)],]
+    set.seed(seed)
+    temp_model_right_bs <- AUCRF(location~., data=train_right_bs, pdel=0.99, ntree=500)
+    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_right_bs$RFopt, test_right_bs, type='prob')[,2]
+  }
+  cv10f_roc_right_bs <- roc(aucrf_data_right_bs$location~cv10f_probs)
+  cv10f_all_pred_right_bs <- c(cv10f_all_pred_right_bs, cv10f_probs)
+  cv10f_all_resp_right_bs <- c(cv10f_all_resp_right_bs, aucrf_data_right_bs$location)
+  cv10f_aucs[j] <- cv10f_roc_right_bs$auc #stores aucs for all iterations, can use to calc IQR
+}
+cv10f_roc_right_bs <- roc(cv10f_all_resp_right_bs~cv10f_all_pred_right_bs)
+
+#10 fold cross validation for L vs R mucosa
+
+iters <- 100
+cv10f_aucs_muc <- c()
+cv10f_all_resp_muc <- c()
+cv10f_all_pred_muc <- c()
+for(j in 1:iters){
+  set.seed(j)
+  sampling_muc <- sample(1:nrow(aucrf_data_LRbowel),nrow(aucrf_data_LRbowel),replace=F)
+  cv10f_probs_muc <- rep(NA,39)
+  for(i in seq(1,36,4)){
+    train_muc <- aucrf_data_LRbowel[sampling_muc[-(i:(i+3))],]
+    test_muc <- aucrf_data_LRbowel[sampling_muc[i:(i+3)],]
+    set.seed(seed)
+    temp_model_muc <- AUCRF(location~., data=train_muc, pdel=0.99, ntree=500)
+    cv10f_probs_muc[sampling_muc[i:(i+3)]] <- predict(temp_model_muc$RFopt, test_muc, type='prob')[,2]
+  }
+  cv10f_roc_muc <- roc(aucrf_data_LRbowel$location~cv10f_probs_muc)
+  cv10f_all_pred_muc <- c(cv10f_all_pred_muc, cv10f_probs_muc)
+  cv10f_all_resp_muc <- c(cv10f_all_resp_muc, aucrf_data_LRbowel$location)
+  cv10f_aucs_muc[j] <- cv10f_roc_muc$auc #stores aucs for all iterations, can use to calc IQR
+}
+cv10f_roc_muc <- roc(cv10f_all_resp_muc~cv10f_all_pred_muc)
+
+#10 fold cross validation for L vs R lumen
+iters <- 100
+cv10f_aucs_lum <- c()
+cv10f_all_resp_lum <- c()
+cv10f_all_pred_lum <- c()
+for(j in 1:iters){
+  set.seed(j)
+  sampling_lum <- sample(1:nrow(aucrf_data_LRlumen),nrow(aucrf_data_LRlumen),replace=F)
+  cv10f_probs_lum <- rep(NA,39)
+  for(i in seq(1,36,4)){
+    train_lum <- aucrf_data_LRlumen[sampling_lum[-(i:(i+3))],]
+    test_lum <- aucrf_data_LRlumen[sampling_lum[i:(i+3)],]
+    set.seed(seed)
+    temp_model_lum <- AUCRF(location~., data=train_lum, pdel=0.99, ntree=500)
+    cv10f_probs_lum[sampling_lum[i:(i+3)]] <- predict(temp_model_lum$RFopt, test_lum, type='prob')[,2]
+  }
+  cv10f_roc_lum <- roc(aucrf_data_LRlumen$location~cv10f_probs_lum)
+  cv10f_all_pred_lum <- c(cv10f_all_pred_lum, cv10f_probs_lum)
+  cv10f_all_resp_lum <- c(cv10f_all_resp_lum, aucrf_data_LRlumen$location)
+  cv10f_aucs_lum[j] <- cv10f_roc_lum$auc #stores aucs for all iterations, can use to calc IQR
+}
+cv10f_roc_lum <- roc(cv10f_all_resp_lum~cv10f_all_pred_lum)
+
+```
+
+```{r important_otus}
+
 ```

### 2017-06-14 code/build_figre3new.R

@@ -35,19 +35,18 @@ exittyc <- subset(tyc, samp2 == 'SS')
 
 #plots
 tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
-ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
+lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
   theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=16)) + 
   scale_x_discrete(limits = tycpositions, breaks = tycpositions,
                    labels=c("R Mucosa vs R Lumen", "L Lumen vs R Lumen", "L Mucosa vs R Mucosa", "L Mucosa vs L Lumen")) +
   ylab("ThetaYC distance") 
 
 exitpositions <- c("RB_SS", "RS_SS", "LB_SS", "LS_SS")
-ggplot(exittyc, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() +
+exit_plot <- ggplot(exittyc, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() +
   theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=16)) +
   scale_x_discrete(limits=exitpositions, breaks=exitpositions,
                    labels=c("R Mucosa vs Stool", "R Lumen vs Stool", "L Mucosa vs Stool", "L Lumen vs Stool")) +
-  theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")+ 
-  stat_summary(aes(x=match, y=thetayc), data = exittyc, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.4)
+  theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")
 
 #trying to get adonis to work for comparisons 
 
@@ -130,13 +129,7 @@ for (i in 1:nrow(alltyc)){
 
 alltyc[10] <- as.factor(alltyc[10])
 
-ggplot(alltyc, aes(x=as.factor(same_pt), y=thetayc)) + geom_jitter(width=0.15, shape=21, size=3, fill='grey', col='black') + theme_bw()+
[…]

### 2017-06-14 code/random_forest.R

@@ -60,7 +60,7 @@ aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
 aucrf_data_allum <- auc_site(rel_meta, "mucosa", "stool")
 
 #need to fix specific function for these to output just the rf object 
-rf_exitlum_aucrf <- auc_site(rel_meta, "stool", "exit") # not working 
+#rf_exitlum_aucrf <- auc_site(rel_meta, "stool", "exit") # not working 
 rf_exitLlum_aucrf <- auc_loc(rel_meta, "LB", "SS")
 rf_exitRlum_aucrf <- auc_loc(rel_meta, "RB", "SS")
 

### 2017-06-15 submission/manuscript.Rmd
rmd updates

@@
 
+```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+opts_chunk$set("tidy" = TRUE)
+opts_chunk$set("echo" = FALSE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("cache" = FALSE)
+opts_chunk$set("message" = FALSE)
+
+```
+
 ```{r dependencies}
@@
 tax <- '../data/mothur/kws_final.an.cons.taxonomy'
-subsample <- '../data/mothur/kws_final.an.0.03.subsample.shared'
+subsample <- read.table(file='../data/mothur/kws_final.an.0.03.subsample.shared', header=T)
 
@@
 
-tyc <- read.table("../data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
+tyc <- read.table(file ="../data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
 
@@
 #use this code to assign phyla to each OTU in the shared file 
-shared_phyla <- get_tax_level_shared(subsample, tax, 2)
+sub <- '../data/mothur/kws_final.an.0.03.subsample.shared'
+shared_phyla <- get_tax_level_shared(sub, tax, 2)
 phyla_met <- merge(meta_file, shared_phyla, by.x='group', by.y='row.names')
@@
 cv10f_roc_lum <- roc(cv10f_all_resp_lum~cv10f_all_pred_lum)
+```
+```{r important_otus}
+#just LB vs LS 
+left_otu_feat <- colnames(aucrf_data_left_bs[2:6])
+otu_taxa_left <- get_tax(1, left_otu_feat, tax_file)
+#Abundance stripchart or most predictive otus
+ls_abunds <- shared_meta[shared_meta$location=='LS', left_otu_feat]/10000 + 1e-4
+lb_abunds <- shared_meta[shared_meta$location=='LB', left_otu_feat]/10000 + 1e-4
+
+right_otu_feat <- colnames(aucrf_data_right_bs[2:6])
+otu_taxa_right <- get_tax(1, right_otu_feat, tax_file)
+#Abundance stripchart or most predictive otus
+rs_abunds <- shared_meta[shared_meta$location=='RS', right_otu_feat]/10000 + 1e-4
+rb_abunds <- shared_meta[shared_meta$location=='RB', right_otu_feat]/10000 + 1e-4
+
+#Lb vs Rb
+LRbowel_otu_feat <- colnames(aucrf_data_LRbowel[2:6])
+otu_taxa_LRbowel <- get_tax(1, LRbowel_otu_feat, tax_file)
+#Abundance stripchart or most predictive otus 
+lb_abunds <- shared_meta[shared_meta$location=='LB', LRbowel_otu_feat]/10000 + 1e-4
+rblb_abunds <- shared_meta[shared_meta$location=='RB', LRbowel_otu_feat]/10000 + 1e-4
+
+#LS vs RS
+LRlumen_otu_feat <- colnames(aucrf_data_LRlumen[2:6])
+otu_taxa_LRlumen <- get_tax(1, LRlumen_otu_feat, tax_file)
+#Abundance stripchart or most predictive otus 
+lsrs_abunds <- shared_meta[shared_meta$location=='LS', LRlumen_otu_feat]/10000 + 1e-4
+rsls_abunds <- shared_meta[shared_meta$location=='RS', LRlumen_otu_feat]/10000 + 1e-4
 
@@
 
-```{r important_otus}
+```{r oncogenic_otus}
+subsampled_meta <- merge(meta_file, subsample, by.x='group', by.y='row.names')
+fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]
 
+fuso179[,5] <- (fuso179[,4]/subsampled_to)*100
+names(fuso179)[5] <- "Otu00179_relAbund"
+
+p_152 <-subsampled_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00152")]
+
+p_152[,4] <- (p_152[,3]/subsampled_to) *100
+names(p_152)[4] <- "Otu152_abund"
 ```

### 2017-06-15 code/Sum_OTU_by_Tax.R

@@ -5,7 +5,7 @@
 ####
 
 sum_OTU_by_tax_level <- function(TAX_LEVEL, OTU_DF, tax_df){
-     source('code/tax_level.R')
+     source('../code/tax_level.R')
      TAX_DF <-get_tax(TAX_LEVEL,names(OTU_DF), tax_df)
      tax_levels <- as.character(unique(TAX_DF$tax))
      OUTPUT_DF <- data.frame(rep(0,length(rownames(OTU_DF))), row.names=rownames(OTU_DF))

### 2017-06-15 code/oncogenic_otus.R

@@ -33,7 +33,7 @@ fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient"
 fuso179[,5] <- (fuso179[,4]/4321)*100
 names(fuso179)[5] <- "Otu00179_relAbund"
 
-ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter() +theme_bw()
+ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(width=0.3) +theme_bw()
 
 #other oncogenic otus - Otu00152 is P. asaccharolytica. Otu00248 is p micra
 
@@ -41,6 +41,13 @@ p_asach <- shared_meta[, colnames(shared_meta) %in% c("Group", "patient", "locat
 
 p_asach[,4] <- (p_asach[,3]/4321) *100
 names(p_asach)[4] <- "Otu00152_abund"
+
+p_152 <-subsampled_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00152")]
+
+p_152[,4] <- (p_152[,3]/4321) *100
+names(p_152)[4] <- "Otu152_abund"
+
+ggplot(p_152, aes(x=location, y=Otu152_abund)) +geom_jitter() +theme_bw()
 
 ggplot(p_asach, aes(x=location, y=Otu00152_abund)) +geom_jitter() +theme_bw()
 

### 2017-06-15 code/random_forest.R

@@ -419,3 +419,33 @@ axis(2, at=seq(1,index-2,2), labels=otu_taxa_LRlumen$tax_label, las=1, line=-0.5
 axis(1, at=c(1e-4, 1e-3, 1e-2, 1e-1, 1), label=c("0", "0.1", "1", "10", "100"), cex.axis=1.2)
 legend('topright', legend=c("Left lumen", "Right lumen"), pch=c(21, 21), pt.bg=c("brown","magenta"), cex=1.2)
 
+
+
+# build and export figure 
+#export as PDF
+
+plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
+pdf(file=plot_file, width=12, height=11)
+layout(matrix(c(1,
+                2,
+                3), 
+              nrow=3, byrow = TRUE))
+
+#Lumen vs mucosa plot 
+par(mar=c(4,4,1,1))
+plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
+plot(cv10f_roc_right_bs, col='blue', lwd=3, add=T, lty=1)
+plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
+mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.5)
+mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.5)
+legend('bottom', legend=c(
+  sprintf('L Lumen vs L Mucosa, 10-fold CV, AUC =0.980'),
+  sprintf('R Lumen vs R Mucosa, 10-fold CV, AUC = 0.797')
+),lty=c(1, 1, 1), lwd=3, col=c('red', 'blue'), bty='n', cex=1.2)
+
+
+
+
+inter_plot
+
+dev.off()

### 2017-06-16 submission/manuscript.Rmd
fixed RF rmd bug

@@
 output:
-  word_document: default
+  html_document: default
   pdf_document:
@@
     keep_tex: yes
-  html_document: default
+  word_document: default
 geometry: margin=1.0in
@@
 
-```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+knitr::opts_chunk$set("tidy" = TRUE)
+knitr::opts_chunk$set("echo" = FALSE)
+knitr::opts_chunk$set("eval" = TRUE)
+knitr::opts_chunk$set("warning" = FALSE)
+knitr::opts_chunk$set("cache" = FALSE)
 
-opts_chunk$set("tidy" = TRUE)
-opts_chunk$set("echo" = FALSE)
-opts_chunk$set("eval" = TRUE)
-opts_chunk$set("warning" = FALSE)
-opts_chunk$set("cache" = FALSE)
-opts_chunk$set("message" = FALSE)
+inline_hook <- function(x){
+	print(x)
 
+	if(is.list(x)){
+		x <- unlist(x)
+	}
+
+	if(is.numeric(x)){
+		if(abs(x - round(x)) < .Machine$double.eps^0.5){
+			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+		} else {
+			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
+		}
+	} else {
+    	paste(x)      
+	}
+}
+knitr::knit_hooks$set(inline=inline_hook)
 ```
 
-```{r dependencies}
+```{r dependencies, echo=FALSE, warning = FALSE}
 pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')
@@
 
-```{r datasets}
+```{r datasets, echo=FALSE}
 meta <- '../data/raw/kws_metadata.tsv'
@@
 
-```{r otu_barplot}
+```{r otu_barplot, echo=FALSE}
 source('../code/Sum_OTU_by_Tax.R')
@@
 
-```{r thetayc}
+```{r thetayc, echo=FALSE}
 tyc <- read.table("../data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
@@
 
-```{r randomforest}
+```{r randomforest, echo=FALSE}
+
+#make OTU abundance file
+#Create df with relative abundances
+shared_file_rf <- subset(shared_file, select = -c(numOtus, label))
+shared_meta_rf <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+
 #create dataframe with relative abundances
-rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+rel_abund <- 100*shared_file_rf/unique(apply(shared_file_rf, 1, sum))
 
@@
 source('../code/random_functions.R')
+source('../code/tax_level.R')
 
@@
 ```
-```{r important_otus}
+```{r important_otus, echo=FALSE}
 #just LB vs LS 
@@
 
-```{r oncogenic_otus}
+```{r oncogenic_otus, echo=FALSE}
 subsampled_meta <- merge(meta_file, subsample, by.x='group', by.y='row.names')

### 2017-06-16 submission/manuscript.tex
fixed RF rmd bug

@@
 \hypersetup{unicode=true,
-            pdftitle={Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota},
+            pdftitle={**Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota*},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{\textbf{Machine learning identifies unique taxa differentiating proximal
-and distal human colonic microbiota}}
+  \title{**Machine learning identifies unique taxa differentiating proximal and
+distal human colonic microbiota*}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
 \linenumbers
+
+\begin{verbatim}
+## randomForest 4.6-12
+\end{verbatim}
+
+\begin{verbatim}
+## Type rfNews() to see new features/changes/bug fixes.
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Attaching package: 'ggplot2'
+\end{verbatim}
+
+\begin{verbatim}
+## The following object is masked from 'package:randomForest':
+## 
+##     margin
+\end{verbatim}
+
+\begin{verbatim}
+## Type 'citation("pROC")' for a citation.
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Attaching package: 'pROC'
+\end{verbatim}
+
+\begin{verbatim}
+## The following objects are masked from 'package:stats':
+## 
+##     cov, smooth, var
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Attaching package: 'dplyr'
+\end{verbatim}
+
+\begin{verbatim}
+## The following object is masked from 'package:randomForest':
+## 
+##     combine
+\end{verbatim}
+
+\begin{verbatim}
+## The following objects are masked from 'package:stats':
+## 
+##     filter, lag
+\end{verbatim}
+
+\begin{verbatim}
+## The following objects are masked from 'package:base':
+## 
+##     intersect, setdiff, setequal, union
+\end{verbatim}
+
+\begin{verbatim}
+## AUCRF 1.1
+\end{verbatim}
+
+\begin{verbatim}
+## Loading required package: lattice
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Attaching package: 'reshape2'
+\end{verbatim}
+
+\begin{verbatim}
+## The following object is masked from 'package:tidyr':
+## 
+##     smiths
+\end{verbatim}
 

### 2017-06-16 code/random_forest.R

@@ -34,6 +34,7 @@ seed <- 1
 n_trees <- 2001
 
 source('code/random_functions.R')
+source('code/tax_level.R')
 
 #####RandomForest###########################################################################################
 #build randomForest model for each location comparison using randomize_loc function 

### 2017-06-18 submission/manuscript.Rmd
text updated with Hannigan edits

@@
 ---
-title: '**Machine learning identifies unique taxa differentiating proximal and distal
-  human colonic microbiota*'
+title: '*Specific taxa differentiate proximal and distal
+  microbiota in the unprepped human colon*'
 csl: mbio.csl
@@
 
-Colorectal cancer (CRC) remains a leading cause of death worldwide.
+Colorectal cancer (CRC) remains a leading cause of death worldwide despite improved preventative and therapeutic measures.
 Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct.
@@
 Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease.
-In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon.
+We extended our observations of the fecal microbiome to include analysis of the proximal and distal human colon.
 We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure.
 This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
-16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and exit stool for 20 healthy individuals.
+16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals.
 Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
-Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites.
+Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key species that distinguish biogeographical sites.
 Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location.
@@
 Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon.
-Taken together, our results have identified distinct bacterial populations of the proximal and distal colon.
+By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides.
 Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.
@@
 Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms.
-In contrast, right-sided colon tumors are bulky and project into the lumen, often causing no symptoms until they have developed into carcinomas (@Benedix2010).
-Because of the absence of symptoms, right-sided tumors, which comprise up to 47% of CRCs, have a significantly higher mortality rate (@Benedix2010).
+In contrast, 47% of CRCs are caused by right-sided colon tumors that are bulky and project into the lumen, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010).
+Due to the absence of symptoms, right-sided tumors have a significantly higher mortality rate (@Benedix2010).
 The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
@@
 Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013).
-Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014).
+Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated (@Dejea2014).
 These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
-Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).
-However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon.
+Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our classification studies (@Baxter2016).
+Because these studies were performed in mice or examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon.
 Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
-Disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.
-However, the contribution of the gut microbiota to these subtypes is largely undefined.
+Due to these limitations, the contribution of the gut microbiota to these subtypes is largely undefined.
+Characterizing these communities could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.
 
-Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods.
+Further, the few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods.
 The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures.
@@
 
-In this study we sought to combine these approaches to characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
-Our design uses unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native flora in 20 healthy volunteers.
-To get around the noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
+We combined these approaches to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
+Our design used unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
+To address noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
 We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes.
@@
 This data combines the samples from all participants in the study.
-Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research.
+Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research (cite).
 Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
-That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on.
+That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa.
 Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.
@@
 To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals.
-Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
+Again, across all patient samples we observed a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
 However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
+
 Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool.
 Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
-Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
+Furthermore, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
 To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005).
+We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005, @deCarcer2010, @Zhang2013).
 Thus interpersonal variation between subjects drives the differences between samples more than sample site or location.
@@
 #### Random Forest classification models identify important OTUs on each side
-To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances.
-The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A).
+To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models built on OTU abundances.
+We built the first model to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A).
 The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left.
@@
 #### Bacterial OTUs associated with cancer are found in healthy individuals 
-Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs.
-Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subsp animalis_.
+Given that specific bacterial species have been associated with colorectal cancer, we probed our sample set for these OTUs.
+Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_.
 It was the only sequence to align with a _F.
@@
 
-When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes.
+When examining the relative abundance of the different phyla at each site, there was a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes.
 This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome.
@@
 Hence, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
-
-We identified a range of Simpson diversity values between patients and between each anatomical site.
-As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon.
-There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite.
-An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (@Eckberg2005).
-We used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other.
-We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants.
-This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (@Eckberg2005, @deCarcer2010, @Zhang2013).
 
@@
 nucleatum_ and _P.
-asacharolytica_ in 7 and 5 of our subjects, respectively.
+asacharolytica_ in 8 and 5 of our subjects, respectively.
 These bacteria have been shown to be predictive of colorectal cancer in humans (@Baxter2016) and have oncogenic properties in cell culture and in mice (@Sears2014).
@@
 nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa.
-That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F.
-nucleatum_ at all (@Mima2016).
-In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F.
-nucleatum_ in mucosal biopsies (@Strauss2011).
-These studies have the added caveat of being sampled from prepared or cleaned colons.
-Thus, the presence of _F.
+The presence of _F.
 nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers.
@@
 nucleatum_ in the healthy colon.
-Indeed, rather than the _F.
-nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (@Flynn2016).
-Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis.
+Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014).
 Finally, _Fusobacteria_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016).
-Thus, further analysis of the distribution and role of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.
+Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.
 

### 2017-06-18 code/oncogenic_otus.R
exported figures

@@ -58,3 +58,21 @@ names(p_micra)[4] <- "Otu00248_abund"
 
 ggplot(p_micra, aes(x=location, y=Otu00248_abund)) +geom_jitter() +theme_bw()
 
+########################################Build figure 6 
+#export as PDF
+
+plot_file <- '~/Documents/Schloss_Lab/Flynn_LRColon_XXXX_2017/submission/figure_3.pdf'
+pdf(file=plot_file, width=12, height=11)
+layout(matrix(c(1,
+                2,
+                3), 
+              nrow=3, byrow = TRUE))
+
+lvr_plot
+
+exit_plot
+
+inter_plot
+
+dev.off()
+

### 2017-06-18 code/random_forest.R

@@ -207,8 +207,8 @@ par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
 plot(cv10f_roc_muc,col = 'green4', lwd=3, add=T, lty=1) #r vs l mucosa cross validation
 plot(cv10f_roc_lum, col = 'orange', lwd=3, add=T, lty=1) #r vs l lumen cross validation
-mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.5)
-mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.5)
+mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.2)
+mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.2)
 legend('bottom', legend=c(sprintf('L mucosa vs R mucosa 10-fold CV, AUC = 0.912'),
                           sprintf('L lumen vs R lumen 10-fold CV, AUC = 0.7551')
                           # sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
@@ -422,11 +422,11 @@ legend('topright', legend=c("Left lumen", "Right lumen"), pch=c(21, 21), pt.bg=c
 
 
 
-# build and export figure 
+######################################################## build and export figure 
 #export as PDF
 
-plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
-pdf(file=plot_file, width=12, height=11)
+plot_file <- '~/Documents/Schloss_Lab/KWS_Project/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
+pdf(file=plot_file, width=9, height=12)
 layout(matrix(c(1,
                 2,
                 3), 
@@ -437,16 +437,123 @@ par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
[…]

### 2017-06-18 code/oncogenic_otus.R
finished exporting figures

@@ -33,7 +33,13 @@ fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient"
 fuso179[,5] <- (fuso179[,4]/4321)*100
 names(fuso179)[5] <- "Otu00179_relAbund"
 
-ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(width=0.3) +theme_bw()
+positions <- c("RB", "RS", "LB", "LS", "SS")
+fuso_plot <- ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(width=0.3) +theme_bw() +
+  ylab("F. nucleatum Relative Abundance") +
+  scale_x_discrete(limits = positions, breaks=positions, 
+                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
+
 
 #other oncogenic otus - Otu00152 is P. asaccharolytica. Otu00248 is p micra
 
@@ -51,6 +57,15 @@ ggplot(p_152, aes(x=location, y=Otu152_abund)) +geom_jitter() +theme_bw()
 
 ggplot(p_asach, aes(x=location, y=Otu00152_abund)) +geom_jitter() +theme_bw()
 
+positions <- c("RB", "RS", "LB", "LS", "SS")
+porphy_plot <- ggplot(p_152, aes(x=location, y=Otu152_abund)) +geom_jitter(width=0.3) +theme_bw() +
+  ylab("P. asacharolytica Relative Abundance") +
+  scale_x_discrete(limits = positions, breaks=positions, 
+                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
+
+
+
 p_micra <- shared_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00248")]
 
 p_micra[,4] <- (p_micra[,3]/4321) *100
@@ -61,18 +76,15 @@ ggplot(p_micra, aes(x=location, y=Otu00248_abund)) +geom_jitter() +theme_bw()
 ########################################Build figure 6 
 #export as PDF
 
[…]

### 2017-06-19 submission/manuscript.Rmd
almost done
final tweaks

@@
 ---
-title: '*Specific taxa differentiate proximal and distal
-  microbiota in the unprepped human colon*'
+title: 'Specific taxa differentiate proximal and distal microbiota in the unprepped
+  human colon'
 csl: mbio.csl
@@
 The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left.
-The models performed well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively).
+The models performed well when classifying these samples (`r cv10f_roc_right_bs$auc` and `r cv10f_roc_left_bs$auc`, respectively).
 The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
@@
 
-### Figure legends
+\newpage
+
+### Figures
 
@@
 
+\newpage
+
 ####Figure 2
@@
 
+\newpage
+
 ####Figure 3
@@
 
+\newpage
+
 ####Figure 4
@@
 (B) Top five OTUs that are most important for the classification model for the left mucosa and lumen (B) and the right mucosa and lumen (C).
+
+\newpage
 
@@
 
+\newpage
+
 ####Figure 6

### 2017-06-19 submission/manuscript.tex
final tweaks

@@
 \hypersetup{unicode=true,
-            pdftitle={**Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota*},
+            pdftitle={Specific taxa differentiate proximal and distal microbiota in the unprepped human colon},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{**Machine learning identifies unique taxa differentiating proximal and
-distal human colonic microbiota*}
+  \title{\emph{Specific taxa differentiate proximal and distal microbiota in the
+unprepped human colon}}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-Colorectal cancer (CRC) remains a leading cause of death worldwide.
-Tumors of the proximal (right) and distal (left) colon are
-morphologically and genetically distinct. Previous work from our group
-found that microbial dysbiosis is associated with the development of CRC
-tumors in studies of both mice and humans. Analysis of the fecal
-microbiota from healthy and CRC patients further revealed different
-microbial signatures associated with disease. In this study, we extended
-our observations of the fecal microbiome to analysis of the proximal and
-distal human colon. We used a two-colonoscope approach on subjects that
-had not undergone standard bowel preparation procedure. This technique
-allowed us to characterize the native proximal and distal luminal and
-mucosal microbiome without prior chemical disruption. 16S rRNA gene
-sequencing was performed on proximal and distal mucosal and luminal
-biopsies and exit stool for 20 healthy individuals. Diversity analysis
-revealed that each site contained a diverse community, and that a
-patient's samples were more similar to each other than to that of other
-individuals. Since we could not differentiate sites along the colon
-based on community structure or community membership alone, we employed
-the machine-learning algorithm Random Forest to identify key species
-that distinguish biogeographical sites. Random Forest classification
-models were built using taxa abundance and sample location and revealed
-distinct populations that were found in each location.
-\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
-\emph{Actinomyces} were most likely to be found in mucosal samples
-versus luminal samples (AUC = 0.925). The classification model performed
-well (AUC = 0.912) when classifying mucosal samples into proximal or
-distal sides, but separating luminal samples from each side proved more
-challenging (AUC = 0.755). The left mucosa was found to have high
-populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
-Proximal and distal luminal samples were comprised of many of the same
-taxa, likely reflecting the fact that stool moves along the colon from
-the proximal to distal end. Finally, comparison of all samples to fecal
-samples taken at exit uncovered that the feces were most similar to
-samples taken from the left lumen, again reflecting the anatomical
-structure of the colon. Taken together, our results have identified
-distinct bacterial populations of the proximal and distal colon. Further
-investigation of these bacteria may elucidate if and how these groups
-contribute to differential oncogenesis processes on the respective sides
-of the colon.
+Colorectal cancer (CRC) remains a leading cause of death worldwide
+despite improved preventative and therapeutic measures. Tumors of the
+proximal (right) and distal (left) colon are morphologically and
+genetically distinct. Previous work from our group found that microbial
+dysbiosis is associated with the development of CRC tumors in studies of
+both mice and humans. Analysis of the fecal microbiota from healthy and
+CRC patients further revealed different microbial signatures associated
+with disease. We extended our observations of the fecal microbiome to
+include analysis of the proximal and distal human colon. We used a
+two-colonoscope approach on subjects that had not undergone standard
+bowel preparation procedure. This technique allowed us to characterize
+the native proximal and distal luminal and mucosal microbiome without
+prior chemical disruption. 16S rRNA gene sequencing was performed on
+proximal and distal mucosal and luminal biopsies and home-collected
+stool for 20 healthy individuals. Diversity analysis revealed that each
+site contained a diverse community, and that a patient's samples were
+more similar to each other than to that of other individuals. Since we
+could not differentiate sites along the colon based on community
+structure or community membership alone, we employed the Random Forest
+machine-learning algorithm to identify key species that distinguish
+biogeographical sites. Random Forest classification models were built
+using taxa abundance and sample location and revealed distinct
+populations that were found in each location. \emph{Peptoniphilus,
+Anaerococcus, Enterobacteraceae, Pseudomonas} and \emph{Actinomyces}
+were most likely to be found in mucosal samples versus luminal samples
+(AUC = 0.925). The classification model performed well (AUC = 0.912)
+when classifying mucosal samples into proximal or distal sides, but
+separating luminal samples from each side proved more challenging (AUC =
+0.755). The left mucosa was found to have high populations of
+\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
+distal luminal samples were comprised of many of the same taxa, likely
+reflecting the fact that stool moves along the colon from the proximal
+to distal end. Finally, comparison of all samples to fecal samples taken
+at exit uncovered that the feces were most similar to samples taken from
+the left lumen, again reflecting the anatomical structure of the colon.
+By sampling the unprepped human colon, our results have identified
+distinct bacterial populations native to the proximal and distal sides.
+Further investigation of these bacteria may elucidate if and how these
+groups contribute to differential oncogenesis processes on the
+respective sides of the colon.
 
@@
 the colon are infiltrating lesions that present with painful symptoms.
-In contrast, right-sided colon tumors are bulky and project into the
-lumen, often causing no symptoms until they have developed into
-carcinomas ((2)). Because of the absence of symptoms, right-sided
-tumors, which comprise up to 47\% of CRCs, have a significantly higher
-mortality rate ((2)). The left and right sides of the colon differ in
-the amount of inflammation present and the genomic instability of
-precancerous cells, respectively, as well as oxygen, pH and the presence
-of antimicrobial peptides ((3), (4), (5)). Microenvironments differ not
-only longitutinally along the colon, but latitudinally from the
-epithelium to mucosa to intestinal lumen, offering several sites for
-different microbial communities to flourish. Given this varied
-physiology of the proximal-distal axis of the colon, symbiotic microbes
-and their metabolites likely vary between sites.
+In contrast, 47\% of CRCs are caused by right-sided colon tumors that
+are bulky and project into the lumen, often remaining asymptomatic until
+advanced carcinogenesis ((2)). Due to the absence of symptoms,
+right-sided tumors have a significantly higher mortality rate ((2)). The
+left and right sides of the colon differ in the amount of inflammation
+present and the genomic instability of precancerous cells, respectively,
+as well as oxygen, pH and the presence of antimicrobial peptides ((3),
+(4), (5)). Microenvironments differ not only longitutinally along the
+colon, but latitudinally from the epithelium to mucosa to intestinal
+lumen, offering several sites for different microbial communities to
+flourish. Given this varied physiology of the proximal-distal axis of
+the colon, symbiotic microbes and their metabolites likely vary between
+sites.
 
@@
 with those found on nearby healthy tissue has also identified specific
-bacterial species that are tumor-associated based on FISH staining and
-16S rRNA sequencing ((8)). These species include the oral pathogens
-\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
-Interestingly, the periodontal pathogens are also highly predictive of
-if a patient has CRC tumors or not in our classification studies ((9)).
-However, these studies have been performed comparing shed stool or in
-mice, which does not allow for analysis of paried samples from the
-proximal and distal sides of the colon. Similarly, comparisons of on- or
-off-tumor bacteria rarely have matched tissue from the other side of the
-colon from the same patient, limiting what conclusions can be drawn
-about the colonic microbiome overall, let alone at that specific site.
-Disruption of the healthy community could promote the initiation or
-proliferation of the distinct left and right CRC tumors. However, the
-contribution of the gut microbiota to these subtypes is largely
-undefined.
+bacterial species that are tumor-associated ((8)). These species include
+the oral pathogens \emph{Fusobacterium nucleatum} and
+\emph{Porphyromonas asacharolytica}. Interestingly, these periodontal
+pathogens have been highly predictive of whether a patient had CRC
+tumors or not in our classification studies ((9)). Because these studies
+were performed in mice or examined only shed human stool, they were
+unable to analyze paried samples from the proximal and distal sides of
+the colon. Similarly, comparisons of on- or off-tumor bacteria rarely
+have matched tissue from the other side of the colon from the same
+patient, limiting what conclusions can be drawn about the colonic
+microbiome overall, let alone at that specific site. Due to these
+limitations, the contribution of the gut microbiota to these subtypes is
+largely undefined. Characterizing these communities could provide needed
+insight into CRC etiology, including how the disruption of the healthy
+community could promote the initiation or proliferation of the distinct
+left and right CRC tumors.
 
-Furthermore, profiles of the microbial biogeography of the gut have been
-limited by sample collection methods. The majority of human gut
-microbiome studies have been performed on whole shed stool or on samples
-collected during colonoscopy procedures. While the latter method allows
-investigators to acquire samples from inside the human colon, typically
-this procedure is preceded by the use of bowel preparation methods such
-as the consumption of laxatives to cleanse the bowel. Bowel preparation
-is essential for detecting cancerous or precancerous lesions in the
-colon, but complicates microbiome profiling as the chemicals strip the
-bowel of contents and disrupt the mucosal layer ((10), (11)). As such,
-what little information we do have about the biogeographical
-distribution of the microbes in the proximal and distal colon is
-confounded by the bowel preparation procedure.
+Further, the few existing profiles of the microbial biogeography of the
+gut have been limited by sample collection methods. The majority of
+human gut microbiome studies have been performed on whole shed stool or
+on samples collected during colonoscopy procedures. While the latter
+method allows investigators to acquire samples from inside the human
+colon, typically this procedure is preceded by the use of bowel
+preparation methods such as the consumption of laxatives to cleanse the
+bowel. Bowel preparation is essential for detecting cancerous or
+precancerous lesions in the colon, but complicates microbiome profiling
+as the chemicals strip the bowel of contents and disrupt the mucosal
+layer ((10), (11)). As such, what little information we do have about
+the biogeographical distribution of the microbes in the proximal and
+distal colon is confounded by the bowel preparation procedure.
 
-In this study we sought to combine these approaches to characterize the
-microbiome in the lumen and mucosa of the proximal and distal healthy
-human colon. Our design uses unprepared colonoscopy techniques to sample
-each location of the gut without prior disruption of the native flora in
-20 healthy volunteers. To get around the noise created by a diverse set
-of human microbiomes, we used a machine-learning classification
-algorithm trained on curated 16S rRNA sequencing reads to identify
-microbes specific to each location. We found that our classification
-models were able to separate mucosal and lumenal samples as well as
-differentiate between sides of the colon based on populations of
-specific microbes. By identifying the specific microbes we are poised to
-ask if and how the presence or disruption of the microbes at each site
-contribute to the development of the specific tumor subtypes of CRC in
-the proximal and distal human colon.
+We combined these approaches to address the limitations of previous
+studies and effectively characterize the microbiome in the lumen and
+mucosa of the proximal and distal healthy human colon. Our design used
+unprepared colonoscopy techniques to sample each location of the gut
+without prior disruption of the native bacteria in 20 healthy
+volunteers. To address noise created by a diverse set of human
+microbiomes, we used a machine-learning classification algorithm trained
+on curated 16S rRNA sequencing reads to identify microbes specific to
+each location. We found that our classification models were able to
+separate mucosal and lumenal samples as well as differentiate between
+sides of the colon based on populations of specific microbes. By
+identifying the specific microbes we are poised to ask if and how the
+presence or disruption of the microbes at each site contribute to the
+development of the specific tumor subtypes of CRC in the proximal and
+distal human colon.
 
@@
 \emph{Bacteriodetes}, consistent with known variability in human
-microbiome research. Likewise, samples had varying levels of diversity
-at each site, irrespective of the individual (Figure 2B). That is, while
-some individuals had a more diverse right mucosa, some had a more
-diverse left mucosa, and so on. Therefore we could not identify a clear
+microbiome research (cite). Likewise, samples had varying levels of
+diversity at each site, irrespective of the individual (Figure 2B). That
+is, while some individuals had a more diverse right mucosa, some had a
+more diverse left mucosa. Therefore we could not identify a clear
 pattern of changes in microbial diversity along the gut axis.
@@
 compared distances for all individuals. Again, across all patient
-samples we saw a range of theta YC distances when comparing sample
+samples we observed a range of theta YC distances when comparing sample
 locations (Figure 3A) and again those ranges did not follow a clear
@@
 the right lumen and mucosa were most similar to each other than the
-other samples (P \textless{} 0.005, Wilcoxon, BH adjustment). Next, we
-calculated thetaYC distances to examine how each sample compared to the
-home-collected exit stool. Amidst variability between patients, we did
-identify significantly smaller thetaYc distance between the left lumenal
-sample and the exit stool (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
-adjustment). Further, there was an even larger difference in the
-comparisons of the left mucosa to the exit stool, indicating that the
-mucosa is quite different from the stool as compared to lumen (P
-\textless{} 0.0005, Wilcoxon, BH adjustment). To determine what factors
-may be driving the differences seen among the samples, we compared
-thetaYC distances between samples from all subjects (interpersonal)
-versus samples from within one subject (intrapersonal). We found that
-samples from one individual were overall much more similar to each other
-than to other study subjects (Figure 3C), consistent with previous human
-microbiome studies that have sampled multiple sites of the human colon
-(\textbf{???}). Thus interpersonal variation between subjects drives the
-differences between samples more than sample site or location. Overall,
-the results of the similarity analysis suggest that the contents of the
-left lumen are most representative of stool at exit, and the microbes
-remaining on the mucosa are much different.
+other samples (P \textless{} 0.005, Wilcoxon, BH adjustment).
+
+Next, we calculated thetaYC distances to examine how each sample
+compared to the home-collected exit stool. Amidst variability between
+patients, we did identify significantly smaller thetaYc distance between
+the left lumenal sample and the exit stool (Figure 3B, P \textless{}
+0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger
+difference in the comparisons of the left mucosa to the exit stool,
+indicating that the mucosa is different from the stool as compared to
+lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To determine what
+factors may be driving the differences seen among the samples, we
+compared thetaYC distances between samples from all subjects
+(interpersonal) versus samples from within one subject (intrapersonal).
+We found that samples from one individual were overall much more similar
+to each other than to other study subjects (Figure 3C), consistent with
+previous human microbiome studies that have sampled multiple sites of
+the human colon (\textbf{???}, (12), (13)). Thus interpersonal variation
+between subjects drives the differences between samples more than sample
+site or location. Overall, the results of the similarity analysis
+suggest that the contents of the left lumen are most representative of
+stool at exit, and the microbes remaining on the mucosa are much
+different.
 
@@
 To identify OTUs that were distinct at each biogeographical site, we
-constructed several Random Forest machine-learning models built on OTU
-abundances. The first model we built to classify the lumen versus
-mucosal samples for the right and left sides, independently (Figure 4A).
-The constructed model used ((Xopt)) features for the right and ((Xopt))
-for the left. The models performed well when classifying these samples
-(AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most
-predictive of each site are identified by their greatest mean decrease
-in accuracy when removed from the model. For distinguishing the right
-lumen and mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
+constructed several Random Forest models built on OTU abundances. We
+built the first model to classify the lumen versus mucosal samples for
+the right and left sides, independently (Figure 4A). The constructed
+model used ((Xopt)) features for the right and ((Xopt)) for the left.
+The models performed well when classifying these samples (AUC = 0.797
+and AUC = 0.980, respectively). The OTUs that were most predictive of
+each site are identified by their greatest mean decrease in accuracy
+when removed from the model. For distinguishing the right lumen and
+mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
 \emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
@@
 Given that specific bacterial species have been associated with
-colorectal cancer we probed our sample set for these OTUs. Among our 100
-samples, the most frequent sequence associated wtih the
-\emph{Fusobacterium} genus was OTU179, which BLASTs to
+colorectal cancer, we probed our sample set for these OTUs. Among our
+100 samples, the most frequent sequence associated wtih the
+\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
 \emph{Fusobacterium nucleatum subsp animalis}. It was the only sequence
@@
 \emph{Fusobacterium} known to have oncogenic properties and be found on
-the surfaces of colorectal cancer tumors. ((12)). The
+the surfaces of colorectal cancer tumors. ((14)). The
 \emph{Fusobacterium} positive samples were located on the right and left
@@
 When examining the relative abundance of the different phyla at each
-site, there is a wide amount of variation for each phyla with
+site, there was a wide amount of variation for each phyla with
 communities primarily dominated by the Bacteriodes and Firmicutes. This
@@
 location can be vastly different when the samples are taken just 1 cm
-away from each other ((13)). Similar patchiness is also observed in
+away from each other ((15)). Similar patchiness is also observed in
 lumenal contents and fecal samples themselves; there is observed
 separation of different interacting microbes along the length of a stool
-sample, for instance ((14)). That said, across our samples the mucosal
+sample, for instance ((16)). That said, across our samples the mucosal
 samples harbor more Proteobacteria, consistent with previous studies
@@
 differences in sampling and patchiness between subjects.
-
-We identified a range of Simpson diversity values between patients and
-between each anatomical site. As this measurement takes both number of
-species and abundance of species into account, it seems that there is a
-varied level of each component along the human colon. There was no clear
-pattern of diversity when comparing patients- some patients had a higher
-level of diversity in the proximal than the distal colon, some patients
-had the opposite. An early study of microbial diversity and membership
-of three human subjects found a varied level of diversity in each
-subject, but that subjects were overall more similar to themselves than
-to each other (\textbf{???}). We used the thetaYC dissimilarity metric
-to compare the communities present at each anatomical location to each
-other. We found that while there was a wide range of similarity of each
-site within a person, the study participants were ultimately more
-similar to themselves than all of the other participants. This finding
-is consistent with numerous studies examining the biogeography of the
-colon in healthy humans (\textbf{???}, (15), (16)).
 
@@
 
-We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 7 and 5
+We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
 of our subjects, respectively. These bacteria have been shown to be
@@
 for \emph{F. nucleatum} in this study, the bacteria was spread across
-the right mucosa, left lumen and left mucosa. That said, in a previous
-study of 1,102 colorectal cancer cases, only 138 (11\%) of them
-contained detectable \emph{F. nucleatum} at all ((18)). In a separate
-study of patients with inflammatory bowel disease (IBD), 50\% of 22 IBD
-patients and 17.6\% of 34 people without disease contained \emph{F.
-nucleatum} in mucosal biopsies ((19)). These studies have the added
-caveat of being sampled from prepared or cleaned colons. Thus, the
-presence of \emph{F. nucleatum} in a healthy individual is not
-necessarily linked to the development of future colorectal cancers.
-Because of the spatial distribution of the \emph{F. nucleatum} in our
-sample set, we cannot develop a model for the role of \emph{F.
-nucleatum} in the healthy colon. Indeed, rather than the \emph{F.
-nucleatum} initiating tumorigenesis, these bacteria may instead arrive
-``late'' to the tumor microenvironment and feed on tumors that have
-alread begun to develop ((20)). Data examining bacterial biofilms on CRC
-tumors suggests that \emph{Fusobacteria} species are more commonly found
-both on proximal tumors and in biofilms, indicating that it is not only
-the presence of the bacteria but the organization of the tumor community
-that contributes to \emph{Fusobacterium's} role in tumorigenesis.
+the right mucosa, left lumen and left mucosa. The presence of \emph{F.
+nucleatum} in a healthy individual is not necessarily linked to the
+development of future colorectal cancers. Because of the spatial
+distribution of the \emph{F. nucleatum} in our sample set, we cannot
+develop a model for the role of \emph{F. nucleatum} in the healthy
+colon. Data examining bacterial biofilms on CRC tumors suggests that
+\emph{Fusobacteria} species are more commonly found both on proximal
+tumors and in biofilms, indicating that it is not only the presence of
+the bacteria but the organization of the tumor community that
+contributes to \emph{Fusobacterium's} role in tumorigenesis ((8)).
 Finally, \emph{Fusobacteria} and \emph{Porphyromonas} species have been
 known to not only co-occur on CRC tumors but also to synergistically
-promote tumorigenesis in an oral cancer model ((21), (20)). Thus,
-further analysis of the distribution and role of these pathogens may
-elucidate a mechanism for development of CRC tumors in the proximal or
-distal colon.
+promote tumorigenesis in an oral cancer model ((19), (20)). Thus,
+further analysis of the distribution and activities of these pathogens
+may elucidate a mechanism for development of CRC tumors in the proximal
+or distal colon.
 
@@
 matched-colonoscopy study that sampled 18 patient's colonic mucosa and
-lumenal contents prior to and after bowel cleansing ((22)). This group
+lumenal contents prior to and after bowel cleansing ((21)). This group
 found that mucosa and lumenal samples were distinguishable prior to
 bowel cleansing, but that bowel preparation resulted in an increase in
-shared OTUs between each site ((22)). Bowel cleansing not only made the
+shared OTUs between each site ((21)). Bowel cleansing not only made the
 samples harder to distinguish, it resulted in decreases in diversity
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described ((23)). Sequences were curated
-using the mothur software as described previously ((24)). The sequences
+Illumina MiSeq as previously described ((22)). Sequences were curated
+using the mothur software as described previously ((23)). The sequences
 were assigned taxonomic classification using a naive Bayesian classifier
 trained using a 16S rRNA gene training set from the Ribosomal Database
-Project (RDP) ((25)) and clustered into operational taxonomic units
+Project (RDP) ((24)) and clustered into operational taxonomic units
 (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis of a
@@
 Diversity analysis was performed using the Simpson diversity calculator
-and theta YC calculator metrics in mothur ((24)). ThetaYC distances were
+and theta YC calculator metrics in mothur ((23)). ThetaYC distances were
 calculated to determine the dissimilarity between two samples. Random
@@
 and resultant models were used to identify the OTUs that were most
-important for classifying each location ((26)). To get species-level
+important for classifying each location ((25)). To get species-level
 information about sequences of interest, sequences were aligned using
@@
 
-\subsubsection{Figure legends}\label{figure-legends}
+\newpage
+
+\subsubsection{Figures}\label{figures}
+
+\includegraphics{submission/figure_1.pdf}
 
@@
 
+\newpage
+
+\includegraphics{submission/figure_2.pdf}
+
 \paragraph{Figure 2}\label{figure-2}
@@
 
+\newpage
+
+\includegraphics{submission/figure_3.pdf}
+
 \paragraph{Figure 3}\label{figure-3}
@@
 each site to the exit stool are shown.
+
+\newpage
+
+\includegraphics{submission/figure_4.pdf}
 
@@
 
+\newpage
+
+\includegraphics{submission/figure_5.pdf}
+
 \paragraph{Figure 5}\label{figure-5}
@@
 left and right mucosa (B) and the left and right lumen (C).
+
+\newpage
+
+\includegraphics{submission/figure_6.pdf}
 
@@
 
+\hypertarget{ref-deCarcer2010}{}
+12. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
+\textbf{Kang S}, \textbf{Worthley D}, \textbf{Whitehall V},
+\textbf{Gordon I}, \textbf{McSweeney C}, \textbf{Leggett B},
+\textbf{Morrison M}. 2010. Numerical ecology validates a biogeographical
+distribution and gender-based effect on mucosa-associated bacteria along
+the human colon. The ISME Journal \textbf{5}:801--809.
+doi:\href{https://doi.org/10.1038/ismej.2010.177}{10.1038/ismej.2010.177}.
+
+\hypertarget{ref-Zhang2013}{}
+13. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
+\textbf{Xu J}, \textbf{Wen X}, \textbf{Ma Z (Sam)}, \textbf{Shi P}.
+2013. Spatial heterogeneity and co-occurrence patterns of human
+mucosal-associated intestinal microbiota. The ISME Journal
+\textbf{8}:881--893.
+doi:\href{https://doi.org/10.1038/ismej.2013.185}{10.1038/ismej.2013.185}.
+
 \hypertarget{ref-Castellarin2011}{}
-12. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
+14. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
 \textbf{Dreolini L}, \textbf{Krzywinski M}, \textbf{Strauss J},
@@
 \hypertarget{ref-Hong2011}{}
-13. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
+15. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
 \textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
@@
 \hypertarget{ref-Stearns2011}{}
-14. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
+16. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
 \textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
@@
 doi:\href{https://doi.org/10.1038/srep00170}{10.1038/srep00170}.
-
-\hypertarget{ref-deCarcer2010}{}
-15. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
-\textbf{Kang S}, \textbf{Worthley D}, \textbf{Whitehall V},
-\textbf{Gordon I}, \textbf{McSweeney C}, \textbf{Leggett B},
-\textbf{Morrison M}. 2010. Numerical ecology validates a biogeographical
-distribution and gender-based effect on mucosa-associated bacteria along
-the human colon. The ISME Journal \textbf{5}:801--809.
-doi:\href{https://doi.org/10.1038/ismej.2010.177}{10.1038/ismej.2010.177}.
-
-\hypertarget{ref-Zhang2013}{}
-16. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
-\textbf{Xu J}, \textbf{Wen X}, \textbf{Ma Z (Sam)}, \textbf{Shi P}.
-2013. Spatial heterogeneity and co-occurrence patterns of human
-mucosal-associated intestinal microbiota. The ISME Journal
-\textbf{8}:881--893.
-doi:\href{https://doi.org/10.1038/ismej.2013.185}{10.1038/ismej.2013.185}.
 
@@
 
-\hypertarget{ref-Strauss2011}{}
-19. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
-\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
-\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
-gut mucosa-derived fusobacterium nucleatum positively correlates with
-IBD status of the host. Inflammatory Bowel Diseases
-\textbf{17}:1971--1978.
-doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.
+\hypertarget{ref-Whitmore2014}{}
+19. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
+cancer. PLoS Pathogens \textbf{10}:e1003933.
+doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.
 
@@
 
-\hypertarget{ref-Whitmore2014}{}
-21. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
-cancer. PLoS Pathogens \textbf{10}:e1003933.
-doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.
-
 \hypertarget{ref-Shobar2016}{}
-22. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
+21. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
 \textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
@@
 \hypertarget{ref-Kozich2013}{}
-23. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+22. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Schloss2009}{}
-24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+23. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Wang2007}{}
-25. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+24. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
@@
 \hypertarget{ref-Liaw2002}{}
-26. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+25. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News: The Newsletter of the R Project

### 2017-06-20 submission/manuscript.Rmd
updated with pdf

@@
 ---
-title: 'Specific taxa differentiate proximal and distal microbiota in the unprepped
-  human colon'
+title: Specific taxa differentiate proximal and distal microbiota in the unprepped
+  human colon
 csl: mbio.csl
@@
 output:
-  html_document: default
   pdf_document:
@@
     keep_tex: yes
+    latex_engine: xelatex
+  html_document: default
   word_document: default
@@
 bibliography: references.bib
+header-includes:
+- \usepackage{setspace}
+- \doublespacing
+- \usepackage{lineno}
+- \linenumbers
 ---
@@
 \newpage
-\linenumbers
 
@@
 
-```{r dependencies, echo=FALSE, warning = FALSE}
+```{r dependencies, echo=FALSE, warning = FALSE, message=FALSE}
 pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')

### 2017-06-20 submission/manuscript.tex
updated with pdf

@@
 \setlength{\droptitle}{-2em}
-  \title{\emph{Specific taxa differentiate proximal and distal microbiota in the
-unprepped human colon}}
+  \title{Specific taxa differentiate proximal and distal microbiota in the
+unprepped human colon}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
   \predate{}\postdate{}
+
+\usepackage{setspace}
+\doublespacing
+\usepackage{lineno}
+\linenumbers
 
@@
 \newpage
-
-\linenumbers
-
-\begin{verbatim}
-## randomForest 4.6-12
-\end{verbatim}
-
-\begin{verbatim}
-## Type rfNews() to see new features/changes/bug fixes.
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'ggplot2'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:randomForest':
-## 
-##     margin
-\end{verbatim}
-
-\begin{verbatim}
-## Type 'citation("pROC")' for a citation.
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'pROC'
-\end{verbatim}
-
-\begin{verbatim}
-## The following objects are masked from 'package:stats':
-## 
-##     cov, smooth, var
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'dplyr'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:randomForest':
-## 
-##     combine
-\end{verbatim}
-
-\begin{verbatim}
-## The following objects are masked from 'package:stats':
-## 
-##     filter, lag
-\end{verbatim}
-
-\begin{verbatim}
-## The following objects are masked from 'package:base':
-## 
-##     intersect, setdiff, setequal, union
-\end{verbatim}
-
-\begin{verbatim}
-## AUCRF 1.1
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: lattice
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'reshape2'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:tidyr':
-## 
-##     smiths
-\end{verbatim}
 
@@
 model used ((Xopt)) features for the right and ((Xopt)) for the left.
-The models performed well when classifying these samples (AUC = 0.797
-and AUC = 0.980, respectively). The OTUs that were most predictive of
-each site are identified by their greatest mean decrease in accuracy
-when removed from the model. For distinguishing the right lumen and
-mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
-\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
-were differentially abundant (Figure 4B). The model classifying the left
-lumen and mucosa identified OTUs from \emph{Turicibacter},
-\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
-\emph{Anaerococcus} genera that could distinguish lumen from mucosal
-samples (Figure 4C). These results indicate that there are fine
-differences between the different sites of the colon, and that these can
-be traced down to specific OTUs on each side.
+The models performed well when classifying these samples (0.8 and 1.0,
+respectively). The OTUs that were most predictive of each site are
+identified by their greatest mean decrease in accuracy when removed from
+the model. For distinguishing the right lumen and mucosa, OTUs from the
+\emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and two OTUs
+from the \emph{Enterobacteraceae} genera were differentially abundant
+(Figure 4B). The model classifying the left lumen and mucosa identified
+OTUs from \emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus}
+and two OTUs from the \emph{Anaerococcus} genera that could distinguish
+lumen from mucosal samples (Figure 4C). These results indicate that
+there are fine differences between the different sites of the colon, and
+that these can be traced down to specific OTUs on each side.
 
@@
 
-\includegraphics{submission/figure_1.pdf}
-
 \paragraph{Figure 1}\label{figure-1}
@@
 
-\includegraphics{submission/figure_2.pdf}
-
 \paragraph{Figure 2}\label{figure-2}
@@
 
-\includegraphics{submission/figure_3.pdf}
-
 \paragraph{Figure 3}\label{figure-3}
@@
 \newpage
-
-\includegraphics{submission/figure_4.pdf}
 
@@
 
-\includegraphics{submission/figure_5.pdf}
-
 \paragraph{Figure 5}\label{figure-5}
@@
 \newpage
-
-\includegraphics{submission/figure_6.pdf}
 

### 2017-06-27 submission/manuscript.Rmd
added in some Pat edits

@@
 stooltyc <- subset(tyc, match=='LB_RB' | match== 'LS_RS')
-leftandrighttyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
-lvsr <- rbind(stooltyc, leftandrighttyc)
+distalandproximaltyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
+lvsr <- rbind(stooltyc, distalandproximaltyc)
 
@@
 #build randomForest model for each location comparison using randomize_loc function 
-rf_left <- randomize_loc(rel_meta, "LB", "LS") #OOB 10.26%
-rf_right <- randomize_loc(rel_meta, "RB", "RS") #OOB 53%
+rf_distal <- randomize_loc(rel_meta, "LB", "LS") #OOB 10.26%
+rf_proximal <- randomize_loc(rel_meta, "RB", "RS") #OOB 53%
 rf_bowel <- randomize_loc(rel_meta, "LB", "RB") #OOB 25.64%
@@
 # create RF model with AUCRF outputs top OTUs
-aucrf_data_left_bs <- auc_loc(rel_meta, "LB", "LS")
+aucrf_data_distal_bs <- auc_loc(rel_meta, "LB", "LS")
 aucrf_data_LRbowel <- auc_loc(rel_meta, "LB", "RB")
-aucrf_data_right_bs <- auc_loc(rel_meta, "RB", "RS")
+aucrf_data_proximal_bs <- auc_loc(rel_meta, "RB", "RS")
 aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
@@
 cv10f_aucs <- c()
-cv10f_all_resp_left_bs <- c()
-cv10f_all_pred_left_bs <- c()
+cv10f_all_resp_distal_bs <- c()
+cv10f_all_pred_distal_bs <- c()
 for(j in 1:iters){
   set.seed(j)
-  sampling <- sample(1:nrow(aucrf_data_left_bs),nrow(aucrf_data_left_bs),replace=F)
+  sampling <- sample(1:nrow(aucrf_data_distal_bs),nrow(aucrf_data_distal_bs),replace=F)
   cv10f_probs <- rep(NA,39)
   for(i in seq(1,36,4)){
-    train_left_bs <- aucrf_data_left_bs[sampling[-(i:(i+3))],]
-    test_left_bs <- aucrf_data_left_bs[sampling[i:(i+3)],]
+    train_distal_bs <- aucrf_data_distal_bs[sampling[-(i:(i+3))],]
+    test_distal_bs <- aucrf_data_distal_bs[sampling[i:(i+3)],]
     set.seed(seed)
-    temp_model_left_bs <- AUCRF(location~., data=train_left_bs, pdel=0.99, ntree=500)
-    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_left_bs$RFopt, test_left_bs, type='prob')[,2]
+    temp_model_distal_bs <- AUCRF(location~., data=train_distal_bs, pdel=0.99, ntree=500)
+    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_distal_bs$RFopt, test_distal_bs, type='prob')[,2]
   }
-  cv10f_roc_left_bs <- roc(aucrf_data_left_bs$location~cv10f_probs)
-  cv10f_all_pred_left_bs <- c(cv10f_all_pred_left_bs, cv10f_probs)
-  cv10f_all_resp_left_bs <- c(cv10f_all_resp_left_bs, aucrf_data_left_bs$location)
-  cv10f_aucs[j] <- cv10f_roc_left_bs$auc #stores aucs for all iterations, can use to calc IQR
+  cv10f_roc_distal_bs <- roc(aucrf_data_distal_bs$location~cv10f_probs)
+  cv10f_all_pred_distal_bs <- c(cv10f_all_pred_distal_bs, cv10f_probs)
+  cv10f_all_resp_distal_bs <- c(cv10f_all_resp_distal_bs, aucrf_data_distal_bs$location)
+  cv10f_aucs[j] <- cv10f_roc_distal_bs$auc #stores aucs for all iterations, can use to calc IQR
 }
-cv10f_roc_left_bs <- roc(cv10f_all_resp_left_bs~cv10f_all_pred_left_bs)
+cv10f_roc_distal_bs <- roc(cv10f_all_resp_distal_bs~cv10f_all_pred_distal_bs)
 
@@
 cv10f_aucs <- c()
-cv10f_all_resp_right_bs <- c()
-cv10f_all_pred_right_bs <- c()
+cv10f_all_resp_proximal_bs <- c()
+cv10f_all_pred_proximal_bs <- c()
 for(j in 1:iters){
   set.seed(j)
-  sampling <- sample(1:nrow(aucrf_data_right_bs),nrow(aucrf_data_right_bs),replace=F)
+  sampling <- sample(1:nrow(aucrf_data_proximal_bs),nrow(aucrf_data_proximal_bs),replace=F)
   cv10f_probs <- rep(NA,39)
   for(i in seq(1,36,4)){
-    train_right_bs <- aucrf_data_right_bs[sampling[-(i:(i+3))],]
-    test_right_bs <- aucrf_data_right_bs[sampling[i:(i+3)],]
+    train_proximal_bs <- aucrf_data_proximal_bs[sampling[-(i:(i+3))],]
+    test_proximal_bs <- aucrf_data_proximal_bs[sampling[i:(i+3)],]
     set.seed(seed)
-    temp_model_right_bs <- AUCRF(location~., data=train_right_bs, pdel=0.99, ntree=500)
-    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_right_bs$RFopt, test_right_bs, type='prob')[,2]
+    temp_model_proximal_bs <- AUCRF(location~., data=train_proximal_bs, pdel=0.99, ntree=500)
+    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_proximal_bs$RFopt, test_proximal_bs, type='prob')[,2]
   }
-  cv10f_roc_right_bs <- roc(aucrf_data_right_bs$location~cv10f_probs)
-  cv10f_all_pred_right_bs <- c(cv10f_all_pred_right_bs, cv10f_probs)
-  cv10f_all_resp_right_bs <- c(cv10f_all_resp_right_bs, aucrf_data_right_bs$location)
-  cv10f_aucs[j] <- cv10f_roc_right_bs$auc #stores aucs for all iterations, can use to calc IQR
+  cv10f_roc_proximal_bs <- roc(aucrf_data_proximal_bs$location~cv10f_probs)
+  cv10f_all_pred_proximal_bs <- c(cv10f_all_pred_proximal_bs, cv10f_probs)
+  cv10f_all_resp_proximal_bs <- c(cv10f_all_resp_proximal_bs, aucrf_data_proximal_bs$location)
+  cv10f_aucs[j] <- cv10f_roc_proximal_bs$auc #stores aucs for all iterations, can use to calc IQR
 }
-cv10f_roc_right_bs <- roc(cv10f_all_resp_right_bs~cv10f_all_pred_right_bs)
+cv10f_roc_proximal_bs <- roc(cv10f_all_resp_proximal_bs~cv10f_all_pred_proximal_bs)
 
@@
 #just LB vs LS 
-left_otu_feat <- colnames(aucrf_data_left_bs[2:6])
-otu_taxa_left <- get_tax(1, left_otu_feat, tax_file)
+distal_otu_feat <- colnames(aucrf_data_distal_bs[2:6])
+otu_taxa_distal <- get_tax(1, distal_otu_feat, tax_file)
 #Abundance stripchart or most predictive otus
-ls_abunds <- shared_meta[shared_meta$location=='LS', left_otu_feat]/10000 + 1e-4
-lb_abunds <- shared_meta[shared_meta$location=='LB', left_otu_feat]/10000 + 1e-4
+ls_abunds <- shared_meta[shared_meta$location=='LS', distal_otu_feat]/10000 + 1e-4
+lb_abunds <- shared_meta[shared_meta$location=='LB', distal_otu_feat]/10000 + 1e-4
 
-right_otu_feat <- colnames(aucrf_data_right_bs[2:6])
-otu_taxa_right <- get_tax(1, right_otu_feat, tax_file)
+proximal_otu_feat <- colnames(aucrf_data_proximal_bs[2:6])
+otu_taxa_proximal <- get_tax(1, proximal_otu_feat, tax_file)
 #Abundance stripchart or most predictive otus
-rs_abunds <- shared_meta[shared_meta$location=='RS', right_otu_feat]/10000 + 1e-4
-rb_abunds <- shared_meta[shared_meta$location=='RB', right_otu_feat]/10000 + 1e-4
+rs_abunds <- shared_meta[shared_meta$location=='RS', proximal_otu_feat]/10000 + 1e-4
+rb_abunds <- shared_meta[shared_meta$location=='RB', proximal_otu_feat]/10000 + 1e-4
 
@@
 Colorectal cancer (CRC) remains a leading cause of death worldwide despite improved preventative and therapeutic measures.
-Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct.
+Tumors of the proximal (proximal) and distal (distal) colon are morphologically and genetically distinct.
 Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans.
@@
 The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755).
-The left mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_.
+The distal mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_.
 Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end.
-Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon.
+Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the distal lumen, again reflecting the anatomical structure of the colon.
 By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides.
@@
 CRC has recently been described to fall into four molecular subtypes (@Guinney2015).
-Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms.
-In contrast, 47% of CRCs are caused by right-sided colon tumors that are bulky and project into the lumen, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010).
-Due to the absence of symptoms, right-sided tumors have a significantly higher mortality rate (@Benedix2010).
-The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
+Of these subtypes, tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms.
+In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are bulky and project into the lumen, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010).
+Due to the absence of symptoms, proximal-sided tumors have a significantly higher mortality rate (@Benedix2010).
+The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
 Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
@@
 Due to these limitations, the contribution of the gut microbiota to these subtypes is largely undefined.
-Characterizing these communities could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.
+Characterizing these communities could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct distal and proximal CRC tumors.
 
@@
 
-We combined these approaches to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
-Our design used unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
-To address noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
+Here we aimed to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
+Our design used unprepared colonoscopy techniques to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
+To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
 We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes.
@@
 #### Microbial membership and diversity of the proximal and distal colon 
-Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation(Figure 1).
+Lumenal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1).
 Participants also collected stool at home one week prior to the procedure.
 To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
-The relative abundance of each phyla for each site is shown in Figure 2A.
-This data combines the samples from all participants in the study.
-Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research (cite).
+Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research (cite).
 Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
-That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa.
+For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others.
 Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.
 
-To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals.
-Again, across all patient samples we observed a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
-However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
+To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated ${\theta}$YC distances from OTU abundances and compared these distances for all individuals.
+Again, across all patient samples we observed a range of ${\theta}$YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
+However, when comparing median distances between the proximal lumen and mucosa, the proximal versus distal lumen, the proximal versus distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
 
-Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool.
-Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
-Furthermore, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
+#### Stool at exit most resembles lumenal samples from the distal colon 
+Next, we calculated ${\theta}$YC distances to examine how each sample compared to the home-collected exit stool.
+Amidst variability between patients, we did identify significantly smaller ${\theta}$YC distance between the distal lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
+Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
 To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005, @deCarcer2010, @Zhang2013).
+We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005, @deCarcer2010, @Zhang2013).
 Thus interpersonal variation between subjects drives the differences between samples more than sample site or location.
-Overall, the results of the similarity analysis suggest that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
+Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
 
 #### Random Forest classification models identify important OTUs on each side
-To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models built on OTU abundances.
-We built the first model to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A).
-The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left.
-The models performed well when classifying these samples (`r cv10f_roc_right_bs$auc` and `r cv10f_roc_left_bs$auc`, respectively).
+To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances.
+We built the first model to classify the lumen versus mucosal samples for the proximal and distal sides, independently (Figure 4A).
+The constructed model used ((Xopt)) features for the proximal and ((Xopt)) for the distal.
+The models performed well when classifying these samples (`r cv10f_roc_proximal_bs$auc` and `r cv10f_roc_distal_bs$auc`, respectively).
 The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
-For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B).
-The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C).
+For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B).
+The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C).
 These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.
 
-Next, we built a model to differentiate the left and right microbiome.
-The model performed best when distinguishing the right versus left mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen (AUC = 0.755).
-The model was able to explain ((X%)) of the variance.
-OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B).
-The model was less-effective at classifying the right and left lumen (AUC = 0.755).
-However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).
-This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also drive sidedness- such as _Anaerococcus_ and _Finegoldia_.
+Next, we built a model to differentiate the proximal and distal lumenal samples.
+The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal versus distal lumen (AUC = 0.755).
+These models were able to explain ((X%)) of the variance, respectively.
+OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B).
+Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).
+This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.
 
 #### Bacterial OTUs associated with cancer are found in healthy individuals 
-Given that specific bacterial species have been associated with colorectal cancer, we probed our sample set for these OTUs.
-Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_.
-It was the only sequence to align with a _F.
-nucleatum_ strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
+Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs.
+Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (XX% over full length).
+This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
 (@Castellarin2011).
-The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A).
-OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors.
-OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B).
-Samples that were positive for _P.
-asacharolytica_ ranged in abundances from 0.01% - 16%.
-Thus, cancer-associated OTUs could be found in our sample set of 20 healthy individuals.
+The _Fusobacterium_ positive samples were located in x% of the the proximal and X% of the distal mucosa and represented as much as 1% of any sample (Figure 6A).
+OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (X% over full length), another bacterium commonly detected and isolated from colorectal tumors.
+OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6B).
+Among the samples that were positive for _P.
+asacharolytica_, the relative abundances for this OTU ranged from 0.01% - 16%.
+Thus, disease-associated OTUs could be found in our sample set of 20 healthy individuals.
 
 ### Discussion
-Here we identified bacterial taxa that are specific to the lumen and mucosa of the right and left human colon from samples collected during unprepared colonoscopy.
+Here we identified bacterial taxa that are specific to the lumen and mucosa of the proximal and distal human colon from samples collected during unprepared colonoscopy.
 We found that all locations contained a range of phyla and a range of diversity, but that there was a wide variability between subjects.
-Pairwise comparisons of each of the sites revealed that the right mucosa and lumen were most similar to each other.
-Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the left lumen is most similar to stool at exit.
+Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other.
+Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the distal lumen is most similar to stool at exit.
 Random Forest algorithms built on OTU abundances from each sample identified microbes that are particular to each location of the colon.
@@
 Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (@Donaldson2015).
-For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon.
-The left mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_.
-The model was less effective at classifying the right and left lumenal contents, probably because the samples are arguably composed of the same material.
+For instance, the proximal mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon.
+The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_.
+The model was less effective at classifying the proximal and distal lumenal contents, probably because the samples are arguably composed of the same material.
 
@@
 nucleatum_ was found on both sides of the colon, _P.
-asacharolytica_ was only detected in the left mucosa.
+asacharolytica_ was only detected in the distal mucosa.
 Not much is known about the distribution of _P.
-asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (@Albenberg2014).
+asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa (@Albenberg2014).
 In studies examining bacteria on colorectal cancer tumors, _F.
-nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F.
+nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F.
 nucleatum_ decreases along the colon to rectum (@Mima2016).
 Of the (8) (40%) individuals positive for _F.
-nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa.
+nucleatum_ in this study, the bacteria was spread across the proximal mucosa, distal lumen and distal mucosa.
 The presence of _F.
@@
 One caveat of the FIT study was that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.
-Given that our results showed that the stool most accurately reflects the community of the left lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
+Given that our results showed that the stool most accurately reflects the community of the distal lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
 
@@
 Three mucosal biopsies were collected and flash-frozen in RNAlater.
-These samples comprised the distal or left colon samples.
+These samples comprised the distal or distal colon samples.
 The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the ascending colon.
@@
 ThetaYC distances are shown for interpersonal similarities between two sites – each point represents one individual.
-In (A), comparisons of the right and left mucosal and lumen are shown.
+In (A), comparisons of the proximal and distal mucosal and lumen are shown.
 In (B), comparisons of each site to the exit stool are shown.
@@
 Random Forest classifies the mucosa and lumen of each side of the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the left and right sides of the colon.
-(B) Top five OTUs that are most important for the classification model for the left mucosa and lumen (B) and the right mucosa and lumen (C).
+A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal and proximal sides of the colon.
+(B) Top five OTUs that are most important for the classification model for the distal mucosa and lumen (B) and the proximal mucosa and lumen (C).
 
@@
 ####Figure 5 
-Random Forest classifies the left and right sides of the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying left lumen versus right lumen (orange) and left mucosa vs right mucosa (green).
-(B) Top five OTUs that are most important for the classification model for the left and right mucosa (B) and the left and right lumen (C).
+Random Forest classifies the distal and proximal sides of the colon.
+A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal lumen versus proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
+(B) Top five OTUs that are most important for the classification model for the distal and proximal mucosa (B) and the distal and proximal lumen (C).
 

### 2017-06-28 submission/manuscript.Rmd
easy Pat edits complete

@@
     keep_tex: yes
-    latex_engine: xelatex
   html_document: default
@@
 ### Discussion
-Here we identified bacterial taxa that are specific to the lumen and mucosa of the proximal and distal human colon from samples collected during unprepared colonoscopy.
+Here we identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon from samples collected during unprepared colonoscopy.
 We found that all locations contained a range of phyla and a range of diversity, but that there was a wide variability between subjects.
@@
 Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the distal lumen is most similar to stool at exit.
-Random Forest algorithms built on OTU abundances from each sample identified microbes that are particular to each location of the colon.
-Finally, we were able to detect some bacterial OTUs associated with colorectal cancer in our healthy patient cohort.
+Random Forest models built on OTU relative abundances from each sample identified microbes that are particular to each location of the colon.
+Finally, we were able to detect some bacterial OTUs associated with colonic disease in our healthy patient cohort.
 Using unprepped colonoscopy and machine learning, we have identified bacterial phyla specific to the healthy proximal and distal human colon.
 
-When examining the relative abundance of the different phyla at each site, there was a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes.
-This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome.
-Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (@Hong2011).
-Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (@Stearns2011).
-That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014).
-Hence, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
+When examining the relative abundance of the different phyla at each site, there was a wide amount of variation for each phyla with communities primarily dominated by the _Bacteriodes_ and _Firmicutes_.
+This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of biogeogrphical "patchiness" in the gut microbiome.
+Several studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011.
+Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011.
+That said, across our samples the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to lumenal content in humans @Albenberg2014.
+Hence, the conclusions we can draw from phyla analysis are likely impacted by patchiness between subjects.
 
-To get around the noisiness created by a diverse set of samples, we built a Random Forest model to identify microbes specific to each side.
+To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side.
 For each comparison we identified top X OTUs that were strongly predictive of one site or another.
-Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (@Donaldson2015).
-For instance, the proximal mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon.
+Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015.
+For instance, the proximal mucosa contains the highest oxygen concentratons of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_.
 The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_.
-The model was less effective at classifying the proximal and distal lumenal contents, probably because the samples are arguably composed of the same material.
+The model was less effective at classifying the proximal and distal lumenal contents, probably because the samples are arguably composed of the same bacteria but differ in water content.
 
@@
 asacharolytica_ in 8 and 5 of our subjects, respectively.
-These bacteria have been shown to be predictive of colorectal cancer in humans (@Baxter2016) and have oncogenic properties in cell culture and in mice (@Sears2014).
+These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014.
 Interestingly, while _F.
@@
 nucleatum_ decreases along the colon to rectum (@Mima2016).
-Of the (8) (40%) individuals positive for _F.
-nucleatum_ in this study, the bacteria was spread across the proximal mucosa, distal lumen and distal mucosa.
+Of the 8 (40%) individuals positive for _F.
+nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa.
 The presence of _F.
-nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers.
-Because of the spatial distribution of the _F.
-nucleatum_ in our sample set, we cannot develop a model for the role of _F.
-nucleatum_ in the healthy colon.
-Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014).
-Finally, _Fusobacteria_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016).
+nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers (cite).
+Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014).
+Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016).
 Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.
@@
 By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect oncogenesis of proximal and distal CRC tumors.
-To this point, 16S community profiling studies do not provide enough information to probe these questions.
+To this point, 16S rRNA gene sequecing community profiling studies do not provide enough information to probe these questions.
 Our sample set of matched proximal, distal, lumenal and mucosal samples from colons that have not undergone bowel preparation presents a unique opportunity to explore further questions about the microbiome along the gut axis.
-Specifically, examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.
 
@@
 We would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman for their roles in study coordination, subject recruitment, sample collection and sample processing.
+<need grant # here>
 
@@
 At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit supplies).
-At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.) and ship the samples to the University.
+At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and ship the samples to the University on ice.
 Notably, subjects did not undergo any bowel preparation method prior to sampling.
-On procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System.
+On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System.
 Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate.
@@
 Sequencing and analysis of a mock community revealed the error rate to be X%.
-Samples were rarefied to 4231 sequences per sample in order to reduce sampling bias.
+Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.
 
-Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur (@Schloss2009).
+Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version <3.2?> (@Schloss2009).
 ThetaYC distances were calculated to determine the dissimilarity between two samples.
@@
 Differences in thetaYC distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
+
+#### Data availability
+
+#### Github repository
 
@@
 ####Figure 3
-Similarity of microbial community structure between sites of the gut.
+Distances of microbial community structure between sites of the gut.
 ThetaYC distances are shown for interpersonal similarities between two sites – each point represents one individual.

### 2017-06-28 submission/manuscript.tex
easy Pat edits complete

@@
 despite improved preventative and therapeutic measures. Tumors of the
-proximal (right) and distal (left) colon are morphologically and
+proximal (proximal) and distal (distal) colon are morphologically and
 genetically distinct. Previous work from our group found that microbial
@@
 separating luminal samples from each side proved more challenging (AUC =
-0.755). The left mucosa was found to have high populations of
+0.755). The distal mucosa was found to have high populations of
 \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
@@
 at exit uncovered that the feces were most similar to samples taken from
-the left lumen, again reflecting the anatomical structure of the colon.
-By sampling the unprepped human colon, our results have identified
-distinct bacterial populations native to the proximal and distal sides.
-Further investigation of these bacteria may elucidate if and how these
-groups contribute to differential oncogenesis processes on the
-respective sides of the colon.
+the distal lumen, again reflecting the anatomical structure of the
+colon. By sampling the unprepped human colon, our results have
+identified distinct bacterial populations native to the proximal and
+distal sides. Further investigation of these bacteria may elucidate if
+and how these groups contribute to differential oncogenesis processes on
+the respective sides of the colon.
 
@@
 the colon. CRC has recently been described to fall into four molecular
-subtypes ((1)). Of these subtypes, tumors that arise on the left side of
-the colon are infiltrating lesions that present with painful symptoms.
-In contrast, 47\% of CRCs are caused by right-sided colon tumors that
-are bulky and project into the lumen, often remaining asymptomatic until
-advanced carcinogenesis ((2)). Due to the absence of symptoms,
-right-sided tumors have a significantly higher mortality rate ((2)). The
-left and right sides of the colon differ in the amount of inflammation
-present and the genomic instability of precancerous cells, respectively,
-as well as oxygen, pH and the presence of antimicrobial peptides ((3),
-(4), (5)). Microenvironments differ not only longitutinally along the
-colon, but latitudinally from the epithelium to mucosa to intestinal
-lumen, offering several sites for different microbial communities to
-flourish. Given this varied physiology of the proximal-distal axis of
-the colon, symbiotic microbes and their metabolites likely vary between
-sites.
+subtypes ((1)). Of these subtypes, tumors that arise on the distal side
+of the colon are infiltrating lesions that present with painful
+symptoms. In contrast, 47\% of CRCs are caused by proximal-sided colon
+tumors that are bulky and project into the lumen, often remaining
+asymptomatic until advanced carcinogenesis ((2)). Due to the absence of
+symptoms, proximal-sided tumors have a significantly higher mortality
+rate ((2)). The distal and proximal sides of the colon differ in the
+amount of inflammation present and the genomic instability of
+precancerous cells, respectively, as well as oxygen, pH and the presence
+of antimicrobial peptides ((3), (4), (5)). Microenvironments differ not
+only longitutinally along the colon, but latitudinally from the
+epithelium to mucosa to intestinal lumen, offering several sites for
+different microbial communities to flourish. Given this varied
+physiology of the proximal-distal axis of the colon, symbiotic microbes
+and their metabolites likely vary between sites.
 
@@
 community could promote the initiation or proliferation of the distinct
-left and right CRC tumors.
+distal and proximal CRC tumors.
 
@@
 
-We combined these approaches to address the limitations of previous
-studies and effectively characterize the microbiome in the lumen and
-mucosa of the proximal and distal healthy human colon. Our design used
-unprepared colonoscopy techniques to sample each location of the gut
-without prior disruption of the native bacteria in 20 healthy
-volunteers. To address noise created by a diverse set of human
+Here we aimed to address the limitations of previous studies and
+effectively characterize the microbiome in the lumen and mucosa of the
+proximal and distal healthy human colon. Our design used unprepared
+colonoscopy techniques to sample the natural community of each location
+of the gut without prior disruption of the native bacteria in 20 healthy
+volunteers. To address the inherent inter-individual variation in human
 microbiomes, we used a machine-learning classification algorithm trained
@@
 
-Proximal and distal lumenal and mucosal samples were collected from 20
-healthy humans that had not undergone bowel preparation(Figure 1).
-Participants also collected stool at home one week prior to the
-procedure. To characterize the bacterial communities present at these
-sites, 16S rRNA gene sequencing was performed on extracted DNA from each
-sample. The relative abundance of each phyla for each site is shown in
-Figure 2A. This data combines the samples from all participants in the
-study. Each site is primarily dominated by \emph{Firmicutes} and
-\emph{Bacteriodetes}, consistent with known variability in human
-microbiome research (cite). Likewise, samples had varying levels of
-diversity at each site, irrespective of the individual (Figure 2B). That
-is, while some individuals had a more diverse right mucosa, some had a
-more diverse left mucosa. Therefore we could not identify a clear
-pattern of changes in microbial diversity along the gut axis.
+Lumenal and mucosal samples were collected from the proximal and distal
+colon of 20 healthy humans that had not undergone bowel preparation
+(Figure 1). Participants also collected stool at home one week prior to
+the procedure. To characterize the bacterial communities present at
+these sites, 16S rRNA gene sequencing was performed on extracted DNA
+from each sample. Each site was primarily dominated by \emph{Firmicutes}
+and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
+in human microbiome research (cite). Likewise, samples had varying
+levels of diversity at each site, irrespective of the individual (Figure
+2B). For example, the proximal mucosa was more diverse than the distal
+for some individuals while the opposite was true for others. Therefore
+we could not identify a clear pattern of changes in microbial diversity
+along the gut axis.
 
-To compare similarity between sides (left or right) or sites (lumen or
-mucosa), we calculated theta YC distances from OTU abundances and
-compared distances for all individuals. Again, across all patient
-samples we observed a range of theta YC distances when comparing sample
-locations (Figure 3A) and again those ranges did not follow a clear
-pattern on an individual basis. However, when comparing median distances
-between the right lumen and mucosa, the left versus right lumen, the
-left versus right mucosa and the left lumen and mucosa, we found that
-the right lumen and mucosa were most similar to each other than the
-other samples (P \textless{} 0.005, Wilcoxon, BH adjustment).
+To compare similarity between sides (proximal or distal) or sites (lumen
+or mucosa), we calculated \({\theta}\)YC distances from OTU abundances
+and compared these distances for all individuals. Again, across all
+patient samples we observed a range of \({\theta}\)YC distances when
+comparing sample locations (Figure 3A) and again those ranges did not
+follow a clear pattern on an individual basis. However, when comparing
+median distances between the proximal lumen and mucosa, the proximal
+versus distal lumen, the proximal versus distal mucosa, and the distal
+lumen and mucosa, we found that the proximal lumen and mucosa were most
+similar to each other than the other samples (P \textless{} 0.005,
+Wilcoxon, BH adjustment).
 
-Next, we calculated thetaYC distances to examine how each sample
+\paragraph{Stool at exit most resembles lumenal samples from the distal
+colon}\label{stool-at-exit-most-resembles-lumenal-samples-from-the-distal-colon}
+
+Next, we calculated \({\theta}\)YC distances to examine how each sample
 compared to the home-collected exit stool. Amidst variability between
-patients, we did identify significantly smaller thetaYc distance between
-the left lumenal sample and the exit stool (Figure 3B, P \textless{}
-0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger
-difference in the comparisons of the left mucosa to the exit stool,
-indicating that the mucosa is different from the stool as compared to
-lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To determine what
-factors may be driving the differences seen among the samples, we
-compared thetaYC distances between samples from all subjects
+patients, we did identify significantly smaller \({\theta}\)YC distance
+between the distal lumenal sample and the exit stool (Figure 3B, P
+\textless{} 0.05, Wilcoxon, BH adjustment). Furthermore, there was an
+even larger difference in the comparisons of the distal mucosa to the
+exit stool, indicating that the mucosa is different from the stool as
+compared to lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To
+determine what factors may be driving the differences seen among the
+samples, we compared thetaYC distances between samples from all subjects
 (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were overall much more similar
-to each other than to other study subjects (Figure 3C), consistent with
-previous human microbiome studies that have sampled multiple sites of
-the human colon (\textbf{???}, (12), (13)). Thus interpersonal variation
-between subjects drives the differences between samples more than sample
-site or location. Overall, the results of the similarity analysis
-suggest that the contents of the left lumen are most representative of
-stool at exit, and the microbes remaining on the mucosa are much
-different.
+We found that samples from one individual were far more similar to each
+other than to other study subjects (Figure 3C), consistent with previous
+human microbiome studies that have sampled multiple sites of the human
+colon (\textbf{???}, (12), (13)). Thus interpersonal variation between
+subjects drives the differences between samples more than sample site or
+location. Overall, the results comparing the structure of the
+communities suggest that the contents of the distal lumen are most
+representative of stool at exit, and the microbes remaining on the
+mucosa are much different.
 
@@
 To identify OTUs that were distinct at each biogeographical site, we
-constructed several Random Forest models built on OTU abundances. We
-built the first model to classify the lumen versus mucosal samples for
-the right and left sides, independently (Figure 4A). The constructed
-model used ((Xopt)) features for the right and ((Xopt)) for the left.
-The models performed well when classifying these samples (0.8 and 1.0,
-respectively). The OTUs that were most predictive of each site are
-identified by their greatest mean decrease in accuracy when removed from
-the model. For distinguishing the right lumen and mucosa, OTUs from the
-\emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and two OTUs
-from the \emph{Enterobacteraceae} genera were differentially abundant
-(Figure 4B). The model classifying the left lumen and mucosa identified
-OTUs from \emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus}
-and two OTUs from the \emph{Anaerococcus} genera that could distinguish
-lumen from mucosal samples (Figure 4C). These results indicate that
-there are fine differences between the different sites of the colon, and
-that these can be traced down to specific OTUs on each side.
+constructed several Random Forest models trained using OTU abundances.
+We built the first model to classify the lumen versus mucosal samples
+for the proximal and distal sides, independently (Figure 4A). The
+constructed model used ((Xopt)) features for the proximal and ((Xopt))
+for the distal. The models performed well when classifying these samples
+(0.8 and 1.0, respectively). The OTUs that were most predictive of each
+site are identified by their greatest mean decrease in accuracy when
+removed from the model. For distinguishing the proximal lumen and
+mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
+\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
+were differentially abundant (Figure 4B). The model classifying the
+distal lumen and mucosa identified OTUs from \emph{Turicibacter},
+\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
+\emph{Anaerococcus} genera that could distinguish lumen from mucosal
+samples (Figure 4C). These results indicate that there are fine
+differences between the different sites of the colon, and that these can
+be traced down to specific OTUs on each side.
 
-Next, we built a model to differentiate the left and right microbiome.
-The model performed best when distinguishing the right versus left
-mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen
-(AUC = 0.755). The model was able to explain ((X\%)) of the variance.
-OTUs that were differentially abundant between the left and right mucosa
-included members of the \emph{Porphyromonas}, \emph{Murdochiella},
-\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} genera
-(Figure 5B). The model was less-effective at classifying the right and
-left lumen (AUC = 0.755). However it did identify differentially
-abundant OTUs such as three members of the \emph{Bacteroides} genus, a
-\emph{Clostridium IV} OTU and an \emph{Oscillibacter} OTU (Figure 5C).
-This analysis found that some of the same OTUs that are distinct between
-the mucosa and lumen also drive sidedness- such as \emph{Anaerococcus}
-and \emph{Finegoldia}.
+Next, we built a model to differentiate the proximal and distal lumenal
+samples. The model performed best when distinguishing the proximal
+versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
+versus distal lumen (AUC = 0.755). These models were able to explain
+((X\%)) of the variance, respectively. OTUs that were differentially
+abundant between the distal and proximal mucosa included members of the
+\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
+\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B).
+Differentially abundant OTUs of the proximal and distal lumen included
+three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
+and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
+some of the same OTUs that are distinct between the mucosa and lumen
+also helped to differentiate between the two sides- such as
+\emph{Anaerococcus} and \emph{Finegoldia}.
 
@@
 Given that specific bacterial species have been associated with
-colorectal cancer, we probed our sample set for these OTUs. Among our
-100 samples, the most frequent sequence associated wtih the
+colorectal cancer and IBD, we probed our sample set for these OTUs.
+Among our 100 samples, the most frequent sequence associated with the
 \emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
-\emph{Fusobacterium nucleatum subsp animalis}. It was the only sequence
-to align with a \emph{F. nucleatum} strain- the only species of
-\emph{Fusobacterium} known to have oncogenic properties and be found on
-the surfaces of colorectal cancer tumors. ((14)). The
-\emph{Fusobacterium} positive samples were located on the right and left
-mucosa and ranged up to 1\% of the sample (Figure 6A). OTU00152 is
-\emph{Porphyromonas} and the most frequent sequence in that OTU aligns
-to \emph{Porphyromonas asacharolytica}, another bacterium commonly
-detected and isolated from colorectal tumors. OTU152 was only detected
-on the left mucosa, and in fact was one of the OTUs the classification
-model identified as separating left and right sides (Figure 6B). Samples
-that were positive for \emph{P. asacharolytica} ranged in abundances
-from 0.01\% - 16\%. Thus, cancer-associated OTUs could be found in our
+\emph{Fusobacterium nucleatum subsp animalis} (XX\% over full length).
+This is the only species of \emph{Fusobacterium} known to have oncogenic
+properties and be found on the surfaces of colorectal cancer tumors.
+((14)). The \emph{Fusobacterium} positive samples were located in x\% of
+the the proximal and X\% of the distal mucosa and represented as much as
+1\% of any sample (Figure 6A). OTU152 was similar to the members of the
+\emph{Porphyromonas} genus and the most frequent sequence in that OTU
+aligned to \emph{Porphyromonas asacharolytica} (X\% over full length),
+another bacterium commonly detected and isolated from colorectal tumors.
+OTU152 was only detected on the distal mucosa, and in fact was one of
+the OTUs the classification model identified as separating distal and
+proximal sides (Figure 6B). Among the samples that were positive for
+\emph{P. asacharolytica}, the relative abundances for this OTU ranged
+from 0.01\% - 16\%. Thus, disease-associated OTUs could be found in our
 sample set of 20 healthy individuals.
@@
 Here we identified bacterial taxa that are specific to the lumen and
-mucosa of the right and left human colon from samples collected during
-unprepared colonoscopy. We found that all locations contained a range of
-phyla and a range of diversity, but that there was a wide variability
-between subjects. Pairwise comparisons of each of the sites revealed
-that the right mucosa and lumen were most similar to each other.
-Further, comparison of colonoscopy-collected samples with samples
-collected from stool at home showed that the left lumen is most similar
-to stool at exit. Random Forest algorithms built on OTU abundances from
-each sample identified microbes that are particular to each location of
-the colon. Finally, we were able to detect some bacterial OTUs
-associated with colorectal cancer in our healthy patient cohort. Using
-unprepped colonoscopy and machine learning, we have identified bacterial
-phyla specific to the healthy proximal and distal human colon.
+mucosa of the proximal and distal human colon from samples collected
+during unprepared colonoscopy. We found that all locations contained a
+range of phyla and a range of diversity, but that there was a wide
+variability between subjects. Pairwise comparisons of each of the sites
+revealed that the proximal mucosa and lumen were most similar to each
+other. Further, comparison of colonoscopy-collected samples with samples
+collected from stool at home showed that the distal lumen is most
+similar to stool at exit. Random Forest algorithms built on OTU
+abundances from each sample identified microbes that are particular to
+each location of the colon. Finally, we were able to detect some
+bacterial OTUs associated with colorectal cancer in our healthy patient
+cohort. Using unprepped colonoscopy and machine learning, we have
+identified bacterial phyla specific to the healthy proximal and distal
+human colon.
 
@@
 location were consistent with known physiological gradients along the
-gut axis ((5)). For instance, the right mucosa harbored
+gut axis ((5)). For instance, the proximal mucosa harbored
 mucosa-associated facultative anaerobes such as \emph{Actinomyces} and
 \emph{Enterobacteraceae} and the aerobic \emph{Psuedomonas} consistent
-with the highest oxygen regions of the colon. The left mucosa was far
+with the highest oxygen regions of the colon. The distal mucosa was far
 more likely to host strictly anaerobic species such as
@@
 \emph{Peptoniphilus}. The model was less effective at classifying the
-right and left lumenal contents, probably because the samples are
+proximal and distal lumenal contents, probably because the samples are
 arguably composed of the same material.
@@
 \emph{F. nucleatum} was found on both sides of the colon, \emph{P.
-asacharolytica} was only detected in the left mucosa. Not much is known
-about the distribution of \emph{P. asacharolytica} but given its
+asacharolytica} was only detected in the distal mucosa. Not much is
+known about the distribution of \emph{P. asacharolytica} but given its
 documented anaerobic characteristics and asacharolytic metabolism, it
 might not be surprising that it resides in the less-oygen-rich and
-proteinaceous left mucosa ((4)). In studies examining bacteria on
+proteinaceous distal mucosa ((4)). In studies examining bacteria on
 colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
-on right-sided tumors, and distribution of \emph{F. nucleatum} decreases
-along the colon to rectum ((18)). Of the (8) (40\%) individuals positive
-for \emph{F. nucleatum} in this study, the bacteria was spread across
-the right mucosa, left lumen and left mucosa. The presence of \emph{F.
-nucleatum} in a healthy individual is not necessarily linked to the
-development of future colorectal cancers. Because of the spatial
-distribution of the \emph{F. nucleatum} in our sample set, we cannot
-develop a model for the role of \emph{F. nucleatum} in the healthy
-colon. Data examining bacterial biofilms on CRC tumors suggests that
-\emph{Fusobacteria} species are more commonly found both on proximal
-tumors and in biofilms, indicating that it is not only the presence of
-the bacteria but the organization of the tumor community that
-contributes to \emph{Fusobacterium's} role in tumorigenesis ((8)).
-Finally, \emph{Fusobacteria} and \emph{Porphyromonas} species have been
-known to not only co-occur on CRC tumors but also to synergistically
-promote tumorigenesis in an oral cancer model ((19), (20)). Thus,
-further analysis of the distribution and activities of these pathogens
-may elucidate a mechanism for development of CRC tumors in the proximal
-or distal colon.
+on proximal-sided tumors, and distribution of \emph{F. nucleatum}
+decreases along the colon to rectum ((18)). Of the (8) (40\%)
+individuals positive for \emph{F. nucleatum} in this study, the bacteria
+was spread across the proximal mucosa, distal lumen and distal mucosa.
+The presence of \emph{F. nucleatum} in a healthy individual is not
+necessarily linked to the development of future colorectal cancers.
+Because of the spatial distribution of the \emph{F. nucleatum} in our
+sample set, we cannot develop a model for the role of \emph{F.
+nucleatum} in the healthy colon. Data examining bacterial biofilms on
+CRC tumors suggests that \emph{Fusobacteria} species are more commonly
+found both on proximal tumors and in biofilms, indicating that it is not
+only the presence of the bacteria but the organization of the tumor
+community that contributes to \emph{Fusobacterium's} role in
+tumorigenesis ((8)). Finally, \emph{Fusobacteria} and
+\emph{Porphyromonas} species have been known to not only co-occur on CRC
+tumors but also to synergistically promote tumorigenesis in an oral
+cancer model ((19), (20)). Thus, further analysis of the distribution
+and activities of these pathogens may elucidate a mechanism for
+development of CRC tumors in the proximal or distal colon.
 
@@
 would not be effective. Given that our results showed that the stool
-most accurately reflects the community of the left lumen, we likely
+most accurately reflects the community of the distal lumen, we likely
 cannot use Random Forest of stool samples to diagnose any changes in the
@@
 were collected and flash-frozen in RNAlater. These samples comprised the
-distal or left colon samples. The sigmoidoscope was then withdrawn and a
-pediatric colonoscope was inserted to reach the ascending colon. Samples
-were then collected as in the distal colon and the colonoscope
+distal or distal colon samples. The sigmoidoscope was then withdrawn and
+a pediatric colonoscope was inserted to reach the ascending colon.
+Samples were then collected as in the distal colon and the colonoscope
 withdrawn. All samples were stored at -80 C until study completion.
@@
 sites -- each point represents one individual. In (A), comparisons of
-the right and left mucosal and lumen are shown. In (B), comparisons of
-each site to the exit stool are shown.
+the proximal and distal mucosal and lumen are shown. In (B), comparisons
+of each site to the exit stool are shown.
 
@@
 cross validation of the Random Forest model classifying lumen and
-mucosal samples for the left and right sides of the colon. (B) Top five
-OTUs that are most important for the classification model for the left
-mucosa and lumen (B) and the right mucosa and lumen (C).
+mucosal samples for the distal and proximal sides of the colon. (B) Top
+five OTUs that are most important for the classification model for the
+distal mucosa and lumen (B) and the proximal mucosa and lumen (C).
 
@@
 
-Random Forest classifies the left and right sides of the colon. A)
+Random Forest classifies the distal and proximal sides of the colon. A)
 Receiver Operator Characteristic curves are shown for the 10-fold cross
-validation of the Random Forest model classifying left lumen versus
-right lumen (orange) and left mucosa vs right mucosa (green). (B) Top
-five OTUs that are most important for the classification model for the
-left and right mucosa (B) and the left and right lumen (C).
+validation of the Random Forest model classifying distal lumen versus
+proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
+(B) Top five OTUs that are most important for the classification model
+for the distal and proximal mucosa (B) and the distal and proximal lumen
+(C).
 

### 2017-07-02 submission/manuscript.Rmd
restructured intro

@@
 
-Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States.
-CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon.
-CRC has recently been described to fall into four molecular subtypes (@Guinney2015).
-Of these subtypes, tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms.
-In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are bulky and project into the lumen, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010).
-Due to the absence of symptoms, proximal-sided tumors have a significantly higher mortality rate (@Benedix2010).
-The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
+The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome.
+Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location.
 Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
-Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.
+The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
+IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012).
+Further, the severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
+For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms.
+In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010).
+The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
+In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically present with disease in the small intestine, ileum and proximal colon (cite).
+UC presents as large and highly inflammed mucosal ulcers, where as Crohn's disease lesions are often smaller and have areas of normal tissue distributed amongst the flare (cite).
+Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC at these locations.
 
@@
 Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014).
-Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013).
 Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated (@Dejea2014).
@@
 As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
+
+Previous profiles of the gut biogeography only focus on prepped bowels and dont identify rare taxa.
+but we know that rare taxa adn small members of the community can drive tumorigenesis and whatnot.
+so identifying specifically UNIQUE bugs
+
+why are L/R M/L interesting questions? because diseases progress differently depending on where they are located.
 

### 2017-07-05 submission/manuscript.Rmd
progress with restructuring intro

@@
 
-Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012).
-Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014).
-Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated (@Dejea2014).
+Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012, IBDREF).
+For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated (@Dejea2014).
 These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
-Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our classification studies (@Baxter2016).
-Because these studies were performed in mice or examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon.
+Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our human stool classification studies (@Baxter2016).
+_F.
+nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls (Strauss2011).
+Furthermore, studies of _F.
+nucleatum_ isolated from mucosal biopsies showed that more invasive  _F.
+nucleatum_ positively correlates with IBD disease level (Strauss2011).
+The bacteria appear to have a high-impact despite being lowly-abundant-.
+Thus the molecular capabilities of these rare taxa may drive the colonic disease state.
+Previous studies examining the microbes associated with CRC and IBD have focused on overall community structure, diversity and membership and not lowly-abundant species.
+Further, because these studies often examined only shed human stool or one biopsy at a time, they were unable to analyze paried samples from the proximal and distal sides of the colon.
 Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
@@
 so identifying specifically UNIQUE bugs
-
-why are L/R M/L interesting questions? because diseases progress differently depending on where they are located.
 

### 2017-07-05 submission/manuscript.Rmd
intro redone

@@
 IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012).
-Further, the severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
+The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur (cite).
 For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms.
@@
 Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC at these locations.
+Because CRC is a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.
 
@@
 nucleatum_ positively correlates with IBD disease level (Strauss2011).
-The bacteria appear to have a high-impact despite being lowly-abundant-.
+Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant (cite).
 Thus the molecular capabilities of these rare taxa may drive the colonic disease state.
 Previous studies examining the microbes associated with CRC and IBD have focused on overall community structure, diversity and membership and not lowly-abundant species.
-Further, because these studies often examined only shed human stool or one biopsy at a time, they were unable to analyze paried samples from the proximal and distal sides of the colon.
-Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
-Due to these limitations, the contribution of the gut microbiota to these subtypes is largely undefined.
-Characterizing these communities could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct distal and proximal CRC tumors.
+Further, because these studies often examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon.
+Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
+Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location is largely undefined.
+Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct distal and proximal CRC tumors or IBD flares.
 
-Further, the few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods.
+The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods.
 The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures.
@@
 
-Previous profiles of the gut biogeography only focus on prepped bowels and dont identify rare taxa.
-but we know that rare taxa adn small members of the community can drive tumorigenesis and whatnot.
-so identifying specifically UNIQUE bugs
-
-Here we aimed to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon.
+Here we aimed to address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon.
 Our design used unprepared colonoscopy techniques to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.

### 2017-07-10 submission/manuscript.Rmd
updated discussion

@@
 The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
-In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically present with disease in the small intestine, ileum and proximal colon (cite).
-UC presents as large and highly inflammed mucosal ulcers, where as Crohn's disease lesions are often smaller and have areas of normal tissue distributed amongst the flare (cite).
-Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC at these locations.
+In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon (cite).
+UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare (cite).
+Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC.
 Because CRC is a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.
@@
 nucleatum_ positively correlates with IBD disease level (Strauss2011).
-Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant (cite).
-Thus the molecular capabilities of these rare taxa may drive the colonic disease state.
+Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community (cite).
+The molecular capabilities of these rare taxa may drive the colonic disease state.
 Previous studies examining the microbes associated with CRC and IBD have focused on overall community structure, diversity and membership and not lowly-abundant species.
@@
 Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location is largely undefined.
-Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct distal and proximal CRC tumors or IBD flares.
+Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.
 
@@
 Here we aimed to address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon.
-Our design used unprepared colonoscopy techniques to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
+Our design used an unprepared colonoscopy technique to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
 To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
@@
 Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016).
-Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.
+Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.
 
 Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies.
-A rare report of a matched-colonoscopy study that sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016).
+A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016).
 This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016).
@@
 Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit and it performed much better than FIT alone.
-Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (@Baxter2016).
 One caveat of the FIT study was that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.

### 2017-07-11 submission/manuscript.Rmd
discussion almost fixed

@@
 
+The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD (cite).
+Laboratory experiments with these isolates have shown that disease-isolated _F.
+nucleatum_ are more invasive and stimulate more TNF-a production than strains from healthy individuals (@Strauss2011), suggesting the bacteria may increase inflammation in the gut as well (cite - Dharmani 2011).
+_F.
+varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model (Ohkusa 2003).
+_F.
+varium_ was only detected in 3 of our study participants and two of those samples were isolated form the proximal mucosa (cite figure).
+Although there is low occurrance in our study, _F.
+varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum (cite), suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites.
+Further work will assess how gut environment may select for species which then cause localized disease.
+
 Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies.

### 2017-07-11 code/oncogenic_otus.R

@@ -40,6 +40,20 @@ fuso_plot <- ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(
                    labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
   theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
 
+#IBD fuso OTU - F. varium 
+
+fusoviv <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00472")]
+fusoviv[,5] <- (fusoviv[,4]/4321)*100
+names(fusoviv)[5] <- "Otu00472_relAbund"
+
+positions <- c("RB", "RS", "LB", "LS", "SS")
+fusoviv_plot <- ggplot(fusoviv, aes(x=location, y=Otu00472_relAbund)) +geom_jitter(width=0.3) +theme_bw() +
+  ylab("F. varium Relative Abundance") +
+  scale_x_discrete(limits = positions, breaks=positions, 
+                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
+
+
 
 #other oncogenic otus - Otu00152 is P. asaccharolytica. Otu00248 is p micra
 

### 2017-07-17 submission/manuscript.Rmd
most text tweaks done

@@
 ---
-title: Specific taxa differentiate proximal and distal microbiota in the unprepped
+title: Specific taxa differentiate proximal and distal microbiota in the unprepared
   human colon
@@
 
-Colorectal cancer (CRC) remains a leading cause of death worldwide despite improved preventative and therapeutic measures.
-Tumors of the proximal (proximal) and distal (distal) colon are morphologically and genetically distinct.
-Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans.
-Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease.
+The human colon contains a number of chemical and nutrient gradients along the proximal to distal gut axis.
+These gradients create microenvironments that affect the distrubution and composition of the gut microbiota.
+The microbiome has been implicated in the colonic diseases colorectal cancer (CRC) and inflammatory bowel disease (IBD).
+Further, these diseases exhibit different symptoms depending on the location of the colon they are found in.
+For instance, CRC tumors of the proximal and distal colon are morphologically and genetically distinct.
+Similarly, bowel diseases such as Crohn's are typically exacerbated in the proximal instestine while ulcerative colitis patients often experience symptoms in the distal colon.
+Previous work from our group and others found that microbial dysbiosis is associated with the development of CRC tumors and IBD flares in studies of both mice and humans.
+Analysis of the fecal microbiota from healthy and CRC or IBD patients further revealed different microbial signatures associated with these diseases.
 We extended our observations of the fecal microbiome to include analysis of the proximal and distal human colon.
@@
 A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016).
-This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016).
+This group found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016).
 Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites.
 Further, the differences were not great enough to overcome interpersonal differences between subjects.
-So, bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel.
-Thus our findings of specific bacteria at each site of the colon are strengthened by the lack of bowel preparation.
+Bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel.
+Thus our findings in this study are strengthened by the lack of bowel preparation.
 
-Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer.
-Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors.
-Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit and it performed much better than FIT alone.
-One caveat of the FIT study was that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.
-Given that our results showed that the stool most accurately reflects the community of the distal lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
-
-By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect oncogenesis of proximal and distal CRC tumors.
-To this point, 16S rRNA gene sequecing community profiling studies do not provide enough information to probe these questions.
-Our sample set of matched proximal, distal, lumenal and mucosal samples from colons that have not undergone bowel preparation presents a unique opportunity to explore further questions about the microbiome along the gut axis.
+By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes.
+To this point, 16S rRNA gene sequecing community profiling studies do not provide enough information to fully probe these questions.
+In particular, 16S sequencing cannot not profile the host characteristics at each site.
+Combining the unprepared colonoscopy approach with analysis of multi-omic sequencing data may be useful in further characterizing host-microbiome interactions along the gut axis for both health and disease.
 

### 2017-07-18 submission/manuscript.Rmd
updated citations

@@
 ---
-title: Specific taxa differentiate proximal and distal microbiota in the unprepared
-  human colon
+title: Unprepared colonoscopy identifies microbiota specific to the proximal and distal human colon
 csl: mbio.csl
@@
 
-Running title: Specific taxa differentiate proximal and distal human colon microbiota
+Running title: Unprepared colonoscopy identifies proximal and distal human colonic microbiota
 
@@
 
-The human colon contains a number of chemical and nutrient gradients along the proximal to distal gut axis.
+The human colon contains chemicals and nutrients that change along the proximal to distal gut axis.
 These gradients create microenvironments that affect the distrubution and composition of the gut microbiota.
@@
 Further, these diseases exhibit different symptoms depending on the location of the colon they are found in.
-For instance, CRC tumors of the proximal and distal colon are morphologically and genetically distinct.
+CRC tumors of the proximal and distal colon are morphologically and genetically distinct.
 Similarly, bowel diseases such as Crohn's are typically exacerbated in the proximal instestine while ulcerative colitis patients often experience symptoms in the distal colon.
-Previous work from our group and others found that microbial dysbiosis is associated with the development of CRC tumors and IBD flares in studies of both mice and humans.
-Analysis of the fecal microbiota from healthy and CRC or IBD patients further revealed different microbial signatures associated with these diseases.
-We extended our observations of the fecal microbiome to include analysis of the proximal and distal human colon.
+Previous analysis of the fecal microbiota from healthy and CRC or IBD patients has revealed different microbial signatures associated with these diseases.
+We extended these observations of the fecal microbiome to include analysis of the proximal and distal healthy human colon.
 We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure.
@@
 Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
+Comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the distal lumen, likely reflecting the anatomical structure of the colon.
 Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key species that distinguish biogeographical sites.
@@
 Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end.
-Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the distal lumen, again reflecting the anatomical structure of the colon.
 By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides.
-Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.
+Further investigation of these bacteria may elucidate if and how these groups contribute to different pathogenesis processes on the respective sides of the colon.
 
@@
 The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
-IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012).
-The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur (cite).
-For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms.
+IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012, @Forbes2016, @Halfvarson2017).
+The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
+For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms (@Benedix2010).
 In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010).
 The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
-In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon (cite).
-UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare (cite).
+In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon (@Forbes2016).
+UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare (@Forbes2016).
 Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC.
@@
 
-Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012, IBDREF).
+Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012, @Halfvarson2017).
 For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated (@Dejea2014).
@@
 nucleatum_ positively correlates with IBD disease level (Strauss2011).
-Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community (cite).
-The molecular capabilities of these rare taxa may drive the colonic disease state.
-Previous studies examining the microbes associated with CRC and IBD have focused on overall community structure, diversity and membership and not lowly-abundant species.
-Further, because these studies often examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon.
+Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community (@Forbes2016).
+The molecular capabilities of these rare taxa may contribute to the colonic disease state.
+These studies often examined only shed human stool or the small intestine, preventing fine-resolution analysis of paired samples from the proximal and distal sides of the colon.
 Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
-Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location is largely undefined.
+Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location in the colon is largely undefined.
 Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.
@@
 To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
-Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research (cite).
+Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research (@LloydPrice2016).
 Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
@@
 The constructed model used ((Xopt)) features for the proximal and ((Xopt)) for the distal.
-The models performed well when classifying these samples (`r cv10f_roc_proximal_bs$auc` and `r cv10f_roc_distal_bs$auc`, respectively).
+The models performed well when classifying these samples (`r round(cv10f_roc_proximal_bs$auc,3)` and `r round(cv10f_roc_distal_bs$auc, 3)`, respectively).
 The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.

### 2017-07-18 submission/manuscript.tex
more edits

@@
 \hypersetup{unicode=true,
-            pdftitle={Specific taxa differentiate proximal and distal microbiota in the unprepped human colon},
+            pdftitle={Unprepared colonoscopy identifies microbiota specific to the proximal and distal human colon},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Specific taxa differentiate proximal and distal microbiota in the
-unprepped human colon}
+  \title{Unprepared colonoscopy identifies microbiota specific to the proximal
+and distal human colon}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-Running title: Specific taxa differentiate proximal and distal human
-colon microbiota
+Running title: Unprepared colonoscopy identifies proximal and distal
+human colonic microbiota
 
@@
 
-Colorectal cancer (CRC) remains a leading cause of death worldwide
-despite improved preventative and therapeutic measures. Tumors of the
-proximal (proximal) and distal (distal) colon are morphologically and
-genetically distinct. Previous work from our group found that microbial
-dysbiosis is associated with the development of CRC tumors in studies of
-both mice and humans. Analysis of the fecal microbiota from healthy and
-CRC patients further revealed different microbial signatures associated
-with disease. We extended our observations of the fecal microbiome to
-include analysis of the proximal and distal human colon. We used a
-two-colonoscope approach on subjects that had not undergone standard
-bowel preparation procedure. This technique allowed us to characterize
-the native proximal and distal luminal and mucosal microbiome without
-prior chemical disruption. 16S rRNA gene sequencing was performed on
-proximal and distal mucosal and luminal biopsies and home-collected
-stool for 20 healthy individuals. Diversity analysis revealed that each
-site contained a diverse community, and that a patient's samples were
-more similar to each other than to that of other individuals. Since we
-could not differentiate sites along the colon based on community
-structure or community membership alone, we employed the Random Forest
-machine-learning algorithm to identify key species that distinguish
-biogeographical sites. Random Forest classification models were built
-using taxa abundance and sample location and revealed distinct
-populations that were found in each location. \emph{Peptoniphilus,
-Anaerococcus, Enterobacteraceae, Pseudomonas} and \emph{Actinomyces}
-were most likely to be found in mucosal samples versus luminal samples
-(AUC = 0.925). The classification model performed well (AUC = 0.912)
-when classifying mucosal samples into proximal or distal sides, but
-separating luminal samples from each side proved more challenging (AUC =
-0.755). The distal mucosa was found to have high populations of
-\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
-distal luminal samples were comprised of many of the same taxa, likely
-reflecting the fact that stool moves along the colon from the proximal
-to distal end. Finally, comparison of all samples to fecal samples taken
-at exit uncovered that the feces were most similar to samples taken from
-the distal lumen, again reflecting the anatomical structure of the
-colon. By sampling the unprepped human colon, our results have
-identified distinct bacterial populations native to the proximal and
-distal sides. Further investigation of these bacteria may elucidate if
-and how these groups contribute to differential oncogenesis processes on
-the respective sides of the colon.
+The human colon contains chemicals and nutrients that change along the
+proximal to distal gut axis. These gradients create microenvironments
+that affect the distrubution and composition of the gut microbiota. The
+microbiome has been implicated in the colonic diseases colorectal cancer
+(CRC) and inflammatory bowel disease (IBD). Further, these diseases
+exhibit different symptoms depending on the location of the colon they
+are found in. CRC tumors of the proximal and distal colon are
+morphologically and genetically distinct. Similarly, bowel diseases such
+as Crohn's are typically exacerbated in the proximal instestine while
+ulcerative colitis patients often experience symptoms in the distal
+colon. Previous analysis of the fecal microbiota from healthy and CRC or
+IBD patients has revealed different microbial signatures associated with
+these diseases. We extended these observations of the fecal microbiome
+to include analysis of the proximal and distal healthy human colon. We
+used a two-colonoscope approach on subjects that had not undergone
+standard bowel preparation procedure. This technique allowed us to
+characterize the native proximal and distal luminal and mucosal
+microbiome without prior chemical disruption. 16S rRNA gene sequencing
+was performed on proximal and distal mucosal and luminal biopsies and
+home-collected stool for 20 healthy individuals. Diversity analysis
+revealed that each site contained a diverse community, and that a
+patient's samples were more similar to each other than to that of other
+individuals. Comparison of all samples to fecal samples taken at exit
+uncovered that the feces were most similar to samples taken from the
+distal lumen, likely reflecting the anatomical structure of the colon.
+Since we could not differentiate sites along the colon based on
+community structure or community membership alone, we employed the
+Random Forest machine-learning algorithm to identify key species that
+distinguish biogeographical sites. Random Forest classification models
+were built using taxa abundance and sample location and revealed
+distinct populations that were found in each location.
+\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
+\emph{Actinomyces} were most likely to be found in mucosal samples
+versus luminal samples (AUC = 0.925). The classification model performed
+well (AUC = 0.912) when classifying mucosal samples into proximal or
+distal sides, but separating luminal samples from each side proved more
+challenging (AUC = 0.755). The distal mucosa was found to have high
+populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
+Proximal and distal luminal samples were comprised of many of the same
+taxa, likely reflecting the fact that stool moves along the colon from
+the proximal to distal end. By sampling the unprepped human colon, our
+results have identified distinct bacterial populations native to the
+proximal and distal sides. Further investigation of these bacteria may
+elucidate if and how these groups contribute to different pathogenesis
+processes on the respective sides of the colon.
 
@@
 
-Colorectal cancer (CRC) is the second-leading cause of cancer-related
-deaths in the United States. CRC tumors vary in structure, size,
-morbidity and symptomology depending upon their geographical location in
-the colon. CRC has recently been described to fall into four molecular
-subtypes ((1)). Of these subtypes, tumors that arise on the distal side
-of the colon are infiltrating lesions that present with painful
-symptoms. In contrast, 47\% of CRCs are caused by proximal-sided colon
-tumors that are bulky and project into the lumen, often remaining
-asymptomatic until advanced carcinogenesis ((2)). Due to the absence of
-symptoms, proximal-sided tumors have a significantly higher mortality
-rate ((2)). The distal and proximal sides of the colon differ in the
-amount of inflammation present and the genomic instability of
-precancerous cells, respectively, as well as oxygen, pH and the presence
-of antimicrobial peptides ((3), (4), (5)). Microenvironments differ not
-only longitutinally along the colon, but latitudinally from the
-epithelium to mucosa to intestinal lumen, offering several sites for
-different microbial communities to flourish. Given this varied
-physiology of the proximal-distal axis of the colon, symbiotic microbes
-and their metabolites likely vary between sites.
+The human colon is an ecosystem comprised of several different
+microenvironments inhabited by resident bacterial members of the
+microbiome. Concentrations of oxygen, water and anti-microbial peptides
+change along the gut axis and influence what populations of microbes
+reside in each location. Microenvironments differ not only
+longitutinally along the colon, but latitudinally from the epithelium to
+mucosa to intestinal lumen, offering several sites for different
+microbial communities to flourish. The identity of these specific
+microbes and communities are important for understanding the etiology of
+complex colon diseases such as Inflammatory Bowel Disease (IBD) and
+Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
+accelerated by perturbations in gut microbes ((1), (2), (3)). The
+severity, symptoms, morbidity and mortality of these diseases is known
+to vary based upon the biogeographical location in which they occur. For
+instance, CRC tumors that arise on the distal side of the colon are
+infiltrating lesions that present with painful symptoms ((4)). In
+contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
+are sessile and form along the wall of the colon, often remaining
+asymptomatic until advanced carcinogenesis ((4)). The distal and
+proximal sides of the colon differ in the amount of inflammation present
+and the genomic instability of precancerous cells, respectively, in
+addition to variation in the previously mentioned chemical gradients
+((1), (5), (6)). In IBD patients, disease flares in the distal colon are
+usually indicative of ulcerative colitis (UC), whereas Crohn's disease
+(CD) patients typically experience disease in the small intestine, ileum
+and proximal colon ((2)). UC presents as large and highly inflammed
+mucosal ulcers, where as CD lesions are often smaller and have areas of
+normal tissue distributed amongst the flare ((2)). Thus, given the
+varied physiology of the proximal-distal axis of the colon and known
+differences in disease patterns at these sites, symbiotic microbes and
+their metabolites likely vary as well, and may influence the
+heterogenous disease prognoses of IBD and CRC. Because CRC is a
+long-term complication of IBD, the distribution of microbes is important
+to understanding the pathophysiology of both diseases.
 
 Several recent findings have shown that development and progression of
-CRC can be attributed to specific molecular events as a result of
-interactions between the gut microbiota and human host ((3)). Our group
-and others have found that the stool microbiome of patients with CRC is
-distinct from that of healthy people ((6)). Further studies manipulating
-the gut microbiome using antibiotics or other chemoagents in mice has
-shown that dysbiosis preceded and accelerated the development of CRC
-tumors ((7)). Comparison of the specific bacteria present on CRC tumors
-with those found on nearby healthy tissue has also identified specific
-bacterial species that are tumor-associated ((8)). These species include
-the oral pathogens \emph{Fusobacterium nucleatum} and
-\emph{Porphyromonas asacharolytica}. Interestingly, these periodontal
-pathogens have been highly predictive of whether a patient had CRC
-tumors or not in our classification studies ((9)). Because these studies
-were performed in mice or examined only shed human stool, they were
-unable to analyze paried samples from the proximal and distal sides of
-the colon. Similarly, comparisons of on- or off-tumor bacteria rarely
-have matched tissue from the other side of the colon from the same
-patient, limiting what conclusions can be drawn about the colonic
-microbiome overall, let alone at that specific site. Due to these
-limitations, the contribution of the gut microbiota to these subtypes is
-largely undefined. Characterizing these communities could provide needed
-insight into CRC etiology, including how the disruption of the healthy
-community could promote the initiation or proliferation of the distinct
-distal and proximal CRC tumors.
+IBD or CRC can be attributed to specific molecular events as a result of
+interactions between the gut microbiota and human host ((1), (3)). For
+instance, comparison of the bacteria present on CRC tumors with those
+found on nearby healthy tissue has identified specific species that are
+tumor-associated ((7)). These species include the oral pathogens
+\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
+Interestingly, these periodontal pathogens have been highly predictive
+of whether a patient had CRC tumors or not in our human stool
+classification studies ((8)). \emph{F. nucleatum} has also been found to
+be elevated in the stool and biopsies of patients with IBD as compared
+to healthy controls (Strauss2011). Furthermore, studies of \emph{F.
+nucleatum} isolated from mucosal biopsies showed that more invasive
+\emph{F. nucleatum} positively correlates with IBD disease level
+(Strauss2011). Like many intestinal pathogens, the bacteria appear to
+have a high-impact despite being lowly-abundant in the community ((2)).
+The molecular capabilities of these rare taxa may contribute to the
+colonic disease state. These studies often examined only shed human
+stool or the small intestine, preventing fine-resolution analysis of
+paired samples from the proximal and distal sides of the colon.
+Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have
+matched tissue from the other side of the colon from the same patient,
+limiting what conclusions can be drawn about the colonic microbiome
+overall, let alone at that specific site. Due to these limitations, the
+contribution of the gut microbiota to IBD and CRC disease location in
+the colon is largely undefined. Characterizing these communities could
+provide needed insight into disease etiology, including how the
+disruption of the healthy community could promote the initiation or
+proliferation of the distinct proximal and distal CRC tumors or IBD
+flares.
 
-Further, the few existing profiles of the microbial biogeography of the
-gut have been limited by sample collection methods. The majority of
-human gut microbiome studies have been performed on whole shed stool or
-on samples collected during colonoscopy procedures. While the latter
-method allows investigators to acquire samples from inside the human
-colon, typically this procedure is preceded by the use of bowel
-preparation methods such as the consumption of laxatives to cleanse the
-bowel. Bowel preparation is essential for detecting cancerous or
-precancerous lesions in the colon, but complicates microbiome profiling
-as the chemicals strip the bowel of contents and disrupt the mucosal
-layer ((10), (11)). As such, what little information we do have about
-the biogeographical distribution of the microbes in the proximal and
-distal colon is confounded by the bowel preparation procedure.
+The few existing profiles of the microbial biogeography of the gut have
+been limited by sample collection methods. The majority of human gut
+microbiome studies have been performed on whole shed stool or on samples
+collected during colonoscopy procedures. While the latter method allows
+investigators to acquire samples from inside the human colon, typically
+this procedure is preceded by the use of bowel preparation methods such
+as the consumption of laxatives to cleanse the bowel. Bowel preparation
+is essential for detecting cancerous or precancerous lesions in the
+colon, but complicates microbiome profiling as the chemicals strip the
+bowel of contents and disrupt the mucosal layer ((9), (10)). As such,
+what little information we do have about the biogeographical
+distribution of the microbes in the proximal and distal colon is
+confounded by the bowel preparation procedure.
 
 Here we aimed to address the limitations of previous studies and
-effectively characterize the microbiome in the lumen and mucosa of the
-proximal and distal healthy human colon. Our design used unprepared
-colonoscopy techniques to sample the natural community of each location
-of the gut without prior disruption of the native bacteria in 20 healthy
-volunteers. To address the inherent inter-individual variation in human
-microbiomes, we used a machine-learning classification algorithm trained
-on curated 16S rRNA sequencing reads to identify microbes specific to
-each location. We found that our classification models were able to
-separate mucosal and lumenal samples as well as differentiate between
-sides of the colon based on populations of specific microbes. By
-identifying the specific microbes we are poised to ask if and how the
-presence or disruption of the microbes at each site contribute to the
-development of the specific tumor subtypes of CRC in the proximal and
-distal human colon.
+effectively identify the microbes specific to the lumen and mucosa of
+the proximal and distal healthy human colon. Our design used an
+unprepared colonoscopy technique to sample the natural community of each
+location of the gut without prior disruption of the native bacteria in
+20 healthy volunteers. To address the inherent inter-individual
+variation in human microbiomes, we used a machine-learning
+classification algorithm trained on curated 16S rRNA sequencing reads to
+identify microbes specific to each location. We found that our
+classification models were able to separate mucosal and lumenal samples
+as well as differentiate between sides of the colon based on populations
+of specific microbes. By identifying the specific microbes we are poised
+to ask if and how the presence or disruption of the microbes at each
+site contribute to the development of the specific tumor subtypes of CRC
+in the proximal and distal human colon.
 
@@
 and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
-in human microbiome research (cite). Likewise, samples had varying
+in human microbiome research ((11)). Likewise, samples had varying
 levels of diversity at each site, irrespective of the individual (Figure
@@
 
-Here we identified bacterial taxa that are specific to the lumen and
-mucosa of the proximal and distal human colon from samples collected
-during unprepared colonoscopy. We found that all locations contained a
-range of phyla and a range of diversity, but that there was a wide
-variability between subjects. Pairwise comparisons of each of the sites
-revealed that the proximal mucosa and lumen were most similar to each
-other. Further, comparison of colonoscopy-collected samples with samples
-collected from stool at home showed that the distal lumen is most
-similar to stool at exit. Random Forest algorithms built on OTU
-abundances from each sample identified microbes that are particular to
-each location of the colon. Finally, we were able to detect some
-bacterial OTUs associated with colorectal cancer in our healthy patient
-cohort. Using unprepped colonoscopy and machine learning, we have
-identified bacterial phyla specific to the healthy proximal and distal
-human colon.
+Here we identified bacterial taxa that were specific to the lumen and
+mucosa of the proximal and distal sides of the human colon from samples
+collected during unprepared colonoscopy. We found that all locations
+contained a range of phyla and a range of diversity, but that there was
+a wide variability between subjects. Pairwise comparisons of each of the
+sites revealed that the proximal mucosa and lumen were most similar to
+each other. Further, comparison of colonoscopy-collected samples with
+samples collected from stool at home showed that the distal lumen is
+most similar to stool at exit. Random Forest models built on OTU
+relative abundances from each sample identified microbes that are
+particular to each location of the colon. Finally, we were able to
+detect some bacterial OTUs associated with colonic disease in our
+healthy patient cohort. Using unprepped colonoscopy and machine
+learning, we have identified bacterial phyla specific to the healthy
+proximal and distal human colon.
 
@@
 site, there was a wide amount of variation for each phyla with
-communities primarily dominated by the Bacteriodes and Firmicutes. This
-likely reflects not only the variability between human subjects, casued
-by differences in age, gender, diet, but also reports of microbial
-``patchiness'' in the gut microbiome. Several previous studies have
-noted that the bacteria recoverable from the same mucosal sample
-location can be vastly different when the samples are taken just 1 cm
-away from each other ((15)). Similar patchiness is also observed in
+communities primarily dominated by the \emph{Bacteriodes} and
+\emph{Firmicutes}. This likely reflects not only the variability between
+human subjects, casued by differences in age, gender, diet, but also
+reports of biogeogrphical ``patchiness'' in the gut microbiome. Several
+studies have noted that the bacteria recoverable from the same mucosal
+sample location can be vastly different when the samples are taken just
+1 cm away from each other (15). Similar patchiness is also observed in
 lumenal contents and fecal samples themselves; there is observed
 separation of different interacting microbes along the length of a stool
-sample, for instance ((16)). That said, across our samples the mucosal
-samples harbor more Proteobacteria, consistent with previous studies
-comparing mucosal swabs to lumenal content in humans ((4)). Hence, the
-conclusions we can draw from phyla analysis are likely confounded by
-differences in sampling and patchiness between subjects.
+sample, for instance (16). That said, across our samples the mucosal
+samples harbor more \emph{Proteobacteria}, consistent with previous
+studies comparing mucosal swabs to lumenal content in humans (5). Hence,
+the conclusions we can draw from phyla analysis are likely impacted by
+patchiness between subjects.
 
-To get around the noisiness created by a diverse set of samples, we
-built a Random Forest model to identify microbes specific to each side.
-For each comparison we identified top X OTUs that were strongly
-predictive of one site or another. Generally, OTUs identified in each
-location were consistent with known physiological gradients along the
-gut axis ((5)). For instance, the proximal mucosa harbored
-mucosa-associated facultative anaerobes such as \emph{Actinomyces} and
-\emph{Enterobacteraceae} and the aerobic \emph{Psuedomonas} consistent
-with the highest oxygen regions of the colon. The distal mucosa was far
-more likely to host strictly anaerobic species such as
-\emph{Porphyromonas}, \emph{Anaerococcus}, \emph{Finegoldia} and
-\emph{Peptoniphilus}. The model was less effective at classifying the
-proximal and distal lumenal contents, probably because the samples are
-arguably composed of the same material.
+To get around the noisiness from a diverse set of samples, we built a
+Random Forest model to identify microbes specific to each side. For each
+comparison we identified top X OTUs that were strongly predictive of one
+site or another. Generally, OTUs identified in each location were
+consistent with known physiological gradients along the gut axis (6).
+For instance, the proximal mucosa contains the highest oxygen
+concentratons of the colon and harbored mucosa-associated facultative
+anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
+aerobic \emph{Psuedomonas}. The distal mucosa was far more likely to
+host strictly anaerobic species such as \emph{Porphyromonas},
+\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. The
+model was less effective at classifying the proximal and distal lumenal
+contents, probably because the samples are arguably composed of the same
+bacteria but differ in water content.
 
@@
 of our subjects, respectively. These bacteria have been shown to be
-predictive of colorectal cancer in humans ((9)) and have oncogenic
-properties in cell culture and in mice ((17)). Interestingly, while
+predictive of colorectal cancer in humans (8) and have oncogenic
+properties in cell culture and in mice (17). Interestingly, while
 \emph{F. nucleatum} was found on both sides of the colon, \emph{P.
@@
 might not be surprising that it resides in the less-oygen-rich and
-proteinaceous distal mucosa ((4)). In studies examining bacteria on
+proteinaceous distal mucosa ((5)). In studies examining bacteria on
 colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
 on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreases along the colon to rectum ((18)). Of the (8) (40\%)
-individuals positive for \emph{F. nucleatum} in this study, the bacteria
-was spread across the proximal mucosa, distal lumen and distal mucosa.
-The presence of \emph{F. nucleatum} in a healthy individual is not
-necessarily linked to the development of future colorectal cancers.
-Because of the spatial distribution of the \emph{F. nucleatum} in our
-sample set, we cannot develop a model for the role of \emph{F.
-nucleatum} in the healthy colon. Data examining bacterial biofilms on
-CRC tumors suggests that \emph{Fusobacteria} species are more commonly
-found both on proximal tumors and in biofilms, indicating that it is not
-only the presence of the bacteria but the organization of the tumor
-community that contributes to \emph{Fusobacterium's} role in
-tumorigenesis ((8)). Finally, \emph{Fusobacteria} and
-\emph{Porphyromonas} species have been known to not only co-occur on CRC
-tumors but also to synergistically promote tumorigenesis in an oral
-cancer model ((19), (20)). Thus, further analysis of the distribution
-and activities of these pathogens may elucidate a mechanism for
-development of CRC tumors in the proximal or distal colon.
+decreases along the colon to rectum ((18)). Of the 8 (40\%) individuals
+positive for \emph{F. nucleatum} in this study, the bacterium was spread
+across the proximal mucosa, distal lumen and distal mucosa. The presence
+of \emph{F. nucleatum} in a healthy individual is not necessarily linked
+to the development of future colorectal cancers (cite). Data examining
+bacterial biofilms on CRC tumors suggests that \emph{Fusobacteria}
+species are more commonly found on proximal tumors and in biofilms,
+indicating that it is not only the presence of the bacteria but the
+organization of the tumor community that contributes to
+\emph{Fusobacterium's} role in tumorigenesis ((7)). Finally,
+\emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
+not only co-occur on CRC tumors but also to synergistically promote
+tumorigenesis in an oral cancer model ((19), (20)). Thus, further
+analysis of the distribution and activities of these pathogens may
+elucidate a mechanism for development of IBD or CRC subtypes in the
+proximal or distal colon.
+
+The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
+been commonly isolated from mucosal biopsies of patients with IBD
+(cite). Laboratory experiments with these isolates have shown that
+disease-isolated \emph{F. nucleatum} are more invasive and stimulate
+more TNF-a production than strains from healthy individuals ((21)),
+suggesting the bacteria may increase inflammation in the gut as well
+(cite - Dharmani 2011). \emph{F. varium} isolated from UC patients
+caused colonic ulcers in an experimental mouse model (Ohkusa 2003).
+\emph{F. varium} was only detected in 3 of our study participants and
+two of those samples were isolated form the proximal mucosa (cite
+figure). Although there is low occurrance in our study, \emph{F. varium}
+is most commonly isolated from UC patient biopsies from the ileum or
+cecum (cite), suggesting this species may exhibit preference for the
+different environmental conditions of these gastrointestinal sites.
+Further work will assess how gut environment may select for species
+which then cause localized disease.
 
@@
 preparation methods in most other studies. A rare report of a
-matched-colonoscopy study that sampled 18 patient's colonic mucosa and
-lumenal contents prior to and after bowel cleansing ((21)). This group
-found that mucosa and lumenal samples were distinguishable prior to
+matched-colonoscopy study sampled 18 patient's colonic mucosa and
+lumenal contents prior to and after bowel cleansing ((22)). This group
+found that mucosal and lumenal samples were distinguishable prior to
 bowel cleansing, but that bowel preparation resulted in an increase in
-shared OTUs between each site ((21)). Bowel cleansing not only made the
+shared OTUs between each site ((22)). Bowel cleansing not only made the
 samples harder to distinguish, it resulted in decreases in diversity
 across sites. Further, the differences were not great enough to overcome
-interpersonal differences between subjects. So, bowel preparation
-clearly induces bias into the microbes recovered from sampling the lumen
-or mucosa of a prepared bowel. Thus our findings of specific bacteria at
-each site of the colon are strengthened by the lack of bowel
-preparation.
-
-Microbiome-based diagnostics are increasingly being explored as
-non-invasive tools to survey for the development of colon cancer. Random
-Forest models have been used by our group and others to increase the
-detection sensitivity of CRC tumors. Indeed, our group found that a
-classification model that used microbiome data in combination with Fecal
-Immunohistochemical Test (FIT) results could correctly identify both
-carcinoma and adenoma lesions from communities of stool at exit and it
-performed much better than FIT alone. Further work from our lab has
-shown that microbiome profiling of the FIT cartridge contents
-sufficiently represented the stool community ((9)). One caveat of the
-FIT study was that there was not sufficient information to test if a
-classification model could differentiate between proximal and distal CRC
-tumors based on exit stool sample alone, but we would hypothesize that
-would not be effective. Given that our results showed that the stool
-most accurately reflects the community of the distal lumen, we likely
-cannot use Random Forest of stool samples to diagnose any changes in the
-proximal or mucosal communities.
+interpersonal differences between subjects. Bowel preparation clearly
+induces bias into the microbes recovered from sampling the lumen or
+mucosa of a prepared bowel. Thus our findings in this study are
+strengthened by the lack of bowel preparation.
 
@@
 form hypothesies about how specific host-microbe interactions can affect
-oncogenesis of proximal and distal CRC tumors. To this point, 16S
-community profiling studies do not provide enough information to probe
-these questions. Our sample set of matched proximal, distal, lumenal and
-mucosal samples from colons that have not undergone bowel preparation
-presents a unique opportunity to explore further questions about the
-microbiome along the gut axis. Specifically, examining metagenomic,
-metabolomic and host interactions at each site will be useful in further
-characterizing the host-microbe interactions in the development of
-proximal and distal colorectal cancer.
+disease progression of proximal and distal CRC and IBD subtypes. To this
+point, 16S rRNA gene sequecing community profiling studies do not
+provide enough information to fully probe these questions. In
+particular, 16S sequencing cannot not profile the host characteristics
+at each site. Combining the unprepared colonoscopy approach with
+analysis of multi-omic sequencing data may be useful in further
+characterizing host-microbiome interactions along the gut axis for both
+health and disease.
 
@@
 scheduled colonoscopy, subjects were to collect whole stool at home and
-two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.)
-and ship the samples to the University. Notably, subjects did not
-undergo any bowel preparation method prior to sampling. On procedure
+ship the samples to the University on ice. Notably, subjects did not
+undergo any bowel preparation method prior to sampling. On the procedure
 day, subjects reported to the Michigan Clinical Research Unit at the
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described ((22)). Sequences were curated
-using the mothur software as described previously ((23)). The sequences
+Illumina MiSeq as previously described ((23)). Sequences were curated
+using the mothur software as described previously ((24)). The sequences
 were assigned taxonomic classification using a naive Bayesian classifier
 trained using a 16S rRNA gene training set from the Ribosomal Database
-Project (RDP) ((24)) and clustered into operational taxonomic units
+Project (RDP) ((25)) and clustered into operational taxonomic units
 (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis of a
 mock community revealed the error rate to be X\%. Samples were rarefied
-to 4231 sequences per sample in order to reduce sampling bias.
+to 4231 sequences per sample in order to reduce uneven sampling bias.
 
 Diversity analysis was performed using the Simpson diversity calculator
-and theta YC calculator metrics in mothur ((23)). ThetaYC distances were
-calculated to determine the dissimilarity between two samples. Random
-Forest classification models were built using the randomForest R package
-and resultant models were used to identify the OTUs that were most
-important for classifying each location ((25)). To get species-level
-information about sequences of interest, sequences were aligned using
-blastn and the species name was only used if the identity score was
-\textgreater{}= 99\%.
+and theta YC calculator metrics in mothur version
+\textless{}3.2?\textgreater{} ((24)). ThetaYC distances were calculated
+to determine the dissimilarity between two samples. Random Forest
+classification models were built using the randomForest R package and
+resultant models were used to identify the OTUs that were most important
+for classifying each location ((26)). To get species-level information
+about sequences of interest, sequences were aligned using blastn and the
+species name was only used if the identity score was \textgreater{}=
+99\%.
 
@@
 procedure.
+
+\paragraph{Data availability}\label{data-availability}
+
+\paragraph{Github repository}\label{github-repository}
 
@@
 
-Similarity of microbial community structure between sites of the gut.
+Distances of microbial community structure between sites of the gut.
 ThetaYC distances are shown for interpersonal similarities between two
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Guinney2015}{}
-1. \textbf{Guinney J}, \textbf{Dienstmann R}, \textbf{Wang X},
-\textbf{Reyniès A de}, \textbf{Schlicker A}, \textbf{Soneson C},
-\textbf{Marisa L}, \textbf{Roepman P}, \textbf{Nyamundanda G},
-\textbf{Angelino P}, \textbf{Bot BM}, \textbf{Morris JS}, \textbf{Simon
-IM}, \textbf{Gerster S}, \textbf{Fessler E}, \textbf{Melo FDSE},
-\textbf{Missiaglia E}, \textbf{Ramay H}, \textbf{Barras D},
-\textbf{Homicsko K}, \textbf{Maru D}, \textbf{Manyam GC}, \textbf{Broom
-B}, \textbf{Boige V}, \textbf{Perez-Villamil B}, \textbf{Laderas T},
-\textbf{Salazar R}, \textbf{Gray JW}, \textbf{Hanahan D},
-\textbf{Tabernero J}, \textbf{Bernards R}, \textbf{Friend SH},
-\textbf{Laurent-Puig P}, \textbf{Medema JP}, \textbf{Sadanandam A},
-\textbf{Wessels L}, \textbf{Delorenzi M}, \textbf{Kopetz S},
-\textbf{Vermeulen L}, \textbf{Tejpar S}. 2015. The consensus molecular
-subtypes of colorectal cancer. Nature Medicine \textbf{21}:1350--1356.
-doi:\href{https://doi.org/10.1038/nm.3967}{10.1038/nm.3967}.
+\hypertarget{ref-Yamauchi2012}{}
+1. \textbf{Yamauchi M}, \textbf{Lochhead P}, \textbf{Morikawa T},
+\textbf{Huttenhower C}, \textbf{Chan AT}, \textbf{Giovannucci E},
+\textbf{Fuchs C}, \textbf{Ogino S}. 2012. Colorectal cancer: A tale of
+two sides or a continuum?: Figure 1. Gut \textbf{61}:794--797.
+doi:\href{https://doi.org/10.1136/gutjnl-2012-302014}{10.1136/gutjnl-2012-302014}.
+
+\hypertarget{ref-Forbes2016}{}
+2. \textbf{Forbes JD}, \textbf{Domselaar GV}, \textbf{Bernstein CN}.
+2016. The gut microbiota in immune-mediated inflammatory diseases.
+Frontiers in Microbiology \textbf{7}.
+doi:\href{https://doi.org/10.3389/fmicb.2016.01081}{10.3389/fmicb.2016.01081}.
+
+\hypertarget{ref-Halfvarson2017}{}
+3. \textbf{Halfvarson J}, \textbf{Brislawn CJ}, \textbf{Lamendella R},
+\textbf{Vázquez-Baeza Y}, \textbf{Walters WA}, \textbf{Bramer LM},
+\textbf{D\textquotesingleAmato M}, \textbf{Bonfiglio F},
+\textbf{McDonald D}, \textbf{Gonzalez A}, \textbf{McClure EE},
+\textbf{Dunklebarger MF}, \textbf{Knight R}, \textbf{Jansson JK}. 2017.
+Dynamics of the human gut microbiome in inflammatory bowel disease.
+Nature Microbiology \textbf{2}:17004.
+doi:\href{https://doi.org/10.1038/nmicrobiol.2017.4}{10.1038/nmicrobiol.2017.4}.
 
 \hypertarget{ref-Benedix2010}{}
-2. \textbf{Benedix F}, \textbf{Kube R}, \textbf{Meyer F},
+4. \textbf{Benedix F}, \textbf{Kube R}, \textbf{Meyer F},
 \textbf{Schmidt U}, \textbf{Gastinger I}, \textbf{Lippert H}. 2010.
@@
 
-\hypertarget{ref-Yamauchi2012}{}
-3. \textbf{Yamauchi M}, \textbf{Lochhead P}, \textbf{Morikawa T},
-\textbf{Huttenhower C}, \textbf{Chan AT}, \textbf{Giovannucci E},
-\textbf{Fuchs C}, \textbf{Ogino S}. 2012. Colorectal cancer: A tale of
-two sides or a continuum?: Figure 1. Gut \textbf{61}:794--797.
-doi:\href{https://doi.org/10.1136/gutjnl-2012-302014}{10.1136/gutjnl-2012-302014}.
-
 \hypertarget{ref-Albenberg2014}{}
-4. \textbf{Albenberg L}, \textbf{Esipova TV}, \textbf{Judge CP},
+5. \textbf{Albenberg L}, \textbf{Esipova TV}, \textbf{Judge CP},
 \textbf{Bittinger K}, \textbf{Chen J}, \textbf{Laughlin A},
@@
 \hypertarget{ref-Donaldson2015}{}
-5. \textbf{Donaldson GP}, \textbf{Lee SM}, \textbf{Mazmanian SK}. 2015.
+6. \textbf{Donaldson GP}, \textbf{Lee SM}, \textbf{Mazmanian SK}. 2015.
 Gut biogeography of the bacterial microbiota. Nature Reviews
@@
 
-\hypertarget{ref-Zackular2014}{}
-6. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
-\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
-for colorectal cancer. Cancer Prevention Research \textbf{7}:1112--1121.
-doi:\href{https://doi.org/10.1158/1940-6207.capr-14-0129}{10.1158/1940-6207.capr-14-0129}.
-
-\hypertarget{ref-Zackular2013}{}
-7. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
-\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
-tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
-doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.
-
 \hypertarget{ref-Dejea2014}{}
-8. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
+7. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
 \textbf{White JR}, \textbf{Welch JLM}, \textbf{Rossetti BJ},
@@
 \hypertarget{ref-Baxter2016}{}
-9. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+8. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-Jalanka2014}{}
-10. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
+9. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
 \textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
@@
 \hypertarget{ref-Harrell2012}{}
-11. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
+10. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
 \textbf{Young V}, \textbf{Lichtenstein L}, \textbf{Huang Y},
@@
 doi:\href{https://doi.org/10.1371/journal.pone.0032545}{10.1371/journal.pone.0032545}.
+
+\hypertarget{ref-LloydPrice2016}{}
+11. \textbf{Lloyd-Price J}, \textbf{Abu-Ali G}, \textbf{Huttenhower C}.
+2016. The healthy human microbiome. Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0307-y}{10.1186/s13073-016-0307-y}.
 
@@
 
+\hypertarget{ref-Strauss2011}{}
+21. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
+\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
+\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
+gut mucosa-derived fusobacterium nucleatum positively correlates with
+IBD status of the host. Inflammatory Bowel Diseases
+\textbf{17}:1971--1978.
+doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.
+
 \hypertarget{ref-Shobar2016}{}
-21. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
+22. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
 \textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
@@
 \hypertarget{ref-Kozich2013}{}
-22. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+23. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Schloss2009}{}
-23. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Wang2007}{}
-24. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+25. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
@@
 \hypertarget{ref-Liaw2002}{}
-25. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+26. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News: The Newsletter of the R Project

### 2017-07-21 submission/test.Rmd
found the bug, this compiles for now

--- 2017-07-18 submission/manuscript.Rmd
@@
 
-```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=TRUE}
 knitr::opts_chunk$set("tidy" = TRUE)
@@
 knitr::opts_chunk$set("warning" = FALSE)
-knitr::opts_chunk$set("cache" = FALSE)
+knitr::opts_chunk$set("cache" = TRUE)
 
@@
 Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
-The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
-IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012, @Forbes2016, @Halfvarson2017).
+
+*2 sentences in here are broken*
+
 The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
@@
 
-Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012, @Halfvarson2017).
-For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated (@Dejea2014).
-These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
-Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our human stool classification studies (@Baxter2016).
-_F.
-nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls (Strauss2011).
-Furthermore, studies of _F.
-nucleatum_ isolated from mucosal biopsies showed that more invasive  _F.
-nucleatum_ positively correlates with IBD disease level (Strauss2011).
-Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community (@Forbes2016).
-The molecular capabilities of these rare taxa may contribute to the colonic disease state.
-These studies often examined only shed human stool or the small intestine, preventing fine-resolution analysis of paired samples from the proximal and distal sides of the colon.
-Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
-Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location in the colon is largely undefined.
-Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.
+Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012).
+
+_missing last half of second paragraph here, hasnt been fully tested_
 
@@
 To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
-Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research (@LloydPrice2016).
+Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research (cite).
 Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
@@
 The constructed model used ((Xopt)) features for the proximal and ((Xopt)) for the distal.
-The models performed well when classifying these samples (`r round(cv10f_roc_proximal_bs$auc,3)` and `r round(cv10f_roc_distal_bs$auc, 3)`, respectively).
+The models performed well when classifying these samples (`r cv10f_roc_proximal_bs$auc` and `r cv10f_roc_distal_bs$auc`, respectively).
 The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
@@
 ### References
+
+##this is file from commit right before broken commit
+
+#testing intro

### 2017-07-21 submission/manuscript.Rmd
fixed the bug

@@
 ---
-title: Unprepared colonoscopy identifies microbiota specific to the proximal and distal human colon
+title: Unprepared colonoscopy identifies microbiota specific to the proximal and distal
+  human colon
 csl: mbio.csl
@@
   word_document: default
-geometry: margin=1.0in
-bibliography: references.bib
 header-includes:
@@
 - \linenumbers
+geometry: margin=1.0in
+bibliography: references.bib
 ---

### 2017-07-26 submission/manuscript.Rmd
fixed some inline values

@@
 We built the first model to classify the lumen versus mucosal samples for the proximal and distal sides, independently (Figure 4A).
-The constructed model used ((Xopt)) features for the proximal and ((Xopt)) for the distal.
+The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal.
 The models performed well when classifying these samples (`r round(cv10f_roc_proximal_bs$auc,3)` and `r round(cv10f_roc_distal_bs$auc, 3)`, respectively).
@@
 Next, we built a model to differentiate the proximal and distal lumenal samples.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal versus distal lumen (AUC = 0.755).
+The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, `r round(cv10f_roc_muc$auc,3)`) compared to the proximal versus distal lumen (`r round(cv10f_roc_lum$auc, 3)`).
 These models were able to explain ((X%)) of the variance, respectively.

### 2017-07-27 submission/manuscript.Rmd
updated inline values and citations for v2

@@
 ```{r oncogenic_otus, echo=FALSE}
-subsampled_meta <- merge(meta_file, subsample, by.x='group', by.y='row.names')
+subsampled_meta <- merge(meta_file, subsampled_file, by.x='group', by.y='row.names')
 fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]
+subsampled_to = 4231
 
@@
 names(p_152)[4] <- "Otu152_abund"
+
+fusoviv <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00472")]
+fusoviv[,5] <- (fusoviv[,4]/4321)*100
+names(fusoviv)[5] <- "Otu00472_relAbund"
 ```
@@
 Next, we built a model to differentiate the proximal and distal lumenal samples.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, `r round(cv10f_roc_muc$auc,3)`) compared to the proximal versus distal lumen (`r round(cv10f_roc_lum$auc, 3)`).
-These models were able to explain ((X%)) of the variance, respectively.
+The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = `r round(cv10f_roc_muc$auc,3)`) compared to the proximal versus distal lumen (AUC = `r round(cv10f_roc_lum$auc, 3)`).
+The model that separated proximal and distal mucosal samples used `r length(2:ncol(aucrf_data_LRlumen))` features.
 OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B).
+The model used `r length(2:ncol(aucrf_data_LRbowel))` features to separate proximal and distal lumenal samples.
 Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).
@@
 Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs.
-Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (XX% over full length).
+Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (100% over full length).
 This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
 (@Castellarin2011).
-The _Fusobacterium_ positive samples were located in x% of the the proximal and X% of the distal mucosa and represented as much as 1% of any sample (Figure 6A).
-OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (X% over full length), another bacterium commonly detected and isolated from colorectal tumors.
-OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6B).
+There were 14 samples with the _F.
+nucleatum subsp.
+animalis_ sequence.
+Four sequences were located in the proximal mucosa and three in the distal mucosal samples (Figure 6A).
+The second most frequent _Fusobacterium_ sequence was OTU472, which aligns with 99% identity to _F.
+varium_.
+In addition to _F.
+nucleatum_, _F.
+varium_ has been associated with IBD (@Lee2016).
+4 study participants harbored _F.
+varium_ and the samples were split evenly between the proximal and distal mucosa (Figure 6B).
+OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (99% over full length), another bacterium commonly detected and isolated from colorectal tumors.
+OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6C).
 Among the samples that were positive for _P.
@@
 To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side.
-For each comparison we identified top X OTUs that were strongly predictive of one site or another.
+For each comparison we identified top 5 OTUs that were strongly predictive of one site or another.
 Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015.
@@
 Not much is known about the distribution of _P.
-asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa (@Albenberg2014).
+asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa @Albenberg2014.
 In studies examining bacteria on colorectal cancer tumors, _F.
 nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F.
-nucleatum_ decreases along the colon to rectum (@Mima2016).
+nucleatum_ decreases along the colon to rectum @Mima2016.
 Of the 8 (40%) individuals positive for _F.
 nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa.
-The presence of _F.
-nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers (cite).
-Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014).
-Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016).
+Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014.
+Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model @Whitmore2014, @Flynn2016.
 Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.
 
-The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD (cite).
+The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD @Lee2016.
 Laboratory experiments with these isolates have shown that disease-isolated _F.
-nucleatum_ are more invasive and stimulate more TNF-a production than strains from healthy individuals (@Strauss2011), suggesting the bacteria may increase inflammation in the gut as well (cite - Dharmani 2011).
+nucleatum_ are more invasive and stimulate more TNF-a production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011.
 _F.
-varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model (Ohkusa 2003).
+varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model @Ohkusa2003.
 _F.
-varium_ was only detected in 3 of our study participants and two of those samples were isolated form the proximal mucosa (cite figure).
+varium_ was only detected in 3 of our study participants and two of those samples were isolated form the proximal mucosa (Figure 6B).
 Although there is low occurrance in our study, _F.
-varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum (cite), suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites.
+varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum @Ohkusa2002, suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites.
 Further work will assess how gut environment may select for species which then cause localized disease.
@@
 Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies.
-A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016).
-This group found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016).
+A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing @Shobar2016.
+This group found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016.
 Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites.
@@
 For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples.
-The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich2013).
-Sequences were curated using the mothur software as described previously (@Schloss2009).
-The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (@Wang2007) and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
+The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013.
+Sequences were curated using the mothur software as described previously @Schloss2009.
+The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
 Sequencing and analysis of a mock community revealed the error rate to be X%.
@@
 
-Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version <3.2?> (@Schloss2009).
+Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version <3.2?> @Schloss2009.
 ThetaYC distances were calculated to determine the dissimilarity between two samples.
-Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location (@Liaw2002).
+Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location @Liaw2002.
 To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.
@@
 Location and abundance of cancer-associated OTUs.
-Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum_ and (B) _Porphyromonas asacharolytica_
+Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum subsp.
+animalis_ (B) _Fusobacterium varium_ and (C) _Porphyromonas asacharolytica_
 

### 2017-07-27 submission/manuscript.tex
fixed the bug
updated inline values and citations for v2

@@
 for the proximal and distal sides, independently (Figure 4A). The
-constructed model used ((Xopt)) features for the proximal and ((Xopt))
-for the distal. The models performed well when classifying these samples
-(0.8 and 1.0, respectively). The OTUs that were most predictive of each
-site are identified by their greatest mean decrease in accuracy when
-removed from the model. For distinguishing the proximal lumen and
-mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
-\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
-were differentially abundant (Figure 4B). The model classifying the
-distal lumen and mucosa identified OTUs from \emph{Turicibacter},
-\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
-\emph{Anaerococcus} genera that could distinguish lumen from mucosal
-samples (Figure 4C). These results indicate that there are fine
-differences between the different sites of the colon, and that these can
-be traced down to specific OTUs on each side.
+constructed model used 14 features for the proximal and 6 for the
+distal. The models performed well when classifying these samples (0.8
+and 1.0, respectively). The OTUs that were most predictive of each site
+are identified by their greatest mean decrease in accuracy when removed
+from the model. For distinguishing the proximal lumen and mucosa, OTUs
+from the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
+two OTUs from the \emph{Enterobacteraceae} genera were differentially
+abundant (Figure 4B). The model classifying the distal lumen and mucosa
+identified OTUs from \emph{Turicibacter}, \emph{Finegoldia},
+\emph{Peptoniphilus} and two OTUs from the \emph{Anaerococcus} genera
+that could distinguish lumen from mucosal samples (Figure 4C). These
+results indicate that there are fine differences between the different
+sites of the colon, and that these can be traced down to specific OTUs
+on each side.
 
@@
 samples. The model performed best when distinguishing the proximal
-versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
-versus distal lumen (AUC = 0.755). These models were able to explain
-((X\%)) of the variance, respectively. OTUs that were differentially
-abundant between the distal and proximal mucosa included members of the
+versus distal mucosa (Figure 5A, AUC = 0.9) compared to the proximal
+versus distal lumen (AUC = 0.8). OTUs that were differentially abundant
+between the distal and proximal mucosa included members of the
 \emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
@@
 3. \textbf{Halfvarson J}, \textbf{Brislawn CJ}, \textbf{Lamendella R},
-\textbf{Vázquez-Baeza Y}, \textbf{Walters WA}, \textbf{Bramer LM},
-\textbf{D\textquotesingleAmato M}, \textbf{Bonfiglio F},
-\textbf{McDonald D}, \textbf{Gonzalez A}, \textbf{McClure EE},
-\textbf{Dunklebarger MF}, \textbf{Knight R}, \textbf{Jansson JK}. 2017.
-Dynamics of the human gut microbiome in inflammatory bowel disease.
-Nature Microbiology \textbf{2}:17004.
+\textbf{Vazquez-Baeza Y}, \textbf{Walters WA}, \textbf{Bramer LM},
+\textbf{DAmato M}, \textbf{Bonfiglio F}, \textbf{McDonald D},
+\textbf{Gonzalez A}, \textbf{McClure EE}, \textbf{Dunklebarger MF},
+\textbf{Knight R}, \textbf{Jansson JK}. 2017. Dynamics of the human gut
+microbiome in inflammatory bowel disease. Nature Microbiology
+\textbf{2}:17004.
 doi:\href{https://doi.org/10.1038/nmicrobiol.2017.4}{10.1038/nmicrobiol.2017.4}.

### 2017-07-27 submission/manuscript.Rmd
v2 tweaks done before final read through

@@
 #### Sample collection
-At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit supplies).
+At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit?).
 At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and ship the samples to the University on ice.
@@
 A flexible sigmoidoscope was first inserted about 25cm into the colon and endoscopy brush used to collect lumenal/stool contents.
-Two lumenal samples were collected and the contents immediately deposited into RNAlater (source) and flash-frozen in liquid nitrogen.
+Two lumenal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen.
 The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
@@
 The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
-Sequencing and analysis of a mock community revealed the error rate to be X%.
+Sequencing and analysis of a mock community revealed the error rate to be 1.8%.
 Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.
 
-Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version <3.2?> @Schloss2009.
+Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version 3.2?
+@Schloss2009.
 ThetaYC distances were calculated to determine the dissimilarity between two samples.
@@
 Differences in thetaYC distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
-
 #### Data availability
-
-#### Github repository
+16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number XXXX.
+A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017.
 

### 2017-07-27 code/build_figre3new.R

@@ -38,14 +38,14 @@ tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
 lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
   theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=16)) + 
   scale_x_discrete(limits = tycpositions, breaks = tycpositions,
-                   labels=c("R Mucosa vs R Lumen", "L Lumen vs R Lumen", "L Mucosa vs R Mucosa", "L Mucosa vs L Lumen")) +
+                   labels=c("P Mucosa vs P Lumen", "D Lumen vs P Lumen", "D Mucosa vs P Mucosa", "D Mucosa vs D Lumen")) +
   ylab("ThetaYC distance") 
 
 exitpositions <- c("RB_SS", "RS_SS", "LB_SS", "LS_SS")
 exit_plot <- ggplot(exittyc, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() +
   theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=16)) +
   scale_x_discrete(limits=exitpositions, breaks=exitpositions,
-                   labels=c("R Mucosa vs Stool", "R Lumen vs Stool", "L Mucosa vs Stool", "L Lumen vs Stool")) +
+                   labels=c("P Mucosa vs Stool", "P Lumen vs Stool", "D Mucosa vs Stool", "D Lumen vs Stool")) +
   theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")
 
 #trying to get adonis to work for comparisons 
@@ -141,7 +141,7 @@ wilcox.test(thetayc ~ same_pt, data = alltyc)
 # build and export figure 
 #export as PDF
 
-plot_file <- '~/Documents/Schloss_Lab/Flynn_LRColon_XXXX_2017/submission/figure_3.pdf'
+plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_3.pdf'
 pdf(file=plot_file, width=12, height=11)
 layout(matrix(c(1,
                 2,

### 2017-07-27 code/build_figure2.R

@@ -42,7 +42,7 @@ phy_plot <- ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(co
   geom_boxplot(aes(fill=variable), outlier.shape=21, outlier.size=2.5) + theme_bw() + 
   theme(axis.text = element_text(size= 16), axis.title= element_text(size=18), legend.text=element_text(size=14), legend.title=element_text(size=16)) +
   scale_x_discrete(limits = positions, breaks=positions, 
-                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+                   labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
   theme(axis.title.x=element_blank(), panel.grid.major.x = element_blank(),panel.grid.minor.x = element_blank()) +
   theme(legend.justification = c(0.999, 0.999), legend.position = c(0.999, 0.999)) + scale_fill_brewer(palette="Dark2", name="Phylum") +
   ylab("% Relative Abundance") 
@@ -57,7 +57,7 @@ simpmeta$location <- factor(simpmeta$location, c("LB","RB", "LS", "RS", "SS"))
 positions <- c("RB", "RS", "LB", "LS", "SS")
 simp_plot <- ggplot(simpmeta, aes(x=location, y=invsimpson, group =1)) +geom_point() +geom_jitter(width=0.2) +theme_bw() + ylab("Inverse Simpson Diversity") +
   scale_x_discrete(limits = positions, breaks=positions, 
-                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+                   labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
   theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 16), axis.title= element_text(size=18)) +
   stat_summary(aes(x=location, y=invsimpson), data = simpmeta, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.4)
 

### 2017-07-27 code/oncogenic_otus.R

@@ -37,7 +37,7 @@ positions <- c("RB", "RS", "LB", "LS", "SS")
 fuso_plot <- ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(width=0.3) +theme_bw() +
   ylab("F. nucleatum Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
-                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+                   labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
   theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
 
 #IBD fuso OTU - F. varium 
@@ -50,7 +50,7 @@ positions <- c("RB", "RS", "LB", "LS", "SS")
 fusoviv_plot <- ggplot(fusoviv, aes(x=location, y=Otu00472_relAbund)) +geom_jitter(width=0.3) +theme_bw() +
   ylab("F. varium Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
-                   labels=c("R Mucosa", "R Lumen", "L Mucosa", "L Lumen", "Stool")) +
+                   labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
   theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
 
 
@@ -62,7 +62,7 @@ p_asach <- shared_meta[, colnames(shared_meta) %in% c("Group", "patient", "locat
 p_asach[,4] <- (p_asach[,3]/4321) *100
 names(p_asach)[4] <- "Otu00152_abund"
 
-p_152 <-subsampled_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00152")]
+p_152 <-subsampled_meta[, colnames(subsampled_meta) %in% c("Group", "patient", "location", "Otu00152")]
 
 p_152[,4] <- (p_152[,3]/4321) *100
 names(p_152)[4] <- "Otu152_abund"
@@ -75,7 +75,7 @@ positions <- c("RB", "RS", "LB", "LS", "SS")
 porphy_plot <- ggplot(p_152, aes(x=location, y=Otu152_abund)) +geom_jitter(width=0.3) +theme_bw() +
[…]

### 2017-07-27 code/random_forest.R

@@ -195,8 +195,8 @@ plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
 mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.5)
 mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.5)
 legend('bottom', legend=c(#sprintf('Lumen vs Mucosa, 10-fold CV, AUC = 0.925'),
-                          sprintf('L Lumen vs L Mucosa, 10-fold CV, AUC =0.980'),
-                          sprintf('R Lumen vs R Mucosa, 10-fold CV, AUC = 0.797')
+                          sprintf('D Lumen vs D Mucosa, 10-fold CV, AUC =0.980'),
+                          sprintf('P Lumen vs P Mucosa, 10-fold CV, AUC = 0.797')
                                #sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
                                #sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
 ),lty=c(1, 1, 1), lwd=3, col=c('red', 'blue'), bty='n', cex=1.2)
@@ -209,8 +209,8 @@ plot(cv10f_roc_muc,col = 'green4', lwd=3, add=T, lty=1) #r vs l mucosa cross val
 plot(cv10f_roc_lum, col = 'orange', lwd=3, add=T, lty=1) #r vs l lumen cross validation
 mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.2)
 mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.2)
-legend('bottom', legend=c(sprintf('L mucosa vs R mucosa 10-fold CV, AUC = 0.912'),
-                          sprintf('L lumen vs R lumen 10-fold CV, AUC = 0.7551')
+legend('bottom', legend=c(sprintf('D mucosa vs P mucosa 10-fold CV, AUC = 0.912'),
+                          sprintf('D lumen vs P lumen 10-fold CV, AUC = 0.7551')
                           # sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
[…]

### 2017-07-27 submission/manuscript.Rmd
code tweaks to make it run

@@
 ```{r oncogenic_otus, echo=FALSE}
-subsampled_meta <- merge(meta_file, subsampled_file, by.x='group', by.y='row.names')
+subsampled_meta <- merge(meta_file, subsample, by.x='group', by.y='row.names')
 fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]
@@
 
-p_152 <-subsampled_meta[, colnames(shared_meta) %in% c("Group", "patient", "location", "Otu00152")]
+p_152 <-subsampled_meta[, colnames(subsampled_meta) %in% c("Group", "patient", "location", "Otu00152")]
 
-p_152[,4] <- (p_152[,3]/subsampled_to) *100
-names(p_152)[4] <- "Otu152_abund"
+p_152[,5] <- (p_152[,4]/subsampled_to) *100
+names(p_152)[5] <- "Otu152_abund"
 

### 2017-07-27 submission/manuscript.tex
code tweaks to make it run

@@
 versus distal mucosa (Figure 5A, AUC = 0.9) compared to the proximal
-versus distal lumen (AUC = 0.8). OTUs that were differentially abundant
-between the distal and proximal mucosa included members of the
+versus distal lumen (AUC = 0.8). The model that separated proximal and
+distal mucosal samples used 6 features. OTUs that were differentially
+abundant between the distal and proximal mucosa included members of the
 \emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
-\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B).
+\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B). The
+model used 44 features to separate proximal and distal lumenal samples.
 Differentially abundant OTUs of the proximal and distal lumen included
@@
 \emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
-\emph{Fusobacterium nucleatum subsp animalis} (XX\% over full length).
+\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
 This is the only species of \emph{Fusobacterium} known to have oncogenic
 properties and be found on the surfaces of colorectal cancer tumors.
-((14)). The \emph{Fusobacterium} positive samples were located in x\% of
-the the proximal and X\% of the distal mucosa and represented as much as
-1\% of any sample (Figure 6A). OTU152 was similar to the members of the
-\emph{Porphyromonas} genus and the most frequent sequence in that OTU
-aligned to \emph{Porphyromonas asacharolytica} (X\% over full length),
-another bacterium commonly detected and isolated from colorectal tumors.
-OTU152 was only detected on the distal mucosa, and in fact was one of
-the OTUs the classification model identified as separating distal and
-proximal sides (Figure 6B). Among the samples that were positive for
-\emph{P. asacharolytica}, the relative abundances for this OTU ranged
-from 0.01\% - 16\%. Thus, disease-associated OTUs could be found in our
-sample set of 20 healthy individuals.
+((14)). There were 14 samples with the \emph{F. nucleatum subsp.
+animalis} sequence. Four sequences were located in the proximal mucosa
+and three in the distal mucosal samples (Figure 6A). The second most
+frequent \emph{Fusobacterium} sequence was OTU472, which aligns with
+99\% identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
+\emph{F. varium} has been associated with IBD ((15)). 4 study
+participants harbored \emph{F. varium} and the samples were split evenly
+between the proximal and distal mucosa (Figure 6B). OTU152 was similar
+to the members of the \emph{Porphyromonas} genus and the most frequent
+sequence in that OTU aligned to \emph{Porphyromonas asacharolytica}
+(99\% over full length), another bacterium commonly detected and
+isolated from colorectal tumors. OTU152 was only detected on the distal
+mucosa, and in fact was one of the OTUs the classification model
+identified as separating distal and proximal sides (Figure 6C). Among
+the samples that were positive for \emph{P. asacharolytica}, the
+relative abundances for this OTU ranged from 0.01\% - 16\%. Thus,
+disease-associated OTUs could be found in our sample set of 20 healthy
+individuals.
 
@@
 sample location can be vastly different when the samples are taken just
-1 cm away from each other (15). Similar patchiness is also observed in
+1 cm away from each other (16). Similar patchiness is also observed in
 lumenal contents and fecal samples themselves; there is observed
 separation of different interacting microbes along the length of a stool
-sample, for instance (16). That said, across our samples the mucosal
+sample, for instance (17). That said, across our samples the mucosal
 samples harbor more \emph{Proteobacteria}, consistent with previous
@@
 Random Forest model to identify microbes specific to each side. For each
-comparison we identified top X OTUs that were strongly predictive of one
+comparison we identified top 5 OTUs that were strongly predictive of one
 site or another. Generally, OTUs identified in each location were
@@
 predictive of colorectal cancer in humans (8) and have oncogenic
-properties in cell culture and in mice (17). Interestingly, while
+properties in cell culture and in mice (18). Interestingly, while
 \emph{F. nucleatum} was found on both sides of the colon, \emph{P.
@@
 might not be surprising that it resides in the less-oygen-rich and
-proteinaceous distal mucosa ((5)). In studies examining bacteria on
+proteinaceous distal mucosa (5). In studies examining bacteria on
 colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
 on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreases along the colon to rectum ((18)). Of the 8 (40\%) individuals
+decreases along the colon to rectum (19). Of the 8 (40\%) individuals
 positive for \emph{F. nucleatum} in this study, the bacterium was spread
-across the proximal mucosa, distal lumen and distal mucosa. The presence
-of \emph{F. nucleatum} in a healthy individual is not necessarily linked
-to the development of future colorectal cancers (cite). Data examining
-bacterial biofilms on CRC tumors suggests that \emph{Fusobacteria}
-species are more commonly found on proximal tumors and in biofilms,
-indicating that it is not only the presence of the bacteria but the
-organization of the tumor community that contributes to
-\emph{Fusobacterium's} role in tumorigenesis ((7)). Finally,
+across the proximal mucosa, distal lumen and distal mucosa. Data
+examining bacterial biofilms on the mucosa of CRC tumors suggests that
+\emph{Fusobacteria} species are more commonly found on proximal tumors
+and in biofilms, indicating that it is not only the presence of the
+bacteria but the organization of the tumor community that contributes to
+\emph{Fusobacterium's} role in tumorigenesis (7). Finally,
 \emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
 not only co-occur on CRC tumors but also to synergistically promote
-tumorigenesis in an oral cancer model ((19), (20)). Thus, further
-analysis of the distribution and activities of these pathogens may
-elucidate a mechanism for development of IBD or CRC subtypes in the
-proximal or distal colon.
+tumorigenesis in an oral cancer model (20), (21). Thus, further analysis
+of the distribution and activities of these pathogens may elucidate a
+mechanism for development of IBD or CRC subtypes in the proximal or
+distal colon.
 
 The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
-been commonly isolated from mucosal biopsies of patients with IBD
-(cite). Laboratory experiments with these isolates have shown that
+been commonly isolated from mucosal biopsies of patients with IBD (15).
+Laboratory experiments with these isolates have shown that
 disease-isolated \emph{F. nucleatum} are more invasive and stimulate
-more TNF-a production than strains from healthy individuals ((21)),
+more TNF-a production than strains from healthy individuals (22),
 suggesting the bacteria may increase inflammation in the gut as well
-(cite - Dharmani 2011). \emph{F. varium} isolated from UC patients
-caused colonic ulcers in an experimental mouse model (Ohkusa 2003).
-\emph{F. varium} was only detected in 3 of our study participants and
-two of those samples were isolated form the proximal mucosa (cite
-figure). Although there is low occurrance in our study, \emph{F. varium}
-is most commonly isolated from UC patient biopsies from the ileum or
-cecum (cite), suggesting this species may exhibit preference for the
-different environmental conditions of these gastrointestinal sites.
-Further work will assess how gut environment may select for species
-which then cause localized disease.
+(23). \emph{F. varium} isolated from UC patients caused colonic ulcers
+in an experimental mouse model (24). \emph{F. varium} was only detected
+in 3 of our study participants and two of those samples were isolated
+form the proximal mucosa (Figure 6B). Although there is low occurrance
+in our study, \emph{F. varium} is most commonly isolated from UC patient
+biopsies from the ileum or cecum (25), suggesting this species may
+exhibit preference for the different environmental conditions of these
+gastrointestinal sites. Further work will assess how gut environment may
+select for species which then cause localized disease.
 
@@
 matched-colonoscopy study sampled 18 patient's colonic mucosa and
-lumenal contents prior to and after bowel cleansing ((22)). This group
+lumenal contents prior to and after bowel cleansing (26). This group
 found that mucosal and lumenal samples were distinguishable prior to
 bowel cleansing, but that bowel preparation resulted in an increase in
-shared OTUs between each site ((22)). Bowel cleansing not only made the
+shared OTUs between each site (26). Bowel cleansing not only made the
 samples harder to distinguish, it resulted in decreases in diversity
@@
 At a baseline visit, subjects were consented and given a home collection
-stool kit (Source of kit supplies). At least one week prior to the
-scheduled colonoscopy, subjects were to collect whole stool at home and
-ship the samples to the University on ice. Notably, subjects did not
-undergo any bowel preparation method prior to sampling. On the procedure
-day, subjects reported to the Michigan Clinical Research Unit at the
+stool kit (Source of kit?). At least one week prior to the scheduled
+colonoscopy, subjects were to collect whole stool at home and ship the
+samples to the University on ice. Notably, subjects did not undergo any
+bowel preparation method prior to sampling. On the procedure day,
+subjects reported to the Michigan Clinical Research Unit at the
 University of Michigan Health System. Patients were consciously sedated
@@
 brush used to collect lumenal/stool contents. Two lumenal samples were
-collected and the contents immediately deposited into RNAlater (source)
+collected and the contents immediately deposited into RNAlater (Fischer)
 and flash-frozen in liquid nitrogen. The brushes were withdrawn and
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described ((23)). Sequences were curated
-using the mothur software as described previously ((24)). The sequences
+Illumina MiSeq as previously described (27). Sequences were curated
+using the mothur software as described previously (28). The sequences
 were assigned taxonomic classification using a naive Bayesian classifier
 trained using a 16S rRNA gene training set from the Ribosomal Database
-Project (RDP) ((25)) and clustered into operational taxonomic units
-(OTUs) based on a 97\% similarity cutoff. Sequencing and analysis of a
-mock community revealed the error rate to be X\%. Samples were rarefied
-to 4231 sequences per sample in order to reduce uneven sampling bias.
+Project (RDP) (29) and clustered into operational taxonomic units (OTUs)
+based on a 97\% similarity cutoff. Sequencing and analysis of a mock
+community revealed the error rate to be 1.8\%. Samples were rarefied to
+4231 sequences per sample in order to reduce uneven sampling bias.
 
 Diversity analysis was performed using the Simpson diversity calculator
-and theta YC calculator metrics in mothur version
-\textless{}3.2?\textgreater{} ((24)). ThetaYC distances were calculated
-to determine the dissimilarity between two samples. Random Forest
-classification models were built using the randomForest R package and
-resultant models were used to identify the OTUs that were most important
-for classifying each location ((26)). To get species-level information
-about sequences of interest, sequences were aligned using blastn and the
-species name was only used if the identity score was \textgreater{}=
-99\%.
+and theta YC calculator metrics in mothur version 3.2? (28). ThetaYC
+distances were calculated to determine the dissimilarity between two
+samples. Random Forest classification models were built using the
+randomForest R package and resultant models were used to identify the
+OTUs that were most important for classifying each location (30). To get
+species-level information about sequences of interest, sequences were
+aligned using blastn and the species name was only used if the identity
+score was \textgreater{}= 99\%.
 
@@
 test adjusted for multiple comparisons using the Benjamini-Hochberg
-procedure.
-
-\paragraph{Data availability}\label{data-availability}
-
-\paragraph{Github repository}\label{github-repository}
+procedure. \#\#\#\# Data availability 16S rRNA gene sequence reads and
+experiment metadata are available on the NCBI Sequence Read Archive
+(SRA) with accession number XXXX. A reproducible data analysis pipeline
+can be found at
+\url{https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017}.
 
@@
 calculated and plotted by sample site for each OTU of interest: (A)
-\emph{Fusobacterium nucleatum} and (B) \emph{Porphyromonas
-asacharolytica}
+\emph{Fusobacterium nucleatum subsp. animalis} (B) \emph{Fusobacterium
+varium} and (C) \emph{Porphyromonas asacharolytica}
 
@@
 
+\hypertarget{ref-Lee2016}{}
+15. \textbf{Lee Y}, \textbf{Eun CS}, \textbf{Lee AR}, \textbf{Park CH},
+\textbf{Han DS}. 2016. FusobacteriumIsolates recovered from colonic
+biopsies of inflammatory bowel disease patients in korea. Annals of
+Laboratory Medicine \textbf{36}:387.
+doi:\href{https://doi.org/10.3343/alm.2016.36.4.387}{10.3343/alm.2016.36.4.387}.
+
 \hypertarget{ref-Hong2011}{}
-15. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
+16. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
 \textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
@@
 \hypertarget{ref-Stearns2011}{}
-16. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
+17. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
 \textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
@@
 \hypertarget{ref-Sears2014}{}
-17. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
+18. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
 and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
@@
 \hypertarget{ref-Mima2016}{}
-18. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
+19. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
 \textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
@@
 \hypertarget{ref-Whitmore2014}{}
-19. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
+20. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
 cancer. PLoS Pathogens \textbf{10}:e1003933.
@@
 \hypertarget{ref-Flynn2016}{}
-20. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
+21. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
 Metabolic and community synergy of oral bacteria in colorectal cancer.
@@
 \hypertarget{ref-Strauss2011}{}
-21. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
+22. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
 \textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
@@
 
+\hypertarget{ref-Dharmani2011}{}
+23. \textbf{Dharmani P}, \textbf{Strauss J}, \textbf{Ambrose C},
+\textbf{Allen-Vercoe E}, \textbf{Chadee K}. 2011. Fusobacterium
+nucleatum infection of colonic cells stimulates MUC2 mucin and tumor
+necrosis factor alpha. Infection and Immunity \textbf{79}:2597--2607.
+doi:\href{https://doi.org/10.1128/iai.05118-11}{10.1128/iai.05118-11}.
+
+\hypertarget{ref-Ohkusa2003}{}
+24. \textbf{Ohkusa T}. 2003. Induction of experimental ulcerative
+colitis by fusobacterium varium isolated from colonic mucosa of patients
+with ulcerative colitis. Gut \textbf{52}:79--83.
+doi:\href{https://doi.org/10.1136/gut.52.1.79}{10.1136/gut.52.1.79}.
+
+\hypertarget{ref-Ohkusa2002}{}
+25. \textbf{Ohkusa T}, \textbf{Sato N}, \textbf{Ogihara T},
+\textbf{Morita K}, \textbf{Ogawa M}, \textbf{Okayasu I}. 2002.
+Fusobacterium varium localized in the colonic mucosa of patients with
+ulcerative colitis stimulates species-specific antibody. Journal of
+Gastroenterology and Hepatology \textbf{17}:849--853.
+doi:\href{https://doi.org/10.1046/j.1440-1746.2002.02834.x}{10.1046/j.1440-1746.2002.02834.x}.
+
 \hypertarget{ref-Shobar2016}{}
-22. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
+26. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
 \textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
@@
 \hypertarget{ref-Kozich2013}{}
-23. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Schloss2009}{}
-24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Wang2007}{}
-25. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+29. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
@@
 \hypertarget{ref-Liaw2002}{}
-26. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+30. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News: The Newsletter of the R Project

### 2017-07-28 submission/manuscript.Rmd
v2 done for Pat

@@
 
-inline_hook <- function(x){
-	print(x)
-
-	if(is.list(x)){
-		x <- unlist(x)
-	}
-
-	if(is.numeric(x)){
-		if(abs(x - round(x)) < .Machine$double.eps^0.5){
-			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
-		} else {
-			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
-		}
-	} else {
-    	paste(x)      
-	}
-}
-knitr::knit_hooks$set(inline=inline_hook)
+#knitr::knit_hooks$set(inline=inline_hook)
 ```
@@
 CRC tumors of the proximal and distal colon are morphologically and genetically distinct.
-Similarly, bowel diseases such as Crohn's are typically exacerbated in the proximal instestine while ulcerative colitis patients often experience symptoms in the distal colon.
+Similarly, inflammatory bowel diseases such as Crohn's are typically exacerbated in the proximal instestine while ulcerative colitis patients often experience symptoms in the distal colon.
 Previous analysis of the fecal microbiota from healthy and CRC or IBD patients has revealed different microbial signatures associated with these diseases.
@@
 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals.
-Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals.
+Diversity analysis revealed that each site contained a diverse community and that a patient’s samples were more similar to each other than to that of other individuals.
 Comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the distal lumen, likely reflecting the anatomical structure of the colon.
@@
 Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location.
-_Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas_ and _Actinomyces_ were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925).
-The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755).
+A model differentiating the proximal mucosa and lumen was built with an AUC of `r signif(cv10f_roc_proximal_bs$auc, digits=3)`.
+The proximal mucosa had a higher abundance of the genera _Enterobacteriaceae_ and _Bacteriodes_.
+_Peptoniphilus, Anaerococcus, Finegoldia,_ and _Turicibacter_ were most likely to be found in distal mucosal samples versus distal luminal samples (AUC = `r signif(cv10f_roc_distal_bs$auc, digits=3)`).
+The classification model performed well (AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`).
 The distal mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_.
@@
 The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
-IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012, @Forbes2016, @Halfvarson2017).
+IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012, @Forbes2016, @Halfvarson2017.
 The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
-For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms (@Benedix2010).
-In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010).
-The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients (@Yamauchi2012, @Albenberg2014, @Donaldson2015).
-In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon (@Forbes2016).
-UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare (@Forbes2016).
+For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms @Benedix2010.
+In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010.
+The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012, @Albenberg2014, @Donaldson2015.
+In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016.
+UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare @Forbes2016.
 Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC.
@@
 
-Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012, @Halfvarson2017).
-For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated (@Dejea2014).
+Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host @Yamauchi2012, @Halfvarson2017.
+For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated @Dejea2014.
 These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
-Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our human stool classification studies (@Baxter2016).
+Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our prior human stool classification studies @Baxter2016.
 _F.
-nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls (Strauss2011).
+nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls @Strauss2011.
 Furthermore, studies of _F.
 nucleatum_ isolated from mucosal biopsies showed that more invasive  _F.
-nucleatum_ positively correlates with IBD disease level (Strauss2011).
-Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community (@Forbes2016).
+nucleatum_ positively correlates with IBD disease level @Strauss2011.
+Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community @Forbes2016.
 The molecular capabilities of these rare taxa may contribute to the colonic disease state.
@@
 While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
-Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (@Jalanka2014, @Harrell2012).
+Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014, @Harrell2012.
 As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
@@
 To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
-Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research (@LloydPrice2016).
+Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research @LloydPrice2016.
 Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
@@
 Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
-To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005, @deCarcer2010, @Zhang2013).
+To determine what factors may be driving the differences seen among the samples, we compared ${\theta}$YC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
+We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon @Eckburg2005, @deCarcer2010, @Zhang2013.
 Thus interpersonal variation between subjects drives the differences between samples more than sample site or location.
@@
 The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal.
-The models performed well when classifying these samples (`r round(cv10f_roc_proximal_bs$auc,3)` and `r round(cv10f_roc_distal_bs$auc, 3)`, respectively).
+The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively).
 The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
@@
 The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C).
-These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.
+These results indicate that there are fine differences between the different sites of the colon, and that these can be traced to specific OTUs on each side.
 
 Next, we built a model to differentiate the proximal and distal lumenal samples.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = `r round(cv10f_roc_muc$auc,3)`) compared to the proximal versus distal lumen (AUC = `r round(cv10f_roc_lum$auc, 3)`).
+The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) compared to the proximal versus distal lumen (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`).
 The model that separated proximal and distal mucosal samples used `r length(2:ncol(aucrf_data_LRlumen))` features.
@@
 This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
-(@Castellarin2011).
+@Castellarin2011.
 There were 14 samples with the _F.
@@
 nucleatum_, _F.
-varium_ has been associated with IBD (@Lee2016).
+varium_ has been associated with IBD @Lee2016.
 4 study participants harbored _F.
@@
 Here we identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon from samples collected during unprepared colonoscopy.
-We found that all locations contained a range of phyla and a range of diversity, but that there was a wide variability between subjects.
+We found that all locations contained a range of phyla abundances and a range of diversity, but that there was a wide variability between subjects.
 Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other.
@@
 Finally, we were able to detect some bacterial OTUs associated with colonic disease in our healthy patient cohort.
-Using unprepped colonoscopy and machine learning, we have identified bacterial phyla specific to the healthy proximal and distal human colon.
+Using unprepped colonoscopy and machine learning, we have identified bacterial taxa specific to the healthy proximal and distal human colon.
 
-When examining the relative abundance of the different phyla at each site, there was a wide amount of variation for each phyla with communities primarily dominated by the _Bacteriodes_ and _Firmicutes_.
+When examining the relative abundance of the different phyla at each site, there was a wide amount of variation with communities primarily dominated by the _Bacteriodes_ and _Firmicutes_.
 This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of biogeogrphical "patchiness" in the gut microbiome.
@@
 _F.
-varium_ was only detected in 3 of our study participants and two of those samples were isolated form the proximal mucosa (Figure 6B).
+varium_ was only detected in three of our study participants and two of those samples were isolated form the proximal mucosa (Figure 6B).
 Although there is low occurrance in our study, _F.
 varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum @Ohkusa2002, suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites.
-Further work will assess how gut environment may select for species which then cause localized disease.
+Further work will assess how gut environment may select for species which may then cause localized disease.
 
@@
 
-Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur version 3.2?
-@Schloss2009.
+Diversity analysis was performed using the Simpson diversity calculator and thetaYC calculator metrics in mothur version 3.2 @Schloss2009.
 ThetaYC distances were calculated to determine the dissimilarity between two samples.

### 2017-07-28 submission/manuscript.tex
v2 done for Pat

@@
 are found in. CRC tumors of the proximal and distal colon are
-morphologically and genetically distinct. Similarly, bowel diseases such
-as Crohn's are typically exacerbated in the proximal instestine while
-ulcerative colitis patients often experience symptoms in the distal
-colon. Previous analysis of the fecal microbiota from healthy and CRC or
-IBD patients has revealed different microbial signatures associated with
-these diseases. We extended these observations of the fecal microbiome
-to include analysis of the proximal and distal healthy human colon. We
-used a two-colonoscope approach on subjects that had not undergone
-standard bowel preparation procedure. This technique allowed us to
-characterize the native proximal and distal luminal and mucosal
-microbiome without prior chemical disruption. 16S rRNA gene sequencing
-was performed on proximal and distal mucosal and luminal biopsies and
-home-collected stool for 20 healthy individuals. Diversity analysis
-revealed that each site contained a diverse community, and that a
-patient's samples were more similar to each other than to that of other
-individuals. Comparison of all samples to fecal samples taken at exit
-uncovered that the feces were most similar to samples taken from the
-distal lumen, likely reflecting the anatomical structure of the colon.
-Since we could not differentiate sites along the colon based on
-community structure or community membership alone, we employed the
-Random Forest machine-learning algorithm to identify key species that
-distinguish biogeographical sites. Random Forest classification models
-were built using taxa abundance and sample location and revealed
-distinct populations that were found in each location.
-\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
-\emph{Actinomyces} were most likely to be found in mucosal samples
-versus luminal samples (AUC = 0.925). The classification model performed
-well (AUC = 0.912) when classifying mucosal samples into proximal or
-distal sides, but separating luminal samples from each side proved more
-challenging (AUC = 0.755). The distal mucosa was found to have high
-populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
-Proximal and distal luminal samples were comprised of many of the same
-taxa, likely reflecting the fact that stool moves along the colon from
-the proximal to distal end. By sampling the unprepped human colon, our
-results have identified distinct bacterial populations native to the
-proximal and distal sides. Further investigation of these bacteria may
-elucidate if and how these groups contribute to different pathogenesis
-processes on the respective sides of the colon.
+morphologically and genetically distinct. Similarly, inflammatory bowel
+diseases such as Crohn's are typically exacerbated in the proximal
+instestine while ulcerative colitis patients often experience symptoms
+in the distal colon. Previous analysis of the fecal microbiota from
+healthy and CRC or IBD patients has revealed different microbial
+signatures associated with these diseases. We extended these
+observations of the fecal microbiome to include analysis of the proximal
+and distal healthy human colon. We used a two-colonoscope approach on
+subjects that had not undergone standard bowel preparation procedure.
+This technique allowed us to characterize the native proximal and distal
+luminal and mucosal microbiome without prior chemical disruption. 16S
+rRNA gene sequencing was performed on proximal and distal mucosal and
+luminal biopsies and home-collected stool for 20 healthy individuals.
+Diversity analysis revealed that each site contained a diverse community
+and that a patient's samples were more similar to each other than to
+that of other individuals. Comparison of all samples to fecal samples
+taken at exit uncovered that the feces were most similar to samples
+taken from the distal lumen, likely reflecting the anatomical structure
+of the colon. Since we could not differentiate sites along the colon
+based on community structure or community membership alone, we employed
+the Random Forest machine-learning algorithm to identify key species
+that distinguish biogeographical sites. Random Forest classification
+models were built using taxa abundance and sample location and revealed
+distinct populations that were found in each location. A model
+differentiating the proximal mucosa and lumen was built with an AUC of
+0.831. The proximal mucosa had a higher abundance of the genera
+\emph{Enterobacteriaceae} and \emph{Bacteriodes}. \emph{Peptoniphilus,
+Anaerococcus, Finegoldia,} and \emph{Turicibacter} were most likely to
+be found in distal mucosal samples versus distal luminal samples (AUC =
+0.98). The classification model performed well (AUC = 0.912) when
+classifying mucosal samples into proximal or distal sides, but
+separating luminal samples from each side proved more challenging (AUC =
+0.755). The distal mucosa was found to have high populations of
+\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
+distal luminal samples were comprised of many of the same taxa, likely
+reflecting the fact that stool moves along the colon from the proximal
+to distal end. By sampling the unprepped human colon, our results have
+identified distinct bacterial populations native to the proximal and
+distal sides. Further investigation of these bacteria may elucidate if
+and how these groups contribute to different pathogenesis processes on
+the respective sides of the colon.
 
@@
 Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
-accelerated by perturbations in gut microbes ((1), (2), (3)). The
+accelerated by perturbations in gut microbes (1), (2), (3). The
 severity, symptoms, morbidity and mortality of these diseases is known
@@
 instance, CRC tumors that arise on the distal side of the colon are
-infiltrating lesions that present with painful symptoms ((4)). In
+infiltrating lesions that present with painful symptoms (4). In
 contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
 are sessile and form along the wall of the colon, often remaining
-asymptomatic until advanced carcinogenesis ((4)). The distal and
-proximal sides of the colon differ in the amount of inflammation present
-and the genomic instability of precancerous cells, respectively, in
-addition to variation in the previously mentioned chemical gradients
-((1), (5), (6)). In IBD patients, disease flares in the distal colon are
-usually indicative of ulcerative colitis (UC), whereas Crohn's disease
-(CD) patients typically experience disease in the small intestine, ileum
-and proximal colon ((2)). UC presents as large and highly inflammed
-mucosal ulcers, where as CD lesions are often smaller and have areas of
-normal tissue distributed amongst the flare ((2)). Thus, given the
-varied physiology of the proximal-distal axis of the colon and known
+asymptomatic until advanced carcinogenesis (4). The distal and proximal
+sides of the colon differ in the amount of inflammation present and the
+genomic instability of precancerous cells, respectively, in addition to
+variation in the previously mentioned chemical gradients (1), (5), (6).
+In IBD patients, disease flares in the distal colon are usually
+indicative of ulcerative colitis (UC), whereas Crohn's disease (CD)
+patients typically experience disease in the small intestine, ileum and
+proximal colon (2). UC presents as large and highly inflammed mucosal
+ulcers, where as CD lesions are often smaller and have areas of normal
+tissue distributed amongst the flare (2). Thus, given the varied
+physiology of the proximal-distal axis of the colon and known
 differences in disease patterns at these sites, symbiotic microbes and
@@
 IBD or CRC can be attributed to specific molecular events as a result of
-interactions between the gut microbiota and human host ((1), (3)). For
+interactions between the gut microbiota and human host (1), (3). For
 instance, comparison of the bacteria present on CRC tumors with those
 found on nearby healthy tissue has identified specific species that are
-tumor-associated ((7)). These species include the oral pathogens
+tumor-associated (7). These species include the oral pathogens
 \emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
 Interestingly, these periodontal pathogens have been highly predictive
-of whether a patient had CRC tumors or not in our human stool
-classification studies ((8)). \emph{F. nucleatum} has also been found to
+of whether a patient had CRC tumors or not in our prior human stool
+classification studies (8). \emph{F. nucleatum} has also been found to
 be elevated in the stool and biopsies of patients with IBD as compared
-to healthy controls (Strauss2011). Furthermore, studies of \emph{F.
-nucleatum} isolated from mucosal biopsies showed that more invasive
-\emph{F. nucleatum} positively correlates with IBD disease level
-(Strauss2011). Like many intestinal pathogens, the bacteria appear to
-have a high-impact despite being lowly-abundant in the community ((2)).
-The molecular capabilities of these rare taxa may contribute to the
-colonic disease state. These studies often examined only shed human
-stool or the small intestine, preventing fine-resolution analysis of
-paired samples from the proximal and distal sides of the colon.
-Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have
-matched tissue from the other side of the colon from the same patient,
-limiting what conclusions can be drawn about the colonic microbiome
-overall, let alone at that specific site. Due to these limitations, the
-contribution of the gut microbiota to IBD and CRC disease location in
-the colon is largely undefined. Characterizing these communities could
-provide needed insight into disease etiology, including how the
-disruption of the healthy community could promote the initiation or
-proliferation of the distinct proximal and distal CRC tumors or IBD
-flares.
+to healthy controls (9). Furthermore, studies of \emph{F. nucleatum}
+isolated from mucosal biopsies showed that more invasive \emph{F.
+nucleatum} positively correlates with IBD disease level (9). Like many
+intestinal pathogens, the bacteria appear to have a high-impact despite
+being lowly-abundant in the community (2). The molecular capabilities of
+these rare taxa may contribute to the colonic disease state. These
+studies often examined only shed human stool or the small intestine,
+preventing fine-resolution analysis of paired samples from the proximal
+and distal sides of the colon. Similarly, comparisons of on- or
+off-tumor/lesion bacteria rarely have matched tissue from the other side
+of the colon from the same patient, limiting what conclusions can be
+drawn about the colonic microbiome overall, let alone at that specific
+site. Due to these limitations, the contribution of the gut microbiota
+to IBD and CRC disease location in the colon is largely undefined.
+Characterizing these communities could provide needed insight into
+disease etiology, including how the disruption of the healthy community
+could promote the initiation or proliferation of the distinct proximal
+and distal CRC tumors or IBD flares.
 
@@
 colon, but complicates microbiome profiling as the chemicals strip the
-bowel of contents and disrupt the mucosal layer ((9), (10)). As such,
+bowel of contents and disrupt the mucosal layer (10), (11). As such,
 what little information we do have about the biogeographical
@@
 and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
-in human microbiome research ((11)). Likewise, samples had varying
-levels of diversity at each site, irrespective of the individual (Figure
-2B). For example, the proximal mucosa was more diverse than the distal
-for some individuals while the opposite was true for others. Therefore
-we could not identify a clear pattern of changes in microbial diversity
+in human microbiome research (12). Likewise, samples had varying levels
+of diversity at each site, irrespective of the individual (Figure 2B).
+For example, the proximal mucosa was more diverse than the distal for
+some individuals while the opposite was true for others. Therefore we
+could not identify a clear pattern of changes in microbial diversity
 along the gut axis.
@@
 determine what factors may be driving the differences seen among the
-samples, we compared thetaYC distances between samples from all subjects
-(interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were far more similar to each
-other than to other study subjects (Figure 3C), consistent with previous
-human microbiome studies that have sampled multiple sites of the human
-colon (\textbf{???}, (12), (13)). Thus interpersonal variation between
-subjects drives the differences between samples more than sample site or
-location. Overall, the results comparing the structure of the
-communities suggest that the contents of the distal lumen are most
-representative of stool at exit, and the microbes remaining on the
-mucosa are much different.
+samples, we compared \({\theta}\)YC distances between samples from all
+subjects (interpersonal) versus samples from within one subject
+(intrapersonal). We found that samples from one individual were far more
+similar to each other than to other study subjects (Figure 3C),
+consistent with previous human microbiome studies that have sampled
+multiple sites of the human colon (13), (14), (15). Thus interpersonal
+variation between subjects drives the differences between samples more
+than sample site or location. Overall, the results comparing the
+structure of the communities suggest that the contents of the distal
+lumen are most representative of stool at exit, and the microbes
+remaining on the mucosa are much different.
 
@@
 constructed model used 14 features for the proximal and 6 for the
-distal. The models performed well when classifying these samples (0.8
-and 1.0, respectively). The OTUs that were most predictive of each site
+distal. The models performed well when classifying these samples (0.831
+and 0.98, respectively). The OTUs that were most predictive of each site
 are identified by their greatest mean decrease in accuracy when removed
@@
 results indicate that there are fine differences between the different
-sites of the colon, and that these can be traced down to specific OTUs
-on each side.
+sites of the colon, and that these can be traced to specific OTUs on
+each side.
 
@@
 samples. The model performed best when distinguishing the proximal
-versus distal mucosa (Figure 5A, AUC = 0.9) compared to the proximal
-versus distal lumen (AUC = 0.8). The model that separated proximal and
+versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
+versus distal lumen (AUC = 0.755). The model that separated proximal and
 distal mucosal samples used 6 features. OTUs that were differentially
@@
 properties and be found on the surfaces of colorectal cancer tumors.
-((14)). There were 14 samples with the \emph{F. nucleatum subsp.
-animalis} sequence. Four sequences were located in the proximal mucosa
-and three in the distal mucosal samples (Figure 6A). The second most
-frequent \emph{Fusobacterium} sequence was OTU472, which aligns with
-99\% identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
-\emph{F. varium} has been associated with IBD ((15)). 4 study
-participants harbored \emph{F. varium} and the samples were split evenly
-between the proximal and distal mucosa (Figure 6B). OTU152 was similar
-to the members of the \emph{Porphyromonas} genus and the most frequent
-sequence in that OTU aligned to \emph{Porphyromonas asacharolytica}
-(99\% over full length), another bacterium commonly detected and
-isolated from colorectal tumors. OTU152 was only detected on the distal
-mucosa, and in fact was one of the OTUs the classification model
-identified as separating distal and proximal sides (Figure 6C). Among
-the samples that were positive for \emph{P. asacharolytica}, the
-relative abundances for this OTU ranged from 0.01\% - 16\%. Thus,
-disease-associated OTUs could be found in our sample set of 20 healthy
-individuals.
+(16). There were 14 samples with the \emph{F. nucleatum subsp. animalis}
+sequence. Four sequences were located in the proximal mucosa and three
+in the distal mucosal samples (Figure 6A). The second most frequent
+\emph{Fusobacterium} sequence was OTU472, which aligns with 99\%
+identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
+\emph{F. varium} has been associated with IBD (17). 4 study participants
+harbored \emph{F. varium} and the samples were split evenly between the
+proximal and distal mucosa (Figure 6B). OTU152 was similar to the
+members of the \emph{Porphyromonas} genus and the most frequent sequence
+in that OTU aligned to \emph{Porphyromonas asacharolytica} (99\% over
+full length), another bacterium commonly detected and isolated from
+colorectal tumors. OTU152 was only detected on the distal mucosa, and in
+fact was one of the OTUs the classification model identified as
+separating distal and proximal sides (Figure 6C). Among the samples that
+were positive for \emph{P. asacharolytica}, the relative abundances for
+this OTU ranged from 0.01\% - 16\%. Thus, disease-associated OTUs could
+be found in our sample set of 20 healthy individuals.
 
@@
 collected during unprepared colonoscopy. We found that all locations
-contained a range of phyla and a range of diversity, but that there was
-a wide variability between subjects. Pairwise comparisons of each of the
-sites revealed that the proximal mucosa and lumen were most similar to
-each other. Further, comparison of colonoscopy-collected samples with
-samples collected from stool at home showed that the distal lumen is
-most similar to stool at exit. Random Forest models built on OTU
-relative abundances from each sample identified microbes that are
+contained a range of phyla abundances and a range of diversity, but that
+there was a wide variability between subjects. Pairwise comparisons of
+each of the sites revealed that the proximal mucosa and lumen were most
+similar to each other. Further, comparison of colonoscopy-collected
+samples with samples collected from stool at home showed that the distal
+lumen is most similar to stool at exit. Random Forest models built on
+OTU relative abundances from each sample identified microbes that are
 particular to each location of the colon. Finally, we were able to
@@
 healthy patient cohort. Using unprepped colonoscopy and machine
-learning, we have identified bacterial phyla specific to the healthy
+learning, we have identified bacterial taxa specific to the healthy
 proximal and distal human colon.
@@
 When examining the relative abundance of the different phyla at each
-site, there was a wide amount of variation for each phyla with
-communities primarily dominated by the \emph{Bacteriodes} and
-\emph{Firmicutes}. This likely reflects not only the variability between
-human subjects, casued by differences in age, gender, diet, but also
-reports of biogeogrphical ``patchiness'' in the gut microbiome. Several
-studies have noted that the bacteria recoverable from the same mucosal
-sample location can be vastly different when the samples are taken just
-1 cm away from each other (16). Similar patchiness is also observed in
-lumenal contents and fecal samples themselves; there is observed
-separation of different interacting microbes along the length of a stool
-sample, for instance (17). That said, across our samples the mucosal
-samples harbor more \emph{Proteobacteria}, consistent with previous
-studies comparing mucosal swabs to lumenal content in humans (5). Hence,
-the conclusions we can draw from phyla analysis are likely impacted by
-patchiness between subjects.
+site, there was a wide amount of variation with communities primarily
+dominated by the \emph{Bacteriodes} and \emph{Firmicutes}. This likely
+reflects not only the variability between human subjects, casued by
+differences in age, gender, diet, but also reports of biogeogrphical
+``patchiness'' in the gut microbiome. Several studies have noted that
+the bacteria recoverable from the same mucosal sample location can be
+vastly different when the samples are taken just 1 cm away from each
+other (18). Similar patchiness is also observed in lumenal contents and
+fecal samples themselves; there is observed separation of different
+interacting microbes along the length of a stool sample, for instance
+(19). That said, across our samples the mucosal samples harbor more
+\emph{Proteobacteria}, consistent with previous studies comparing
+mucosal swabs to lumenal content in humans (5). Hence, the conclusions
+we can draw from phyla analysis are likely impacted by patchiness
+between subjects.
 
@@
 predictive of colorectal cancer in humans (8) and have oncogenic
-properties in cell culture and in mice (18). Interestingly, while
+properties in cell culture and in mice (20). Interestingly, while
 \emph{F. nucleatum} was found on both sides of the colon, \emph{P.
@@
 on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreases along the colon to rectum (19). Of the 8 (40\%) individuals
+decreases along the colon to rectum (21). Of the 8 (40\%) individuals
 positive for \emph{F. nucleatum} in this study, the bacterium was spread
@@
 not only co-occur on CRC tumors but also to synergistically promote
-tumorigenesis in an oral cancer model (20), (21). Thus, further analysis
+tumorigenesis in an oral cancer model (22), (23). Thus, further analysis
 of the distribution and activities of these pathogens may elucidate a
@@
 The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
-been commonly isolated from mucosal biopsies of patients with IBD (15).
+been commonly isolated from mucosal biopsies of patients with IBD (17).
 Laboratory experiments with these isolates have shown that
 disease-isolated \emph{F. nucleatum} are more invasive and stimulate
-more TNF-a production than strains from healthy individuals (22),
+more TNF-a production than strains from healthy individuals (9),
 suggesting the bacteria may increase inflammation in the gut as well
-(23). \emph{F. varium} isolated from UC patients caused colonic ulcers
-in an experimental mouse model (24). \emph{F. varium} was only detected
-in 3 of our study participants and two of those samples were isolated
-form the proximal mucosa (Figure 6B). Although there is low occurrance
-in our study, \emph{F. varium} is most commonly isolated from UC patient
-biopsies from the ileum or cecum (25), suggesting this species may
-exhibit preference for the different environmental conditions of these
-gastrointestinal sites. Further work will assess how gut environment may
-select for species which then cause localized disease.
+(24). \emph{F. varium} isolated from UC patients caused colonic ulcers
+in an experimental mouse model (25). \emph{F. varium} was only detected
+in three of our study participants and two of those samples were
+isolated form the proximal mucosa (Figure 6B). Although there is low
+occurrance in our study, \emph{F. varium} is most commonly isolated from
+UC patient biopsies from the ileum or cecum (26), suggesting this
+species may exhibit preference for the different environmental
+conditions of these gastrointestinal sites. Further work will assess how
+gut environment may select for species which may then cause localized
+disease.
 
@@
 matched-colonoscopy study sampled 18 patient's colonic mucosa and
-lumenal contents prior to and after bowel cleansing (26). This group
+lumenal contents prior to and after bowel cleansing (27). This group
 found that mucosal and lumenal samples were distinguishable prior to
 bowel cleansing, but that bowel preparation resulted in an increase in
-shared OTUs between each site (26). Bowel cleansing not only made the
+shared OTUs between each site (27). Bowel cleansing not only made the
 samples harder to distinguish, it resulted in decreases in diversity
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described (27). Sequences were curated
-using the mothur software as described previously (28). The sequences
+Illumina MiSeq as previously described (28). Sequences were curated
+using the mothur software as described previously (29). The sequences
 were assigned taxonomic classification using a naive Bayesian classifier
 trained using a 16S rRNA gene training set from the Ribosomal Database
-Project (RDP) (29) and clustered into operational taxonomic units (OTUs)
+Project (RDP) (30) and clustered into operational taxonomic units (OTUs)
 based on a 97\% similarity cutoff. Sequencing and analysis of a mock
@@
 Diversity analysis was performed using the Simpson diversity calculator
-and theta YC calculator metrics in mothur version 3.2? (28). ThetaYC
+and thetaYC calculator metrics in mothur version 3.2 (29). ThetaYC
 distances were calculated to determine the dissimilarity between two
@@
 randomForest R package and resultant models were used to identify the
-OTUs that were most important for classifying each location (30). To get
+OTUs that were most important for classifying each location (31). To get
 species-level information about sequences of interest, sequences were
@@
 
+\hypertarget{ref-Strauss2011}{}
+9. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
+\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
+\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
+gut mucosa-derived fusobacterium nucleatum positively correlates with
+IBD status of the host. Inflammatory Bowel Diseases
+\textbf{17}:1971--1978.
+doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.
+
 \hypertarget{ref-Jalanka2014}{}
-9. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
+10. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
 \textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
@@
 \hypertarget{ref-Harrell2012}{}
-10. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
+11. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
 \textbf{Young V}, \textbf{Lichtenstein L}, \textbf{Huang Y},
@@
 \hypertarget{ref-LloydPrice2016}{}
-11. \textbf{Lloyd-Price J}, \textbf{Abu-Ali G}, \textbf{Huttenhower C}.
+12. \textbf{Lloyd-Price J}, \textbf{Abu-Ali G}, \textbf{Huttenhower C}.
 2016. The healthy human microbiome. Genome Medicine \textbf{8}.
@@
 
+\hypertarget{ref-Eckburg2005}{}
+13. \textbf{Eckburg PB}. 2005. Diversity of the human intestinal
+microbial flora. Science \textbf{308}:1635--1638.
+doi:\href{https://doi.org/10.1126/science.1110591}{10.1126/science.1110591}.
+
 \hypertarget{ref-deCarcer2010}{}
-12. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
+14. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
 \textbf{Kang S}, \textbf{Worthley D}, \textbf{Whitehall V},
@@
 \hypertarget{ref-Zhang2013}{}
-13. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
+15. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
 \textbf{Xu J}, \textbf{Wen X}, \textbf{Ma Z (Sam)}, \textbf{Shi P}.
@@
 \hypertarget{ref-Castellarin2011}{}
-14. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
+16. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
 \textbf{Dreolini L}, \textbf{Krzywinski M}, \textbf{Strauss J},
@@
 \hypertarget{ref-Lee2016}{}
-15. \textbf{Lee Y}, \textbf{Eun CS}, \textbf{Lee AR}, \textbf{Park CH},
+17. \textbf{Lee Y}, \textbf{Eun CS}, \textbf{Lee AR}, \textbf{Park CH},
 \textbf{Han DS}. 2016. FusobacteriumIsolates recovered from colonic
@@
 \hypertarget{ref-Hong2011}{}
-16. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
+18. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
 \textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
@@
 \hypertarget{ref-Stearns2011}{}
-17. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
+19. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
 \textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
@@
 \hypertarget{ref-Sears2014}{}
-18. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
+20. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
 and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
@@
 \hypertarget{ref-Mima2016}{}
-19. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
+21. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
 \textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
@@
 \hypertarget{ref-Whitmore2014}{}
-20. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
+22. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
 cancer. PLoS Pathogens \textbf{10}:e1003933.
@@
 \hypertarget{ref-Flynn2016}{}
-21. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
+23. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
 Metabolic and community synergy of oral bacteria in colorectal cancer.
@@
 
-\hypertarget{ref-Strauss2011}{}
-22. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
-\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
-\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
-gut mucosa-derived fusobacterium nucleatum positively correlates with
-IBD status of the host. Inflammatory Bowel Diseases
-\textbf{17}:1971--1978.
-doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.
-
 \hypertarget{ref-Dharmani2011}{}
-23. \textbf{Dharmani P}, \textbf{Strauss J}, \textbf{Ambrose C},
+24. \textbf{Dharmani P}, \textbf{Strauss J}, \textbf{Ambrose C},
 \textbf{Allen-Vercoe E}, \textbf{Chadee K}. 2011. Fusobacterium
@@
 \hypertarget{ref-Ohkusa2003}{}
-24. \textbf{Ohkusa T}. 2003. Induction of experimental ulcerative
+25. \textbf{Ohkusa T}. 2003. Induction of experimental ulcerative
 colitis by fusobacterium varium isolated from colonic mucosa of patients
@@
 \hypertarget{ref-Ohkusa2002}{}
-25. \textbf{Ohkusa T}, \textbf{Sato N}, \textbf{Ogihara T},
+26. \textbf{Ohkusa T}, \textbf{Sato N}, \textbf{Ogihara T},
 \textbf{Morita K}, \textbf{Ogawa M}, \textbf{Okayasu I}. 2002.
@@
 \hypertarget{ref-Shobar2016}{}
-26. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
+27. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
 \textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
@@
 \hypertarget{ref-Kozich2013}{}
-27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+28. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Schloss2009}{}
-28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+29. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Wang2007}{}
-29. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+30. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
@@
 \hypertarget{ref-Liaw2002}{}
-30. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+31. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News: The Newsletter of the R Project

### 2017-08-16 submission/manuscript.Rmd
some Pat edits made

@@
 ---
-title: Unprepared colonoscopy identifies microbiota specific to the proximal and distal
-  human colon
+title: The microbiota of the proximal and distal human colon
 csl: mbio.csl
@@
 
-Running title: Unprepared colonoscopy identifies proximal and distal human colonic microbiota
+Running title: The microbiome of the proximal and distal human colon
 
@@
 
-Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3^, and Patrick D. Schloss^1$\dagger$^
+Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^
 
@@
 
-$\dagger$ Corresponding author: pschloss@umich.edu
+$\dagger$ Corresponding authors: kturgeon@med.umich.edu and pschloss@umich.edu
 
@@
 
-The human colon contains chemicals and nutrients that change along the proximal to distal gut axis.
-These gradients create microenvironments that affect the distrubution and composition of the gut microbiota.
-The microbiome has been implicated in the colonic diseases colorectal cancer (CRC) and inflammatory bowel disease (IBD).
+Chemical and nutrient gradients along the human colon create microenvironments that affect the distrubution and composition of the gut microbiota.
+The microbiome has been implicated in colorectal cancer (CRC) and inflammatory bowel disease (IBD).
 Further, these diseases exhibit different symptoms depending on the location of the colon they are found in.
@@
 We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure.
-This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
+In contrast to previous efforts using prepared colonoscopy, this technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals.
 Diversity analysis revealed that each site contained a diverse community and that a patient’s samples were more similar to each other than to that of other individuals.
-Comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the distal lumen, likely reflecting the anatomical structure of the colon.
-Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key species that distinguish biogeographical sites.
+A patient's feces were most similar to samples taken from the distal lumen, likely reflecting the anatomical structure of the colon.
+Since we could not differentiate the communities in the proximal and distal colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key taxa that the two sites in the lumen and mucosa.
 Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location.
@@
 Further investigation of these bacteria may elucidate if and how these groups contribute to different pathogenesis processes on the respective sides of the colon.
+
+#### wordcount?
 
@@
 Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC.
-Because CRC is a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.
+Because CRC can be a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.
 
-Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host @Yamauchi2012, @Halfvarson2017.
+Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host @Yamauchi2012 @Halfvarson2017.
 For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated @Dejea2014.
@@
 Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community @Forbes2016.
-The molecular capabilities of these rare taxa may contribute to the colonic disease state.
+The physiology of these rare taxa may contribute to the colonic disease state.
 These studies often examined only shed human stool or the small intestine, preventing fine-resolution analysis of paired samples from the proximal and distal sides of the colon.
@@
 The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods.
-The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures.
+The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy procedures.
 While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.

### 2017-08-18 submission/manuscript.Rmd
pat edits made

@@
 The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
-IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012, @Forbes2016, @Halfvarson2017.
+IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012 @Forbes2016 @Halfvarson2017.
 The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
@@
 In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010.
-The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012, @Albenberg2014, @Donaldson2015.
+The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012 @Albenberg2014 @Donaldson2015.
 In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016.
@@
 While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
-Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014, @Harrell2012.
+Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014 @Harrell2012.
 As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
 
-Here we aimed to address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon.
-Our design used an unprepared colonoscopy technique to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
+Here we address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon.
+We used an unprepared colonoscopy technique to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
 To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
@@
 To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
-Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research @LloydPrice2016.
-Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
+As expected, each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A) @LloydPrice2016.
+Samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
 For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others.
@@
 
-To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated ${\theta}$YC distances from OTU abundances and compared these distances for all individuals.
-Again, across all patient samples we observed a range of ${\theta}$YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
+To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated distances from Operational Taxonomic Unit (OTU) abundances and compared these distances for all individuals.
+Again, across all patient samples we observed a range of distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
 However, when comparing median distances between the proximal lumen and mucosa, the proximal versus distal lumen, the proximal versus distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
 
-#### Stool at exit most resembles lumenal samples from the distal colon 
-Next, we calculated ${\theta}$YC distances to examine how each sample compared to the home-collected exit stool.
-Amidst variability between patients, we did identify significantly smaller ${\theta}$YC distance between the distal lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
-Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
-To determine what factors may be driving the differences seen among the samples, we compared ${\theta}$YC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon @Eckburg2005, @deCarcer2010, @Zhang2013.
-Thus interpersonal variation between subjects drives the differences between samples more than sample site or location.
-Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.
+#### Fecal samples resemble lumenal samples from the distal colon 
+Next, we calculated distances to examine how each sample compared to the home-collected feces.
+Amidst variability between patients, we did identify significantly smaller distance between the distal lumenal sample and the feces (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
+Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the feces, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
+To determine what factors may be driving the differences seen among the samples, we compared distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
+We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon @Eckburg2005 @deCarcer2010 @Zhang2013.
+Thus interpersonal variation between subjects drove the differences between samples more than sample site or location.
+Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of the patient's feces, and the microbes remaining on the mucosa are more distinct.
 
-#### Random Forest classification models identify important OTUs on each side
+#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
 To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances.
-We built the first model to classify the lumen versus mucosal samples for the proximal and distal sides, independently (Figure 4A).
+We built the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure 4A).
 The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal.
 The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively).
-The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model.
+The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
 For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B).
-The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C).
+The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4C).
 These results indicate that there are fine differences between the different sites of the colon, and that these can be traced to specific OTUs on each side.

### 2017-08-21 submission/manuscript.Rmd
more edits on v2

@@
 To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances.
-We built the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure 4A).
+We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure 4A).
 The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal.
@@
 The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4C).
-These results indicate that there are fine differences between the different sites of the colon, and that these can be traced to specific OTUs on each side.
+These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.
 
-Next, we built a model to differentiate the proximal and distal lumenal samples.
+Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation.
 The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) compared to the proximal versus distal lumen (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`).
@@
 @Castellarin2011.
-There were 14 samples with the _F.
+There were 14 patient samples with the _F.
 nucleatum subsp.
-animalis_ sequence.
-Four sequences were located in the proximal mucosa and three in the distal mucosal samples (Figure 6A).
-The second most frequent _Fusobacterium_ sequence was OTU472, which aligns with 99% identity to _F.
+animalis_ sequences.
+Of the samples with the highest abundance of _F.
+nucleatum subsp.
+animalis_, four of the samples were from the proximal mucosa and three from the distal mucosa (Figure 6A).
+The second most frequent _Fusobacterium_ sequence was OTU472, which aligned with 99% identity to _F.
 varium_.
@@
 varium_ has been associated with IBD @Lee2016.
-4 study participants harbored _F.
+Four study participants harbored _F.
 varium_ and the samples were split evenly between the proximal and distal mucosa (Figure 6B).
@@
 OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6C).
-Among the samples that were positive for _P.
+Among the 11 distal mucosa samples that were positive for _P.
 asacharolytica_, the relative abundances for this OTU ranged from 0.01% - 16%.
@@
 Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other.
-Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the distal lumen is most similar to stool at exit.
-Random Forest models built on OTU relative abundances from each sample identified microbes that are particular to each location of the colon.
+Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the distal lumen was most similar to feces.
+Random Forest models built on OTU relative abundances from each sample identified microbes that were particular to each location of the colon.
 Finally, we were able to detect some bacterial OTUs associated with colonic disease in our healthy patient cohort.
@@
 
-When examining the relative abundance of the different phyla at each site, there was a wide amount of variation with communities primarily dominated by the _Bacteriodes_ and _Firmicutes_.
-This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of biogeogrphical "patchiness" in the gut microbiome.
+When examining the relative abundance of the dominant phyla at each site (i.e.
+_Bacteriodes_ and _Firmicutes_), there was a wide amount of variation.
+This likely reflects not only the variability between human subjects, casued by differences in age, sex, diet, but also biogeogrphical "patchiness" in the gut microbiome.
 Several studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011.

### 2017-08-22 submission/manuscript.Rmd
most of Pat edits made

@@
 That said, across our samples the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to lumenal content in humans @Albenberg2014.
-Hence, the conclusions we can draw from phyla analysis are likely impacted by patchiness between subjects.
+Hence, the conclusions we can draw from phyla analysis may be impacted by patchiness between subjects.
 
@@
 Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014.
-Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model @Whitmore2014, @Flynn2016.
+Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model @Whitmore2014 @Flynn2016.
 Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.
@@
 Laboratory experiments with these isolates have shown that disease-isolated _F.
-nucleatum_ are more invasive and stimulate more TNF-a production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011.
+nucleatum_ are more invasive and stimulate more TNF-\alpha production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011.
 _F.
@@
 varium_ was only detected in three of our study participants and two of those samples were isolated form the proximal mucosa (Figure 6B).
-Although there is low occurrance in our study, _F.
+_F.
 varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum @Ohkusa2002, suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites.
@@
 
-Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies.
+Specific comparisons of our findings to previously published gut biogeography studies are confounded by the use of bowel preparation methods in most other studies.
 A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing @Shobar2016.
-This group found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016.
-Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites.
-Further, the differences were not great enough to overcome interpersonal differences between subjects.
-Bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel.
-Thus our findings in this study are strengthened by the lack of bowel preparation.
+This study found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016.
+After seven days, bowel cleansing not only made the samples harder to distinguish, but it also decreased in diversity across sites.
+Bowel preparation clearly biases into the microbes recovered from sampling the lumen or mucosa of a prepared bowel.
 
@@
 At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit?).
-At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and ship the samples to the University on ice.
+At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and ship the samples to a research coordinator on ice.
 Notably, subjects did not undergo any bowel preparation method prior to sampling.
@@
 Samples were then collected as in the distal colon and the colonoscope withdrawn.
-All samples were stored at -80 C until study completion.
+All samples were stored at -80$^\circ$C.
 
@@
 The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
-Sequencing and analysis of a mock community revealed the error rate to be 1.8%.
+Sequencing and analysis of a mock community revealed the error rate to be 0.018%.
 Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.
 
-Diversity analysis was performed using the Simpson diversity calculator and thetaYC calculator metrics in mothur version 3.2 @Schloss2009.
+Diversity analysis was performed using the Simpson diversity calculator and $\theta$<sub>YC</sub> calculator metrics in mothur version 1.39.5 @Schloss2009.
 ThetaYC distances were calculated to determine the dissimilarity between two samples.
 Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location @Liaw2002.
-To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.
+To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.
 
@@
 Differences in thetaYC distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
+
 #### Data availability

### 2017-08-23 submission/manuscript.Rmd
starting working on feature reduction

@@
 Laboratory experiments with these isolates have shown that disease-isolated _F.
-nucleatum_ are more invasive and stimulate more TNF-\alpha production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011.
+nucleatum_ are more invasive and stimulate more TNF-$\alpha$ production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011.
 _F.
@@
 Differences in community membership at the phyla level were tested using the analysis of molecular variance (AMOVA) metric in mothur.
-Differences in thetaYC distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
+Differences in $\theta$<sub>YC</sub> distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
 

### 2017-08-23 submission/manuscript.tex
starting working on feature reduction

@@
 \hypersetup{unicode=true,
-            pdftitle={Unprepared colonoscopy identifies microbiota specific to the proximal and distal human colon},
+            pdftitle={The microbiota of the proximal and distal human colon},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Unprepared colonoscopy identifies microbiota specific to the proximal
-and distal human colon}
+  \title{The microbiota of the proximal and distal human colon}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-Running title: Unprepared colonoscopy identifies proximal and distal
-human colonic microbiota
+Running title: The microbiome of the proximal and distal human colon
 
@@
 Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2},
-Danielle Kimberly Turgeon\textsuperscript{3}, and Patrick D.
+Danielle Kimberly Turgeon\textsuperscript{3\(\dagger\)}, and Patrick D.
 Schloss\textsuperscript{1\(\dagger\)}
@@
 
-\(\dagger\) Corresponding author:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+\(\dagger\) Corresponding authors:
+\href{mailto:kturgeon@med.umich.edu}{\nolinkurl{kturgeon@med.umich.edu}}
+and \href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
 
@@
 
-The human colon contains chemicals and nutrients that change along the
-proximal to distal gut axis. These gradients create microenvironments
-that affect the distrubution and composition of the gut microbiota. The
-microbiome has been implicated in the colonic diseases colorectal cancer
+Chemical and nutrient gradients along the human colon create
+microenvironments that affect the distrubution and composition of the
+gut microbiota. The microbiome has been implicated in colorectal cancer
 (CRC) and inflammatory bowel disease (IBD). Further, these diseases
@@
 and distal healthy human colon. We used a two-colonoscope approach on
-subjects that had not undergone standard bowel preparation procedure.
-This technique allowed us to characterize the native proximal and distal
-luminal and mucosal microbiome without prior chemical disruption. 16S
-rRNA gene sequencing was performed on proximal and distal mucosal and
-luminal biopsies and home-collected stool for 20 healthy individuals.
-Diversity analysis revealed that each site contained a diverse community
-and that a patient's samples were more similar to each other than to
-that of other individuals. Comparison of all samples to fecal samples
-taken at exit uncovered that the feces were most similar to samples
-taken from the distal lumen, likely reflecting the anatomical structure
-of the colon. Since we could not differentiate sites along the colon
-based on community structure or community membership alone, we employed
-the Random Forest machine-learning algorithm to identify key species
-that distinguish biogeographical sites. Random Forest classification
-models were built using taxa abundance and sample location and revealed
-distinct populations that were found in each location. A model
-differentiating the proximal mucosa and lumen was built with an AUC of
-0.831. The proximal mucosa had a higher abundance of the genera
-\emph{Enterobacteriaceae} and \emph{Bacteriodes}. \emph{Peptoniphilus,
-Anaerococcus, Finegoldia,} and \emph{Turicibacter} were most likely to
-be found in distal mucosal samples versus distal luminal samples (AUC =
-0.98). The classification model performed well (AUC = 0.912) when
-classifying mucosal samples into proximal or distal sides, but
-separating luminal samples from each side proved more challenging (AUC =
-0.755). The distal mucosa was found to have high populations of
-\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
-distal luminal samples were comprised of many of the same taxa, likely
-reflecting the fact that stool moves along the colon from the proximal
-to distal end. By sampling the unprepped human colon, our results have
-identified distinct bacterial populations native to the proximal and
-distal sides. Further investigation of these bacteria may elucidate if
-and how these groups contribute to different pathogenesis processes on
-the respective sides of the colon.
+subjects that had not undergone standard bowel preparation procedure. In
+contrast to previous efforts using prepared colonoscopy, this technique
+allowed us to characterize the native proximal and distal luminal and
+mucosal microbiome without prior chemical disruption. 16S rRNA gene
+sequencing was performed on proximal and distal mucosal and luminal
+biopsies and home-collected stool for 20 healthy individuals. Diversity
+analysis revealed that each site contained a diverse community and that
+a patient's samples were more similar to each other than to that of
+other individuals. A patient's feces were most similar to samples taken
+from the distal lumen, likely reflecting the anatomical structure of the
+colon. Since we could not differentiate the communities in the proximal
+and distal colon based on community structure or community membership
+alone, we employed the Random Forest machine-learning algorithm to
+identify key taxa that the two sites in the lumen and mucosa. Random
+Forest classification models were built using taxa abundance and sample
+location and revealed distinct populations that were found in each
+location. A model differentiating the proximal mucosa and lumen was
+built with an AUC of 0.831. The proximal mucosa had a higher abundance
+of the genera \emph{Enterobacteriaceae} and \emph{Bacteriodes}.
+\emph{Peptoniphilus, Anaerococcus, Finegoldia,} and \emph{Turicibacter}
+were most likely to be found in distal mucosal samples versus distal
+luminal samples (AUC = 0.98). The classification model performed well
+(AUC = 0.912) when classifying mucosal samples into proximal or distal
+sides, but separating luminal samples from each side proved more
+challenging (AUC = 0.755). The distal mucosa was found to have high
+populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
+Proximal and distal luminal samples were comprised of many of the same
+taxa, likely reflecting the fact that stool moves along the colon from
+the proximal to distal end. By sampling the unprepped human colon, our
+results have identified distinct bacterial populations native to the
+proximal and distal sides. Further investigation of these bacteria may
+elucidate if and how these groups contribute to different pathogenesis
+processes on the respective sides of the colon.
+
+\paragraph{wordcount?}\label{wordcount}
 
@@
 Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
-accelerated by perturbations in gut microbes (1), (2), (3). The
-severity, symptoms, morbidity and mortality of these diseases is known
-to vary based upon the biogeographical location in which they occur. For
+accelerated by perturbations in gut microbes (1) (2) (3). The severity,
+symptoms, morbidity and mortality of these diseases is known to vary
+based upon the biogeographical location in which they occur. For
 instance, CRC tumors that arise on the distal side of the colon are
@@
 genomic instability of precancerous cells, respectively, in addition to
-variation in the previously mentioned chemical gradients (1), (5), (6).
-In IBD patients, disease flares in the distal colon are usually
-indicative of ulcerative colitis (UC), whereas Crohn's disease (CD)
-patients typically experience disease in the small intestine, ileum and
-proximal colon (2). UC presents as large and highly inflammed mucosal
-ulcers, where as CD lesions are often smaller and have areas of normal
-tissue distributed amongst the flare (2). Thus, given the varied
-physiology of the proximal-distal axis of the colon and known
-differences in disease patterns at these sites, symbiotic microbes and
-their metabolites likely vary as well, and may influence the
-heterogenous disease prognoses of IBD and CRC. Because CRC is a
-long-term complication of IBD, the distribution of microbes is important
-to understanding the pathophysiology of both diseases.
+variation in the previously mentioned chemical gradients (1) (5) (6). In
+IBD patients, disease flares in the distal colon are usually indicative
+of ulcerative colitis (UC), whereas Crohn's disease (CD) patients
+typically experience disease in the small intestine, ileum and proximal
+colon (2). UC presents as large and highly inflammed mucosal ulcers,
+where as CD lesions are often smaller and have areas of normal tissue
+distributed amongst the flare (2). Thus, given the varied physiology of
+the proximal-distal axis of the colon and known differences in disease
+patterns at these sites, symbiotic microbes and their metabolites likely
+vary as well, and may influence the heterogenous disease prognoses of
+IBD and CRC. Because CRC can be a long-term complication of IBD, the
+distribution of microbes is important to understanding the
+pathophysiology of both diseases.
 
@@
 IBD or CRC can be attributed to specific molecular events as a result of
-interactions between the gut microbiota and human host (1), (3). For
+interactions between the gut microbiota and human host (1) (3). For
 instance, comparison of the bacteria present on CRC tumors with those
@@
 intestinal pathogens, the bacteria appear to have a high-impact despite
-being lowly-abundant in the community (2). The molecular capabilities of
-these rare taxa may contribute to the colonic disease state. These
-studies often examined only shed human stool or the small intestine,
-preventing fine-resolution analysis of paired samples from the proximal
-and distal sides of the colon. Similarly, comparisons of on- or
-off-tumor/lesion bacteria rarely have matched tissue from the other side
-of the colon from the same patient, limiting what conclusions can be
-drawn about the colonic microbiome overall, let alone at that specific
-site. Due to these limitations, the contribution of the gut microbiota
-to IBD and CRC disease location in the colon is largely undefined.
-Characterizing these communities could provide needed insight into
-disease etiology, including how the disruption of the healthy community
-could promote the initiation or proliferation of the distinct proximal
-and distal CRC tumors or IBD flares.
+being lowly-abundant in the community (2). The physiology of these rare
+taxa may contribute to the colonic disease state. These studies often
+examined only shed human stool or the small intestine, preventing
+fine-resolution analysis of paired samples from the proximal and distal
+sides of the colon. Similarly, comparisons of on- or off-tumor/lesion
+bacteria rarely have matched tissue from the other side of the colon
+from the same patient, limiting what conclusions can be drawn about the
+colonic microbiome overall, let alone at that specific site. Due to
+these limitations, the contribution of the gut microbiota to IBD and CRC
+disease location in the colon is largely undefined. Characterizing these
+communities could provide needed insight into disease etiology,
+including how the disruption of the healthy community could promote the
+initiation or proliferation of the distinct proximal and distal CRC
+tumors or IBD flares.
 
@@
 been limited by sample collection methods. The majority of human gut
-microbiome studies have been performed on whole shed stool or on samples
+microbiome studies have been performed on whole shed feces or on samples
 collected during colonoscopy procedures. While the latter method allows
@@
 colon, but complicates microbiome profiling as the chemicals strip the
-bowel of contents and disrupt the mucosal layer (10), (11). As such,
-what little information we do have about the biogeographical
-distribution of the microbes in the proximal and distal colon is
-confounded by the bowel preparation procedure.
+bowel of contents and disrupt the mucosal layer (10) (11). As such, what
+little information we do have about the biogeographical distribution of
+the microbes in the proximal and distal colon is confounded by the bowel
+preparation procedure.
 
-Here we aimed to address the limitations of previous studies and
-effectively identify the microbes specific to the lumen and mucosa of
-the proximal and distal healthy human colon. Our design used an
-unprepared colonoscopy technique to sample the natural community of each
-location of the gut without prior disruption of the native bacteria in
-20 healthy volunteers. To address the inherent inter-individual
-variation in human microbiomes, we used a machine-learning
-classification algorithm trained on curated 16S rRNA sequencing reads to
-identify microbes specific to each location. We found that our
-classification models were able to separate mucosal and lumenal samples
-as well as differentiate between sides of the colon based on populations
-of specific microbes. By identifying the specific microbes we are poised
-to ask if and how the presence or disruption of the microbes at each
-site contribute to the development of the specific tumor subtypes of CRC
-in the proximal and distal human colon.
+Here we address the limitations of previous studies and effectively
+identify the microbes specific to the lumen and mucosa of the proximal
+and distal healthy human colon. We used an unprepared colonoscopy
+technique to sample the natural community of each location of the gut
+without prior disruption of the native bacteria in 20 healthy
+volunteers. To address the inherent inter-individual variation in human
+microbiomes, we used a machine-learning classification algorithm trained
+on curated 16S rRNA sequencing reads to identify microbes specific to
+each location. We found that our classification models were able to
+separate mucosal and lumenal samples as well as differentiate between
+sides of the colon based on populations of specific microbes. By
+identifying the specific microbes we are poised to ask if and how the
+presence or disruption of the microbes at each site contribute to the
+development of the specific tumor subtypes of CRC in the proximal and
+distal human colon.
 
@@
 these sites, 16S rRNA gene sequencing was performed on extracted DNA
-from each sample. Each site was primarily dominated by \emph{Firmicutes}
-and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
-in human microbiome research (12). Likewise, samples had varying levels
-of diversity at each site, irrespective of the individual (Figure 2B).
-For example, the proximal mucosa was more diverse than the distal for
-some individuals while the opposite was true for others. Therefore we
-could not identify a clear pattern of changes in microbial diversity
-along the gut axis.
+from each sample. As expected, each site was primarily dominated by
+\emph{Firmicutes} and \emph{Bacteriodetes} (Figure 2A) (12). Samples had
+varying levels of diversity at each site, irrespective of the individual
+(Figure 2B). For example, the proximal mucosa was more diverse than the
+distal for some individuals while the opposite was true for others.
+Therefore we could not identify a clear pattern of changes in microbial
+diversity along the gut axis.
 
 To compare similarity between sides (proximal or distal) or sites (lumen
-or mucosa), we calculated \({\theta}\)YC distances from OTU abundances
-and compared these distances for all individuals. Again, across all
-patient samples we observed a range of \({\theta}\)YC distances when
+or mucosa), we calculated distances from Operational Taxonomic Unit
+(OTU) abundances and compared these distances for all individuals.
+Again, across all patient samples we observed a range of distances when
 comparing sample locations (Figure 3A) and again those ranges did not
@@
 
-\paragraph{Stool at exit most resembles lumenal samples from the distal
-colon}\label{stool-at-exit-most-resembles-lumenal-samples-from-the-distal-colon}
+\paragraph{Fecal samples resemble lumenal samples from the distal
+colon}\label{fecal-samples-resemble-lumenal-samples-from-the-distal-colon}
 
-Next, we calculated \({\theta}\)YC distances to examine how each sample
-compared to the home-collected exit stool. Amidst variability between
-patients, we did identify significantly smaller \({\theta}\)YC distance
-between the distal lumenal sample and the exit stool (Figure 3B, P
-\textless{} 0.05, Wilcoxon, BH adjustment). Furthermore, there was an
-even larger difference in the comparisons of the distal mucosa to the
-exit stool, indicating that the mucosa is different from the stool as
-compared to lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To
-determine what factors may be driving the differences seen among the
-samples, we compared \({\theta}\)YC distances between samples from all
-subjects (interpersonal) versus samples from within one subject
-(intrapersonal). We found that samples from one individual were far more
-similar to each other than to other study subjects (Figure 3C),
-consistent with previous human microbiome studies that have sampled
-multiple sites of the human colon (13), (14), (15). Thus interpersonal
-variation between subjects drives the differences between samples more
-than sample site or location. Overall, the results comparing the
-structure of the communities suggest that the contents of the distal
-lumen are most representative of stool at exit, and the microbes
-remaining on the mucosa are much different.
+Next, we calculated distances to examine how each sample compared to the
+home-collected feces. Amidst variability between patients, we did
+identify significantly smaller distance between the distal lumenal
+sample and the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
+adjustment). Furthermore, there was an even larger difference in the
+comparisons of the distal mucosa to the feces, indicating that the
+mucosa is different from the stool as compared to lumen (P \textless{}
+0.0005, Wilcoxon, BH adjustment). To determine what factors may be
+driving the differences seen among the samples, we compared distances
+between samples from all subjects (interpersonal) versus samples from
+within one subject (intrapersonal). We found that samples from one
+individual were far more similar to each other than to other study
+subjects (Figure 3C), consistent with previous human microbiome studies
+that have sampled multiple sites of the human colon (13) (14) (15). Thus
+interpersonal variation between subjects drove the differences between
+samples more than sample site or location. Overall, the results
+comparing the structure of the communities suggest that the contents of
+the distal lumen are most representative of the patient's feces, and the
+microbes remaining on the mucosa are more distinct.
 
-\paragraph{Random Forest classification models identify important OTUs
-on each
-side}\label{random-forest-classification-models-identify-important-otus-on-each-side}
+\paragraph{Random Forest classification models identify important
+Operational Taxonomic Units (OTUs) on each
+side}\label{random-forest-classification-models-identify-important-operational-taxonomic-units-otus-on-each-side}
 
@@
 constructed several Random Forest models trained using OTU abundances.
-We built the first model to classify the lumen versus mucosal samples
-for the proximal and distal sides, independently (Figure 4A). The
-constructed model used 14 features for the proximal and 6 for the
-distal. The models performed well when classifying these samples (0.831
-and 0.98, respectively). The OTUs that were most predictive of each site
-are identified by their greatest mean decrease in accuracy when removed
-from the model. For distinguishing the proximal lumen and mucosa, OTUs
-from the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
-two OTUs from the \emph{Enterobacteraceae} genera were differentially
-abundant (Figure 4B). The model classifying the distal lumen and mucosa
-identified OTUs from \emph{Turicibacter}, \emph{Finegoldia},
-\emph{Peptoniphilus} and two OTUs from the \emph{Anaerococcus} genera
-that could distinguish lumen from mucosal samples (Figure 4C). These
-results indicate that there are fine differences between the different
-sites of the colon, and that these can be traced to specific OTUs on
-each side.
+We used 10-fold cross validation to build the first model to classify
+the lumenal versus mucosal samples for the proximal and distal sides,
+independently (Figure 4A). The constructed model used 14 features for
+the proximal and 6 for the distal. The models performed well when
+classifying these samples (0.831 and 0.98, respectively). The OTUs that
+were most predictive of each site were identified by their greatest mean
+decrease in accuracy when removed from the model. For distinguishing the
+proximal lumen and mucosa, OTUs from the \emph{Bacteriodes},
+\emph{Actinomyces}, \emph{Psuedomonas} and two OTUs from the
+\emph{Enterobacteraceae} genera were differentially abundant (Figure
+4B). The model classifying the distal lumen and mucosa identified OTUs
+from \emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
+two OTUs from the \emph{Anaerococcus} genera (Figure 4C). These results
+indicated that there were fine differences between the different sites
+of the colon, and that these could be traced to specific OTUs on each
+side.
 
 Next, we built a model to differentiate the proximal and distal lumenal
-samples. The model performed best when distinguishing the proximal
-versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
-versus distal lumen (AUC = 0.755). The model that separated proximal and
-distal mucosal samples used 6 features. OTUs that were differentially
-abundant between the distal and proximal mucosa included members of the
-\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
-\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B). The
-model used 44 features to separate proximal and distal lumenal samples.
-Differentially abundant OTUs of the proximal and distal lumen included
-three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
-and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
-some of the same OTUs that are distinct between the mucosa and lumen
-also helped to differentiate between the two sides- such as
-\emph{Anaerococcus} and \emph{Finegoldia}.
+samples using 10-fold cross validation. The model performed best when
+distinguishing the proximal versus distal mucosa (Figure 5A, AUC =
+0.912) compared to the proximal versus distal lumen (AUC = 0.755). The
+model that separated proximal and distal mucosal samples used 6
+features. OTUs that were differentially abundant between the distal and
+proximal mucosa included members of the \emph{Porphyromonas},
+\emph{Murdochiella}, \emph{Finegoldia}, \emph{Anaerococcus} and
+\emph{Peptoniphilus} genera (Figure 5B). The model used 44 features to
+separate proximal and distal lumenal samples. Differentially abundant
+OTUs of the proximal and distal lumen included three OTUs of the
+\emph{Bacteroides} genus, a \emph{Clostridium IV} OTU and an
+\emph{Oscillibacter} OTU (Figure 5C). This analysis found that some of
+the same OTUs that are distinct between the mucosa and lumen also helped
+to differentiate between the two sides- such as \emph{Anaerococcus} and
+\emph{Finegoldia}.
 
@@
 properties and be found on the surfaces of colorectal cancer tumors.
-(16). There were 14 samples with the \emph{F. nucleatum subsp. animalis}
-sequence. Four sequences were located in the proximal mucosa and three
-in the distal mucosal samples (Figure 6A). The second most frequent
-\emph{Fusobacterium} sequence was OTU472, which aligns with 99\%
-identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
-\emph{F. varium} has been associated with IBD (17). 4 study participants
-harbored \emph{F. varium} and the samples were split evenly between the
-proximal and distal mucosa (Figure 6B). OTU152 was similar to the
-members of the \emph{Porphyromonas} genus and the most frequent sequence
-in that OTU aligned to \emph{Porphyromonas asacharolytica} (99\% over
-full length), another bacterium commonly detected and isolated from
-colorectal tumors. OTU152 was only detected on the distal mucosa, and in
-fact was one of the OTUs the classification model identified as
-separating distal and proximal sides (Figure 6C). Among the samples that
-were positive for \emph{P. asacharolytica}, the relative abundances for
-this OTU ranged from 0.01\% - 16\%. Thus, disease-associated OTUs could
-be found in our sample set of 20 healthy individuals.
+(16). There were 14 patient samples with the \emph{F. nucleatum subsp.
+animalis} sequences. Of the samples with the highest abundance of
+\emph{F. nucleatum subsp. animalis}, four of the samples were from the
+proximal mucosa and three from the distal mucosa (Figure 6A). The second
+most frequent \emph{Fusobacterium} sequence was OTU472, which aligned
+with 99\% identity to \emph{F. varium}. In addition to \emph{F.
+nucleatum}, \emph{F. varium} has been associated with IBD (17). Four
+study participants harbored \emph{F. varium} and the samples were split
+evenly between the proximal and distal mucosa (Figure 6B). OTU152 was
+similar to the members of the \emph{Porphyromonas} genus and the most
+frequent sequence in that OTU aligned to \emph{Porphyromonas
+asacharolytica} (99\% over full length), another bacterium commonly
+detected and isolated from colorectal tumors. OTU152 was only detected
+on the distal mucosa, and in fact was one of the OTUs the classification
+model identified as separating distal and proximal sides (Figure 6C).
+Among the 11 distal mucosa samples that were positive for \emph{P.
+asacharolytica}, the relative abundances for this OTU ranged from 0.01\%
+- 16\%. Thus, disease-associated OTUs could be found in our sample set
+of 20 healthy individuals.
 
@@
 samples with samples collected from stool at home showed that the distal
-lumen is most similar to stool at exit. Random Forest models built on
-OTU relative abundances from each sample identified microbes that are
+lumen was most similar to feces. Random Forest models built on OTU
+relative abundances from each sample identified microbes that were
 particular to each location of the colon. Finally, we were able to
@@
 
-When examining the relative abundance of the different phyla at each
-site, there was a wide amount of variation with communities primarily
-dominated by the \emph{Bacteriodes} and \emph{Firmicutes}. This likely
-reflects not only the variability between human subjects, casued by
-differences in age, gender, diet, but also reports of biogeogrphical
-``patchiness'' in the gut microbiome. Several studies have noted that
-the bacteria recoverable from the same mucosal sample location can be
-vastly different when the samples are taken just 1 cm away from each
-other (18). Similar patchiness is also observed in lumenal contents and
-fecal samples themselves; there is observed separation of different
-interacting microbes along the length of a stool sample, for instance
-(19). That said, across our samples the mucosal samples harbor more
-\emph{Proteobacteria}, consistent with previous studies comparing
-mucosal swabs to lumenal content in humans (5). Hence, the conclusions
-we can draw from phyla analysis are likely impacted by patchiness
-between subjects.
+When examining the relative abundance of the dominant phyla at each site
+(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
+of variation. This likely reflects not only the variability between
+human subjects, casued by differences in age, sex, diet, but also
+biogeogrphical ``patchiness'' in the gut microbiome. Several studies
+have noted that the bacteria recoverable from the same mucosal sample
+location can be vastly different when the samples are taken just 1 cm
+away from each other (18). Similar patchiness is also observed in
+lumenal contents and fecal samples themselves; there is observed
+separation of different interacting microbes along the length of a stool
+sample, for instance (19). That said, across our samples the mucosal
+samples harbor more \emph{Proteobacteria}, consistent with previous
+studies comparing mucosal swabs to lumenal content in humans (5). Hence,
+the conclusions we can draw from phyla analysis may be impacted by
+patchiness between subjects.
 
@@
 not only co-occur on CRC tumors but also to synergistically promote
-tumorigenesis in an oral cancer model (22), (23). Thus, further analysis
+tumorigenesis in an oral cancer model (22) (23). Thus, further analysis
 of the distribution and activities of these pathogens may elucidate a
@@
 disease-isolated \emph{F. nucleatum} are more invasive and stimulate
-more TNF-a production than strains from healthy individuals (9),
-suggesting the bacteria may increase inflammation in the gut as well
-(24). \emph{F. varium} isolated from UC patients caused colonic ulcers
-in an experimental mouse model (25). \emph{F. varium} was only detected
-in three of our study participants and two of those samples were
-isolated form the proximal mucosa (Figure 6B). Although there is low
-occurrance in our study, \emph{F. varium} is most commonly isolated from
-UC patient biopsies from the ileum or cecum (26), suggesting this
-species may exhibit preference for the different environmental
-conditions of these gastrointestinal sites. Further work will assess how
-gut environment may select for species which may then cause localized
-disease.
+more TNF-\(\alpha\) production than strains from healthy individuals
+(9), suggesting the bacteria may increase inflammation in the gut as
+well (24). \emph{F. varium} isolated from UC patients caused colonic
+ulcers in an experimental mouse model (25). \emph{F. varium} was only
+detected in three of our study participants and two of those samples
+were isolated form the proximal mucosa (Figure 6B). \emph{F. varium} is
+most commonly isolated from UC patient biopsies from the ileum or cecum
+(26), suggesting this species may exhibit preference for the different
+environmental conditions of these gastrointestinal sites. Further work
+will assess how gut environment may select for species which may then
+cause localized disease.
 
 Specific comparisons of our findings to previously published gut
-biogeography studies are additionally confounded by the use of bowel
-preparation methods in most other studies. A rare report of a
-matched-colonoscopy study sampled 18 patient's colonic mucosa and
-lumenal contents prior to and after bowel cleansing (27). This group
-found that mucosal and lumenal samples were distinguishable prior to
-bowel cleansing, but that bowel preparation resulted in an increase in
-shared OTUs between each site (27). Bowel cleansing not only made the
-samples harder to distinguish, it resulted in decreases in diversity
-across sites. Further, the differences were not great enough to overcome
-interpersonal differences between subjects. Bowel preparation clearly
-induces bias into the microbes recovered from sampling the lumen or
-mucosa of a prepared bowel. Thus our findings in this study are
-strengthened by the lack of bowel preparation.
+biogeography studies are confounded by the use of bowel preparation
+methods in most other studies. A rare report of a matched-colonoscopy
+study sampled 18 patient's colonic mucosa and lumenal contents prior to
+and after bowel cleansing (27). This study found that mucosal and
+lumenal samples were distinguishable prior to bowel cleansing, but that
+bowel preparation resulted in an increase in shared OTUs between each
+site (27). After seven days, bowel cleansing not only made the samples
+harder to distinguish, but it also decreased in diversity across sites.
+Bowel preparation clearly biases into the microbes recovered from
+sampling the lumen or mucosa of a prepared bowel.
 
@@
 colonoscopy, subjects were to collect whole stool at home and ship the
-samples to the University on ice. Notably, subjects did not undergo any
-bowel preparation method prior to sampling. On the procedure day,
-subjects reported to the Michigan Clinical Research Unit at the
+samples to a research coordinator on ice. Notably, subjects did not
+undergo any bowel preparation method prior to sampling. On the procedure
+day, subjects reported to the Michigan Clinical Research Unit at the
 University of Michigan Health System. Patients were consciously sedated
@@
 Samples were then collected as in the distal colon and the colonoscope
-withdrawn. All samples were stored at -80 C until study completion.
+withdrawn. All samples were stored at -80\(^\circ\)C.
 
@@
 based on a 97\% similarity cutoff. Sequencing and analysis of a mock
-community revealed the error rate to be 1.8\%. Samples were rarefied to
-4231 sequences per sample in order to reduce uneven sampling bias.
+community revealed the error rate to be 0.018\%. Samples were rarefied
+to 4231 sequences per sample in order to reduce uneven sampling bias.
 
 Diversity analysis was performed using the Simpson diversity calculator
-and thetaYC calculator metrics in mothur version 3.2 (29). ThetaYC
-distances were calculated to determine the dissimilarity between two
-samples. Random Forest classification models were built using the
+and \(\theta\)YC calculator metrics in mothur version 1.39.5 (29).
+ThetaYC distances were calculated to determine the dissimilarity between
+two samples. Random Forest classification models were built using the
 randomForest R package and resultant models were used to identify the
@@
 aligned using blastn and the species name was only used if the identity
-score was \textgreater{}= 99\%.
+score was \(\ge\) 99\% over the full-length of the contig and matched a
+single reference.
 
@@
 the analysis of molecular variance (AMOVA) metric in mothur. Differences
-in thetaYC distances by location were tested using the Wilcoxon rank-sum
-test adjusted for multiple comparisons using the Benjamini-Hochberg
-procedure. \#\#\#\# Data availability 16S rRNA gene sequence reads and
-experiment metadata are available on the NCBI Sequence Read Archive
-(SRA) with accession number XXXX. A reproducible data analysis pipeline
-can be found at
+in \(\theta\)YC distances by location were tested using the Wilcoxon
+rank-sum test adjusted for multiple comparisons using the
+Benjamini-Hochberg procedure.
+
+\paragraph{Data availability}\label{data-availability}
+
+16S rRNA gene sequence reads and experiment metadata are available on
+the NCBI Sequence Read Archive (SRA) with accession number XXXX. A
+reproducible data analysis pipeline can be found at
 \url{https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017}.

### 2017-08-23 code/random_forest.R

@@ -15,13 +15,19 @@ for (dep in pack_used){
 meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
 shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
 tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+filter_shared <- shared_file <- read.table(file='data/mothur/kws_final.an.0.03.filter.shared', sep = '\t', header=T, row.names=2)
 
 #make OTU abundance file
 #Create df with relative abundances
 shared_file <- subset(shared_file, select = -c(numOtus, label))
 shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
 
+filter_shared <- subset(shared_file, select = -c(numOtus, label))
+filter_meta <- merge(meta_file, filter_shared, by.x='group', by.y='row.names')
+
 rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+
+filter_relabund <- 100*filter_shared/unique(apply(filter_shared, 1, sum))
 
 #Create vector of OTUs with median abundances >1%
 OTUs_1 <- apply(rel_abund, 2, max) > 1
@@ -29,6 +35,12 @@ OTU_list <- colnames(rel_abund)[OTUs_1]
 #get df of just top OTUs
 rel_abund_top <- rel_abund[, OTUs_1]
 rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+OTUs_filter <- apply(filter_relabund, 2, max) >1
+OTU_filter_list <- colnames(filter_relabund)[OTUs_filter]
+
+filter_top <- filter_relabund[, OTUs_filter]
+filter_abund_meta <- merge(meta_file, filter_top, by.x='group', by.y='row.names')
 
 seed <- 1
 n_trees <- 2001
@@ -59,6 +71,9 @@ aucrf_data_LRbowel <- auc_loc(rel_meta, "LB", "RB")
 aucrf_data_right_bs <- auc_loc(rel_meta, "RB", "RS")
 aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
 aucrf_data_allum <- auc_site(rel_meta, "mucosa", "stool")
+
+#testing building the model on filtered data
[…]

### 2017-08-23 code/random_forest.R
feature selection first go works

@@ -73,7 +73,18 @@ aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
 aucrf_data_allum <- auc_site(rel_meta, "mucosa", "stool")
 
 #testing building the model on filtered data
-aucrf_filter_left_bs <- auc_loc(filter_abund_meta, "LB", "LS") # this is not working
+aucrf_filter_left_bs <- auc_loc(filter_abund_meta, "LB", "LS") # this works but we need it to output an aucrf object. change function?
+
+#testing the AUCRFcv approach for optimalset feature selection 
+aucrf_cv_left_bs <- AUCRFcv(rf_aucrf, nCV=10, M=20)
+optimal_leftbs <- OptimalSet(aucrf_cv_left_bs)
+
+#take top ten OTUs from optimal set, put in a list, then use list to subset filtered data
+
+#put new subsetted data in to randomForest, get auc, do CV
+
+#then do that for all models and plot 
+
 
 #need to fix specific function for these to output just the rf object 
 #rf_exitlum_aucrf <- auc_site(rel_meta, "stool", "exit") # not working 

### 2017-08-24 code/random_forest.R
making progress w feature selection

@@ -79,11 +79,41 @@ aucrf_filter_left_bs <- auc_loc(filter_abund_meta, "LB", "LS") # this works but 
 aucrf_cv_left_bs <- AUCRFcv(rf_aucrf, nCV=10, M=20)
 optimal_leftbs <- OptimalSet(aucrf_cv_left_bs)
 
+#testing AUCRFcv approach on unfiltered data
+left_test <- subset(rel_meta, location %in% c("LB", "LS"))
+left_test$location <- factor(left_test$location)
+#change levels of variable of interest to 0/1
+levels(left_test$location) <- c(1:length(levels(left_test$location))-1)
+# create RF model
+set.seed(seed)
+rf_left_test <- AUCRF(location ~ ., data = select(left_test, location, contains("Otu")),
+                  ntree = n_trees, pdel = 0.05, ranking = 'MDA')
+
+aucrf_cv_left_unfiltered <- AUCRFcv(rf_left_test, nCV=10, M=20)
+optimal_left_unfiltered <- OptimalSet(aucrf_cv_left_unfiltered)
+
 #take top ten OTUs from optimal set, put in a list, then use list to subset filtered data
+
+left_top6 <- as.list(optimal_left_unfiltered[,1])
+#but this is the same list as the Xopt from the rf object. so do i even need to do the aucrfcv with optimim set?? this is still main issue to figure out
 
 #put new subsetted data in to randomForest, get auc, do CV
 
 #then do that for all models and plot 
+
+#lets try testing with a bigger dataset 
+bowel_test <- subset(rel_meta, location %in% c("LB", "RB"))
+bowel_test$location <- factor(bowel_test$location)
+#change levels of variable of interest to 0/1
+levels(bowel_test$location) <- c(1:length(levels(bowel_test$location))-1)
+# create RF model
+set.seed(seed)
+rf_bowel_test <- AUCRF(location ~ ., data = select(bowel_test, location, contains("Otu")),
+                      ntree = n_trees, pdel = 0.05, ranking = 'MDA')
+
+aucrf_cv_bowel_unfiltered <- AUCRFcv(rf_bowel_test, nCV=10, M=20)
+optimal_bowel_unfiltered <- OptimalSet(aucrf_cv_bowel_unfiltered)
+
 
 
 #need to fix specific function for these to output just the rf object 

### 2017-08-25 code/random_forest.R
one round of feature reduction model stuff working

@@ -94,9 +94,9 @@ optimal_left_unfiltered <- OptimalSet(aucrf_cv_left_unfiltered)
 
 #take top ten OTUs from optimal set, put in a list, then use list to subset filtered data
 
-left_top6 <- as.list(optimal_left_unfiltered[,1])
+left_top6 <- as.vector(optimal_left_unfiltered$Name)
 #but this is the same list as the Xopt from the rf object. so do i even need to do the aucrfcv with optimim set?? this is still main issue to figure out
-
+#maybe need to do 100 iters??
 #put new subsetted data in to randomForest, get auc, do CV
 
 #then do that for all models and plot 
@@ -111,8 +111,40 @@ set.seed(seed)
 rf_bowel_test <- AUCRF(location ~ ., data = select(bowel_test, location, contains("Otu")),
                       ntree = n_trees, pdel = 0.05, ranking = 'MDA')
 
-aucrf_cv_bowel_unfiltered <- AUCRFcv(rf_bowel_test, nCV=10, M=20)
+aucrf_cv_bowel_unfiltered <- AUCRFcv(rf_bowel_test, nCV=10, M=100)
 optimal_bowel_unfiltered <- OptimalSet(aucrf_cv_bowel_unfiltered)
+
+bowel_top10 <- as.vector(optimal_bowel_unfiltered$Name[1:10])
+
+#subset shared to just be these columns then run randomForest again 
+
+bowel_top10_shared <- subset(rel_meta, select=colnames(rel_meta) %in% c('location',bowel_top10))
+rf_bowel_top10 <- randomize_loc(bowel_top10_shared, "LB", "RB")
+auc_bowel_top10 <- auc_loc(bowel_top10_shared, "LB", "RB")
+
+
+######doing the feat selection thing for the LS and RS model, see if it changes AUC or model, then decide if all of this is worth it.
+
+lumen_test <- subset(rel_meta, location %in% c("LS", "RS"))
+lumen_test$location <- factor(lumen_test$location)
+levels(lumen_test$location) <- c(1:length(levels(lumen_test$location))-1)
+set.seed(seed)
[…]

### 2017-09-05 rf_featurereduction.R
working on new plots

@@ -62,13 +62,19 @@ optimal_bowel_unfiltered <- OptimalSet(aucrf_cv_bowel_unfiltered)
 
 bowel_top10 <- as.vector(optimal_bowel_unfiltered$Name[1:10])
 
+bowel_top5 <- as.vector(optimal_bowel_unfiltered$Name[1:6])
+
 #subset shared to just be these columns then run model again 
 
 bowel_top10_shared <- subset(rel_meta, select=colnames(rel_meta) %in% c('location',bowel_top10))
 auc_bowel_top10 <- auc_loc(bowel_top10_shared, "LB", "RB")
 
+bowel_top5_shared <- subset(rel_meta, select=colnames(rel_meta) %in% c('location',bowel_top5))
+auc_bowel_top5 <- auc_loc(bowel_top5_shared, "LB", "RB")
+
+
 ######10fold cross v with reduced feature input
-#10 fold cross validation for L vs R mucosa
+#10 fold cross validation for L vs R mucosa - 10 features
 
 iters <- 100
 cv10f_aucs_muc10 <- c()
@@ -92,7 +98,34 @@ for(j in 1:iters){
 }
 cv10f_roc_muc10 <- roc(cv10f_all_resp_muc10~cv10f_all_pred_muc10)
 
-one_run <- roc(auc_bowel_top10$location~)
+#one_run <- roc(auc_bowel_top10$location~)
+
+###### 10 fold CV for mucosa - 5 features as input 
+iters <- 100
+cv10f_aucs_muc5 <- c()
+cv10f_all_resp_muc5 <- c()
+cv10f_all_pred_muc5 <- c()
+for(j in 1:iters){
+  set.seed(j)
+  sampling_muc <- sample(1:nrow(auc_bowel_top5),nrow(auc_bowel_top5),replace=F)
+  cv10f_probs_muc5 <- rep(NA,39)
+  for(i in seq(1,36,4)){
+    train_muc <- auc_bowel_top5[sampling_muc[-(i:(i+3))],]
+    test_muc <- auc_bowel_top5[sampling_muc[i:(i+3)],]
+    set.seed(seed)
+    temp_model_muc <- AUCRF(location~., data=train_muc, pdel=0.99, ntree=500)
+    cv10f_probs_muc5[sampling_muc[i:(i+3)]] <- predict(temp_model_muc$RFopt, test_muc, type='prob')[,2]
+  }
+  cv10f_roc_muc5 <- roc(auc_bowel_top5$location~cv10f_probs_muc5)
[…]

### 2017-09-05 code/build_figre3new.R
made lots of figure tweaks to get ready for DKT review

@@ -3,7 +3,7 @@
 
 ###########################################################
 
-pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson')
+pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson', 'cowplot')
 for (dep in pack_used){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
@@ -36,14 +36,16 @@ exittyc <- subset(tyc, samp2 == 'SS')
 #plots
 tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
 lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
-  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=16)) + 
+  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12)) + 
   scale_x_discrete(limits = tycpositions, breaks = tycpositions,
                    labels=c("P Mucosa vs P Lumen", "D Lumen vs P Lumen", "D Mucosa vs P Mucosa", "D Mucosa vs D Lumen")) +
   ylab("ThetaYC distance") 
 
+
+
 exitpositions <- c("RB_SS", "RS_SS", "LB_SS", "LS_SS")
 exit_plot <- ggplot(exittyc, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() +
-  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=16)) +
+  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12)) +
   scale_x_discrete(limits=exitpositions, breaks=exitpositions,
                    labels=c("P Mucosa vs Stool", "P Lumen vs Stool", "D Mucosa vs Stool", "D Lumen vs Stool")) +
   theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")
@@ -130,8 +132,8 @@ for (i in 1:nrow(alltyc)){
[…]

### 2017-09-05 code/build_figure2.R

@@ -40,7 +40,7 @@ positions <- c("RB", "RS", "LB", "LS", "SS")
 phy_plot <- ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(color=variable)) + 
   scale_color_discrete(guide=FALSE)+
   geom_boxplot(aes(fill=variable), outlier.shape=21, outlier.size=2.5) + theme_bw() + 
-  theme(axis.text = element_text(size= 16), axis.title= element_text(size=18), legend.text=element_text(size=14), legend.title=element_text(size=16)) +
+  theme(axis.text = element_text(size= 10), axis.title= element_text(size=12), legend.text=element_text(size=10), legend.title=element_text(size=12)) +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
   theme(axis.title.x=element_blank(), panel.grid.major.x = element_blank(),panel.grid.minor.x = element_blank()) +
@@ -58,7 +58,7 @@ positions <- c("RB", "RS", "LB", "LS", "SS")
 simp_plot <- ggplot(simpmeta, aes(x=location, y=invsimpson, group =1)) +geom_point() +geom_jitter(width=0.2) +theme_bw() + ylab("Inverse Simpson Diversity") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 16), axis.title= element_text(size=18)) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12)) +
   stat_summary(aes(x=location, y=invsimpson), data = simpmeta, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.4)
 
 #export as PDF
@@ -74,3 +74,8 @@ phy_plot
 simp_plot 
 
 dev.off()
+
+##### Cowplot way to save the plots! 
+fig2 <- plot_grid(phy_plot, simp_plot, labels = c("A", "B"), ncol = 1, align = "v")  
+save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_2.pdf', fig2, ncol=1, nrow=2, base_width=12, base_height = 7)
+

### 2017-09-05 code/oncogenic_otus.R

@@ -38,7 +38,7 @@ fuso_plot <- ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(
   ylab("F. nucleatum Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12))
 
 #IBD fuso OTU - F. varium 
 
@@ -51,7 +51,7 @@ fusoviv_plot <- ggplot(fusoviv, aes(x=location, y=Otu00472_relAbund)) +geom_jitt
   ylab("F. varium Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12))
 
 
 
@@ -76,7 +76,7 @@ porphy_plot <- ggplot(p_152, aes(x=location, y=Otu152_abund)) +geom_jitter(width
   ylab("P. asacharolytica Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 14), axis.title= element_text(size=14))
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12))
 
 
 
@@ -103,4 +103,10 @@ fusoviv_plot
 porphy_plot
 
 dev.off()
+
+####Cowplot way!
+
+fig6 <- plot_grid(fuso_plot, fusoviv_plot, porphy_plot, labels = c("A", "B", "C"), ncol = 1, align = "v")  
[…]

### 2017-09-05 code/random_forest.R

@@ -592,23 +592,10 @@ legend('topright', legend=c("Left lumen", "Right lumen"), pch=c(21, 21), pt.bg=c
 #export as PDF
 
 plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
-pdf(file=plot_file, width=9, height=12)
+pdf(file=plot_file, width=11, height=12)
 layout(matrix(c(1,
-                2,
-                3), 
-              nrow=3, byrow = TRUE))
-
-#Lumen vs mucosa plot 
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
-plot(cv10f_roc_right_bs, col='blue', lwd=3, add=T, lty=1)
-plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
-mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.2)
-mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.2)
-legend('bottom', legend=c(
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, AUC =0.980'),
-  sprintf('P Lumen vs P Mucosa, 10-fold CV, AUC = 0.797')
-),lty=c(1, 1, 1), lwd=3, col=c('red', 'blue'), bty='n', cex=1.2)
+                2), 
+              nrow=2, byrow = TRUE))
 
 #RB vs RS
 
@@ -622,16 +609,17 @@ par(mar=c(5, 15, 1, 1))
 plot(1, type="n", ylim=c(0,length(right_otu_feat)*2), xlim=c(1e-4,3), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n", cex.lab=1.5)
 index <- 1
 for(i in right_otu_feat){
-  stripchart(at=index-0.35, jitter(rs_abunds[,i], amount=1e-5), pch=21, bg="purple", method="jitter", jitter=0.2, add=T, cex=1.2, lwd=0.5)
-  stripchart(at=index+0.35, jitter(rb_abunds[,i], amount=1e-5), pch=21, bg="orange", method="jitter", jitter=0.2, add=T, cex=1.2, lwd=0.5)
[…]

### 2017-09-05 rf_featurereduction.R

@@ -346,20 +346,39 @@ muc_vs_lum_pval <- roc.test(cv10f_roc_muc10, cv10f_roc_lum10)
 plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_S1.pdf'
 pdf(file=plot_file, width=9, height=12)
 layout(matrix(c(1,
-                2,
-                3), 
-              nrow=3, byrow = TRUE))
+                2), 
+              nrow=2, byrow = TRUE))
+
+#Lumen vs mucosa plot - 5 features 
+par(mar=c(4,5,1,1))
+plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.2)
+plot(cv10f_roc_right_bs6, col='blue', lwd=3, add=T, lty=1)
+plot(cv10f_roc_left_bs6, col = 'red', lwd=3, add=T, lty=1)
+mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.5)
+mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.5)
+legend('bottom', legend=c(
+  sprintf('D Lumen vs D Mucosa, AUC = 0.975'),
+  sprintf('P Lumen vs P Mucosa, AUC = 0.856')
+  #sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
+),lty=c(1, 1), lwd=3, col=c('red', 'blue'), bty='n', cex=1.2)
+
+mtext('A.', side=2, line=1.5, las=1, adj=2, padj=-10, cex=2, font=2)
+
 
 #####10fold plot left vs right mucosa and lumen plot - 5 features 
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
+par(mar=c(4,5,1,1))
+plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.2)
 plot(cv10f_roc_muc5,col = 'green4', lwd=3, add=T, lty=1) #r vs l mucosa cross validation
[…]

### 2017-09-05 submission/manuscript.Rmd
manuscript with tweaks for DKT review

@@
 To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances.
-We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure 4A).
-The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal.
+We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure S1A).
+The models were constructed to only use the five most predictive OTUs as input to reduce overfitting between samples.
 The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively).
 The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
-For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B).
-The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4C).
+For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4A).
+The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4B).
 These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.
@@
 Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) compared to the proximal versus distal lumen (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`).
-The model that separated proximal and distal mucosal samples used `r length(2:ncol(aucrf_data_LRlumen))` features.
-OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B).
-The model used `r length(2:ncol(aucrf_data_LRbowel))` features to separate proximal and distal lumenal samples.
-Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).
+The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) compared to the proximal versus distal lumen (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`).
+These models were also constructed to use only the five most predictive OTUs as input to reduce overfitting.
+OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5A).
+Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5B).
 This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.
@@
 To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side.
-For each comparison we identified top 5 OTUs that were strongly predictive of one site or another.
+For each comparison we identified the top five OTUs that were strongly predictive of one site or another.
 Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015.
@@
 The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_.
-The model was less effective at classifying the proximal and distal lumenal contents, probably because the samples are arguably composed of the same bacteria but differ in water content.
+Thus the gut microenvironment of each location likely enriches for these specific microbes.
+
+In addition to identifying features that are specific to each side of the gut, the ability of the Random Forest to classify samples can serve as a proxy for similarity.
+That is, a higher AUC value means the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value.
+For instance, the model separating the proximal and distal mucosa has an AUC of `r signif(cv10f_roc_muc$auc, digits=3)` whereas the model for classifying the proximal and distal lumen has a much lower AUC of `r signif(cv10f_roc_lum$auc, digits =3)`.
+Further, prior to reducing the input to five OTUs, the latter model required 44 OTUs to best separate the samples.
+The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested.
+We speculate that the model was least effective at classifying the proximal and distal lumenal contents because the samples are arguably composed of the same bacteria but differ in water content.
 

### 2017-09-05 submission/manuscript.Rmd
manuscript version sent to kim

@@
 - \linenumbers
+- \renewcommand{\familydefault}{\sfdefault}
+- \usepackage{graphicx}
 geometry: margin=1.0in
@@
 Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location.
-A model differentiating the proximal mucosa and lumen was built with an AUC of `r signif(cv10f_roc_proximal_bs$auc, digits=3)`.
+A model differentiating the proximal mucosa and lumen was built with an AUC of 0.856.
 The proximal mucosa had a higher abundance of the genera _Enterobacteriaceae_ and _Bacteriodes_.
-_Peptoniphilus, Anaerococcus, Finegoldia,_ and _Turicibacter_ were most likely to be found in distal mucosal samples versus distal luminal samples (AUC = `r signif(cv10f_roc_distal_bs$auc, digits=3)`).
-The classification model performed well (AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`).
+_Peptoniphilus, Anaerococcus, Finegoldia,_ and _Turicibacter_ were most likely to be found in distal mucosal samples versus distal luminal samples (AUC = 0.975).
+The classification model performed well (AUC = 0.920) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.575).
 The distal mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_.
@@
 The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
-IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012 @Forbes2016 @Halfvarson2017.
+IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes [@Yamauchi2012; @Forbes2016; @Halfvarson2017].
 The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
@@
 In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010.
-The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012 @Albenberg2014 @Donaldson2015.
+The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients [@Yamauchi2012; @Albenberg2014; @Donaldson2015].
 In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016.
@@
 
-Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host @Yamauchi2012 @Halfvarson2017.
+Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host [@Yamauchi2012; @Halfvarson2017].
 For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated @Dejea2014.
@@
 While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
-Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014 @Harrell2012.
+Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer [@Jalanka2014; @Harrell2012].
 As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
@@
 To determine what factors may be driving the differences seen among the samples, we compared distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon @Eckburg2005 @deCarcer2010 @Zhang2013.
+We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon [@Eckburg2005; @deCarcer2010; @Zhang2013].
 Thus interpersonal variation between subjects drove the differences between samples more than sample site or location.
@@
 The models were constructed to only use the five most predictive OTUs as input to reduce overfitting between samples.
-The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively).
+The models performed well when classifying these samples (AUC = 0.856 and AUC = 0.975, respectively).
 The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
@@
 Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) compared to the proximal versus distal lumen (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`).
+The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.920) compared to the proximal versus distal lumen (AUC = 0.575).
 These models were also constructed to use only the five most predictive OTUs as input to reduce overfitting.
@@
 That is, a higher AUC value means the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value.
-For instance, the model separating the proximal and distal mucosa has an AUC of `r signif(cv10f_roc_muc$auc, digits=3)` whereas the model for classifying the proximal and distal lumen has a much lower AUC of `r signif(cv10f_roc_lum$auc, digits =3)`.
-Further, prior to reducing the input to five OTUs, the latter model required 44 OTUs to best separate the samples.
+For instance, the model separating the proximal and distal mucosa has an AUC of 0.925 whereas the model for classifying the proximal and distal lumen has a much lower AUC of 0.575.
+Further, prior to reducing the input to five OTUs, the latter model required 44 OTUs to best separate the samples (AUC = 0.755).
 The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested.
@@
 ####Figure 1
+
+\includegraphics{../submission/figure_1.pdf}
+
 Sampling strategy.
@@
 \newpage
+\includegraphics{../submission/figure_2.pdf}
 
@@
 \newpage
+\includegraphics{../submission/figure_3.pdf}
 
@@
 \newpage
+\includegraphics{../submission/figure_4.pdf}
 
 ####Figure 4
-Random Forest classifies the mucosa and lumen of each side of the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal and proximal sides of the colon.
-(B) Top five OTUs that are most important for the classification model for the distal mucosa and lumen (B) and the proximal mucosa and lumen (C).
+Taxa specific to the distal and proximal sides of the colon.
+Top five OTUs that are most important for the classification model for the distal mucosa and lumen (A) and the proximal mucosa and lumen (B).
 
 \newpage
+\includegraphics{../submission/figure_5.pdf}
 
 ####Figure 5 
-Random Forest classifies the distal and proximal sides of the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal lumen versus proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
-(B) Top five OTUs that are most important for the classification model for the distal and proximal mucosa (B) and the distal and proximal lumen (C).
+Taxa specific to the distal and proximal mucosa and lumen.
+Top five OTUs that are most important for the classification model for the distal and proximal mucosa (A) and the distal and proximal lumen (B).
 
 \newpage
+\includegraphics{../submission/figure_6.pdf}
 
@@
 
+\newpage
+\includegraphics{../submission/figure_S1.pdf}
+
+####Figure S1
+Random Forest classifies locations in the colon.
+A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal lumen versus proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
+(B) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal (red) and proximal (blue) sides of the colon.
+
 ### References

### 2017-09-05 submission/manuscript.tex
manuscript version sent to kim

@@
 \linenumbers
+\renewcommand{\familydefault}{\sfdefault}
+\usepackage{graphicx}
 
@@
 location. A model differentiating the proximal mucosa and lumen was
-built with an AUC of 0.831. The proximal mucosa had a higher abundance
+built with an AUC of 0.856. The proximal mucosa had a higher abundance
 of the genera \emph{Enterobacteriaceae} and \emph{Bacteriodes}.
@@
 were most likely to be found in distal mucosal samples versus distal
-luminal samples (AUC = 0.98). The classification model performed well
-(AUC = 0.912) when classifying mucosal samples into proximal or distal
+luminal samples (AUC = 0.975). The classification model performed well
+(AUC = 0.920) when classifying mucosal samples into proximal or distal
 sides, but separating luminal samples from each side proved more
-challenging (AUC = 0.755). The distal mucosa was found to have high
+challenging (AUC = 0.575). The distal mucosa was found to have high
 populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
@@
 Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
-accelerated by perturbations in gut microbes (1) (2) (3). The severity,
+accelerated by perturbations in gut microbes (1--3). The severity,
 symptoms, morbidity and mortality of these diseases is known to vary
@@
 genomic instability of precancerous cells, respectively, in addition to
-variation in the previously mentioned chemical gradients (1) (5) (6). In
+variation in the previously mentioned chemical gradients (1, 5, 6). In
 IBD patients, disease flares in the distal colon are usually indicative
@@
 IBD or CRC can be attributed to specific molecular events as a result of
-interactions between the gut microbiota and human host (1) (3). For
+interactions between the gut microbiota and human host (1, 3). For
 instance, comparison of the bacteria present on CRC tumors with those
@@
 colon, but complicates microbiome profiling as the chemicals strip the
-bowel of contents and disrupt the mucosal layer (10) (11). As such, what
+bowel of contents and disrupt the mucosal layer (10, 11). As such, what
 little information we do have about the biogeographical distribution of
@@
 subjects (Figure 3C), consistent with previous human microbiome studies
-that have sampled multiple sites of the human colon (13) (14) (15). Thus
+that have sampled multiple sites of the human colon (13--15). Thus
 interpersonal variation between subjects drove the differences between
@@
 the lumenal versus mucosal samples for the proximal and distal sides,
-independently (Figure 4A). The constructed model used 14 features for
-the proximal and 6 for the distal. The models performed well when
-classifying these samples (0.831 and 0.98, respectively). The OTUs that
-were most predictive of each site were identified by their greatest mean
-decrease in accuracy when removed from the model. For distinguishing the
-proximal lumen and mucosa, OTUs from the \emph{Bacteriodes},
-\emph{Actinomyces}, \emph{Psuedomonas} and two OTUs from the
-\emph{Enterobacteraceae} genera were differentially abundant (Figure
-4B). The model classifying the distal lumen and mucosa identified OTUs
-from \emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
-two OTUs from the \emph{Anaerococcus} genera (Figure 4C). These results
-indicated that there were fine differences between the different sites
-of the colon, and that these could be traced to specific OTUs on each
-side.
+independently (Figure S1A). The models were constructed to only use the
+five most predictive OTUs as input to reduce overfitting between
+samples. The models performed well when classifying these samples (AUC =
+0.856 and AUC = 0.975, respectively). The OTUs that were most predictive
+of each site were identified by their greatest mean decrease in accuracy
+when removed from the model. For distinguishing the proximal lumen and
+mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
+\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
+were differentially abundant (Figure 4A). The model classifying the
+distal lumen and mucosa identified OTUs from \emph{Turicibacter},
+\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
+\emph{Anaerococcus} genera (Figure 4B). These results indicated that
+there were fine differences between the different sites of the colon,
+and that these could be traced to specific OTUs on each side.
 
@@
 samples using 10-fold cross validation. The model performed best when
-distinguishing the proximal versus distal mucosa (Figure 5A, AUC =
-0.912) compared to the proximal versus distal lumen (AUC = 0.755). The
-model that separated proximal and distal mucosal samples used 6
-features. OTUs that were differentially abundant between the distal and
-proximal mucosa included members of the \emph{Porphyromonas},
-\emph{Murdochiella}, \emph{Finegoldia}, \emph{Anaerococcus} and
-\emph{Peptoniphilus} genera (Figure 5B). The model used 44 features to
-separate proximal and distal lumenal samples. Differentially abundant
-OTUs of the proximal and distal lumen included three OTUs of the
-\emph{Bacteroides} genus, a \emph{Clostridium IV} OTU and an
-\emph{Oscillibacter} OTU (Figure 5C). This analysis found that some of
-the same OTUs that are distinct between the mucosa and lumen also helped
-to differentiate between the two sides- such as \emph{Anaerococcus} and
-\emph{Finegoldia}.
+distinguishing the proximal versus distal mucosa (Figure S1B, AUC =
+0.920) compared to the proximal versus distal lumen (AUC = 0.575). These
+models were also constructed to use only the five most predictive OTUs
+as input to reduce overfitting. OTUs that were differentially abundant
+between the distal and proximal mucosa included members of the
+\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
+\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5A).
+Differentially abundant OTUs of the proximal and distal lumen included
+three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
+and an \emph{Oscillibacter} OTU (Figure 5B). This analysis found that
+some of the same OTUs that are distinct between the mucosa and lumen
+also helped to differentiate between the two sides- such as
+\emph{Anaerococcus} and \emph{Finegoldia}.
 
@@
 Random Forest model to identify microbes specific to each side. For each
-comparison we identified top 5 OTUs that were strongly predictive of one
-site or another. Generally, OTUs identified in each location were
+comparison we identified the top five OTUs that were strongly predictive
+of one site or another. Generally, OTUs identified in each location were
 consistent with known physiological gradients along the gut axis (6).
@@
 host strictly anaerobic species such as \emph{Porphyromonas},
-\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. The
-model was less effective at classifying the proximal and distal lumenal
-contents, probably because the samples are arguably composed of the same
-bacteria but differ in water content.
+\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. Thus
+the gut microenvironment of each location likely enriches for these
+specific microbes.
+
+In addition to identifying features that are specific to each side of
+the gut, the ability of the Random Forest to classify samples can serve
+as a proxy for similarity. That is, a higher AUC value means the samples
+are more efficiently classfied (and thus more different) than a model
+with a lower AUC value. For instance, the model separating the proximal
+and distal mucosa has an AUC of 0.925 whereas the model for classifying
+the proximal and distal lumen has a much lower AUC of 0.575. Further,
+prior to reducing the input to five OTUs, the latter model required 44
+OTUs to best separate the samples (AUC = 0.755). The much lower AUC and
+need for a high number of features compared to other models suggest
+these locations are the most similar of the comparisons tested. We
+speculate that the model was least effective at classifying the proximal
+and distal lumenal contents because the samples are arguably composed of
+the same bacteria but differ in water content.
 
@@
 
+\includegraphics{../submission/figure_1.pdf}
+
 Sampling strategy. A flexible sigmoidoscope was used to sample the
@@
 
+\includegraphics{../submission/figure_2.pdf}
+
 \paragraph{Figure 2}\label{figure-2}
@@
 \newpage
+
+\includegraphics{../submission/figure_3.pdf}
 
@@
 
+\includegraphics{../submission/figure_4.pdf}
+
 \paragraph{Figure 4}\label{figure-4}
 
-Random Forest classifies the mucosa and lumen of each side of the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold
-cross validation of the Random Forest model classifying lumen and
-mucosal samples for the distal and proximal sides of the colon. (B) Top
-five OTUs that are most important for the classification model for the
-distal mucosa and lumen (B) and the proximal mucosa and lumen (C).
+Taxa specific to the distal and proximal sides of the colon. Top five
+OTUs that are most important for the classification model for the distal
+mucosa and lumen (A) and the proximal mucosa and lumen (B).
 
@@
 
+\includegraphics{../submission/figure_5.pdf}
+
 \paragraph{Figure 5}\label{figure-5}
 
-Random Forest classifies the distal and proximal sides of the colon. A)
-Receiver Operator Characteristic curves are shown for the 10-fold cross
-validation of the Random Forest model classifying distal lumen versus
-proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
-(B) Top five OTUs that are most important for the classification model
-for the distal and proximal mucosa (B) and the distal and proximal lumen
-(C).
+Taxa specific to the distal and proximal mucosa and lumen. Top five OTUs
+that are most important for the classification model for the distal and
+proximal mucosa (A) and the distal and proximal lumen (B).
 
 \newpage
+
+\includegraphics{../submission/figure_6.pdf}
 
@@
 varium} and (C) \emph{Porphyromonas asacharolytica}
+
+\newpage
+
+\includegraphics{../submission/figure_S1.pdf}
+
+\paragraph{Figure S1}\label{figure-s1}
+
+Random Forest classifies locations in the colon. A) Receiver Operator
+Characteristic curves are shown for the 10-fold cross validation of the
+Random Forest model classifying distal lumen versus proximal lumen
+(orange) and distal mucosa vs proximal mucosa (green). (B) Receiver
+Operator Characteristic curves are shown for the 10-fold cross
+validation of the Random Forest model classifying lumen and mucosal
+samples for the distal (red) and proximal (blue) sides of the colon.
 

### 2017-09-26 submission/manuscript.Rmd
updated with new fig 1

@@
 ---
-title: The microbiota of the proximal and distal human colon
+title: "The microbiota of the proximal and distal human colon"
 csl: mbio.csl
@@
 output:
+  word_document: default
   pdf_document:
@@
   html_document: default
-  word_document: default
 header-includes:

### 2017-09-26 code/rf_optimized.R
added RF hold one out code

@@ -1,0 +1,70 @@
+### optimizing RF models by doing leave-one-out, per reviewers suggestions on marc's papers
+
+#Kaitlin Flynn, Schloss lab, updated 9-23-17
+
+
+#6) aggregate all of the models from the left-out-list to get the AUC table to plot a curve and get an AUC value for the whole model
+# do that for all of the models we have
+
+#Load packages, files, make OTU table 
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret')
+for (dep in pack_used){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
+                     quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+#load in all of the files, get rel abund of >1% and/or >10%
+
+meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+
+#make OTU abundance file
+#Create df with relative abundances
+shared_file <- subset(shared_file, select = -c(numOtus, label))
+shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+
+rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+
+#Create vector of OTUs with median abundances >1%
+OTUs_1 <- apply(rel_abund, 2, max) > 1
+OTU_list <- colnames(rel_abund)[OTUs_1]
+#get df of just top OTUs
+rel_abund_top <- rel_abund[, OTUs_1]
+rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+seed <- 1
+n_trees <- 2001
+
+source('code/random_functions.R')
+source('code/tax_level.R')
+
+#iteratively, do the following:
+#1) select 19 samples (patients) from the 20, hold out the other one
+#2) train the model using AUCRF and cross validation.
+#3) from that model, will have an AUCRF object that i can use to predict/test on the held out sample
[…]

### 2017-09-26 code/build_figre3new.R
figure tweaks

@@ -36,7 +36,7 @@ exittyc <- subset(tyc, samp2 == 'SS')
 #plots
 tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
 lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
-  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12)) + 
+  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
   scale_x_discrete(limits = tycpositions, breaks = tycpositions,
                    labels=c("P Mucosa vs P Lumen", "D Lumen vs P Lumen", "D Mucosa vs P Mucosa", "D Mucosa vs D Lumen")) +
   ylab("ThetaYC distance") 
@@ -45,7 +45,7 @@ lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +the
 
 exitpositions <- c("RB_SS", "RS_SS", "LB_SS", "LS_SS")
 exit_plot <- ggplot(exittyc, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() +
-  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12)) +
+  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
   scale_x_discrete(limits=exitpositions, breaks=exitpositions,
                    labels=c("P Mucosa vs Stool", "P Lumen vs Stool", "D Mucosa vs Stool", "D Lumen vs Stool")) +
   theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")

### 2017-09-26 code/build_figure2.R

@@ -3,6 +3,7 @@
 #
 #
 ##########################################
+library(reshape2)
 
 meta <- 'data/raw/kws_metadata.tsv'
 shared <- 'data/mothur/kws_final.an.shared'
@@ -43,7 +44,7 @@ phy_plot <- ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(co
   theme(axis.text = element_text(size= 10), axis.title= element_text(size=12), legend.text=element_text(size=10), legend.title=element_text(size=12)) +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(axis.title.x=element_blank(), panel.grid.major.x = element_blank(),panel.grid.minor.x = element_blank()) +
+  theme(axis.title.x=element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
   theme(legend.justification = c(0.999, 0.999), legend.position = c(0.999, 0.999)) + scale_fill_brewer(palette="Dark2", name="Phylum") +
   ylab("% Relative Abundance") 
 
@@ -58,7 +59,8 @@ positions <- c("RB", "RS", "LB", "LS", "SS")
 simp_plot <- ggplot(simpmeta, aes(x=location, y=invsimpson, group =1)) +geom_point() +geom_jitter(width=0.2) +theme_bw() + ylab("Inverse Simpson Diversity") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12)) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.border = element_blank(), panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
   stat_summary(aes(x=location, y=invsimpson), data = simpmeta, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.4)
 
 #export as PDF

### 2017-09-26 code/build_figure3.R

@@ -74,7 +74,7 @@ wilcox.test(thetayc ~ match, data=leftandrighttyc, paired = T)
 #lvsr$match <- factor(lvsr$match, levels=lvsr$match[order(lvsr$samp1)])
 ggplot(lvsr, aes(x=match, y=thetayc)) + geom_point() + geom_jitter(width= 0.2) +theme_bw() +
   scale_x_discrete(labels=c("L Mucosa vs L Lumen", "L Mucosa vs R Lumen", "L Lumen vs R Lumen", "R Mucosa vs R Lumen")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 16), axis.title= element_text(size=18)) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 16), axis.title= element_text(size=18), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
   ylab("ThetaYC distance") + stat_summary(aes(x=match, y=thetayc), data = lvsr, fun.y=median, fun.ymin=median, fun.ymax=median, geom="crossbar", width=0.4)
 
 

### 2017-09-26 code/oncogenic_otus.R

@@ -3,6 +3,7 @@
 #first gotta find the OTU numbers for the cancer bugs
 
 #fuso OTU numbers: Otu00179, Otu00184, Otu00314, Otu00472, Otu01440
+library(cowplot)
 
 meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
 shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
@@ -38,7 +39,7 @@ fuso_plot <- ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(
   ylab("F. nucleatum Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12))
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
 
 #IBD fuso OTU - F. varium 
 
@@ -51,7 +52,7 @@ fusoviv_plot <- ggplot(fusoviv, aes(x=location, y=Otu00472_relAbund)) +geom_jitt
   ylab("F. varium Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12))
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
 
 
 
@@ -76,7 +77,7 @@ porphy_plot <- ggplot(p_152, aes(x=location, y=Otu152_abund)) +geom_jitter(width
   ylab("P. asacharolytica Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
[…]

### 2017-09-26 code/random_forest.R

@@ -343,8 +343,8 @@ plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i
 plot(cv10f_roc_right_bs, col='blue', lwd=3, add=T, lty=1)
 #plot(cv10f_roc, col = 'purple', lwd=3, add=T, lty=1)
 plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
-mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.5)
-mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.5)
+mtext(side=2, text="Sensitivity", line=2.5, cex=1.5)
+mtext(side=1, text="Specificity", line=2.5, cex=1.5)
 legend('bottom', legend=c(#sprintf('Lumen vs Mucosa, 10-fold CV, AUC = 0.925'),
                           sprintf('D Lumen vs D Mucosa, 10-fold CV, AUC =0.980'),
                           sprintf('P Lumen vs P Mucosa, 10-fold CV, AUC = 0.797')
@@ -358,8 +358,8 @@ par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
 plot(cv10f_roc_muc,col = 'green4', lwd=3, add=T, lty=1) #r vs l mucosa cross validation
 plot(cv10f_roc_lum, col = 'orange', lwd=3, add=T, lty=1) #r vs l lumen cross validation
-mtext(side=2, text="True Positive (Sensitivity)", line=2.5, cex=1.2)
-mtext(side=1, text="True Negative (Specificity)", line=2.5, cex=1.2)
+mtext(side=2, text="Sensitivity", line=2.5, cex=1.2)
+mtext(side=1, text="Specificity", line=2.5, cex=1.2)
 legend('bottom', legend=c(sprintf('D mucosa vs P mucosa 10-fold CV, AUC = 0.9159'),
                           sprintf('D lumen vs P lumen 10-fold CV, AUC = 0.7551')
                           # sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
@@ -372,8 +372,8 @@ par(mar=c(4,4,1,1))
[…]

### 2017-10-09 submission/manuscript.Rmd
kim edits and optimization code

@@
 
-Running title: The microbiome of the proximal and distal human colon
+Running title: The microbiota of the proximal and distal human colon
 
@@
 
-Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^
+Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, D. Kim Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^
 
@@
 
-$\dagger$ Corresponding authors: kturgeon@med.umich.edu and pschloss@umich.edu
+$\dagger$ Corresponding authors: kturgeon@umich.edu and pschloss@umich.edu
 
@@
 
-2\. Pennslyvania State University, Hershey, Pennslyvania ??
+2\. Pennsylvania State University, Hershey, Pennsylvania ??
 
@@
 
-Chemical and nutrient gradients along the human colon create microenvironments that affect the distrubution and composition of the gut microbiota.
+Chemical and nutrient gradients along the human colon create microenvironments that affect the distribution and composition of the gut microbiota.
 The microbiome has been implicated in colorectal cancer (CRC) and inflammatory bowel disease (IBD).
-Further, these diseases exhibit different symptoms depending on the location of the colon they are found in.
+Further, these diseases can exhibit different findings depending on the location of the colon they are found in.
 CRC tumors of the proximal and distal colon are morphologically and genetically distinct.
-Similarly, inflammatory bowel diseases such as Crohn's are typically exacerbated in the proximal instestine while ulcerative colitis patients often experience symptoms in the distal colon.
+Similarly, Crohn's and Ulcerative Colitis (UC) manifest different morphological characteristics in different parts of the GI tract.
 Previous analysis of the fecal microbiota from healthy and CRC or IBD patients has revealed different microbial signatures associated with these diseases.
 We extended these observations of the fecal microbiome to include analysis of the proximal and distal healthy human colon.
-We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure.
-In contrast to previous efforts using prepared colonoscopy, this technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
+We used a two-colonoscope approach on subjects that had not undergone any bowel preparation procedure.
+In contrast, previous studies have obtained samples from colonoscopy or surgical specimens where the subjects had undergone colonic preparation.
+Our technique allowed characterization of native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals.
@@
 
-#### wordcount?
+Keywords:
 
@@
 Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location.
-Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
+Microenvironments differ not only longitudinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
 The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
 IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes [@Yamauchi2012; @Forbes2016; @Halfvarson2017].
-The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur.
-For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms @Benedix2010.
-In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010.
+The manifestations of these diseases are known to vary based upon the biogeographical location in which they occur.
+For instance, CRC that arise in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012.
+In contrast, CRCs arising in the proximal (right) colon tumors are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012.
 The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients [@Yamauchi2012; @Albenberg2014; @Donaldson2015].
-In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016.
-UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare @Forbes2016.
+In IBD patients, disease occurring in the distal colon extending proximally is usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur anywhere along the GI tract, most commonly in the ileum and the cecum @Forbes2016.
+UC presents as continuous disease with only mucosal involvement, where as CD has skip lesions and full thickness involvement that may cause abscesses, strictures and fistulas @Forbes2016.
 Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC.
@@
 The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods.
-The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy procedures.
-While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
+The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy or surgery.
+While the latter method allows investigators to acquire samples from inside the human colon, typically these procedures are preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
 Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer [@Jalanka2014; @Harrell2012].

### 2017-10-09 code/rf_optimized.R

@@ -62,9 +62,9 @@ test_input <- subset(testsub, patient %in% c('11', '15', '17'))
 #create empty list to store in
 test_results <- data.frame(Patient = character(), zero = double(), one = double())
 held_out_results <- data.frame()
-for(p in test_input$patient){
-  test_set <- subset(test_input, test_input$patient != p)
-  held_out <- subset(test_input, test_input$patient == p)
+for(p in unique(testsub$patient)){
+  test_set <- subset(testsub, testsub$patient != p)
+  held_out <- subset(testsub, testsub$patient == p)
   rf_testset <- AUCRF(location~., data=select(test_set, location, contains("Otu")), ntree=n_trees, pdel=0.05, ranking="MDA")
   aucrf_test <- AUCRFcv(rf_testset, nCV=10, M=20)
   test_held_out <- predict(aucrf_test$RFopt, held_out, type='prob')

### 2017-10-11 code/bowel_optimized.R
got optimized RF going

@@ -49,7 +49,7 @@ source('code/tax_level.R')
 #6) aggregate all of the models from the left-out-list to get the AUC table to plot a curve and get an AUC value for the whole model
 # do that for all of the models we have
 
-testsub <- subset(rel_meta, location %in% c("LB", "LS"))
+testsub <- subset(rel_meta, location %in% c("LB", "RB"))
 testsub$location <- factor(testsub$location)
 levels(testsub$location) <- c(1:length(levels(testsub$location))-1)
 
@@ -68,53 +68,6 @@ for(p in unique(testsub$patient)){
   held_out_results$Patient <- p
   colnames(held_out_results) <- c("zero", "one", "Patient")
   test_results <- rbind(test_results, held_out_results)
-  }
+}
 
-left_optimized <- write.table(test_results, file = 'data/process/left_optimized.tsv', sep = '\t')
-
-
-#then plot curve outside of the loop 
-
-#do for all models!
-
-left_optimized_results <- read.table(file = 'data/process/left_optimized.tsv', sep = '\t')
-
-left_roc <- roc(testsub$location ~ left_optimized_results$one)
-
-
-#Lumen vs mucosa plot 
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
-plot(left_roc, col='pink', lwd=3, add=T, lty=1)
-#plot(cv10f_roc, col = 'purple', lwd=3, add=T, lty=1)
-plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
-mtext(side=2, text="Sensitivity", line=2.5, cex=1.5)
-mtext(side=1, text="Specificity", line=2.5, cex=1.5)
-legend('bottom', legend=c(#sprintf('Lumen vs Mucosa, 10-fold CV, AUC = 0.925'),
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, hold-one-out, AUC = 0.9079'),
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, unoptimized, AUC = 0.980')
-  #sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
-  #sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
[…]

### 2017-10-11 code/random_forest.R

@@ -346,10 +346,10 @@ plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
 mtext(side=2, text="Sensitivity", line=2.5, cex=1.5)
 mtext(side=1, text="Specificity", line=2.5, cex=1.5)
 legend('bottom', legend=c(#sprintf('Lumen vs Mucosa, 10-fold CV, AUC = 0.925'),
-                          sprintf('D Lumen vs D Mucosa, 10-fold CV, AUC =0.980'),
-                          sprintf('P Lumen vs P Mucosa, 10-fold CV, AUC = 0.797')
-                               #sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
-                               #sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
+  sprintf('D Lumen vs D Mucosa, 10-fold CV, AUC =0.980'),
+  sprintf('P Lumen vs P Mucosa, 10-fold CV, AUC = 0.797')
+  #sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
+  #sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
 ),lty=c(1, 1, 1), lwd=3, col=c('red', 'blue'), bty='n', cex=1.2)
 
 

### 2017-10-11 code/right_optimized.R

@@ -49,7 +49,7 @@ source('code/tax_level.R')
 #6) aggregate all of the models from the left-out-list to get the AUC table to plot a curve and get an AUC value for the whole model
 # do that for all of the models we have
 
-testsub <- subset(rel_meta, location %in% c("LB", "LS"))
+testsub <- subset(rel_meta, location %in% c("RB", "RS"))
 testsub$location <- factor(testsub$location)
 levels(testsub$location) <- c(1:length(levels(testsub$location))-1)
 
@@ -68,53 +68,6 @@ for(p in unique(testsub$patient)){
   held_out_results$Patient <- p
   colnames(held_out_results) <- c("zero", "one", "Patient")
   test_results <- rbind(test_results, held_out_results)
-  }
+}
 
-left_optimized <- write.table(test_results, file = 'data/process/left_optimized.tsv', sep = '\t')
-
-
-#then plot curve outside of the loop 
-
-#do for all models!
-
-left_optimized_results <- read.table(file = 'data/process/left_optimized.tsv', sep = '\t')
-
-left_roc <- roc(testsub$location ~ left_optimized_results$one)
-
-
-#Lumen vs mucosa plot 
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
-plot(left_roc, col='pink', lwd=3, add=T, lty=1)
-#plot(cv10f_roc, col = 'purple', lwd=3, add=T, lty=1)
-plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
-mtext(side=2, text="Sensitivity", line=2.5, cex=1.5)
-mtext(side=1, text="Specificity", line=2.5, cex=1.5)
-legend('bottom', legend=c(#sprintf('Lumen vs Mucosa, 10-fold CV, AUC = 0.925'),
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, hold-one-out, AUC = 0.9079'),
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, unoptimized, AUC = 0.980')
-  #sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
-  #sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
[…]

### 2017-10-11 code/stool_optimized.R

@@ -49,7 +49,7 @@ source('code/tax_level.R')
 #6) aggregate all of the models from the left-out-list to get the AUC table to plot a curve and get an AUC value for the whole model
 # do that for all of the models we have
 
-testsub <- subset(rel_meta, location %in% c("LB", "LS"))
+testsub <- subset(rel_meta, location %in% c("LS", "RS"))
 testsub$location <- factor(testsub$location)
 levels(testsub$location) <- c(1:length(levels(testsub$location))-1)
 
@@ -68,53 +68,6 @@ for(p in unique(testsub$patient)){
   held_out_results$Patient <- p
   colnames(held_out_results) <- c("zero", "one", "Patient")
   test_results <- rbind(test_results, held_out_results)
-  }
+}
 
-left_optimized <- write.table(test_results, file = 'data/process/left_optimized.tsv', sep = '\t')
-
-
-#then plot curve outside of the loop 
-
-#do for all models!
-
-left_optimized_results <- read.table(file = 'data/process/left_optimized.tsv', sep = '\t')
-
-left_roc <- roc(testsub$location ~ left_optimized_results$one)
-
-
-#Lumen vs mucosa plot 
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
-plot(left_roc, col='pink', lwd=3, add=T, lty=1)
-#plot(cv10f_roc, col = 'purple', lwd=3, add=T, lty=1)
-plot(cv10f_roc_left_bs, col = 'red', lwd=3, add=T, lty=1)
-mtext(side=2, text="Sensitivity", line=2.5, cex=1.5)
-mtext(side=1, text="Specificity", line=2.5, cex=1.5)
-legend('bottom', legend=c(#sprintf('Lumen vs Mucosa, 10-fold CV, AUC = 0.925'),
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, hold-one-out, AUC = 0.9079'),
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, unoptimized, AUC = 0.980')
-  #sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
-  #sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
[…]

### 2017-10-12 code/rf_optimized.R
hold one out model plots made

@@ -74,15 +74,33 @@ left_optimized <- write.table(test_results, file = 'data/process/left_optimized.
 
 
 #then plot curve outside of the loop 
+testsubR <- subset(rel_meta, location %in% c("RB", "RS"))
+testsubR$location <- factor(testsubR$location)
+levels(testsubR$location) <- c(1:length(levels(testsubR$location))-1)
+
+
+testsubM <- subset(rel_meta, location %in% c("LB", "RB"))
+testsubM$location <- factor(testsubM$location)
+levels(testsubM$location) <- c(1:length(levels(testsubM$location))-1)
+
+testsubS <- subset(rel_meta, location %in% c("LS", "RS"))
+testsubS$location <- factor(testsubS$location)
+levels(testsubS$location) <- c(1:length(levels(testsubS$location))-1)
 
 #do for all models!
 
 left_optimized_results <- read.table(file = 'data/process/left_optimized.tsv', sep = '\t')
+right_optimized_results <- read.table(file = 'data/process/right_optimized.tsv', sep = '\t')
+
+bowel_optimized_results <- read.table(file = 'data/process/bowel_optimized.tsv', sep = '\t')
+stool_optimized_results <- read.table(file = 'data/process/stool_optimized.tsv', sep = '\t')
 
 left_roc <- roc(testsub$location ~ left_optimized_results$one)
+right_roc <- roc(testsubR$location ~ right_optimized_results$one)
 
-
-#Lumen vs mucosa plot 
+muc_roc <- roc(testsubM$location ~ bowel_optimized_results$one)
+stool_roc <- roc(testsubS$location ~ stool_optimized_results$one)
+#Left old vs new method plot 
 par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
 plot(left_roc, col='pink', lwd=3, add=T, lty=1)
@@ -96,6 +114,49 @@ legend('bottom', legend=c(#sprintf('Lumen vs Mucosa, 10-fold CV, AUC = 0.925'),
[…]

### 2017-10-12 code/stool_optimized.R

@@ -63,7 +63,7 @@ for(p in unique(testsub$patient)){
   held_out <- subset(testsub, testsub$patient == p)
   rf_testset <- AUCRF(location~., data=select(test_set, location, contains("Otu")), ntree=n_trees, pdel=0.05, ranking="MDA")
   aucrf_test <- AUCRFcv(rf_testset, nCV=10, M=20)
-  test_held_out <- predict(aucrf_test$RFopt, held_out, type='prob')[,2]
+  test_held_out <- predict(aucrf_test$RFopt, held_out, type='prob')
   held_out_results <- as.data.frame(test_held_out)
   held_out_results$Patient <- p
   colnames(held_out_results) <- c("zero", "one", "Patient")

### 2017-10-16 code/rf_featurereduction.R
fixed fig 1 and rf figure

@@ -344,7 +344,7 @@ muc_vs_lum_pval <- roc.test(cv10f_roc_muc10, cv10f_roc_lum10)
 #############PLOTS
 
 plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_S1.pdf'
-pdf(file=plot_file, width=9, height=12)
+pdf(file=plot_file, width=7, height=9)
 layout(matrix(c(1,
                 2), 
               nrow=2, byrow = TRUE))

### 2017-10-16 code/rf_optimized.R

@@ -146,56 +146,42 @@ legend('bottom', legend=c(sprintf('P lumen vs D lumen, hold-one-out, AUC = 0.580
                           sprintf('P lumen vs D lumen, unoptimized, AUC = 0.7551')
 ),lty=c(1, 1), lwd=2, col=c('lightblue', 'darkblue'), bty='n', cex=1.2)
 
+###########saving figures as proper size
+
+plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_S1.pdf'
+pdf(file=plot_file, width=7, height=9)
+layout(matrix(c(1,
+                2), 
+              nrow=2, byrow = TRUE))
+
 #plots for figures 
 #Left and right new method plot
 par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='', cex.axis=1.5)
+plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
 plot(left_roc, col='red', lwd=3, add=T, lty=1)
 plot(right_roc, col = 'blue', lwd=3, add=T, lty=1)
-mtext(side=2, text="Sensitivity", line=2.5, cex=1.5)
-mtext(side=1, text="Specificity", line=2.5, cex=1.5)
+mtext(side=2, text="Sensitivity", line=2.5)
+mtext(side=1, text="Specificity", line=2.5)
 legend('bottom', legend=c(
-  sprintf('D Lumen vs D Mucosa, 10-fold CV, hold-one-out, AUC = 0.9079'),
-  sprintf('P Lumen vs P Mucosa, 10-fold CV, hold-one-out, AUC = 0.7645')
-),lty=c(1, 1, 1), lwd=3, col=c('red', 'blue'), bty='n', cex=1.2)
+  sprintf('D Lumen vs D Mucosa, AUC = 0.9079'),
+  sprintf('P Lumen vs P Mucosa, AUC = 0.7645')
+),lty=c(1, 1, 1), lwd=3, col=c('red', 'blue'), bty='n')
+
+mtext('A', side=2, line=1.5, las=1, adj=1.7, padj=-7, cex=2, font=2)
 
 #stool and mucosa new method plot 
 par(mar=c(4,4,1,1))
[…]

### 2017-10-17 code/build_figre3new.R
figures updated for publication

@@ -36,7 +36,7 @@ exittyc <- subset(tyc, samp2 == 'SS')
 #plots
 tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
 lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
-  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
+  theme(legend.position="none", axis.title.x=element_blank(), axis.line = element_line(colour = "black"), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
   scale_x_discrete(limits = tycpositions, breaks = tycpositions,
                    labels=c("P Mucosa vs P Lumen", "D Lumen vs P Lumen", "D Mucosa vs P Mucosa", "D Mucosa vs D Lumen")) +
   ylab("ThetaYC distance") 
@@ -45,7 +45,7 @@ lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +the
 
 exitpositions <- c("RB_SS", "RS_SS", "LB_SS", "LS_SS")
 exit_plot <- ggplot(exittyc, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() +
-  theme(legend.position="none", axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
+  theme(legend.position="none", axis.title.x=element_blank(), axis.line = element_line(colour = "black"), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
   scale_x_discrete(limits=exitpositions, breaks=exitpositions,
                    labels=c("P Mucosa vs Stool", "P Lumen vs Stool", "D Mucosa vs Stool", "D Lumen vs Stool")) +
   theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")
@@ -132,7 +132,7 @@ for (i in 1:nrow(alltyc)){
[…]

### 2017-10-17 code/build_figure2.R

@@ -40,12 +40,11 @@ phyla_RAmelt <- melt(phyla_RA[, phylaRAnames], id.vars=1)
 positions <- c("RB", "RS", "LB", "LS", "SS")
 phy_plot <- ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(color=variable)) + 
   scale_color_discrete(guide=FALSE)+
-  geom_boxplot(aes(fill=variable), outlier.shape=21, outlier.size=2.5) + theme_bw() + 
-  theme(axis.text = element_text(size= 10), axis.title= element_text(size=12), legend.text=element_text(size=10), legend.title=element_text(size=12)) +
+  geom_boxplot(aes(fill=variable), outlier.shape=21) + theme_bw() + 
+  theme(axis.text = element_text(size= 10), axis.title= element_text(size=12)) +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(axis.title.x=element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
-  theme(legend.justification = c(0.999, 0.999), legend.position = c(0.999, 0.999)) + scale_fill_brewer(palette="Dark2", name="Phylum") +
+  theme(axis.title.x=element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + scale_fill_brewer(palette="Dark2", name="Phylum") +
   ylab("% Relative Abundance") 
 
 
@@ -78,6 +77,6 @@ simp_plot
 dev.off()
 
 ##### Cowplot way to save the plots! 
-fig2 <- plot_grid(phy_plot, simp_plot, labels = c("A", "B"), ncol = 1, align = "v")  
-save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_2.pdf', fig2, ncol=1, nrow=2, base_width=12, base_height = 7)
+fig2 <- plot_grid(phy_plot, simp_plot, labels = c("A", "B"), label_size=16, ncol = 1, align = "v")  
+save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_2.pdf', fig2, ncol=1, nrow=2, base_width=7, base_height = 4)
 

### 2017-10-17 code/random_forest.R

@@ -592,7 +592,7 @@ legend('topright', legend=c("Left lumen", "Right lumen"), pch=c(21, 21), pt.bg=c
 #export as PDF
 
 plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
-pdf(file=plot_file, width=11, height=12)
+pdf(file=plot_file, width=7, height=9)
 layout(matrix(c(1,
                 2), 
               nrow=2, byrow = TRUE))
@@ -605,21 +605,21 @@ otu_taxa_right <- get_tax(1, right_otu_feat, tax_file)
 rs_abunds <- shared_meta[shared_meta$location=='RS', right_otu_feat]/10000 + 1e-4
 rb_abunds <- shared_meta[shared_meta$location=='RB', right_otu_feat]/10000 + 1e-4
 
-par(mar=c(5, 15, 1, 1))
-plot(1, type="n", ylim=c(0,length(right_otu_feat)*2), xlim=c(1e-4,3), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n", cex.lab=1.5)
+par(mar=c(5, 11, 1, 1))
+plot(1, type="n", ylim=c(0,length(right_otu_feat)*2), xlim=c(1e-4,3), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
 index <- 1
 for(i in right_otu_feat){
-  stripchart(at=index-0.35, jitter(rs_abunds[,i], amount=1e-5), pch=21, bg="darkgray", method="jitter", jitter=0.2, add=T, cex=1.2, lwd=0.5)
-  stripchart(at=index+0.35, jitter(rb_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, cex=1.2, lwd=0.5)
+  stripchart(at=index-0.35, jitter(rs_abunds[,i], amount=1e-5), pch=21, bg="black", method="jitter", jitter=0.2, add=T, lwd=0.5)
+  stripchart(at=index+0.35, jitter(rb_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, lwd=0.5)
   segments(median(rs_abunds[,i]),index-0.7,median(rs_abunds[,i]),index, lwd=3)
[…]

### 2017-10-17 submission/manuscript.Rmd
text edits made, manuscript getting closer

@@
 title: "The microbiota of the proximal and distal human colon"
-csl: mbio.csl
+csl: gastro.csl
 fontsize: 11pt
@@
 We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure S1A).
-The models were constructed to only use the five most predictive OTUs as input to reduce overfitting between samples.
-The models performed well when classifying these samples (AUC = 0.856 and AUC = 0.975, respectively).
+The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908).
 The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
@@
 Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.920) compared to the proximal versus distal lumen (AUC = 0.575).
-These models were also constructed to use only the five most predictive OTUs as input to reduce overfitting.
+The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) compared to the proximal versus distal lumen (AUC = 0.580).
 OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5A).
@@
 nucleatum subsp.
-animalis_, four of the samples were from the proximal mucosa and three from the distal mucosa (Figure 6A).
+animalis_, four of the samples were from the proximal mucosa and three from the distal mucosa (Figure S2A).
 The second most frequent _Fusobacterium_ sequence was OTU472, which aligned with 99% identity to _F.
@@
 Four study participants harbored _F.
-varium_ and the samples were split evenly between the proximal and distal mucosa (Figure 6B).
+varium_ and the samples were split evenly between the proximal and distal mucosa (Figure S2B).
 OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (99% over full length), another bacterium commonly detected and isolated from colorectal tumors.
-OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6C).
+OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure S2C).
 Among the 11 distal mucosa samples that were positive for _P.
@@
 In addition to identifying features that are specific to each side of the gut, the ability of the Random Forest to classify samples can serve as a proxy for similarity.
-That is, a higher AUC value means the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value.
-For instance, the model separating the proximal and distal mucosa has an AUC of 0.925 whereas the model for classifying the proximal and distal lumen has a much lower AUC of 0.575.
-Further, prior to reducing the input to five OTUs, the latter model required 44 OTUs to best separate the samples (AUC = 0.755).
+That is, a higher AUC value indicates the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value.
+For instance, the model separating the proximal and distal mucosa has an AUC of 0.850 whereas the model for classifying the proximal and distal lumen has a much lower AUC of 0.580.
+Further, the latter model required 44 OTUs to best separate the samples whereas the models separating the mucosa only needed 10 OTUs.
 The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested.
@@
 _F.
-varium_ was only detected in three of our study participants and two of those samples were isolated form the proximal mucosa (Figure 6B).
+varium_ was only detected in three of our study participants and two of those samples were isolated form the proximal mucosa (Figure S2B).
 _F.
@@
 
-\includegraphics{../submission/figure_1.pdf}
-
 Sampling strategy.
@@
 \newpage
-\includegraphics{../submission/figure_2.pdf}
 
@@
 \newpage
-\includegraphics{../submission/figure_3.pdf}
 
@@
 \newpage
-\includegraphics{../submission/figure_4.pdf}
 
@@
 \newpage
-\includegraphics{../submission/figure_5.pdf}
 
@@
 \newpage
-\includegraphics{../submission/figure_6.pdf}
-
-####Figure 6
-Location and abundance of cancer-associated OTUs.
-Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum subsp.
-animalis_ (B) _Fusobacterium varium_ and (C) _Porphyromonas asacharolytica_
-
-\newpage
-\includegraphics{../submission/figure_S1.pdf}
 
@@
 
+\newpage
+
+####Figure S2
+Location and abundance of cancer-associated OTUs.
+Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum subsp.
+animalis_ (B) _Fusobacterium varium_ and (C) _Porphyromonas asacharolytica_
+
 ### References

### 2017-10-17 submission/old_sub_docs/test.tex
text edits made, manuscript getting closer

--- 2017-07-28 submission/manuscript.tex
@@
 
+\begin{verbatim}
+## Warning in roc.default(response, m[[predictors]], ...): 'response' has
+## more than two levels. Consider setting 'levels' explicitly or using
+## 'multiclass.roc' instead
+\end{verbatim}
+
 \subsubsection{Abstract}\label{abstract}
@@
 are found in. CRC tumors of the proximal and distal colon are
-morphologically and genetically distinct. Similarly, inflammatory bowel
-diseases such as Crohn's are typically exacerbated in the proximal
-instestine while ulcerative colitis patients often experience symptoms
-in the distal colon. Previous analysis of the fecal microbiota from
-healthy and CRC or IBD patients has revealed different microbial
-signatures associated with these diseases. We extended these
-observations of the fecal microbiome to include analysis of the proximal
-and distal healthy human colon. We used a two-colonoscope approach on
-subjects that had not undergone standard bowel preparation procedure.
-This technique allowed us to characterize the native proximal and distal
-luminal and mucosal microbiome without prior chemical disruption. 16S
-rRNA gene sequencing was performed on proximal and distal mucosal and
-luminal biopsies and home-collected stool for 20 healthy individuals.
-Diversity analysis revealed that each site contained a diverse community
-and that a patient's samples were more similar to each other than to
-that of other individuals. Comparison of all samples to fecal samples
-taken at exit uncovered that the feces were most similar to samples
-taken from the distal lumen, likely reflecting the anatomical structure
-of the colon. Since we could not differentiate sites along the colon
-based on community structure or community membership alone, we employed
-the Random Forest machine-learning algorithm to identify key species
-that distinguish biogeographical sites. Random Forest classification
-models were built using taxa abundance and sample location and revealed
-distinct populations that were found in each location. A model
-differentiating the proximal mucosa and lumen was built with an AUC of
-0.831. The proximal mucosa had a higher abundance of the genera
-\emph{Enterobacteriaceae} and \emph{Bacteriodes}. \emph{Peptoniphilus,
-Anaerococcus, Finegoldia,} and \emph{Turicibacter} were most likely to
-be found in distal mucosal samples versus distal luminal samples (AUC =
-0.98). The classification model performed well (AUC = 0.912) when
-classifying mucosal samples into proximal or distal sides, but
-separating luminal samples from each side proved more challenging (AUC =
-0.755). The distal mucosa was found to have high populations of
-\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
-distal luminal samples were comprised of many of the same taxa, likely
-reflecting the fact that stool moves along the colon from the proximal
-to distal end. By sampling the unprepped human colon, our results have
-identified distinct bacterial populations native to the proximal and
-distal sides. Further investigation of these bacteria may elucidate if
-and how these groups contribute to different pathogenesis processes on
-the respective sides of the colon.
+morphologically and genetically distinct. Similarly, bowel diseases such
+as Crohn's are typically exacerbated in the proximal instestine while
+ulcerative colitis patients often experience symptoms in the distal
+colon. Previous analysis of the fecal microbiota from healthy and CRC or
+IBD patients has revealed different microbial signatures associated with
+these diseases. We extended these observations of the fecal microbiome
+to include analysis of the proximal and distal healthy human colon. We
+used a two-colonoscope approach on subjects that had not undergone
+standard bowel preparation procedure. This technique allowed us to
+characterize the native proximal and distal luminal and mucosal
+microbiome without prior chemical disruption. 16S rRNA gene sequencing
+was performed on proximal and distal mucosal and luminal biopsies and
+home-collected stool for 20 healthy individuals. Diversity analysis
+revealed that each site contained a diverse community, and that a
+patient's samples were more similar to each other than to that of other
+individuals. Comparison of all samples to fecal samples taken at exit
+uncovered that the feces were most similar to samples taken from the
+distal lumen, likely reflecting the anatomical structure of the colon.
+Since we could not differentiate sites along the colon based on
+community structure or community membership alone, we employed the
+Random Forest machine-learning algorithm to identify key species that
+distinguish biogeographical sites. Random Forest classification models
+were built using taxa abundance and sample location and revealed
+distinct populations that were found in each location.
+\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
+\emph{Actinomyces} were most likely to be found in mucosal samples
+versus luminal samples (AUC = 0.925). The classification model performed
+well (AUC = 0.912) when classifying mucosal samples into proximal or
+distal sides, but separating luminal samples from each side proved more
+challenging (AUC = 0.755). The distal mucosa was found to have high
+populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
+Proximal and distal luminal samples were comprised of many of the same
+taxa, likely reflecting the fact that stool moves along the colon from
+the proximal to distal end. By sampling the unprepped human colon, our
+results have identified distinct bacterial populations native to the
+proximal and distal sides. Further investigation of these bacteria may
+elucidate if and how these groups contribute to different pathogenesis
+processes on the respective sides of the colon.
 
@@
 mucosa to intestinal lumen, offering several sites for different
-microbial communities to flourish. The identity of these specific
-microbes and communities are important for understanding the etiology of
-complex colon diseases such as Inflammatory Bowel Disease (IBD) and
-Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
-accelerated by perturbations in gut microbes (1), (2), (3). The
-severity, symptoms, morbidity and mortality of these diseases is known
-to vary based upon the biogeographical location in which they occur. For
-instance, CRC tumors that arise on the distal side of the colon are
-infiltrating lesions that present with painful symptoms (4). In
-contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
+microbial communities to flourish.
+
+\emph{2 sentences in here are broken}
+
+The severity, symptoms, morbidity and mortality of these diseases is
+known to vary based upon the biogeographical location in which they
+occur. For instance, CRC tumors that arise on the distal side of the
+colon are infiltrating lesions that present with painful symptoms ((1)).
+In contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
 are sessile and form along the wall of the colon, often remaining
-asymptomatic until advanced carcinogenesis (4). The distal and proximal
-sides of the colon differ in the amount of inflammation present and the
-genomic instability of precancerous cells, respectively, in addition to
-variation in the previously mentioned chemical gradients (1), (5), (6).
-In IBD patients, disease flares in the distal colon are usually
-indicative of ulcerative colitis (UC), whereas Crohn's disease (CD)
-patients typically experience disease in the small intestine, ileum and
-proximal colon (2). UC presents as large and highly inflammed mucosal
-ulcers, where as CD lesions are often smaller and have areas of normal
-tissue distributed amongst the flare (2). Thus, given the varied
-physiology of the proximal-distal axis of the colon and known
+asymptomatic until advanced carcinogenesis ((1)). The distal and
+proximal sides of the colon differ in the amount of inflammation present
+and the genomic instability of precancerous cells, respectively, in
+addition to variation in the previously mentioned chemical gradients
+((2), (3), (4)). In IBD patients, disease flares in the distal colon are
+usually indicative of ulcerative colitis (UC), whereas Crohn's disease
+(CD) patients typically experience disease in the small intestine, ileum
+and proximal colon ((5)). UC presents as large and highly inflammed
+mucosal ulcers, where as CD lesions are often smaller and have areas of
+normal tissue distributed amongst the flare ((5)). Thus, given the
+varied physiology of the proximal-distal axis of the colon and known
 differences in disease patterns at these sites, symbiotic microbes and
@@
 IBD or CRC can be attributed to specific molecular events as a result of
-interactions between the gut microbiota and human host (1), (3). For
-instance, comparison of the bacteria present on CRC tumors with those
-found on nearby healthy tissue has identified specific species that are
-tumor-associated (7). These species include the oral pathogens
-\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
-Interestingly, these periodontal pathogens have been highly predictive
-of whether a patient had CRC tumors or not in our prior human stool
-classification studies (8). \emph{F. nucleatum} has also been found to
-be elevated in the stool and biopsies of patients with IBD as compared
-to healthy controls (9). Furthermore, studies of \emph{F. nucleatum}
-isolated from mucosal biopsies showed that more invasive \emph{F.
-nucleatum} positively correlates with IBD disease level (9). Like many
-intestinal pathogens, the bacteria appear to have a high-impact despite
-being lowly-abundant in the community (2). The molecular capabilities of
-these rare taxa may contribute to the colonic disease state. These
-studies often examined only shed human stool or the small intestine,
-preventing fine-resolution analysis of paired samples from the proximal
-and distal sides of the colon. Similarly, comparisons of on- or
-off-tumor/lesion bacteria rarely have matched tissue from the other side
-of the colon from the same patient, limiting what conclusions can be
-drawn about the colonic microbiome overall, let alone at that specific
-site. Due to these limitations, the contribution of the gut microbiota
-to IBD and CRC disease location in the colon is largely undefined.
-Characterizing these communities could provide needed insight into
-disease etiology, including how the disruption of the healthy community
-could promote the initiation or proliferation of the distinct proximal
-and distal CRC tumors or IBD flares.
+interactions between the gut microbiota and human host ((2)).
+
+\emph{missing last half of second paragraph here, hasnt been fully
+tested}
 
@@
 colon, but complicates microbiome profiling as the chemicals strip the
-bowel of contents and disrupt the mucosal layer (10), (11). As such,
+bowel of contents and disrupt the mucosal layer ((6), (7)). As such,
 what little information we do have about the biogeographical
@@
 and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
-in human microbiome research (12). Likewise, samples had varying levels
-of diversity at each site, irrespective of the individual (Figure 2B).
-For example, the proximal mucosa was more diverse than the distal for
-some individuals while the opposite was true for others. Therefore we
-could not identify a clear pattern of changes in microbial diversity
+in human microbiome research (cite). Likewise, samples had varying
+levels of diversity at each site, irrespective of the individual (Figure
+2B). For example, the proximal mucosa was more diverse than the distal
+for some individuals while the opposite was true for others. Therefore
+we could not identify a clear pattern of changes in microbial diversity
 along the gut axis.
@@
 determine what factors may be driving the differences seen among the
-samples, we compared \({\theta}\)YC distances between samples from all
-subjects (interpersonal) versus samples from within one subject
-(intrapersonal). We found that samples from one individual were far more
-similar to each other than to other study subjects (Figure 3C),
-consistent with previous human microbiome studies that have sampled
-multiple sites of the human colon (13), (14), (15). Thus interpersonal
-variation between subjects drives the differences between samples more
-than sample site or location. Overall, the results comparing the
-structure of the communities suggest that the contents of the distal
-lumen are most representative of stool at exit, and the microbes
-remaining on the mucosa are much different.
+samples, we compared thetaYC distances between samples from all subjects
+(interpersonal) versus samples from within one subject (intrapersonal).
+We found that samples from one individual were far more similar to each
+other than to other study subjects (Figure 3C), consistent with previous
+human microbiome studies that have sampled multiple sites of the human
+colon (\textbf{???}, (8), (9)). Thus interpersonal variation between
+subjects drives the differences between samples more than sample site or
+location. Overall, the results comparing the structure of the
+communities suggest that the contents of the distal lumen are most
+representative of stool at exit, and the microbes remaining on the
+mucosa are much different.
 
@@
 for the proximal and distal sides, independently (Figure 4A). The
-constructed model used 14 features for the proximal and 6 for the
-distal. The models performed well when classifying these samples (0.831
-and 0.98, respectively). The OTUs that were most predictive of each site
-are identified by their greatest mean decrease in accuracy when removed
-from the model. For distinguishing the proximal lumen and mucosa, OTUs
-from the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
-two OTUs from the \emph{Enterobacteraceae} genera were differentially
-abundant (Figure 4B). The model classifying the distal lumen and mucosa
-identified OTUs from \emph{Turicibacter}, \emph{Finegoldia},
-\emph{Peptoniphilus} and two OTUs from the \emph{Anaerococcus} genera
-that could distinguish lumen from mucosal samples (Figure 4C). These
-results indicate that there are fine differences between the different
-sites of the colon, and that these can be traced to specific OTUs on
-each side.
+constructed model used ((Xopt)) features for the proximal and ((Xopt))
+for the distal. The models performed well when classifying these samples
+(0.8312912 and 0.9804871, respectively). The OTUs that were most
+predictive of each site are identified by their greatest mean decrease
+in accuracy when removed from the model. For distinguishing the proximal
+lumen and mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
+\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
+were differentially abundant (Figure 4B). The model classifying the
+distal lumen and mucosa identified OTUs from \emph{Turicibacter},
+\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
+\emph{Anaerococcus} genera that could distinguish lumen from mucosal
+samples (Figure 4C). These results indicate that there are fine
+differences between the different sites of the colon, and that these can
+be traced down to specific OTUs on each side.
 
@@
 versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
-versus distal lumen (AUC = 0.755). The model that separated proximal and
-distal mucosal samples used 6 features. OTUs that were differentially
+versus distal lumen (AUC = 0.755). These models were able to explain
+((X\%)) of the variance, respectively. OTUs that were differentially
 abundant between the distal and proximal mucosa included members of the
 \emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
-\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B). The
-model used 44 features to separate proximal and distal lumenal samples.
+\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B).
 Differentially abundant OTUs of the proximal and distal lumen included
@@
 \emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
-\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
+\emph{Fusobacterium nucleatum subsp animalis} (XX\% over full length).
 This is the only species of \emph{Fusobacterium} known to have oncogenic
 properties and be found on the surfaces of colorectal cancer tumors.
-(16). There were 14 samples with the \emph{F. nucleatum subsp. animalis}
-sequence. Four sequences were located in the proximal mucosa and three
-in the distal mucosal samples (Figure 6A). The second most frequent
-\emph{Fusobacterium} sequence was OTU472, which aligns with 99\%
-identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
-\emph{F. varium} has been associated with IBD (17). 4 study participants
-harbored \emph{F. varium} and the samples were split evenly between the
-proximal and distal mucosa (Figure 6B). OTU152 was similar to the
-members of the \emph{Porphyromonas} genus and the most frequent sequence
-in that OTU aligned to \emph{Porphyromonas asacharolytica} (99\% over
-full length), another bacterium commonly detected and isolated from
-colorectal tumors. OTU152 was only detected on the distal mucosa, and in
-fact was one of the OTUs the classification model identified as
-separating distal and proximal sides (Figure 6C). Among the samples that
-were positive for \emph{P. asacharolytica}, the relative abundances for
-this OTU ranged from 0.01\% - 16\%. Thus, disease-associated OTUs could
-be found in our sample set of 20 healthy individuals.
+((10)). The \emph{Fusobacterium} positive samples were located in x\% of
+the the proximal and X\% of the distal mucosa and represented as much as
+1\% of any sample (Figure 6A). OTU152 was similar to the members of the
+\emph{Porphyromonas} genus and the most frequent sequence in that OTU
+aligned to \emph{Porphyromonas asacharolytica} (X\% over full length),
+another bacterium commonly detected and isolated from colorectal tumors.
+OTU152 was only detected on the distal mucosa, and in fact was one of
+the OTUs the classification model identified as separating distal and
+proximal sides (Figure 6B). Among the samples that were positive for
+\emph{P. asacharolytica}, the relative abundances for this OTU ranged
+from 0.01\% - 16\%. Thus, disease-associated OTUs could be found in our
+sample set of 20 healthy individuals.
 
@@
 collected during unprepared colonoscopy. We found that all locations
-contained a range of phyla abundances and a range of diversity, but that
-there was a wide variability between subjects. Pairwise comparisons of
-each of the sites revealed that the proximal mucosa and lumen were most
-similar to each other. Further, comparison of colonoscopy-collected
-samples with samples collected from stool at home showed that the distal
-lumen is most similar to stool at exit. Random Forest models built on
-OTU relative abundances from each sample identified microbes that are
+contained a range of phyla and a range of diversity, but that there was
+a wide variability between subjects. Pairwise comparisons of each of the
+sites revealed that the proximal mucosa and lumen were most similar to
+each other. Further, comparison of colonoscopy-collected samples with
+samples collected from stool at home showed that the distal lumen is
+most similar to stool at exit. Random Forest models built on OTU
+relative abundances from each sample identified microbes that are
 particular to each location of the colon. Finally, we were able to
@@
 healthy patient cohort. Using unprepped colonoscopy and machine
-learning, we have identified bacterial taxa specific to the healthy
+learning, we have identified bacterial phyla specific to the healthy
 proximal and distal human colon.
@@
 When examining the relative abundance of the different phyla at each
-site, there was a wide amount of variation with communities primarily
-dominated by the \emph{Bacteriodes} and \emph{Firmicutes}. This likely
-reflects not only the variability between human subjects, casued by
-differences in age, gender, diet, but also reports of biogeogrphical
-``patchiness'' in the gut microbiome. Several studies have noted that
-the bacteria recoverable from the same mucosal sample location can be
-vastly different when the samples are taken just 1 cm away from each
-other (18). Similar patchiness is also observed in lumenal contents and
-fecal samples themselves; there is observed separation of different
-interacting microbes along the length of a stool sample, for instance
-(19). That said, across our samples the mucosal samples harbor more
-\emph{Proteobacteria}, consistent with previous studies comparing
-mucosal swabs to lumenal content in humans (5). Hence, the conclusions
-we can draw from phyla analysis are likely impacted by patchiness
-between subjects.
+site, there was a wide amount of variation for each phyla with
+communities primarily dominated by the \emph{Bacteriodes} and
+\emph{Firmicutes}. This likely reflects not only the variability between
+human subjects, casued by differences in age, gender, diet, but also
+reports of biogeogrphical ``patchiness'' in the gut microbiome. Several
+studies have noted that the bacteria recoverable from the same mucosal
+sample location can be vastly different when the samples are taken just
+1 cm away from each other (11). Similar patchiness is also observed in
+lumenal contents and fecal samples themselves; there is observed
+separation of different interacting microbes along the length of a stool
+sample, for instance (12). That said, across our samples the mucosal
+samples harbor more \emph{Proteobacteria}, consistent with previous
+studies comparing mucosal swabs to lumenal content in humans (3). Hence,
+the conclusions we can draw from phyla analysis are likely impacted by
+patchiness between subjects.
 
@@
 Random Forest model to identify microbes specific to each side. For each
-comparison we identified top 5 OTUs that were strongly predictive of one
+comparison we identified top X OTUs that were strongly predictive of one
 site or another. Generally, OTUs identified in each location were
-consistent with known physiological gradients along the gut axis (6).
+consistent with known physiological gradients along the gut axis (4).
 For instance, the proximal mucosa contains the highest oxygen
@@
 of our subjects, respectively. These bacteria have been shown to be
-predictive of colorectal cancer in humans (8) and have oncogenic
-properties in cell culture and in mice (20). Interestingly, while
+predictive of colorectal cancer in humans (13) and have oncogenic
+properties in cell culture and in mice (14). Interestingly, while
 \emph{F. nucleatum} was found on both sides of the colon, \emph{P.
@@
 might not be surprising that it resides in the less-oygen-rich and
-proteinaceous distal mucosa (5). In studies examining bacteria on
+proteinaceous distal mucosa ((3)). In studies examining bacteria on
 colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
 on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreases along the colon to rectum (21). Of the 8 (40\%) individuals
+decreases along the colon to rectum ((15)). Of the 8 (40\%) individuals
 positive for \emph{F. nucleatum} in this study, the bacterium was spread
-across the proximal mucosa, distal lumen and distal mucosa. Data
-examining bacterial biofilms on the mucosa of CRC tumors suggests that
-\emph{Fusobacteria} species are more commonly found on proximal tumors
-and in biofilms, indicating that it is not only the presence of the
-bacteria but the organization of the tumor community that contributes to
-\emph{Fusobacterium's} role in tumorigenesis (7). Finally,
+across the proximal mucosa, distal lumen and distal mucosa. The presence
+of \emph{F. nucleatum} in a healthy individual is not necessarily linked
+to the development of future colorectal cancers (cite). Data examining
+bacterial biofilms on CRC tumors suggests that \emph{Fusobacteria}
+species are more commonly found on proximal tumors and in biofilms,
+indicating that it is not only the presence of the bacteria but the
+organization of the tumor community that contributes to
+\emph{Fusobacterium's} role in tumorigenesis ((16)). Finally,
 \emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
 not only co-occur on CRC tumors but also to synergistically promote
-tumorigenesis in an oral cancer model (22), (23). Thus, further analysis
-of the distribution and activities of these pathogens may elucidate a
-mechanism for development of IBD or CRC subtypes in the proximal or
-distal colon.
+tumorigenesis in an oral cancer model ((17), (18)). Thus, further
+analysis of the distribution and activities of these pathogens may
+elucidate a mechanism for development of IBD or CRC subtypes in the
+proximal or distal colon.
 
 The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
-been commonly isolated from mucosal biopsies of patients with IBD (17).
-Laboratory experiments with these isolates have shown that
+been commonly isolated from mucosal biopsies of patients with IBD
+(cite). Laboratory experiments with these isolates have shown that
 disease-isolated \emph{F. nucleatum} are more invasive and stimulate
-more TNF-a production than strains from healthy individuals (9),
+more TNF-a production than strains from healthy individuals ((19)),
 suggesting the bacteria may increase inflammation in the gut as well
-(24). \emph{F. varium} isolated from UC patients caused colonic ulcers
-in an experimental mouse model (25). \emph{F. varium} was only detected
-in three of our study participants and two of those samples were
-isolated form the proximal mucosa (Figure 6B). Although there is low
-occurrance in our study, \emph{F. varium} is most commonly isolated from
-UC patient biopsies from the ileum or cecum (26), suggesting this
-species may exhibit preference for the different environmental
-conditions of these gastrointestinal sites. Further work will assess how
-gut environment may select for species which may then cause localized
-disease.
+(cite - Dharmani 2011). \emph{F. varium} isolated from UC patients
+caused colonic ulcers in an experimental mouse model (Ohkusa 2003).
+\emph{F. varium} was only detected in 3 of our study participants and
+two of those samples were isolated form the proximal mucosa (cite
+figure). Although there is low occurrance in our study, \emph{F. varium}
+is most commonly isolated from UC patient biopsies from the ileum or
+cecum (cite), suggesting this species may exhibit preference for the
+different environmental conditions of these gastrointestinal sites.
+Further work will assess how gut environment may select for species
+which then cause localized disease.
 
@@
 matched-colonoscopy study sampled 18 patient's colonic mucosa and
-lumenal contents prior to and after bowel cleansing (27). This group
+lumenal contents prior to and after bowel cleansing ((20)). This group
 found that mucosal and lumenal samples were distinguishable prior to
 bowel cleansing, but that bowel preparation resulted in an increase in
-shared OTUs between each site (27). Bowel cleansing not only made the
+shared OTUs between each site ((20)). Bowel cleansing not only made the
 samples harder to distinguish, it resulted in decreases in diversity
@@
 At a baseline visit, subjects were consented and given a home collection
-stool kit (Source of kit?). At least one week prior to the scheduled
-colonoscopy, subjects were to collect whole stool at home and ship the
-samples to the University on ice. Notably, subjects did not undergo any
-bowel preparation method prior to sampling. On the procedure day,
-subjects reported to the Michigan Clinical Research Unit at the
+stool kit (Source of kit supplies). At least one week prior to the
+scheduled colonoscopy, subjects were to collect whole stool at home and
+ship the samples to the University on ice. Notably, subjects did not
+undergo any bowel preparation method prior to sampling. On the procedure
+day, subjects reported to the Michigan Clinical Research Unit at the
 University of Michigan Health System. Patients were consciously sedated
@@
 brush used to collect lumenal/stool contents. Two lumenal samples were
-collected and the contents immediately deposited into RNAlater (Fischer)
+collected and the contents immediately deposited into RNAlater (source)
 and flash-frozen in liquid nitrogen. The brushes were withdrawn and
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described (28). Sequences were curated
-using the mothur software as described previously (29). The sequences
+Illumina MiSeq as previously described ((21)). Sequences were curated
+using the mothur software as described previously ((22)). The sequences
 were assigned taxonomic classification using a naive Bayesian classifier
 trained using a 16S rRNA gene training set from the Ribosomal Database
-Project (RDP) (30) and clustered into operational taxonomic units (OTUs)
-based on a 97\% similarity cutoff. Sequencing and analysis of a mock
-community revealed the error rate to be 1.8\%. Samples were rarefied to
-4231 sequences per sample in order to reduce uneven sampling bias.
+Project (RDP) ((23)) and clustered into operational taxonomic units
+(OTUs) based on a 97\% similarity cutoff. Sequencing and analysis of a
+mock community revealed the error rate to be X\%. Samples were rarefied
+to 4231 sequences per sample in order to reduce uneven sampling bias.
 
 Diversity analysis was performed using the Simpson diversity calculator
-and thetaYC calculator metrics in mothur version 3.2 (29). ThetaYC
-distances were calculated to determine the dissimilarity between two
-samples. Random Forest classification models were built using the
-randomForest R package and resultant models were used to identify the
-OTUs that were most important for classifying each location (31). To get
-species-level information about sequences of interest, sequences were
-aligned using blastn and the species name was only used if the identity
-score was \textgreater{}= 99\%.
+and theta YC calculator metrics in mothur version
+\textless{}3.2?\textgreater{} ((22)). ThetaYC distances were calculated
+to determine the dissimilarity between two samples. Random Forest
+classification models were built using the randomForest R package and
+resultant models were used to identify the OTUs that were most important
+for classifying each location ((24)). To get species-level information
+about sequences of interest, sequences were aligned using blastn and the
+species name was only used if the identity score was \textgreater{}=
+99\%.
 
@@
 test adjusted for multiple comparisons using the Benjamini-Hochberg
-procedure. \#\#\#\# Data availability 16S rRNA gene sequence reads and
-experiment metadata are available on the NCBI Sequence Read Archive
-(SRA) with accession number XXXX. A reproducible data analysis pipeline
-can be found at
-\url{https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017}.
+procedure.
+
+\paragraph{Data availability}\label{data-availability}
+
+\paragraph{Github repository}\label{github-repository}
 
@@
 calculated and plotted by sample site for each OTU of interest: (A)
-\emph{Fusobacterium nucleatum subsp. animalis} (B) \emph{Fusobacterium
-varium} and (C) \emph{Porphyromonas asacharolytica}
+\emph{Fusobacterium nucleatum} and (B) \emph{Porphyromonas
+asacharolytica}
 
-\subsubsection*{References}\label{references}
-\addcontentsline{toc}{subsubsection}{References}
+\subsubsection{References}\label{references}
+
+\subsection{this is file from commit right before broken
+commit}\label{this-is-file-from-commit-right-before-broken-commit}
+
+\section*{testing intro}\label{testing-intro}
+\addcontentsline{toc}{section}{testing intro}
 
 \hypertarget{refs}{}
-\hypertarget{ref-Yamauchi2012}{}
-1. \textbf{Yamauchi M}, \textbf{Lochhead P}, \textbf{Morikawa T},
-\textbf{Huttenhower C}, \textbf{Chan AT}, \textbf{Giovannucci E},
-\textbf{Fuchs C}, \textbf{Ogino S}. 2012. Colorectal cancer: A tale of
-two sides or a continuum?: Figure 1. Gut \textbf{61}:794--797.
-doi:\href{https://doi.org/10.1136/gutjnl-2012-302014}{10.1136/gutjnl-2012-302014}.
-
-\hypertarget{ref-Forbes2016}{}
-2. \textbf{Forbes JD}, \textbf{Domselaar GV}, \textbf{Bernstein CN}.
-2016. The gut microbiota in immune-mediated inflammatory diseases.
-Frontiers in Microbiology \textbf{7}.
-doi:\href{https://doi.org/10.3389/fmicb.2016.01081}{10.3389/fmicb.2016.01081}.
-
-\hypertarget{ref-Halfvarson2017}{}
-3. \textbf{Halfvarson J}, \textbf{Brislawn CJ}, \textbf{Lamendella R},
-\textbf{Vazquez-Baeza Y}, \textbf{Walters WA}, \textbf{Bramer LM},
-\textbf{DAmato M}, \textbf{Bonfiglio F}, \textbf{McDonald D},
-\textbf{Gonzalez A}, \textbf{McClure EE}, \textbf{Dunklebarger MF},
-\textbf{Knight R}, \textbf{Jansson JK}. 2017. Dynamics of the human gut
-microbiome in inflammatory bowel disease. Nature Microbiology
-\textbf{2}:17004.
-doi:\href{https://doi.org/10.1038/nmicrobiol.2017.4}{10.1038/nmicrobiol.2017.4}.
-
 \hypertarget{ref-Benedix2010}{}
-4. \textbf{Benedix F}, \textbf{Kube R}, \textbf{Meyer F},
+1. \textbf{Benedix F}, \textbf{Kube R}, \textbf{Meyer F},
 \textbf{Schmidt U}, \textbf{Gastinger I}, \textbf{Lippert H}. 2010.
@@
 
+\hypertarget{ref-Yamauchi2012}{}
+2. \textbf{Yamauchi M}, \textbf{Lochhead P}, \textbf{Morikawa T},
+\textbf{Huttenhower C}, \textbf{Chan AT}, \textbf{Giovannucci E},
+\textbf{Fuchs C}, \textbf{Ogino S}. 2012. Colorectal cancer: A tale of
+two sides or a continuum?: Figure 1. Gut \textbf{61}:794--797.
+doi:\href{https://doi.org/10.1136/gutjnl-2012-302014}{10.1136/gutjnl-2012-302014}.
+
 \hypertarget{ref-Albenberg2014}{}
-5. \textbf{Albenberg L}, \textbf{Esipova TV}, \textbf{Judge CP},
+3. \textbf{Albenberg L}, \textbf{Esipova TV}, \textbf{Judge CP},
 \textbf{Bittinger K}, \textbf{Chen J}, \textbf{Laughlin A},
@@
 \hypertarget{ref-Donaldson2015}{}
-6. \textbf{Donaldson GP}, \textbf{Lee SM}, \textbf{Mazmanian SK}. 2015.
+4. \textbf{Donaldson GP}, \textbf{Lee SM}, \textbf{Mazmanian SK}. 2015.
 Gut biogeography of the bacterial microbiota. Nature Reviews
@@
 
+\hypertarget{ref-Forbes2016}{}
+5. \textbf{Forbes JD}, \textbf{Domselaar GV}, \textbf{Bernstein CN}.
+2016. The gut microbiota in immune-mediated inflammatory diseases.
+Frontiers in Microbiology \textbf{7}.
+doi:\href{https://doi.org/10.3389/fmicb.2016.01081}{10.3389/fmicb.2016.01081}.
+
+\hypertarget{ref-Jalanka2014}{}
+6. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
+\textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
+\textbf{Gowland P}, \textbf{Hoad C}, \textbf{Garsed K}, \textbf{Lam C},
+\textbf{Palva A}, \textbf{Spiller RC}, \textbf{Vos WM de}. 2014. Effects
+of bowel cleansing on the intestinal microbiota. Gut
+\textbf{64}:1562--1568.
+doi:\href{https://doi.org/10.1136/gutjnl-2014-307240}{10.1136/gutjnl-2014-307240}.
+
+\hypertarget{ref-Harrell2012}{}
+7. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
+\textbf{Young V}, \textbf{Lichtenstein L}, \textbf{Huang Y},
+\textbf{Hanauer S}, \textbf{Chang E}. 2012. Standard colonic lavage
+alters the natural state of mucosal-associated microbiota in the human
+colon. PLoS ONE \textbf{7}:e32545.
+doi:\href{https://doi.org/10.1371/journal.pone.0032545}{10.1371/journal.pone.0032545}.
+
+\hypertarget{ref-deCarcer2010}{}
+8. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
+\textbf{Kang S}, \textbf{Worthley D}, \textbf{Whitehall V},
+\textbf{Gordon I}, \textbf{McSweeney C}, \textbf{Leggett B},
+\textbf{Morrison M}. 2010. Numerical ecology validates a biogeographical
+distribution and gender-based effect on mucosa-associated bacteria along
+the human colon. The ISME Journal \textbf{5}:801--809.
+doi:\href{https://doi.org/10.1038/ismej.2010.177}{10.1038/ismej.2010.177}.
+
+\hypertarget{ref-Zhang2013}{}
+9. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
+\textbf{Xu J}, \textbf{Wen X}, \textbf{Ma Z (Sam)}, \textbf{Shi P}.
+2013. Spatial heterogeneity and co-occurrence patterns of human
+mucosal-associated intestinal microbiota. The ISME Journal
+\textbf{8}:881--893.
+doi:\href{https://doi.org/10.1038/ismej.2013.185}{10.1038/ismej.2013.185}.
+
+\hypertarget{ref-Castellarin2011}{}
+10. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
+\textbf{Dreolini L}, \textbf{Krzywinski M}, \textbf{Strauss J},
+\textbf{Barnes R}, \textbf{Watson P}, \textbf{Allen-Vercoe E},
+\textbf{Moore RA}, \textbf{Holt RA}. 2011. Fusobacterium nucleatum
+infection is prevalent in human colorectal carcinoma. Genome Research
+\textbf{22}:299--306.
+doi:\href{https://doi.org/10.1101/gr.126516.111}{10.1101/gr.126516.111}.
+
+\hypertarget{ref-Hong2011}{}
+11. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
+\textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
+analysis of the mucosal microbiota in healthy individuals reveals
+ubiquitous bacterial groups and micro-heterogeneity. PLoS ONE
+\textbf{6}:e25042.
+doi:\href{https://doi.org/10.1371/journal.pone.0025042}{10.1371/journal.pone.0025042}.
+
+\hypertarget{ref-Stearns2011}{}
+12. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
+\textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
+\textbf{Croitoru K}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}.
+2011. Bacterial biogeography of the human digestive tract. Scientific
+Reports \textbf{1}.
+doi:\href{https://doi.org/10.1038/srep00170}{10.1038/srep00170}.
+
+\hypertarget{ref-Baxter2016}{}
+13. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\hypertarget{ref-Sears2014}{}
+14. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
+and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
+doi:\href{https://doi.org/10.1016/j.chom.2014.02.007}{10.1016/j.chom.2014.02.007}.
+
+\hypertarget{ref-Mima2016}{}
+15. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
+\textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
+\textbf{Silva A da}, \textbf{Gu M}, \textbf{Li W}, \textbf{Hamada T},
+\textbf{Kosumi K}, \textbf{Hanyuda A}, \textbf{Liu L}, \textbf{Kostic
+AD}, \textbf{Giannakis M}, \textbf{Bullman S}, \textbf{Brennan CA},
+\textbf{Milner DA}, \textbf{Baba H}, \textbf{Garraway LA},
+\textbf{Meyerhardt JA}, \textbf{Garrett WS}, \textbf{Huttenhower C},
+\textbf{Meyerson M}, \textbf{Giovannucci EL}, \textbf{Fuchs CS},
+\textbf{Nishihara R}, \textbf{Ogino S}. 2016. Fusobacterium nucleatum in
+colorectal carcinoma tissue according to tumor location. Clinical and
+Translational Gastroenterology \textbf{7}:e200.
+doi:\href{https://doi.org/10.1038/ctg.2016.53}{10.1038/ctg.2016.53}.
+
 \hypertarget{ref-Dejea2014}{}
-7. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
+16. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
 \textbf{White JR}, \textbf{Welch JLM}, \textbf{Rossetti BJ},
@@
 
-\hypertarget{ref-Baxter2016}{}
-8. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+\hypertarget{ref-Whitmore2014}{}
+17. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
+cancer. PLoS Pathogens \textbf{10}:e1003933.
+doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.
+
+\hypertarget{ref-Flynn2016}{}
+18. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
+Metabolic and community synergy of oral bacteria in colorectal cancer.
+mSphere \textbf{1}:e00102--16.
+doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
 
 \hypertarget{ref-Strauss2011}{}
-9. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
+19. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
 \textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
@@
 
-\hypertarget{ref-Jalanka2014}{}
-10. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
-\textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
-\textbf{Gowland P}, \textbf{Hoad C}, \textbf{Garsed K}, \textbf{Lam C},
-\textbf{Palva A}, \textbf{Spiller RC}, \textbf{Vos WM de}. 2014. Effects
-of bowel cleansing on the intestinal microbiota. Gut
-\textbf{64}:1562--1568.
-doi:\href{https://doi.org/10.1136/gutjnl-2014-307240}{10.1136/gutjnl-2014-307240}.
-
-\hypertarget{ref-Harrell2012}{}
-11. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
-\textbf{Young V}, \textbf{Lichtenstein L}, \textbf{Huang Y},
-\textbf{Hanauer S}, \textbf{Chang E}. 2012. Standard colonic lavage
-alters the natural state of mucosal-associated microbiota in the human
-colon. PLoS ONE \textbf{7}:e32545.
-doi:\href{https://doi.org/10.1371/journal.pone.0032545}{10.1371/journal.pone.0032545}.
-
-\hypertarget{ref-LloydPrice2016}{}
-12. \textbf{Lloyd-Price J}, \textbf{Abu-Ali G}, \textbf{Huttenhower C}.
-2016. The healthy human microbiome. Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0307-y}{10.1186/s13073-016-0307-y}.
-
-\hypertarget{ref-Eckburg2005}{}
-13. \textbf{Eckburg PB}. 2005. Diversity of the human intestinal
-microbial flora. Science \textbf{308}:1635--1638.
-doi:\href{https://doi.org/10.1126/science.1110591}{10.1126/science.1110591}.
-
-\hypertarget{ref-deCarcer2010}{}
-14. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
-\textbf{Kang S}, \textbf{Worthley D}, \textbf{Whitehall V},
-\textbf{Gordon I}, \textbf{McSweeney C}, \textbf{Leggett B},
-\textbf{Morrison M}. 2010. Numerical ecology validates a biogeographical
-distribution and gender-based effect on mucosa-associated bacteria along
-the human colon. The ISME Journal \textbf{5}:801--809.
-doi:\href{https://doi.org/10.1038/ismej.2010.177}{10.1038/ismej.2010.177}.
-
-\hypertarget{ref-Zhang2013}{}
-15. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
-\textbf{Xu J}, \textbf{Wen X}, \textbf{Ma Z (Sam)}, \textbf{Shi P}.
-2013. Spatial heterogeneity and co-occurrence patterns of human
-mucosal-associated intestinal microbiota. The ISME Journal
-\textbf{8}:881--893.
-doi:\href{https://doi.org/10.1038/ismej.2013.185}{10.1038/ismej.2013.185}.
-
-\hypertarget{ref-Castellarin2011}{}
-16. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
-\textbf{Dreolini L}, \textbf{Krzywinski M}, \textbf{Strauss J},
-\textbf{Barnes R}, \textbf{Watson P}, \textbf{Allen-Vercoe E},
-\textbf{Moore RA}, \textbf{Holt RA}. 2011. Fusobacterium nucleatum
-infection is prevalent in human colorectal carcinoma. Genome Research
-\textbf{22}:299--306.
-doi:\href{https://doi.org/10.1101/gr.126516.111}{10.1101/gr.126516.111}.
-
-\hypertarget{ref-Lee2016}{}
-17. \textbf{Lee Y}, \textbf{Eun CS}, \textbf{Lee AR}, \textbf{Park CH},
-\textbf{Han DS}. 2016. FusobacteriumIsolates recovered from colonic
-biopsies of inflammatory bowel disease patients in korea. Annals of
-Laboratory Medicine \textbf{36}:387.
-doi:\href{https://doi.org/10.3343/alm.2016.36.4.387}{10.3343/alm.2016.36.4.387}.
-
-\hypertarget{ref-Hong2011}{}
-18. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
-\textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
-analysis of the mucosal microbiota in healthy individuals reveals
-ubiquitous bacterial groups and micro-heterogeneity. PLoS ONE
-\textbf{6}:e25042.
-doi:\href{https://doi.org/10.1371/journal.pone.0025042}{10.1371/journal.pone.0025042}.
-
-\hypertarget{ref-Stearns2011}{}
-19. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
-\textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
-\textbf{Croitoru K}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}.
-2011. Bacterial biogeography of the human digestive tract. Scientific
-Reports \textbf{1}.
-doi:\href{https://doi.org/10.1038/srep00170}{10.1038/srep00170}.
-
-\hypertarget{ref-Sears2014}{}
-20. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
-and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
-doi:\href{https://doi.org/10.1016/j.chom.2014.02.007}{10.1016/j.chom.2014.02.007}.
-
-\hypertarget{ref-Mima2016}{}
-21. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
-\textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
-\textbf{Silva A da}, \textbf{Gu M}, \textbf{Li W}, \textbf{Hamada T},
-\textbf{Kosumi K}, \textbf{Hanyuda A}, \textbf{Liu L}, \textbf{Kostic
-AD}, \textbf{Giannakis M}, \textbf{Bullman S}, \textbf{Brennan CA},
-\textbf{Milner DA}, \textbf{Baba H}, \textbf{Garraway LA},
-\textbf{Meyerhardt JA}, \textbf{Garrett WS}, \textbf{Huttenhower C},
-\textbf{Meyerson M}, \textbf{Giovannucci EL}, \textbf{Fuchs CS},
-\textbf{Nishihara R}, \textbf{Ogino S}. 2016. Fusobacterium nucleatum in
-colorectal carcinoma tissue according to tumor location. Clinical and
-Translational Gastroenterology \textbf{7}:e200.
-doi:\href{https://doi.org/10.1038/ctg.2016.53}{10.1038/ctg.2016.53}.
-
-\hypertarget{ref-Whitmore2014}{}
-22. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
-cancer. PLoS Pathogens \textbf{10}:e1003933.
-doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.
-
-\hypertarget{ref-Flynn2016}{}
-23. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
-Metabolic and community synergy of oral bacteria in colorectal cancer.
-mSphere \textbf{1}:e00102--16.
-doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
-
-\hypertarget{ref-Dharmani2011}{}
-24. \textbf{Dharmani P}, \textbf{Strauss J}, \textbf{Ambrose C},
-\textbf{Allen-Vercoe E}, \textbf{Chadee K}. 2011. Fusobacterium
-nucleatum infection of colonic cells stimulates MUC2 mucin and tumor
-necrosis factor alpha. Infection and Immunity \textbf{79}:2597--2607.
-doi:\href{https://doi.org/10.1128/iai.05118-11}{10.1128/iai.05118-11}.
-
-\hypertarget{ref-Ohkusa2003}{}
-25. \textbf{Ohkusa T}. 2003. Induction of experimental ulcerative
-colitis by fusobacterium varium isolated from colonic mucosa of patients
-with ulcerative colitis. Gut \textbf{52}:79--83.
-doi:\href{https://doi.org/10.1136/gut.52.1.79}{10.1136/gut.52.1.79}.
-
-\hypertarget{ref-Ohkusa2002}{}
-26. \textbf{Ohkusa T}, \textbf{Sato N}, \textbf{Ogihara T},
-\textbf{Morita K}, \textbf{Ogawa M}, \textbf{Okayasu I}. 2002.
-Fusobacterium varium localized in the colonic mucosa of patients with
-ulcerative colitis stimulates species-specific antibody. Journal of
-Gastroenterology and Hepatology \textbf{17}:849--853.
-doi:\href{https://doi.org/10.1046/j.1440-1746.2002.02834.x}{10.1046/j.1440-1746.2002.02834.x}.
-
 \hypertarget{ref-Shobar2016}{}
-27. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
+20. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
 \textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
@@
 \hypertarget{ref-Kozich2013}{}
-28. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+21. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Schloss2009}{}
-29. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+22. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Wang2007}{}
-30. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+23. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
@@
 \hypertarget{ref-Liaw2002}{}
-31. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+24. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News: The Newsletter of the R Project

### 2017-10-17 code/build_figure3.R

@@ -3,7 +3,7 @@
 
 ###########################################################
 
-pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson')
+pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson', 'cowplot')
 for (dep in pack_used){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
@@ -27,59 +27,27 @@ tyc <- subset(tyc, select = -c(row.names, X))
 tyc <- subset(tyc, pt1==pt2)
 tyc <- unite_(tyc, "match", from=c('samp1', 'samp2'), sep="_", remove = F)
 
-#now plot- this is a plot of thetayc distances in all pairings for all patients 
-ggplot(tyc, aes(x=match, y=thetayc)) + geom_point(aes(col=match)) +facet_wrap(~pt1) +theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-ggplot(tyc, aes(x=match, y=thetayc)) + geom_boxplot() +theme_bw()
-
-lefttyc <- subset(tyc, samp1== 'LB' | samp1== 'LS')
-lefttyc <- subset(lefttyc, samp2== 'LB' | samp2== 'LS' | samp2=='SS')
-
-righttyc <- subset(tyc, samp1=='RB' | samp1 == 'RS')
-righttyc <- subset(righttyc, samp2=='RB' | samp2 == 'RS' | samp2 == 'SS')
-
-mucosatyc <- subset(tyc, samp1=='RB' | samp1 == 'LB')
-mucosatyc <- subset(mucosatyc, samp2=='RB' | samp2 == 'LB')
-
 stooltyc <- subset(tyc, match=='LB_RB' | match== 'LS_RS')
-
-exittyc <- subset(tyc, samp2 == 'SS')
-exitLtyc <- subset(exittyc, samp1 == 'LS' | samp1 == 'LB')
-exitRtyc <- subset(exittyc, samp1 == 'LS' | samp1 == 'RS')
-
-
 leftandrighttyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
-
 lvsr <- rbind(stooltyc, leftandrighttyc)
 
[…]

### 2017-10-17 code/old_code/build_figure3_old.R

@@ -3,7 +3,7 @@
 
 ###########################################################
 
-pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson', 'cowplot')
+pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson')
 for (dep in pack_used){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
@@ -27,27 +27,59 @@ tyc <- subset(tyc, select = -c(row.names, X))
 tyc <- subset(tyc, pt1==pt2)
 tyc <- unite_(tyc, "match", from=c('samp1', 'samp2'), sep="_", remove = F)
 
+#now plot- this is a plot of thetayc distances in all pairings for all patients 
+ggplot(tyc, aes(x=match, y=thetayc)) + geom_point(aes(col=match)) +facet_wrap(~pt1) +theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggplot(tyc, aes(x=match, y=thetayc)) + geom_boxplot() +theme_bw()
+
+lefttyc <- subset(tyc, samp1== 'LB' | samp1== 'LS')
+lefttyc <- subset(lefttyc, samp2== 'LB' | samp2== 'LS' | samp2=='SS')
+
+righttyc <- subset(tyc, samp1=='RB' | samp1 == 'RS')
+righttyc <- subset(righttyc, samp2=='RB' | samp2 == 'RS' | samp2 == 'SS')
+
+mucosatyc <- subset(tyc, samp1=='RB' | samp1 == 'LB')
+mucosatyc <- subset(mucosatyc, samp2=='RB' | samp2 == 'LB')
+
 stooltyc <- subset(tyc, match=='LB_RB' | match== 'LS_RS')
-leftandrighttyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
-lvsr <- rbind(stooltyc, leftandrighttyc)
 
 exittyc <- subset(tyc, samp2 == 'SS')
-
-#plots
-tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
-lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
[…]

### 2017-10-17 code/old_code/rf_featurereduction.R

@@ -1,0 +1,386 @@
+####optimizing Random Forest models by standardizing input OTUs
+
+#Kaitlin Flynn, Schloss lab, updated 8 28 17
+#load packages
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret')
+for (dep in pack_used){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
+                     quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+#load in all of the files, get rel abund of >1% and/or >10%
+
+meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
+
+#make OTU abundance file
+#Create df with relative abundances
+shared_file <- subset(shared_file, select = -c(numOtus, label))
+shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+
+rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+
+
+#Create vector of OTUs with median abundances >1%
+OTUs_1 <- apply(rel_abund, 2, max) > 1
+OTU_list <- colnames(rel_abund)[OTUs_1]
+#get df of just top OTUs
+rel_abund_top <- rel_abund[, OTUs_1]
+rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+seed <- 1
+n_trees <- 2001
+
+source('code/random_functions.R')
+source('code/tax_level.R')
+
+
+#####AUCRF#####################################################################################
+###########Left mucosa (bowel) vs Right mucosa (bowel)
+#testing AUCRFcv approach on unfiltered data
[…]

### 2017-10-17 code/oncogenic_otus.R

@@ -39,7 +39,7 @@ fuso_plot <- ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(
   ylab("F. nucleatum Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=10), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
 
 #IBD fuso OTU - F. varium 
 
@@ -52,7 +52,7 @@ fusoviv_plot <- ggplot(fusoviv, aes(x=location, y=Otu00472_relAbund)) +geom_jitt
   ylab("F. varium Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
+  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=10), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
 
 
 
@@ -77,7 +77,7 @@ porphy_plot <- ggplot(p_152, aes(x=location, y=Otu152_abund)) +geom_jitter(width
   ylab("P. asacharolytica Relative Abundance") +
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
[…]

### 2017-10-20 submission/manuscript.Rmd
abstract trimmed for Gastro

@@
 csl: gastro.csl
-fontsize: 11pt
+fontsize: 12pt
 output:
@@
 
-Chemical and nutrient gradients along the human colon create microenvironments that affect the distribution and composition of the gut microbiota.
-The microbiome has been implicated in colorectal cancer (CRC) and inflammatory bowel disease (IBD).
-Further, these diseases can exhibit different findings depending on the location of the colon they are found in.
-CRC tumors of the proximal and distal colon are morphologically and genetically distinct.
-Similarly, Crohn's and Ulcerative Colitis (UC) manifest different morphological characteristics in different parts of the GI tract.
-Previous analysis of the fecal microbiota from healthy and CRC or IBD patients has revealed different microbial signatures associated with these diseases.
-We extended these observations of the fecal microbiome to include analysis of the proximal and distal healthy human colon.
-We used a two-colonoscope approach on subjects that had not undergone any bowel preparation procedure.
-In contrast, previous studies have obtained samples from colonoscopy or surgical specimens where the subjects had undergone colonic preparation.
-Our technique allowed characterization of native proximal and distal luminal and mucosal microbiome without prior chemical disruption.
-16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals.
-Diversity analysis revealed that each site contained a diverse community and that a patient’s samples were more similar to each other than to that of other individuals.
-A patient's feces were most similar to samples taken from the distal lumen, likely reflecting the anatomical structure of the colon.
-Since we could not differentiate the communities in the proximal and distal colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key taxa that the two sites in the lumen and mucosa.
-Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location.
-A model differentiating the proximal mucosa and lumen was built with an AUC of 0.856.
-The proximal mucosa had a higher abundance of the genera _Enterobacteriaceae_ and _Bacteriodes_.
-_Peptoniphilus, Anaerococcus, Finegoldia,_ and _Turicibacter_ were most likely to be found in distal mucosal samples versus distal luminal samples (AUC = 0.975).
-The classification model performed well (AUC = 0.920) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.575).
-The distal mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_.
-Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end.
-By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides.
-Further investigation of these bacteria may elucidate if and how these groups contribute to different pathogenesis processes on the respective sides of the colon.
+**Background and Aims:**
+The microbiome has been implicated in the development of colorectal cancer (CRC).
+CRC tumors of the proximal and distal colon are morphologically and genetically distinct but the microbes that reside in each location are not well described.
+Prior studies have used fecal samples from healthy and CRC patients to examine disease-associated microbes, which cannot reflect different biogeographic sites in the colon.
+Here we profile the microbes of the healthy proximal and distal mucosa and lumen.
 
-Keywords:
+**Methods:**
+We used a two-colonoscope sampling approach on 20 healthy subjects that had not undergone any bowel preparation procedure.
+16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool.
+Alpha and beta diversity metrics for each site were calculated.
+Random Forest classification models were built using taxa abundance and sample location.
+
+**Results:**
+Beta diversity analysis showed that the right mucosa and lumen were the most similar to each other.
+The distal mucosa had higher abundance of _Finegoldia, Murdochiella_, _Peptoniphilus_, _Porphyromonas_ and  _Anaerococcus_.
+The proximal mucosa had more of the genera _Enterobacteriaceae_, _Bacteroides_ and _Psuedomonas_.
+The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC = 0.8500).
+Separating proximal and distal luminal samples proved more challenging (AUC = 0.5803) and specific microbiota that differentiated the two were hard to identify.
+
+**Conclusions:**
+By sampling the unprepped human colon, we identified distinct bacterial populations native to the proximal and distal sides.
+Further investigation of these bacteria may elucidate if and how these groups contribute to different oncogenic processes on the respective sides of the colon.
+
+Words: 260/260
+
+Keywords: microbiome; colon cancer; proximal and distal colon
 

### 2017-10-20 code/random_forest.R

@@ -609,8 +609,8 @@ par(mar=c(5, 11, 1, 1))
 plot(1, type="n", ylim=c(0,length(right_otu_feat)*2), xlim=c(1e-4,3), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
 index <- 1
 for(i in right_otu_feat){
-  stripchart(at=index-0.35, jitter(rs_abunds[,i], amount=1e-5), pch=21, bg="black", method="jitter", jitter=0.2, add=T, lwd=0.5)
-  stripchart(at=index+0.35, jitter(rb_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, lwd=0.5)
+  stripchart(at=index-0.35, jitter(rs_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, lwd=0.5)
+  stripchart(at=index+0.35, jitter(rb_abunds[,i], amount=1e-5), pch=21, bg="black", method="jitter", jitter=0.2, add=T, lwd=0.5)
   segments(median(rs_abunds[,i]),index-0.7,median(rs_abunds[,i]),index, lwd=3)
   segments(median(rb_abunds[,i]),index+0.7,median(rb_abunds[,i]),index, lwd=3)
   index <- index + 2
@@ -632,8 +632,8 @@ par(mar=c(5, 11, 1, 1))
 plot(1, type="n", ylim=c(0,length(left_otu_feat)*2), xlim=c(1e-4,3), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
 index <- 1
 for(i in left_otu_feat){
-  stripchart(at=index-0.35, jitter(ls_abunds[,i], amount=1e-5), pch=21, bg="black", method="jitter", jitter=0.2, add=T, lwd=0.5)
-  stripchart(at=index+0.35, jitter(lb_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, lwd=0.5)
+  stripchart(at=index-0.35, jitter(ls_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, lwd=0.5)
[…]

### 2017-10-20 submission/manuscript.Rmd
manuscript for Pat

@@
 #### Human subjects
-The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number XXXX.
+The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number HUM00082721.
 Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study.
@@
 #### Sample collection
-At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit?).
+At a baseline visit, subjects were consented and given a home collection stool kit.
 At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and ship the samples to a research coordinator on ice.

### 2017-10-30 submission/manuscript.Rmd
10 31 pat edits made intro

@@
 ---
-title: "The microbiota of the proximal and distal human colon"
+title: "Spatial variation of the native colon microbiota in healthy adults"
 csl: gastro.csl
@@
 
-Running title: The microbiota of the proximal and distal human colon
+Running title: Spatial variation of native colon microbiota
 
@@
 ```{r dependencies, echo=FALSE, warning = FALSE, message=FALSE}
-pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson')
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret', 'RColorBrewer', 'reshape2', 'wesanderson', 'vegan', 'cowplot')
 for (dep in pack_used){
@@
 
-#make OTU abundance file
-#Create df with relative abundances
-shared_file_rf <- subset(shared_file, select = -c(numOtus, label))
-shared_meta_rf <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
-
-#create dataframe with relative abundances
-rel_abund <- 100*shared_file_rf/unique(apply(shared_file_rf, 1, sum))
-
-#Create vector of OTUs with median abundances >1%
-OTUs_1 <- apply(rel_abund, 2, max) > 1
-OTU_list <- colnames(rel_abund)[OTUs_1]
-#get df of just top OTUs
-rel_abund_top <- rel_abund[, OTUs_1]
-rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
-
-seed <- 1
-n_trees <- 2001
-
-source('../code/random_functions.R')
-source('../code/tax_level.R')
-
-#####RandomForest###########################################################################################
-#build randomForest model for each location comparison using randomize_loc function 
-rf_distal <- randomize_loc(rel_meta, "LB", "LS") #OOB 10.26%
-rf_proximal <- randomize_loc(rel_meta, "RB", "RS") #OOB 53%
-rf_bowel <- randomize_loc(rel_meta, "LB", "RB") #OOB 25.64%
-rf_lumen <- randomize_loc(rel_meta, "LS", "RS") #OOB 69.23%
-
-#and for each site
-rf_all <- randomize_site(rel_meta, "mucosa", "stool")
-
-#####AUCRF#####################################################################################
-
-# create RF model with AUCRF outputs top OTUs
-aucrf_data_distal_bs <- auc_loc(rel_meta, "LB", "LS")
-aucrf_data_LRbowel <- auc_loc(rel_meta, "LB", "RB")
-aucrf_data_proximal_bs <- auc_loc(rel_meta, "RB", "RS")
-aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
-aucrf_data_allum <- auc_site(rel_meta, "mucosa", "stool")
-
-########Cross-Validation#################################################################
-#10 fold cross validation for all lumen vs mucosa 
-iters <- 100
-cv10f_aucs <- c()
-cv10f_all_resp <- c()
-cv10f_all_pred <- c()
-for(j in 1:iters){
-  set.seed(j)
-  sampling <- sample(1:nrow(aucrf_data_allum),nrow(aucrf_data_allum),replace=F)
-  cv10f_probs <- rep(NA,78)
-  for(i in seq(1,77,7)){
-    train <- aucrf_data_allum[sampling[-(i:(i+6))],]
-    test <- aucrf_data_allum[sampling[i:(i+6)],]
-    set.seed(seed)
-    temp_model <- AUCRF(site~., data=train, pdel=0.99, ntree=500)
-    cv10f_probs[sampling[i:(i+6)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
-  }
-  cv10f_roc <- roc(aucrf_data_allum$site~cv10f_probs)
-  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
-  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data_allum$site)
-  cv10f_aucs[j] <- cv10f_roc$auc #stores aucs for all iterations, can use to calc IQR
-}
-cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
-
-#10fold CV for L lumen vs L mucosa
-iters <- 100
-cv10f_aucs <- c()
-cv10f_all_resp_distal_bs <- c()
-cv10f_all_pred_distal_bs <- c()
-for(j in 1:iters){
-  set.seed(j)
-  sampling <- sample(1:nrow(aucrf_data_distal_bs),nrow(aucrf_data_distal_bs),replace=F)
-  cv10f_probs <- rep(NA,39)
-  for(i in seq(1,36,4)){
-    train_distal_bs <- aucrf_data_distal_bs[sampling[-(i:(i+3))],]
-    test_distal_bs <- aucrf_data_distal_bs[sampling[i:(i+3)],]
-    set.seed(seed)
-    temp_model_distal_bs <- AUCRF(location~., data=train_distal_bs, pdel=0.99, ntree=500)
-    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_distal_bs$RFopt, test_distal_bs, type='prob')[,2]
-  }
-  cv10f_roc_distal_bs <- roc(aucrf_data_distal_bs$location~cv10f_probs)
-  cv10f_all_pred_distal_bs <- c(cv10f_all_pred_distal_bs, cv10f_probs)
-  cv10f_all_resp_distal_bs <- c(cv10f_all_resp_distal_bs, aucrf_data_distal_bs$location)
-  cv10f_aucs[j] <- cv10f_roc_distal_bs$auc #stores aucs for all iterations, can use to calc IQR
-}
-cv10f_roc_distal_bs <- roc(cv10f_all_resp_distal_bs~cv10f_all_pred_distal_bs)
-
-#10fold CV for R lumen vs R mucosa
-iters <- 100
-cv10f_aucs <- c()
-cv10f_all_resp_proximal_bs <- c()
-cv10f_all_pred_proximal_bs <- c()
-for(j in 1:iters){
-  set.seed(j)
-  sampling <- sample(1:nrow(aucrf_data_proximal_bs),nrow(aucrf_data_proximal_bs),replace=F)
-  cv10f_probs <- rep(NA,39)
-  for(i in seq(1,36,4)){
-    train_proximal_bs <- aucrf_data_proximal_bs[sampling[-(i:(i+3))],]
-    test_proximal_bs <- aucrf_data_proximal_bs[sampling[i:(i+3)],]
-    set.seed(seed)
-    temp_model_proximal_bs <- AUCRF(location~., data=train_proximal_bs, pdel=0.99, ntree=500)
-    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_proximal_bs$RFopt, test_proximal_bs, type='prob')[,2]
-  }
-  cv10f_roc_proximal_bs <- roc(aucrf_data_proximal_bs$location~cv10f_probs)
-  cv10f_all_pred_proximal_bs <- c(cv10f_all_pred_proximal_bs, cv10f_probs)
-  cv10f_all_resp_proximal_bs <- c(cv10f_all_resp_proximal_bs, aucrf_data_proximal_bs$location)
-  cv10f_aucs[j] <- cv10f_roc_proximal_bs$auc #stores aucs for all iterations, can use to calc IQR
-}
-cv10f_roc_proximal_bs <- roc(cv10f_all_resp_proximal_bs~cv10f_all_pred_proximal_bs)
-
-#10 fold cross validation for L vs R mucosa
-
-iters <- 100
-cv10f_aucs_muc <- c()
-cv10f_all_resp_muc <- c()
-cv10f_all_pred_muc <- c()
-for(j in 1:iters){
-  set.seed(j)
-  sampling_muc <- sample(1:nrow(aucrf_data_LRbowel),nrow(aucrf_data_LRbowel),replace=F)
-  cv10f_probs_muc <- rep(NA,39)
-  for(i in seq(1,36,4)){
-    train_muc <- aucrf_data_LRbowel[sampling_muc[-(i:(i+3))],]
-    test_muc <- aucrf_data_LRbowel[sampling_muc[i:(i+3)],]
-    set.seed(seed)
-    temp_model_muc <- AUCRF(location~., data=train_muc, pdel=0.99, ntree=500)
-    cv10f_probs_muc[sampling_muc[i:(i+3)]] <- predict(temp_model_muc$RFopt, test_muc, type='prob')[,2]
-  }
-  cv10f_roc_muc <- roc(aucrf_data_LRbowel$location~cv10f_probs_muc)
-  cv10f_all_pred_muc <- c(cv10f_all_pred_muc, cv10f_probs_muc)
-  cv10f_all_resp_muc <- c(cv10f_all_resp_muc, aucrf_data_LRbowel$location)
-  cv10f_aucs_muc[j] <- cv10f_roc_muc$auc #stores aucs for all iterations, can use to calc IQR
-}
-cv10f_roc_muc <- roc(cv10f_all_resp_muc~cv10f_all_pred_muc)
-
-#10 fold cross validation for L vs R lumen
-iters <- 100
-cv10f_aucs_lum <- c()
-cv10f_all_resp_lum <- c()
-cv10f_all_pred_lum <- c()
-for(j in 1:iters){
-  set.seed(j)
-  sampling_lum <- sample(1:nrow(aucrf_data_LRlumen),nrow(aucrf_data_LRlumen),replace=F)
-  cv10f_probs_lum <- rep(NA,39)
-  for(i in seq(1,36,4)){
-    train_lum <- aucrf_data_LRlumen[sampling_lum[-(i:(i+3))],]
-    test_lum <- aucrf_data_LRlumen[sampling_lum[i:(i+3)],]
-    set.seed(seed)
-    temp_model_lum <- AUCRF(location~., data=train_lum, pdel=0.99, ntree=500)
-    cv10f_probs_lum[sampling_lum[i:(i+3)]] <- predict(temp_model_lum$RFopt, test_lum, type='prob')[,2]
-  }
-  cv10f_roc_lum <- roc(aucrf_data_LRlumen$location~cv10f_probs_lum)
-  cv10f_all_pred_lum <- c(cv10f_all_pred_lum, cv10f_probs_lum)
-  cv10f_all_resp_lum <- c(cv10f_all_resp_lum, aucrf_data_LRlumen$location)
-  cv10f_aucs_lum[j] <- cv10f_roc_lum$auc #stores aucs for all iterations, can use to calc IQR
-}
-cv10f_roc_lum <- roc(cv10f_all_resp_lum~cv10f_all_pred_lum)
+#update this to just be importing end files off of optimization scripts to pull AUCs from
 ```
@@
 
-```{r oncogenic_otus, echo=FALSE}
-subsampled_meta <- merge(meta_file, subsample, by.x='group', by.y='row.names')
-fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00179")]
-subsampled_to = 4231
-
-fuso179[,5] <- (fuso179[,4]/subsampled_to)*100
-names(fuso179)[5] <- "Otu00179_relAbund"
-
-p_152 <-subsampled_meta[, colnames(subsampled_meta) %in% c("Group", "patient", "location", "Otu00152")]
-
-p_152[,5] <- (p_152[,4]/subsampled_to) *100
-names(p_152)[5] <- "Otu152_abund"
-
-fusoviv <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient", "location", "Otu00472")]
-fusoviv[,5] <- (fusoviv[,4]/4321)*100
-names(fusoviv)[5] <- "Otu00472_relAbund"
-```
-
 ### Abstract
 
-**Background and Aims:**
-The microbiome has been implicated in the development of colorectal cancer (CRC).
-CRC tumors of the proximal and distal colon are morphologically and genetically distinct but the microbes that reside in each location are not well described.
-Prior studies have used fecal samples from healthy and CRC patients to examine disease-associated microbes, which cannot reflect different biogeographic sites in the colon.
-Here we profile the microbes of the healthy proximal and distal mucosa and lumen.
+**Background and Aims:** The microbiome has been implicated in the development of colorectal cancer (CRC) and inflammatory bowel diseases (IBD).
+The specific traits of these diseases vary along the axis of the digestive tract.
+Further, variation in the structure of the gut microbiota has been associated with both diseases.
+Here we profiled the microbiota of the healthy proximal and distal mucosa and lumen to better understand how bacterial populations vary along the colon.
 
-**Methods:**
-We used a two-colonoscope sampling approach on 20 healthy subjects that had not undergone any bowel preparation procedure.
-16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool.
-Alpha and beta diversity metrics for each site were calculated.
-Random Forest classification models were built using taxa abundance and sample location.
+**Methods:** We used a two-colonoscope approach to sample proximal and distal mucosal and luminal contents from the colons of 20 healthy subjects that had not undergone any bowel preparation procedure.
+The biopsies and home-collected stool were subjected to 16S rRNA gene sequencing and Random Forest classification models were built using taxa abundance and location to identify microbiota specific to each site.
 
-**Results:**
-Beta diversity analysis showed that the right mucosa and lumen were the most similar to each other.
-The distal mucosa had higher abundance of _Finegoldia, Murdochiella_, _Peptoniphilus_, _Porphyromonas_ and  _Anaerococcus_.
-The proximal mucosa had more of the genera _Enterobacteriaceae_, _Bacteroides_ and _Psuedomonas_.
-The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC = 0.8500).
-Separating proximal and distal luminal samples proved more challenging (AUC = 0.5803) and specific microbiota that differentiated the two were hard to identify.
+**Results:** : The right mucosa and lumen had the most similar community structures of the five sites we considered from each subject.
+The distal mucosa had higher relative abundance of Finegoldia, Murdochiella, Peptoniphilus, Porphyromonas and Anaerococcus.
+The proximal mucosa had more of the genera Enterobacteriaceae, Bacteroides and Pseudomonas.
+The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC=0.850).
+Separating proximal and distal luminal samples proved more challenging (AUC=0.580) and specific microbiota that differentiated the two were hard to identify.
 
-**Conclusions:**
-By sampling the unprepped human colon, we identified distinct bacterial populations native to the proximal and distal sides.
-Further investigation of these bacteria may elucidate if and how these groups contribute to different oncogenic processes on the respective sides of the colon.
+**Conclusions:** By sampling the unprepped colon, we identified distinct bacterial populations native to the proximal and distal sides.
+Further investigation of these bacteria may elucidate if and how these groups contribute to different disease processes on their respective sides of the colon.
 
-Words: 260/260
+Words: 259/260
 
-Keywords: microbiome; colon cancer; proximal and distal colon
+**Keywords**: microbiome; colon cancer; proximal and distal colon
 
@@
 
-The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome.
-Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location.
-Microenvironments differ not only longitudinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
-The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC).
-IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes [@Yamauchi2012; @Forbes2016; @Halfvarson2017].
-The manifestations of these diseases are known to vary based upon the biogeographical location in which they occur.
-For instance, CRC that arise in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012.
-In contrast, CRCs arising in the proximal (right) colon tumors are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012.
-The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients [@Yamauchi2012; @Albenberg2014; @Donaldson2015].
+The human colon is an ecosystem comprised of numerous microenvironments that select for different microbiota.
+Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence which microbiota reside in each location.
+Microenvironments differ not only longitudinally along the colon, but also radially from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
+The identity of these specific microbiota and communities are important for understanding the etiology of complex colon diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD).
+CRC and IBD can be preceded or accelerated by perturbations the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017].
+The manifestations of these diseases are known to vary based upon the location in which they occur.
+For instance, CRC that arises in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012.
+In contrast, CRC arising in the proximal (right) colon tumors are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012.
+In addition to the environmental gradients within the colon, the distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively [@Yamauchi2012; @Albenberg2014; @Donaldson2015].
 In IBD patients, disease occurring in the distal colon extending proximally is usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur anywhere along the GI tract, most commonly in the ileum and the cecum @Forbes2016.
 UC presents as continuous disease with only mucosal involvement, where as CD has skip lesions and full thickness involvement that may cause abscesses, strictures and fistulas @Forbes2016.
-Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC.
-Because CRC can be a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.
+Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbiota and their metabolites likely vary as well, and may influence the heterogeneous disease prognoses of IBD and CRC.
+Because CRC can be a long-term complication of IBD, the distribution of microbiota is important to understanding the pathophysiology of both diseases.
 
@@
 For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated @Dejea2014.
+Specific bacteria have also been identified in fecal samples of patients with varying stages of colon tumorigenesis @Baxter2016.
 These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
-Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our prior human stool classification studies @Baxter2016.
 _F.
@@
 These studies often examined only shed human stool or the small intestine, preventing fine-resolution analysis of paired samples from the proximal and distal sides of the colon.
-Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
-Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location in the colon is largely undefined.
-Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.
+Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same, disease-baring patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
+Due to these limitations, the contribution of the gut microbiota to CRC and IBD disease location in the colon is largely undefined.
+Characterizing these communities in healthy individuals could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.
 
-The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods.
+The few existing profiles of the microbial spatial variation of the colon have been limited by sample collection methods.
 The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy or surgery.
@@
 Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer [@Jalanka2014; @Harrell2012].
-As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
+As such, what little information we do have about the spatial distribution of the microbiota in the proximal and distal colon is confounded by the bowel preparation procedure.
 
-Here we address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon.
+Here we address the limitations of previous studies and identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon.
 We used an unprepared colonoscopy technique to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers.
-To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location.
-We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes.
-By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.
+To address the inherent inter-individual variation in microbiota, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify the microbiota that were specific to each location.
+We found that our classification models were able to separate mucosal and luminal samples as well as differentiate between sides of the colon based on populations of particular microbiota.
+By identifying the distinguishing microbiota we are poised to ask if and how the presence or disruption of the microbiota at each site contribute to the development of the tumor subtypes of CRC in the proximal and distal human colon.
 

### 2017-10-30 code/build_figure3.R

@@ -3,7 +3,7 @@
 
 ###########################################################
 
-pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson', 'cowplot')
+pack_used <- c('ggplot2','dplyr', 'tidyr', 'RColorBrewer', 'reshape2', 'wesanderson', 'cowplot', 'vegan')
 for (dep in pack_used){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 

### 2017-10-30 submission/manuscript.Rmd
10 31 pat edits results

@@
 #### Microbial membership and diversity of the proximal and distal colon 
-Lumenal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1).
-Participants also collected stool at home one week prior to the procedure.
-To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample.
+Luminal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1).
+Subjects also collected stool at home one week prior to the procedure.
+To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on DNA extracted from each sample.
 As expected, each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A) @LloydPrice2016.
@@
 
-To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated distances from Operational Taxonomic Unit (OTU) abundances and compared these distances for all individuals.
-Again, across all patient samples we observed a range of distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis.
-However, when comparing median distances between the proximal lumen and mucosa, the proximal versus distal lumen, the proximal versus distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).
+To compare similarity between the proximal and distal sides and within the lumen and mucosa, we compared the community structure of these sites based on the relative abundances of individual Operational Taxonomic Units (OTUs).
+Across all subjects we observed wide variation when comparing sample locations (Figure 3A).
+Those ranges did not follow a clear pattern on an individual basis.
+However, when comparing median dissimilarity between the communities found in the proximal lumen and mucosa, the proximal and distal lumen, the proximal and distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than to the other samples (P < 0.005, Wilcoxon, BH adjustment).
 
-#### Fecal samples resemble lumenal samples from the distal colon 
-Next, we calculated distances to examine how each sample compared to the home-collected feces.
-Amidst variability between patients, we did identify significantly smaller distance between the distal lumenal sample and the feces (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
-Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the feces, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
-To determine what factors may be driving the differences seen among the samples, we compared distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon [@Eckburg2005; @deCarcer2010; @Zhang2013].
-Thus interpersonal variation between subjects drove the differences between samples more than sample site or location.
-Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of the patient's feces, and the microbes remaining on the mucosa are more distinct.
+#### Fecal samples resemble luminal samples from the distal colon 
+Next, we compared the luminal and mucosal samples to the fecal sample of each subject.
+Amidst the large inter-subject variation, we did identify significantly less dissimilarity between the distal luminal sample and the feces (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
+Furthermore, there was an even larger difference in the communities found in the distal mucosa compared to the fecal communities, indicating that the mucosa is as different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
+These results suggest that the contents of the distal lumen were most representative of the subjects' feces, and the mucosal microbiota are distinct from the fecal and luminal communities.
+
+#### Interpersonal community variation is greater than the variation between sites
+To determine what factors may have driven the differences seen among the samples, we compared the community dissimilarity between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
+We found that samples from one individual were far more similar to each other than to matched samples from the other subjects (Figure 3C); this is consistent with previous human microbiome studies that have sampled multiple sites of the human colon [@Eckburg2005; @deCarcer2010; @Zhang2013].
+Thus interpersonal variation drove the differences between samples more than whether the sample came from the proximal or distal side of the colon or from the lumen or mucosa.
 
 #### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
-To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances.
-We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure S1A).
+To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances.
+We used 10-fold cross validation to build the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Figure S1A).
 The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908).
 The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
-For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4A).
-The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4B).
+For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ families were included in the best model (Figure 4A).
+The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ families (Figure 4B).
 These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.
 
-Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation.
+Next, we built a Random Forest model to differentiate the proximal and distal lumenal samples using 10-fold cross validation.
 The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) compared to the proximal versus distal lumen (AUC = 0.580).

### 2017-10-31 submission/manuscript.Rmd
almost all of pat edits made

@@
 The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
-For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ families were included in the best model (Figure 4A).
-The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ families (Figure 4B).
+For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ were included in the best model (Figure 4A).
+The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ (Figure 4B).
 These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.
 
-Next, we built a Random Forest model to differentiate the proximal and distal lumenal samples using 10-fold cross validation.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) compared to the proximal versus distal lumen (AUC = 0.580).
-OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5A).
-Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5B).
-This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.
+Next, we built a Random Forest model to differentiate the proximal and distal luminal samples using 10-fold cross validation.
+The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = 0.580).
+The OTUs included in the model differentiating the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ (Figure 5A).
+The model that attempted to separate the the proximal and distal lumen included OTUs affiliated with the _Bacteroides_, _Clostridium IV_ and _Oscillibacter_ (Figure 5B).
+Interestingly, _Anaerococcus_ and _Finegoldia_ were distinct between the mucosa and lumen and also helped to differentiate between the proximal and distal sides.
 
-#### Bacterial OTUs associated with cancer are found in healthy individuals 
+#### Bacterial OTUs associated with CRC and IBD are found in healthy individuals 
 Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs.
-Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (100% over full length).
-This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors.
-@Castellarin2011.
-There were 14 patient samples with the _F.
+Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligned via blastn to _Fusobacterium nucleatum subsp animalis_ (100% over full length).
+This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors @Castellarin2011.
+There were 14 samples from 8 subjects with the _F.
 nucleatum subsp.
 animalis_ sequences.
-Of the samples with the highest abundance of _F.
+Of the samples with the highest relative abundance of _F.
 nucleatum subsp.
@@
 varium_ has been associated with IBD @Lee2016.
-Four study participants harbored _F.
+Four subjects harbored _F.
 varium_ and the samples were split evenly between the proximal and distal mucosa (Figure S2B).
@@
 Among the 11 distal mucosa samples that were positive for _P.
-asacharolytica_, the relative abundances for this OTU ranged from 0.01% - 16%.
+asacharolytica_, the relative abundances for this OTU ranged from 0.01% to 16%.
 Thus, disease-associated OTUs could be found in our sample set of 20 healthy individuals.
@@
 ### Discussion
-Here we identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon from samples collected during unprepared colonoscopy.
-We found that all locations contained a range of phyla abundances and a range of diversity, but that there was a wide variability between subjects.
+Here we identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon using samples collected during an unprepared colonoscopy of healthy subjects.
+We found that all locations contained a range of phyla relative abundances and a range of diversity, but that there was a wide variability between subjects.
 Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other.
-Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the distal lumen was most similar to feces.
-Random Forest models built on OTU relative abundances from each sample identified microbes that were particular to each location of the colon.
-Finally, we were able to detect some bacterial OTUs associated with colonic disease in our healthy patient cohort.
-Using unprepped colonoscopy and machine learning, we have identified bacterial taxa specific to the healthy proximal and distal human colon.
+Further, comparison of colonoscopy-collected samples with fecal samples demonstrated that the distal lumen was most similar to feces.
+Random Forest models built using OTU relative abundances from each sample identified microbiota that were particular to each location of the colon.
+Finally, we were able to detect some bacterial OTUs associated with colonic disease in our healthy cohort.
+Using unprepped colonoscopies and machine learning, we have identified bacterial taxa specific to the healthy proximal and distal human colon.
 
@@
 _Bacteriodes_ and _Firmicutes_), there was a wide amount of variation.
-This likely reflects not only the variability between human subjects, casued by differences in age, sex, diet, but also biogeogrphical "patchiness" in the gut microbiome.
-Several studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011.
-Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011.
-That said, across our samples the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to lumenal content in humans @Albenberg2014.
-Hence, the conclusions we can draw from phyla analysis may be impacted by patchiness between subjects.
+This likely reflects not only the variability between human subjects, casued by differences in age, sex, diet, but also spatial "patchiness" in the gut microbiome.
+One study noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011.
+Similar patchiness was also observed in luminal contents and fecal samples themselves; there was separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011.
+That said, across our samples, the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to luminal content in humans @Albenberg2014.
+Hence, the conclusions we were able to draw from phyla analysis may have been impacted by inter-subject patchiness.
 
-To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side.
+To get around the noisiness from a diverse set of samples, we built Random Forest classification models to identify the microbiota that were specific to each side and in the lumen and mucosa.
 For each comparison we identified the top five OTUs that were strongly predictive of one site or another.
 Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015.
-For instance, the proximal mucosa contains the highest oxygen concentratons of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_.
+For instance, the proximal mucosa contains the highest oxygen concentrations of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_.
 The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_.
-Thus the gut microenvironment of each location likely enriches for these specific microbes.
+Thus the gut microenvironment of each location likely enriches for these specific microbiota.
 
@@
 That is, a higher AUC value indicates the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value.
-For instance, the model separating the proximal and distal mucosa has an AUC of 0.850 whereas the model for classifying the proximal and distal lumen has a much lower AUC of 0.580.
+For instance, the model separating the proximal and distal mucosa had an AUC of 0.850 whereas the model for classifying the proximal and distal lumen had a much lower AUC of 0.580.
 Further, the latter model required 44 OTUs to best separate the samples whereas the models separating the mucosa only needed 10 OTUs.
 The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested.
-We speculate that the model was least effective at classifying the proximal and distal lumenal contents because the samples are arguably composed of the same bacteria but differ in water content.
+We speculate that the model was less effective at classifying the proximal and distal luminal contents because the mucosal microenvironments have variable selective pressure along the colon than the luminal microenvironments.
 
@@
 Not much is known about the distribution of _P.
-asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa @Albenberg2014.
+asacharolytica_ along the colon, but given its anaerobic lifestyle and asacharolytic metabolism, it is not surprising that it resides in the less-oygen-rich and protein-rich distal mucosa @Albenberg2014.
 In studies examining bacteria on colorectal cancer tumors, _F.
@@
 nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa.
-Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014.
-Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model @Whitmore2014 @Flynn2016.
-Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.
+Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacterium but the structure of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014.
+Finally, _Fusobacterium_ and _Porphyromonas_ populations not only co-occur on CRC tumors but also to synergize to promote tumorigenesis in an oral cancer model @Whitmore2014 @Flynn2016.
+Further analysis of the distribution and activities of these pathogens along the colon is needed to elucidate a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.
 
@@
 _F.
-varium_ was only detected in three of our study participants and two of those samples were isolated form the proximal mucosa (Figure S2B).
+varium_ was only detected in three subjects and two of those samples were isolated form the proximal mucosa (Figure S2B).
 _F.
@@
 
-Specific comparisons of our findings to previously published gut biogeography studies are confounded by the use of bowel preparation methods in most other studies.
-A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing @Shobar2016.
-This study found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016.
-After seven days, bowel cleansing not only made the samples harder to distinguish, but it also decreased in diversity across sites.
-Bowel preparation clearly biases into the microbes recovered from sampling the lumen or mucosa of a prepared bowel.
+Specific comparisons of our findings to previously published studies of spatial variation are confounded by the use of bowel preparation methods.
+A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and luminal contents prior to and after bowel cleansing @Shobar2016.
+This study found that mucosal and luminal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016.
+After seven days, bowel cleansing not only made the samples more difficult to distinguish, but it also decreased the diversity observed across sites.
+Bowel preparation clearly biases the representation of microbiota recovered from sampling the lumen or mucosa.
 
-By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes.
-To this point, 16S rRNA gene sequecing community profiling studies do not provide enough information to fully probe these questions.
-In particular, 16S sequencing cannot not profile the host characteristics at each site.
-Combining the unprepared colonoscopy approach with analysis of multi-omic sequencing data may be useful in further characterizing host-microbiome interactions along the gut axis for both health and disease.
+By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypotheses about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes.
+A better understanding of microbial activities in the gut can enhance microbiome-based screening and treatment modalities for these colon diseases.
 
@@
 #### Sample collection
-At a baseline visit, subjects were consented and given a home collection stool kit.
-At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and ship the samples to a research coordinator on ice.
+At a baseline visit, subjects gave consent and were given a home collection stool kit (Zymo).
+At least one week prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice.
 Notably, subjects did not undergo any bowel preparation method prior to sampling.
 On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System.
-Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate.
-A flexible sigmoidoscope was first inserted about 25cm into the colon and endoscopy brush used to collect lumenal/stool contents.
-Two lumenal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen.
-The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
+Subjects were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate.
+A flexible sigmoidoscope was first inserted about 25cm into the colon and jumbo biopsy forceps used to collect the luminal contents.
+Two luminal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen.
+The forceps were withdrawn and new biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
 Three mucosal biopsies were collected and flash-frozen in RNAlater.
-These samples comprised the distal or distal colon samples.
-The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the ascending colon.
-Samples were then collected as in the distal colon and the colonoscope withdrawn.
+These samples comprised the distal colon samples.
+The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the proximal colon.
+Samples were then collected in the same manner as was done in the distal colon and the colonoscope withdrawn.
 All samples were stored at -80$^\circ$C.
@@
 Sequences were curated using the mothur software as described previously @Schloss2009.
-The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
+The sequences were assigned a taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
 Sequencing and analysis of a mock community revealed the error rate to be 0.018%.
@@
 Diversity analysis was performed using the Simpson diversity calculator and $\theta$<sub>YC</sub> calculator metrics in mothur version 1.39.5 @Schloss2009.
-ThetaYC distances were calculated to determine the dissimilarity between two samples.
-Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location @Liaw2002.
-To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.
+$\theta$<sub>YC</sub> distances were calculated to determine the dissimilarity between two samples.
+Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011.
+For each model the data was split into a 19-subject training set and a 1-subject test set.
+The model was built and cross-validated using AUCRFcv on the training set.
+The model was then tested on the left-out patient.
+This process was repeated iteratively for all subjects and results plotted as Reciever Operator Characteristic curves using the pROC R package @Robin_2011.
+Resultant models were used to identify the OTUs that were most important for classifying each location.
+Species-level information for sequences of interest was obtained by aligning the sequences to the GenBank nucleotide databse using blastn.
+The species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.
 

### 2017-10-31 submission/manuscript.Rmd
all Pat text edits made

@@
 A flexible sigmoidoscope was used to sample the distal colonic luminal contents and mucosa.
-The scope was inserted ~ 25cm into the subject and endoscopy brushes were used to sample the luminal contents (green star).
-A separate set of biopsy forceps was used to sample the distal mucosa (blue star).
+The scope was inserted ~ 25cm into the subject and biopsy forceps were used to sample the luminal contents (D, inset).
+A separate set of biopsy forceps was used to sample the distal mucosa (D, inset).
 The sigmoidoscope was removed.
-A pediatric colonoscope was inserted and used to access the proximal colon.
+A pediatric colonoscope was inserted and used to access the proximal colon (P, inset).
 Biopsies were taken of the proximal luminal contents and mucosa as described.
@@
 ####Figure 2
-Microbial membership and diversity of the proximal and distal human colon.
+Phylum-level relative abundance and diversity in the proximal and distal human colon.
 A) Relative abundance of the top five bacterial phyla in each sampling site.
-Each box represents the median and confidence intervals.
+Each box represents the median and interquartile range.
 B) Simpson diversity of the microbial communities at each location.
-The lines represent the median values.
+The horizontal lines represent the median values.
 
@@
 ####Figure 3
-Distances of microbial community structure between sites of the gut.
-ThetaYC distances are shown for interpersonal similarities between two sites – each point represents one individual.
+Comparison of microbial community structure between sites of the gut.
+$\theta$<sub>YC</sub> distances are shown to indicate the interpersonal dissimilarities between two sites – each point represents one individual.
 In (A), comparisons of the proximal and distal mucosal and lumen are shown.
 In (B), comparisons of each site to the exit stool are shown.
+In (C), comparisons of samples from all subjects to each other (interpersonal) or within one subject (intrapersonal) are shown.
 
@@
 Top five OTUs that are most important for the classification model for the distal mucosa and lumen (A) and the proximal mucosa and lumen (B).
+The vertical lines represent the median values for each OTU.
 
@@
 Taxa specific to the distal and proximal mucosa and lumen.
-Top five OTUs that are most important for the classification model for the distal and proximal mucosa (A) and the distal and proximal lumen (B).
-
+The five OTUs that were most important differentiating the distal and proximal mucosa (A) and the distal and proximal lumen (B).
+The vertical lines represent the median values for each OTU.
 \newpage
@@
 Random Forest classifies locations in the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal lumen versus proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
-(B) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal (red) and proximal (blue) sides of the colon.
+A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal (red) and proximal (blue) sides of the colon.
+(B) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal mucosa vs proximal mucosa (green) and distal lumen versus proximal lumen (purple).
 
@@
 ####Figure S2
-Location and abundance of cancer-associated OTUs.
+Location and relative abundance of cancer-associated OTUs.
 Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum subsp.

### 2017-10-31 submission/manuscript.Rmd
updated manuscript EOD tuesday

@@
 title: "Spatial variation of the native colon microbiota in healthy adults"
-csl: gastro.csl
+csl: aacr.csl
 fontsize: 12pt
 output:
-  word_document: default
   pdf_document:
@@
   html_document: default
+  word_document: default
 header-includes:
@@
 ---
-\vspace{35mm}
+\vspace{15mm}
 
@@
 
-\vspace{35mm}
+\vspace{15mm}
 
@@
 
-2\. Pennsylvania State University, Hershey, Pennsylvania ??
+2\. Department of Family and Community Medicine, College of Medicine, Pennsylvania State University, Hershey, Pennsylvania 17033
 
 3\. Department of Internal Medicine, Division of Gastroenterology, University of Michigan Medical School, Ann Arbor, Michigan
+
+Grant support: Rose and Lawrence C. Page Foundation
+
+Disclosures: nothing to disclose
+
+Word count:
+
+Figures:
+
+Supplementary figures:
 
@@
 ```
-```{r important_otus, echo=FALSE}
-#just LB vs LS 
-distal_otu_feat <- colnames(aucrf_data_distal_bs[2:6])
-otu_taxa_distal <- get_tax(1, distal_otu_feat, tax_file)
-#Abundance stripchart or most predictive otus
-ls_abunds <- shared_meta[shared_meta$location=='LS', distal_otu_feat]/10000 + 1e-4
-lb_abunds <- shared_meta[shared_meta$location=='LB', distal_otu_feat]/10000 + 1e-4
-
-proximal_otu_feat <- colnames(aucrf_data_proximal_bs[2:6])
-otu_taxa_proximal <- get_tax(1, proximal_otu_feat, tax_file)
-#Abundance stripchart or most predictive otus
-rs_abunds <- shared_meta[shared_meta$location=='RS', proximal_otu_feat]/10000 + 1e-4
-rb_abunds <- shared_meta[shared_meta$location=='RB', proximal_otu_feat]/10000 + 1e-4
-
-#Lb vs Rb
-LRbowel_otu_feat <- colnames(aucrf_data_LRbowel[2:6])
-otu_taxa_LRbowel <- get_tax(1, LRbowel_otu_feat, tax_file)
-#Abundance stripchart or most predictive otus 
-lb_abunds <- shared_meta[shared_meta$location=='LB', LRbowel_otu_feat]/10000 + 1e-4
-rblb_abunds <- shared_meta[shared_meta$location=='RB', LRbowel_otu_feat]/10000 + 1e-4
-
-#LS vs RS
-LRlumen_otu_feat <- colnames(aucrf_data_LRlumen[2:6])
-otu_taxa_LRlumen <- get_tax(1, LRlumen_otu_feat, tax_file)
-#Abundance stripchart or most predictive otus 
-lsrs_abunds <- shared_meta[shared_meta$location=='LS', LRlumen_otu_feat]/10000 + 1e-4
-rsls_abunds <- shared_meta[shared_meta$location=='RS', LRlumen_otu_feat]/10000 + 1e-4
-
-```
 
@@
 
-Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host [@Yamauchi2012; @Halfvarson2017].
+Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host [@Yamauchi2012; @Kostic_2013; @Halfvarson2017].
 For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated @Dejea2014.
-Specific bacteria have also been identified in fecal samples of patients with varying stages of colon tumorigenesis @Baxter2016.
+Specific bacteria have also been identified in fecal samples of patients with varying stages of colon tumorigenesis [@McCoy_2013; @Baxter2016].
 These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_.
 _F.
-nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls @Strauss2011.
+nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls [@Strauss2011; @Brennan_2016].
 Furthermore, studies of _F.
@@
 In studies examining bacteria on colorectal cancer tumors, _F.
-nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F.
-nucleatum_ decreases along the colon to rectum @Mima2016.
+nucleatum_ was more commonly detected on proximal-sided tumors, and distribution of _F.
+nucleatum_ decreased along the colon to rectum @Mima2016.
+In another study, _Fusobacterium_ was associated with MSI with BRAF and KRAS mutations, molecular features of proximal CRC @Tahara_2014.
 Of the 8 (40%) individuals positive for _F.
-nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa.
+nucleatum_ in the present study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa.
 Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacterium but the structure of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014.

### 2017-10-31 submission/manuscript.tex
updated manuscript EOD tuesday

@@
-\documentclass[11pt,]{article}
+\documentclass[12pt,]{article}
 \usepackage{lmodern}
@@
 \hypersetup{unicode=true,
-            pdftitle={The microbiota of the proximal and distal human colon},
+            pdftitle={Spatial variation of the native colon microbiota in healthy adults},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{The microbiota of the proximal and distal human colon}
+  \title{Spatial variation of the native colon microbiota in healthy adults}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-\vspace{35mm}
+\vspace{15mm}
 
-Running title: The microbiome of the proximal and distal human colon
+Running title: Spatial variation of native colon microbiota
 
-\vspace{35mm}
+\vspace{15mm}
 
 Kaitlin J. Flynn\textsuperscript{1}, Charles C.
-Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2},
-Danielle Kimberly Turgeon\textsuperscript{3\(\dagger\)}, and Patrick D.
+Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2}, D.
+Kim Turgeon\textsuperscript{3\(\dagger\)}, and Patrick D.
 Schloss\textsuperscript{1\(\dagger\)}
 
-\vspace{35mm}
+\vspace{15mm}
 
 \(\dagger\) Corresponding authors:
-\href{mailto:kturgeon@med.umich.edu}{\nolinkurl{kturgeon@med.umich.edu}}
-and \href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+\href{mailto:kturgeon@umich.edu}{\nolinkurl{kturgeon@umich.edu}} and
+\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
 
@@
 
-2. Pennslyvania State University, Hershey, Pennslyvania ??
+2. Pennsylvania State University, Hershey, Pennsylvania ??
 
@@
 
-Chemical and nutrient gradients along the human colon create
-microenvironments that affect the distrubution and composition of the
-gut microbiota. The microbiome has been implicated in colorectal cancer
-(CRC) and inflammatory bowel disease (IBD). Further, these diseases
-exhibit different symptoms depending on the location of the colon they
-are found in. CRC tumors of the proximal and distal colon are
-morphologically and genetically distinct. Similarly, inflammatory bowel
-diseases such as Crohn's are typically exacerbated in the proximal
-instestine while ulcerative colitis patients often experience symptoms
-in the distal colon. Previous analysis of the fecal microbiota from
-healthy and CRC or IBD patients has revealed different microbial
-signatures associated with these diseases. We extended these
-observations of the fecal microbiome to include analysis of the proximal
-and distal healthy human colon. We used a two-colonoscope approach on
-subjects that had not undergone standard bowel preparation procedure. In
-contrast to previous efforts using prepared colonoscopy, this technique
-allowed us to characterize the native proximal and distal luminal and
-mucosal microbiome without prior chemical disruption. 16S rRNA gene
-sequencing was performed on proximal and distal mucosal and luminal
-biopsies and home-collected stool for 20 healthy individuals. Diversity
-analysis revealed that each site contained a diverse community and that
-a patient's samples were more similar to each other than to that of
-other individuals. A patient's feces were most similar to samples taken
-from the distal lumen, likely reflecting the anatomical structure of the
-colon. Since we could not differentiate the communities in the proximal
-and distal colon based on community structure or community membership
-alone, we employed the Random Forest machine-learning algorithm to
-identify key taxa that the two sites in the lumen and mucosa. Random
-Forest classification models were built using taxa abundance and sample
-location and revealed distinct populations that were found in each
-location. A model differentiating the proximal mucosa and lumen was
-built with an AUC of 0.856. The proximal mucosa had a higher abundance
-of the genera \emph{Enterobacteriaceae} and \emph{Bacteriodes}.
-\emph{Peptoniphilus, Anaerococcus, Finegoldia,} and \emph{Turicibacter}
-were most likely to be found in distal mucosal samples versus distal
-luminal samples (AUC = 0.975). The classification model performed well
-(AUC = 0.920) when classifying mucosal samples into proximal or distal
-sides, but separating luminal samples from each side proved more
-challenging (AUC = 0.575). The distal mucosa was found to have high
-populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
-Proximal and distal luminal samples were comprised of many of the same
-taxa, likely reflecting the fact that stool moves along the colon from
-the proximal to distal end. By sampling the unprepped human colon, our
-results have identified distinct bacterial populations native to the
-proximal and distal sides. Further investigation of these bacteria may
-elucidate if and how these groups contribute to different pathogenesis
-processes on the respective sides of the colon.
+\textbf{Background and Aims:} The microbiome has been implicated in the
+development of colorectal cancer (CRC) and inflammatory bowel diseases
+(IBD). The specific traits of these diseases vary along the axis of the
+digestive tract. Further, variation in the structure of the gut
+microbiota has been associated with both diseases. Here we profiled the
+microbiota of the healthy proximal and distal mucosa and lumen to better
+understand how bacterial populations vary along the colon.
 
-\paragraph{wordcount?}\label{wordcount}
+\textbf{Methods:} We used a two-colonoscope approach to sample proximal
+and distal mucosal and luminal contents from the colons of 20 healthy
+subjects that had not undergone any bowel preparation procedure. The
+biopsies and home-collected stool were subjected to 16S rRNA gene
+sequencing and Random Forest classification models were built using taxa
+abundance and location to identify microbiota specific to each site.
+
+\textbf{Results:} : The right mucosa and lumen had the most similar
+community structures of the five sites we considered from each subject.
+The distal mucosa had higher relative abundance of Finegoldia,
+Murdochiella, Peptoniphilus, Porphyromonas and Anaerococcus. The
+proximal mucosa had more of the genera Enterobacteriaceae, Bacteroides
+and Pseudomonas. The classification model performed well when
+classifying mucosal samples into proximal or distal sides (AUC=0.850).
+Separating proximal and distal luminal samples proved more challenging
+(AUC=0.580) and specific microbiota that differentiated the two were
+hard to identify.
+
+\textbf{Conclusions:} By sampling the unprepped colon, we identified
+distinct bacterial populations native to the proximal and distal sides.
+Further investigation of these bacteria may elucidate if and how these
+groups contribute to different disease processes on their respective
+sides of the colon.
+
+Words: 259/260
+
+\textbf{Keywords}: microbiome; colon cancer; proximal and distal colon
 
@@
 
-The human colon is an ecosystem comprised of several different
-microenvironments inhabited by resident bacterial members of the
-microbiome. Concentrations of oxygen, water and anti-microbial peptides
-change along the gut axis and influence what populations of microbes
-reside in each location. Microenvironments differ not only
-longitutinally along the colon, but latitudinally from the epithelium to
-mucosa to intestinal lumen, offering several sites for different
-microbial communities to flourish. The identity of these specific
-microbes and communities are important for understanding the etiology of
-complex colon diseases such as Inflammatory Bowel Disease (IBD) and
-Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
-accelerated by perturbations in gut microbes (1--3). The severity,
-symptoms, morbidity and mortality of these diseases is known to vary
-based upon the biogeographical location in which they occur. For
-instance, CRC tumors that arise on the distal side of the colon are
-infiltrating lesions that present with painful symptoms (4). In
-contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
-are sessile and form along the wall of the colon, often remaining
-asymptomatic until advanced carcinogenesis (4). The distal and proximal
-sides of the colon differ in the amount of inflammation present and the
-genomic instability of precancerous cells, respectively, in addition to
-variation in the previously mentioned chemical gradients (1, 5, 6). In
-IBD patients, disease flares in the distal colon are usually indicative
-of ulcerative colitis (UC), whereas Crohn's disease (CD) patients
-typically experience disease in the small intestine, ileum and proximal
-colon (2). UC presents as large and highly inflammed mucosal ulcers,
-where as CD lesions are often smaller and have areas of normal tissue
-distributed amongst the flare (2). Thus, given the varied physiology of
-the proximal-distal axis of the colon and known differences in disease
-patterns at these sites, symbiotic microbes and their metabolites likely
-vary as well, and may influence the heterogenous disease prognoses of
-IBD and CRC. Because CRC can be a long-term complication of IBD, the
-distribution of microbes is important to understanding the
-pathophysiology of both diseases.
+The human colon is an ecosystem comprised of numerous microenvironments
+that select for different microbiota. Concentrations of oxygen, water
+and anti-microbial peptides change along the gut axis and influence
+which microbiota reside in each location. Microenvironments differ not
+only longitudinally along the colon, but also radially from the
+epithelium to mucosa to intestinal lumen, offering several sites for
+different microbial communities to flourish. The identity of these
+specific microbiota and communities are important for understanding the
+etiology of complex colon diseases such as Colorectal Cancer (CRC) and
+Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
+accelerated by perturbations the structure of the gut microbiota (1--3).
+The manifestations of these diseases are known to vary based upon the
+location in which they occur. For instance, CRC that arises in the
+distal (left) colon are of hindgut origin and tend to have large
+chromosomal alterations indicative of chromosomal instability (1). In
+contrast, CRC arising in the proximal (right) colon tumors are of midgut
+origin and tend to be sessile and microsatellite instable (MSI with BRAF
+and KRAS mutations) (1). In addition to the environmental gradients
+within the colon, the distal and proximal sides of the colon differ in
+the amount of inflammation present and the genomic instability of
+precancerous cells, respectively (1,4,5). In IBD patients, disease
+occurring in the distal colon extending proximally is usually indicative
+of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur
+anywhere along the GI tract, most commonly in the ileum and the cecum
+(2). UC presents as continuous disease with only mucosal involvement,
+where as CD has skip lesions and full thickness involvement that may
+cause abscesses, strictures and fistulas (2). Thus, given the varied
+physiology of the proximal-distal axis of the colon and known
+differences in disease patterns at these sites, symbiotic microbiota and
+their metabolites likely vary as well, and may influence the
+heterogeneous disease prognoses of IBD and CRC. Because CRC can be a
+long-term complication of IBD, the distribution of microbiota is
+important to understanding the pathophysiology of both diseases.
 
@@
 IBD or CRC can be attributed to specific molecular events as a result of
-interactions between the gut microbiota and human host (1, 3). For
+interactions between the gut microbiota and human host (1,3,6). For
 instance, comparison of the bacteria present on CRC tumors with those
 found on nearby healthy tissue has identified specific species that are
-tumor-associated (7). These species include the oral pathogens
-\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
-Interestingly, these periodontal pathogens have been highly predictive
-of whether a patient had CRC tumors or not in our prior human stool
-classification studies (8). \emph{F. nucleatum} has also been found to
-be elevated in the stool and biopsies of patients with IBD as compared
-to healthy controls (9). Furthermore, studies of \emph{F. nucleatum}
-isolated from mucosal biopsies showed that more invasive \emph{F.
-nucleatum} positively correlates with IBD disease level (9). Like many
-intestinal pathogens, the bacteria appear to have a high-impact despite
-being lowly-abundant in the community (2). The physiology of these rare
-taxa may contribute to the colonic disease state. These studies often
-examined only shed human stool or the small intestine, preventing
-fine-resolution analysis of paired samples from the proximal and distal
-sides of the colon. Similarly, comparisons of on- or off-tumor/lesion
-bacteria rarely have matched tissue from the other side of the colon
-from the same patient, limiting what conclusions can be drawn about the
-colonic microbiome overall, let alone at that specific site. Due to
-these limitations, the contribution of the gut microbiota to IBD and CRC
-disease location in the colon is largely undefined. Characterizing these
-communities could provide needed insight into disease etiology,
+tumor-associated (7). Specific bacteria have also been identified in
+fecal samples of patients with varying stages of colon tumorigenesis
+(8,9). These species include the oral pathogens \emph{Fusobacterium
+nucleatum} and \emph{Porphyromonas asacharolytica}. \emph{F. nucleatum}
+has also been found to be elevated in the stool and biopsies of patients
+with IBD as compared to healthy controls (10,11). Furthermore, studies
+of \emph{F. nucleatum} isolated from mucosal biopsies showed that more
+invasive \emph{F. nucleatum} positively correlates with IBD disease
+level (10). Like many intestinal pathogens, the bacteria appear to have
+a high-impact despite being lowly-abundant in the community (2). The
+physiology of these rare taxa may contribute to the colonic disease
+state. These studies often examined only shed human stool or the small
+intestine, preventing fine-resolution analysis of paired samples from
+the proximal and distal sides of the colon. Similarly, comparisons of
+on- or off-tumor/lesion bacteria rarely have matched tissue from the
+other side of the colon from the same, disease-baring patient, limiting
+what conclusions can be drawn about the colonic microbiome overall, let
+alone at that specific site. Due to these limitations, the contribution
+of the gut microbiota to CRC and IBD disease location in the colon is
+largely undefined. Characterizing these communities in healthy
+individuals could provide needed insight into disease etiology,
 including how the disruption of the healthy community could promote the
@@
 
-The few existing profiles of the microbial biogeography of the gut have
-been limited by sample collection methods. The majority of human gut
-microbiome studies have been performed on whole shed feces or on samples
-collected during colonoscopy procedures. While the latter method allows
-investigators to acquire samples from inside the human colon, typically
-this procedure is preceded by the use of bowel preparation methods such
-as the consumption of laxatives to cleanse the bowel. Bowel preparation
-is essential for detecting cancerous or precancerous lesions in the
-colon, but complicates microbiome profiling as the chemicals strip the
-bowel of contents and disrupt the mucosal layer (10, 11). As such, what
-little information we do have about the biogeographical distribution of
-the microbes in the proximal and distal colon is confounded by the bowel
-preparation procedure.
+The few existing profiles of the microbial spatial variation of the
+colon have been limited by sample collection methods. The majority of
+human gut microbiome studies have been performed on whole shed feces or
+on samples collected during colonoscopy or surgery. While the latter
+method allows investigators to acquire samples from inside the human
+colon, typically these procedures are preceded by the use of bowel
+preparation methods such as the consumption of laxatives to cleanse the
+bowel. Bowel preparation is essential for detecting cancerous or
+precancerous lesions in the colon, but complicates microbiome profiling
+as the chemicals strip the bowel of contents and disrupt the mucosal
+layer (12,13). As such, what little information we do have about the
+spatial distribution of the microbiota in the proximal and distal colon
+is confounded by the bowel preparation procedure.
 
-Here we address the limitations of previous studies and effectively
-identify the microbes specific to the lumen and mucosa of the proximal
-and distal healthy human colon. We used an unprepared colonoscopy
-technique to sample the natural community of each location of the gut
-without prior disruption of the native bacteria in 20 healthy
-volunteers. To address the inherent inter-individual variation in human
-microbiomes, we used a machine-learning classification algorithm trained
-on curated 16S rRNA sequencing reads to identify microbes specific to
-each location. We found that our classification models were able to
-separate mucosal and lumenal samples as well as differentiate between
-sides of the colon based on populations of specific microbes. By
-identifying the specific microbes we are poised to ask if and how the
-presence or disruption of the microbes at each site contribute to the
-development of the specific tumor subtypes of CRC in the proximal and
-distal human colon.
+Here we address the limitations of previous studies and identify the
+microbes specific to the lumen and mucosa of the proximal and distal
+healthy human colon. We used an unprepared colonoscopy technique to
+sample the natural community of each location of the gut without prior
+disruption of the native bacteria in 20 healthy volunteers. To address
+the inherent inter-individual variation in microbiota, we used a
+machine-learning classification algorithm trained on curated 16S rRNA
+sequencing reads to identify the microbiota that were specific to each
+location. We found that our classification models were able to separate
+mucosal and luminal samples as well as differentiate between sides of
+the colon based on populations of particular microbiota. By identifying
+the distinguishing microbiota we are poised to ask if and how the
+presence or disruption of the microbiota at each site contribute to the
+development of the tumor subtypes of CRC in the proximal and distal
+human colon.
 
@@
 
-Lumenal and mucosal samples were collected from the proximal and distal
+Luminal and mucosal samples were collected from the proximal and distal
 colon of 20 healthy humans that had not undergone bowel preparation
-(Figure 1). Participants also collected stool at home one week prior to
-the procedure. To characterize the bacterial communities present at
-these sites, 16S rRNA gene sequencing was performed on extracted DNA
-from each sample. As expected, each site was primarily dominated by
-\emph{Firmicutes} and \emph{Bacteriodetes} (Figure 2A) (12). Samples had
+(Figure 1). Subjects also collected stool at home one week prior to the
+procedure. To characterize the bacterial communities present at these
+sites, 16S rRNA gene sequencing was performed on DNA extracted from each
+sample. As expected, each site was primarily dominated by
+\emph{Firmicutes} and \emph{Bacteriodetes} (Figure 2A) (14). Samples had
 varying levels of diversity at each site, irrespective of the individual
@@
 
-To compare similarity between sides (proximal or distal) or sites (lumen
-or mucosa), we calculated distances from Operational Taxonomic Unit
-(OTU) abundances and compared these distances for all individuals.
-Again, across all patient samples we observed a range of distances when
-comparing sample locations (Figure 3A) and again those ranges did not
-follow a clear pattern on an individual basis. However, when comparing
-median distances between the proximal lumen and mucosa, the proximal
-versus distal lumen, the proximal versus distal mucosa, and the distal
-lumen and mucosa, we found that the proximal lumen and mucosa were most
-similar to each other than the other samples (P \textless{} 0.005,
-Wilcoxon, BH adjustment).
+To compare similarity between the proximal and distal sides and within
+the lumen and mucosa, we compared the community structure of these sites
+based on the relative abundances of individual Operational Taxonomic
+Units (OTUs). Across all subjects we observed wide variation when
+comparing sample locations (Figure 3A). Those ranges did not follow a
+clear pattern on an individual basis. However, when comparing median
+dissimilarity between the communities found in the proximal lumen and
+mucosa, the proximal and distal lumen, the proximal and distal mucosa,
+and the distal lumen and mucosa, we found that the proximal lumen and
+mucosa were most similar to each other than to the other samples (P
+\textless{} 0.005, Wilcoxon, BH adjustment).
 
-\paragraph{Fecal samples resemble lumenal samples from the distal
-colon}\label{fecal-samples-resemble-lumenal-samples-from-the-distal-colon}
+\paragraph{Fecal samples resemble luminal samples from the distal
+colon}\label{fecal-samples-resemble-luminal-samples-from-the-distal-colon}
 
-Next, we calculated distances to examine how each sample compared to the
-home-collected feces. Amidst variability between patients, we did
-identify significantly smaller distance between the distal lumenal
-sample and the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
-adjustment). Furthermore, there was an even larger difference in the
-comparisons of the distal mucosa to the feces, indicating that the
-mucosa is different from the stool as compared to lumen (P \textless{}
-0.0005, Wilcoxon, BH adjustment). To determine what factors may be
-driving the differences seen among the samples, we compared distances
-between samples from all subjects (interpersonal) versus samples from
-within one subject (intrapersonal). We found that samples from one
-individual were far more similar to each other than to other study
-subjects (Figure 3C), consistent with previous human microbiome studies
-that have sampled multiple sites of the human colon (13--15). Thus
-interpersonal variation between subjects drove the differences between
-samples more than sample site or location. Overall, the results
-comparing the structure of the communities suggest that the contents of
-the distal lumen are most representative of the patient's feces, and the
-microbes remaining on the mucosa are more distinct.
+Next, we compared the luminal and mucosal samples to the fecal sample of
+each subject. Amidst the large inter-subject variation, we did identify
+significantly less dissimilarity between the distal luminal sample and
+the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH adjustment).
+Furthermore, there was an even larger difference in the communities
+found in the distal mucosa compared to the fecal communities, indicating
+that the mucosa is as different from the stool as compared to lumen (P
+\textless{} 0.0005, Wilcoxon, BH adjustment). These results suggest that
+the contents of the distal lumen were most representative of the
+subjects' feces, and the mucosal microbiota are distinct from the fecal
+and luminal communities.
+
+\paragraph{Interpersonal community variation is greater than the
+variation between
+sites}\label{interpersonal-community-variation-is-greater-than-the-variation-between-sites}
+
+To determine what factors may have driven the differences seen among the
+samples, we compared the community dissimilarity between samples from
+all subjects (interpersonal) versus samples from within one subject
+(intrapersonal). We found that samples from one individual were far more
+similar to each other than to matched samples from the other subjects
+(Figure 3C); this is consistent with previous human microbiome studies
+that have sampled multiple sites of the human colon (15--17). Thus
+interpersonal variation drove the differences between samples more than
+whether the sample came from the proximal or distal side of the colon or
+from the lumen or mucosa.
 
@@
 
-To identify OTUs that were distinct at each biogeographical site, we
-constructed several Random Forest models trained using OTU abundances.
-We used 10-fold cross validation to build the first model to classify
-the lumenal versus mucosal samples for the proximal and distal sides,
-independently (Figure S1A). The models were constructed to only use the
-five most predictive OTUs as input to reduce overfitting between
-samples. The models performed well when classifying these samples (AUC =
-0.856 and AUC = 0.975, respectively). The OTUs that were most predictive
-of each site were identified by their greatest mean decrease in accuracy
-when removed from the model. For distinguishing the proximal lumen and
-mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
-\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
-were differentially abundant (Figure 4A). The model classifying the
-distal lumen and mucosa identified OTUs from \emph{Turicibacter},
-\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
-\emph{Anaerococcus} genera (Figure 4B). These results indicated that
-there were fine differences between the different sites of the colon,
-and that these could be traced to specific OTUs on each side.
+To identify OTUs that were distinct at each site, we constructed several
+Random Forest models trained using OTU relative abundances. We used
+10-fold cross validation to build the first model to classify the
+luminal versus mucosal samples for the proximal and distal sides,
+independently (Figure S1A). The models performed well when classifying
+these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
+were most predictive of each site were identified by their greatest mean
+decrease in accuracy when removed from the model. For distinguishing the
+proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
+\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
+included in the best model (Figure 4A). The model to differentiate
+between the distal lumen and mucosa included OTUs affiliated with the
+\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
+\emph{Anaerococcus} (Figure 4B). These results indicated that there were
+fine differences between the different sites of the colon, and that
+these could be traced to specific OTUs on each side.
 
-Next, we built a model to differentiate the proximal and distal lumenal
-samples using 10-fold cross validation. The model performed best when
-distinguishing the proximal versus distal mucosa (Figure S1B, AUC =
-0.920) compared to the proximal versus distal lumen (AUC = 0.575). These
-models were also constructed to use only the five most predictive OTUs
-as input to reduce overfitting. OTUs that were differentially abundant
-between the distal and proximal mucosa included members of the
-\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
-\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5A).
-Differentially abundant OTUs of the proximal and distal lumen included
-three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
-and an \emph{Oscillibacter} OTU (Figure 5B). This analysis found that
-some of the same OTUs that are distinct between the mucosa and lumen
-also helped to differentiate between the two sides- such as
-\emph{Anaerococcus} and \emph{Finegoldia}.
+Next, we built a Random Forest model to differentiate the proximal and
+distal luminal samples using 10-fold cross validation. The model
+performed best when distinguishing the proximal versus distal mucosa
+(Figure S1B, AUC = 0.850) whereas the model to differentiate between the
+proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
+included in the model differentiating the distal and proximal mucosa
+included members of the \emph{Porphyromonas}, \emph{Murdochiella},
+\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Figure
+5A). The model that attempted to separate the the proximal and distal
+lumen included OTUs affiliated with the \emph{Bacteroides},
+\emph{Clostridium IV} and \emph{Oscillibacter} (Figure 5B).
+Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were distinct
+between the mucosa and lumen and also helped to differentiate between
+the proximal and distal sides.
 
-\paragraph{Bacterial OTUs associated with cancer are found in healthy
-individuals}\label{bacterial-otus-associated-with-cancer-are-found-in-healthy-individuals}
+\paragraph{Bacterial OTUs associated with CRC and IBD are found in
+healthy
+individuals}\label{bacterial-otus-associated-with-crc-and-ibd-are-found-in-healthy-individuals}
 
@@
 Among our 100 samples, the most frequent sequence associated with the
-\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
+\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
 \emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
 This is the only species of \emph{Fusobacterium} known to have oncogenic
-properties and be found on the surfaces of colorectal cancer tumors.
-(16). There were 14 patient samples with the \emph{F. nucleatum subsp.
-animalis} sequences. Of the samples with the highest abundance of
-\emph{F. nucleatum subsp. animalis}, four of the samples were from the
-proximal mucosa and three from the distal mucosa (Figure 6A). The second
-most frequent \emph{Fusobacterium} sequence was OTU472, which aligned
-with 99\% identity to \emph{F. varium}. In addition to \emph{F.
-nucleatum}, \emph{F. varium} has been associated with IBD (17). Four
-study participants harbored \emph{F. varium} and the samples were split
-evenly between the proximal and distal mucosa (Figure 6B). OTU152 was
+properties and be found on the surfaces of colorectal cancer tumors
+(18). There were 14 samples from 8 subjects with the \emph{F. nucleatum
+subsp. animalis} sequences. Of the samples with the highest relative
+abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
+were from the proximal mucosa and three from the distal mucosa (Figure
+S2A). The second most frequent \emph{Fusobacterium} sequence was OTU472,
+which aligned with 99\% identity to \emph{F. varium}. In addition to
+\emph{F. nucleatum}, \emph{F. varium} has been associated with IBD (19).
+Four subjects harbored \emph{F. varium} and the samples were split
+evenly between the proximal and distal mucosa (Figure S2B). OTU152 was
 similar to the members of the \emph{Porphyromonas} genus and the most
@@
 on the distal mucosa, and in fact was one of the OTUs the classification
-model identified as separating distal and proximal sides (Figure 6C).
+model identified as separating distal and proximal sides (Figure S2C).
 Among the 11 distal mucosa samples that were positive for \emph{P.
 asacharolytica}, the relative abundances for this OTU ranged from 0.01\%
-- 16\%. Thus, disease-associated OTUs could be found in our sample set
+to 16\%. Thus, disease-associated OTUs could be found in our sample set
 of 20 healthy individuals.
@@
 Here we identified bacterial taxa that were specific to the lumen and
-mucosa of the proximal and distal sides of the human colon from samples
-collected during unprepared colonoscopy. We found that all locations
-contained a range of phyla abundances and a range of diversity, but that
-there was a wide variability between subjects. Pairwise comparisons of
-each of the sites revealed that the proximal mucosa and lumen were most
-similar to each other. Further, comparison of colonoscopy-collected
-samples with samples collected from stool at home showed that the distal
-lumen was most similar to feces. Random Forest models built on OTU
-relative abundances from each sample identified microbes that were
-particular to each location of the colon. Finally, we were able to
-detect some bacterial OTUs associated with colonic disease in our
-healthy patient cohort. Using unprepped colonoscopy and machine
-learning, we have identified bacterial taxa specific to the healthy
-proximal and distal human colon.
+mucosa of the proximal and distal sides of the human colon using samples
+collected during an unprepared colonoscopy of healthy subjects. We found
+that all locations contained a range of phyla relative abundances and a
+range of diversity, but that there was a wide variability between
+subjects. Pairwise comparisons of each of the sites revealed that the
+proximal mucosa and lumen were most similar to each other. Further,
+comparison of colonoscopy-collected samples with fecal samples
+demonstrated that the distal lumen was most similar to feces. Random
+Forest models built using OTU relative abundances from each sample
+identified microbiota that were particular to each location of the
+colon. Finally, we were able to detect some bacterial OTUs associated
+with colonic disease in our healthy cohort. Using unprepped
+colonoscopies and machine learning, we have identified bacterial taxa
+specific to the healthy proximal and distal human colon.
 
@@
 human subjects, casued by differences in age, sex, diet, but also
-biogeogrphical ``patchiness'' in the gut microbiome. Several studies
-have noted that the bacteria recoverable from the same mucosal sample
-location can be vastly different when the samples are taken just 1 cm
-away from each other (18). Similar patchiness is also observed in
-lumenal contents and fecal samples themselves; there is observed
-separation of different interacting microbes along the length of a stool
-sample, for instance (19). That said, across our samples the mucosal
-samples harbor more \emph{Proteobacteria}, consistent with previous
-studies comparing mucosal swabs to lumenal content in humans (5). Hence,
-the conclusions we can draw from phyla analysis may be impacted by
-patchiness between subjects.
+spatial ``patchiness'' in the gut microbiome. One study noted that the
+bacteria recoverable from the same mucosal sample location can be vastly
+different when the samples are taken just 1 cm away from each other
+(20). Similar patchiness was also observed in luminal contents and fecal
+samples themselves; there was separation of different interacting
+microbes along the length of a stool sample, for instance (21). That
+said, across our samples, the mucosal samples harbor more
+\emph{Proteobacteria}, consistent with previous studies comparing
+mucosal swabs to luminal content in humans (4). Hence, the conclusions
+we were able to draw from phyla analysis may have been impacted by
+inter-subject patchiness.
 
-To get around the noisiness from a diverse set of samples, we built a
-Random Forest model to identify microbes specific to each side. For each
-comparison we identified the top five OTUs that were strongly predictive
-of one site or another. Generally, OTUs identified in each location were
-consistent with known physiological gradients along the gut axis (6).
+To get around the noisiness from a diverse set of samples, we built
+Random Forest classification models to identify the microbiota that were
+specific to each side and in the lumen and mucosa. For each comparison
+we identified the top five OTUs that were strongly predictive of one
+site or another. Generally, OTUs identified in each location were
+consistent with known physiological gradients along the gut axis (5).
 For instance, the proximal mucosa contains the highest oxygen
-concentratons of the colon and harbored mucosa-associated facultative
+concentrations of the colon and harbored mucosa-associated facultative
 anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
@@
 the gut microenvironment of each location likely enriches for these
-specific microbes.
+specific microbiota.
 
@@
 the gut, the ability of the Random Forest to classify samples can serve
-as a proxy for similarity. That is, a higher AUC value means the samples
-are more efficiently classfied (and thus more different) than a model
-with a lower AUC value. For instance, the model separating the proximal
-and distal mucosa has an AUC of 0.925 whereas the model for classifying
-the proximal and distal lumen has a much lower AUC of 0.575. Further,
-prior to reducing the input to five OTUs, the latter model required 44
-OTUs to best separate the samples (AUC = 0.755). The much lower AUC and
-need for a high number of features compared to other models suggest
-these locations are the most similar of the comparisons tested. We
-speculate that the model was least effective at classifying the proximal
-and distal lumenal contents because the samples are arguably composed of
-the same bacteria but differ in water content.
+as a proxy for similarity. That is, a higher AUC value indicates the
+samples are more efficiently classfied (and thus more different) than a
+model with a lower AUC value. For instance, the model separating the
+proximal and distal mucosa had an AUC of 0.850 whereas the model for
+classifying the proximal and distal lumen had a much lower AUC of 0.580.
+Further, the latter model required 44 OTUs to best separate the samples
+whereas the models separating the mucosa only needed 10 OTUs. The much
+lower AUC and need for a high number of features compared to other
+models suggest these locations are the most similar of the comparisons
+tested. We speculate that the model was less effective at classifying
+the proximal and distal luminal contents because the mucosal
+microenvironments have variable selective pressure along the colon than
+the luminal microenvironments.
 
@@
 of our subjects, respectively. These bacteria have been shown to be
-predictive of colorectal cancer in humans (8) and have oncogenic
-properties in cell culture and in mice (20). Interestingly, while
+predictive of colorectal cancer in humans (9) and have oncogenic
+properties in cell culture and in mice (22). Interestingly, while
 \emph{F. nucleatum} was found on both sides of the colon, \emph{P.
 asacharolytica} was only detected in the distal mucosa. Not much is
-known about the distribution of \emph{P. asacharolytica} but given its
-documented anaerobic characteristics and asacharolytic metabolism, it
-might not be surprising that it resides in the less-oygen-rich and
-proteinaceous distal mucosa (5). In studies examining bacteria on
-colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
+known about the distribution of \emph{P. asacharolytica} along the
+colon, but given its anaerobic lifestyle and asacharolytic metabolism,
+it is not surprising that it resides in the less-oygen-rich and
+protein-rich distal mucosa (4). In studies examining bacteria on
+colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
 on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreases along the colon to rectum (21). Of the 8 (40\%) individuals
-positive for \emph{F. nucleatum} in this study, the bacterium was spread
-across the proximal mucosa, distal lumen and distal mucosa. Data
-examining bacterial biofilms on the mucosa of CRC tumors suggests that
-\emph{Fusobacteria} species are more commonly found on proximal tumors
-and in biofilms, indicating that it is not only the presence of the
-bacteria but the organization of the tumor community that contributes to
-\emph{Fusobacterium's} role in tumorigenesis (7). Finally,
-\emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
-not only co-occur on CRC tumors but also to synergistically promote
-tumorigenesis in an oral cancer model (22) (23). Thus, further analysis
-of the distribution and activities of these pathogens may elucidate a
-mechanism for development of IBD or CRC subtypes in the proximal or
-distal colon.
+decreased along the colon to rectum (23). In another study,
+\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
+mutations, molecular features of proximal CRC (24). Of the 8 (40\%)
+individuals positive for \emph{F. nucleatum} in the present study, the
+bacterium was spread across the proximal mucosa, distal lumen and distal
+mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
+suggests that \emph{Fusobacteria} species are more commonly found on
+proximal tumors and in biofilms, indicating that it is not only the
+presence of the bacterium but the structure of the tumor community that
+contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
+Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
+only co-occur on CRC tumors but also to synergize to promote
+tumorigenesis in an oral cancer model (25) (26). Further analysis of the
+distribution and activities of these pathogens along the colon is needed
+to elucidate a mechanism for development of CRC or IBD subtypes in the
+proximal or distal colon.
 
 The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
-been commonly isolated from mucosal biopsies of patients with IBD (17).
+been commonly isolated from mucosal biopsies of patients with IBD (19).
 Laboratory experiments with these isolates have shown that
@@
 more TNF-\(\alpha\) production than strains from healthy individuals
-(9), suggesting the bacteria may increase inflammation in the gut as
-well (24). \emph{F. varium} isolated from UC patients caused colonic
-ulcers in an experimental mouse model (25). \emph{F. varium} was only
-detected in three of our study participants and two of those samples
-were isolated form the proximal mucosa (Figure 6B). \emph{F. varium} is
-most commonly isolated from UC patient biopsies from the ileum or cecum
-(26), suggesting this species may exhibit preference for the different
+(10), suggesting the bacteria may increase inflammation in the gut as
+well (27). \emph{F. varium} isolated from UC patients caused colonic
+ulcers in an experimental mouse model (28). \emph{F. varium} was only
+detected in three subjects and two of those samples were isolated form
+the proximal mucosa (Figure S2B). \emph{F. varium} is most commonly
+isolated from UC patient biopsies from the ileum or cecum (29),
+suggesting this species may exhibit preference for the different
 environmental conditions of these gastrointestinal sites. Further work
@@
 
-Specific comparisons of our findings to previously published gut
-biogeography studies are confounded by the use of bowel preparation
-methods in most other studies. A rare report of a matched-colonoscopy
-study sampled 18 patient's colonic mucosa and lumenal contents prior to
-and after bowel cleansing (27). This study found that mucosal and
-lumenal samples were distinguishable prior to bowel cleansing, but that
-bowel preparation resulted in an increase in shared OTUs between each
-site (27). After seven days, bowel cleansing not only made the samples
-harder to distinguish, but it also decreased in diversity across sites.
-Bowel preparation clearly biases into the microbes recovered from
-sampling the lumen or mucosa of a prepared bowel.
+Specific comparisons of our findings to previously published studies of
+spatial variation are confounded by the use of bowel preparation
+methods. A rare report of a matched-colonoscopy study sampled 18
+patient's colonic mucosa and luminal contents prior to and after bowel
+cleansing (30). This study found that mucosal and luminal samples were
+distinguishable prior to bowel cleansing, but that bowel preparation
+resulted in an increase in shared OTUs between each site (30). After
+seven days, bowel cleansing not only made the samples more difficult to
+distinguish, but it also decreased the diversity observed across sites.
+Bowel preparation clearly biases the representation of microbiota
+recovered from sampling the lumen or mucosa.
 
@@
 location in the gut via sampling an unprepared bowel, we can begin to
-form hypothesies about how specific host-microbe interactions can affect
-disease progression of proximal and distal CRC and IBD subtypes. To this
-point, 16S rRNA gene sequecing community profiling studies do not
-provide enough information to fully probe these questions. In
-particular, 16S sequencing cannot not profile the host characteristics
-at each site. Combining the unprepared colonoscopy approach with
-analysis of multi-omic sequencing data may be useful in further
-characterizing host-microbiome interactions along the gut axis for both
-health and disease.
+form hypotheses about how specific host-microbe interactions can affect
+disease progression of proximal and distal CRC and IBD subtypes. A
+better understanding of microbial activities in the gut can enhance
+microbiome-based screening and treatment modalities for these colon
+diseases.
 
@@
 Institutional Review Board at the University of Michigan Health System
-with protocol number XXXX. Subjects were recruited using the online
-recruitment platform and were pre-screened prior to enrollment in the
-study. Exclusion criteria included: use of asprin or NSAIDs within 7
+with protocol number HUM00082721. Subjects were recruited using the
+online recruitment platform and were pre-screened prior to enrollment in
+the study. Exclusion criteria included: use of asprin or NSAIDs within 7
 days, use of antibiotics within 3 months, current use of anticoagulants,
@@
 
-At a baseline visit, subjects were consented and given a home collection
-stool kit (Source of kit?). At least one week prior to the scheduled
-colonoscopy, subjects were to collect whole stool at home and ship the
+At a baseline visit, subjects gave consent and were given a home
+collection stool kit (Zymo). At least one week prior to the scheduled
+colonoscopy, subjects collected whole stool at home and shipped the
 samples to a research coordinator on ice. Notably, subjects did not
@@
 day, subjects reported to the Michigan Clinical Research Unit at the
-University of Michigan Health System. Patients were consciously sedated
+University of Michigan Health System. Subjects were consciously sedated
 using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
-sigmoidoscope was first inserted about 25cm into the colon and endoscopy
-brush used to collect lumenal/stool contents. Two lumenal samples were
-collected and the contents immediately deposited into RNAlater (Fischer)
-and flash-frozen in liquid nitrogen. The brushes were withdrawn and
-biopsy forceps were used to collect mucosal biopsies on sections of the
-colon that were pink and free of stool matter. Three mucosal biopsies
-were collected and flash-frozen in RNAlater. These samples comprised the
-distal or distal colon samples. The sigmoidoscope was then withdrawn and
-a pediatric colonoscope was inserted to reach the ascending colon.
-Samples were then collected as in the distal colon and the colonoscope
-withdrawn. All samples were stored at -80\(^\circ\)C.
+sigmoidoscope was first inserted about 25cm into the colon and jumbo
+biopsy forceps used to collect the luminal contents. Two luminal samples
+were collected and the contents immediately deposited into RNAlater
+(Fischer) and flash-frozen in liquid nitrogen. The forceps were
+withdrawn and new biopsy forceps were used to collect mucosal biopsies
+on sections of the colon that were pink and free of stool matter. Three
+mucosal biopsies were collected and flash-frozen in RNAlater. These
+samples comprised the distal colon samples. The sigmoidoscope was then
+withdrawn and a pediatric colonoscope was inserted to reach the proximal
+colon. Samples were then collected in the same manner as was done in the
+distal colon and the colonoscope withdrawn. All samples were stored at
+-80\(^\circ\)C.
 
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described (28). Sequences were curated
-using the mothur software as described previously (29). The sequences
-were assigned taxonomic classification using a naive Bayesian classifier
-trained using a 16S rRNA gene training set from the Ribosomal Database
-Project (RDP) (30) and clustered into operational taxonomic units (OTUs)
-based on a 97\% similarity cutoff. Sequencing and analysis of a mock
-community revealed the error rate to be 0.018\%. Samples were rarefied
-to 4231 sequences per sample in order to reduce uneven sampling bias.
+Illumina MiSeq as previously described (31). Sequences were curated
+using the mothur software as described previously (32). The sequences
+were assigned a taxonomic classification using a naive Bayesian
+classifier trained using a 16S rRNA gene training set from the Ribosomal
+Database Project (RDP) (33) and clustered into operational taxonomic
+units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
+of a mock community revealed the error rate to be 0.018\%. Samples were
+rarefied to 4231 sequences per sample in order to reduce uneven sampling
+bias.
 
 Diversity analysis was performed using the Simpson diversity calculator
-and \(\theta\)YC calculator metrics in mothur version 1.39.5 (29).
-ThetaYC distances were calculated to determine the dissimilarity between
-two samples. Random Forest classification models were built using the
-randomForest R package and resultant models were used to identify the
-OTUs that were most important for classifying each location (31). To get
-species-level information about sequences of interest, sequences were
-aligned using blastn and the species name was only used if the identity
-score was \(\ge\) 99\% over the full-length of the contig and matched a
-single reference.
+and \(\theta\)YC calculator metrics in mothur version 1.39.5 (32).
+\(\theta\)YC distances were calculated to determine the dissimilarity
+between two samples. Random Forest classification models were built
+using the AUCRF R package using a leave-one-subject out approach (34).
+For each model the data was split into a 19-subject training set and a
+1-subject test set. The model was built and cross-validated using
+AUCRFcv on the training set. The model was then tested on the left-out
+patient. This process was repeated iteratively for all subjects and
+results plotted as Reciever Operator Characteristic curves using the
+pROC R package (35). Resultant models were used to identify the OTUs
+that were most important for classifying each location. Species-level
+information for sequences of interest was obtained by aligning the
+sequences to the GenBank nucleotide databse using blastn. The species
+name was only used if the identity score was \(\ge\) 99\% over the
+full-length of the contig and matched a single reference.
 
@@
 
-\includegraphics{../submission/figure_1.pdf}
-
 Sampling strategy. A flexible sigmoidoscope was used to sample the
 distal colonic luminal contents and mucosa. The scope was inserted
-\textasciitilde{} 25cm into the subject and endoscopy brushes were used
-to sample the luminal contents (green star). A separate set of biopsy
-forceps was used to sample the distal mucosa (blue star). The
-sigmoidoscope was removed. A pediatric colonoscope was inserted and used
-to access the proximal colon. Biopsies were taken of the proximal
-luminal contents and mucosa as described. One week prior to the
-procedure stool was collected at home and sent into the laboratory.
-Representative images from one individual are shown.
+\textasciitilde{} 25cm into the subject and biopsy forceps were used to
+sample the luminal contents (D, inset). A separate set of biopsy forceps
+was used to sample the distal mucosa (D, inset). The sigmoidoscope was
+removed. A pediatric colonoscope was inserted and used to access the
+proximal colon (P, inset). Biopsies were taken of the proximal luminal
+contents and mucosa as described. One week prior to the procedure stool
+was collected at home and sent into the laboratory. Representative
+images from one individual are shown.
 
@@
 
-\includegraphics{../submission/figure_2.pdf}
-
 \paragraph{Figure 2}\label{figure-2}
 
-Microbial membership and diversity of the proximal and distal human
-colon. A) Relative abundance of the top five bacterial phyla in each
-sampling site. Each box represents the median and confidence intervals.
-B) Simpson diversity of the microbial communities at each location. The
-lines represent the median values.
+Phylum-level relative abundance and diversity in the proximal and distal
+human colon. A) Relative abundance of the top five bacterial phyla in
+each sampling site. Each box represents the median and interquartile
+range. B) Simpson diversity of the microbial communities at each
+location. The horizontal lines represent the median values.
 
@@
 
-\includegraphics{../submission/figure_3.pdf}
-
 \paragraph{Figure 3}\label{figure-3}
 
-Distances of microbial community structure between sites of the gut.
-ThetaYC distances are shown for interpersonal similarities between two
-sites -- each point represents one individual. In (A), comparisons of
-the proximal and distal mucosal and lumen are shown. In (B), comparisons
-of each site to the exit stool are shown.
+Comparison of microbial community structure between sites of the gut.
+\(\theta\)YC distances are shown to indicate the interpersonal
+dissimilarities between two sites -- each point represents one
+individual. In (A), comparisons of the proximal and distal mucosal and
+lumen are shown. In (B), comparisons of each site to the exit stool are
+shown. In (C), comparisons of samples from all subjects to each other
+(interpersonal) or within one subject (intrapersonal) are shown.
 
 \newpage
-
-\includegraphics{../submission/figure_4.pdf}
 
@@
 OTUs that are most important for the classification model for the distal
-mucosa and lumen (A) and the proximal mucosa and lumen (B).
+mucosa and lumen (A) and the proximal mucosa and lumen (B). The vertical
+lines represent the median values for each OTU.
 
@@
 
-\includegraphics{../submission/figure_5.pdf}
-
 \paragraph{Figure 5}\label{figure-5}
 
-Taxa specific to the distal and proximal mucosa and lumen. Top five OTUs
-that are most important for the classification model for the distal and
-proximal mucosa (A) and the distal and proximal lumen (B).
-
-\newpage
-
-\includegraphics{../submission/figure_6.pdf}
-
-\paragraph{Figure 6}\label{figure-6}
-
-Location and abundance of cancer-associated OTUs. Relative abundance was
-calculated and plotted by sample site for each OTU of interest: (A)
-\emph{Fusobacterium nucleatum subsp. animalis} (B) \emph{Fusobacterium
-varium} and (C) \emph{Porphyromonas asacharolytica}
-
-\newpage
-
-\includegraphics{../submission/figure_S1.pdf}
+Taxa specific to the distal and proximal mucosa and lumen. The five OTUs
+that were most important differentiating the distal and proximal mucosa
+(A) and the distal and proximal lumen (B). The vertical lines represent
+the median values for each OTU. \newpage
 
@@
 Characteristic curves are shown for the 10-fold cross validation of the
-Random Forest model classifying distal lumen versus proximal lumen
-(orange) and distal mucosa vs proximal mucosa (green). (B) Receiver
-Operator Characteristic curves are shown for the 10-fold cross
-validation of the Random Forest model classifying lumen and mucosal
-samples for the distal (red) and proximal (blue) sides of the colon.
+Random Forest model classifying lumen and mucosal samples for the distal
+(red) and proximal (blue) sides of the colon. (B) Receiver Operator
+Characteristic curves are shown for the 10-fold cross validation of the
+Random Forest model classifying distal mucosa vs proximal mucosa (green)
+and distal lumen versus proximal lumen (purple).
+
+\newpage
+
+\paragraph{Figure S2}\label{figure-s2}
+
+Location and relative abundance of cancer-associated OTUs. Relative
+abundance was calculated and plotted by sample site for each OTU of
+interest: (A) \emph{Fusobacterium nucleatum subsp. animalis} (B)
+\emph{Fusobacterium varium} and (C) \emph{Porphyromonas asacharolytica}
 
@@
 \hypertarget{ref-Yamauchi2012}{}
-1. \textbf{Yamauchi M}, \textbf{Lochhead P}, \textbf{Morikawa T},
-\textbf{Huttenhower C}, \textbf{Chan AT}, \textbf{Giovannucci E},
-\textbf{Fuchs C}, \textbf{Ogino S}. 2012. Colorectal cancer: A tale of
-two sides or a continuum?: Figure 1. Gut \textbf{61}:794--797.
-doi:\href{https://doi.org/10.1136/gutjnl-2012-302014}{10.1136/gutjnl-2012-302014}.
+1. Yamauchi M, Lochhead P, Morikawa T, Huttenhower C, Chan AT,
+Giovannucci E, et al. Colorectal cancer: A tale of two sides or a
+continuum?: Figure 1. Gut {[}Internet{]}. BMJ; 2012;61:794--7. Available
+from: \url{https://doi.org/10.1136/gutjnl-2012-302014}
 
 \hypertarget{ref-Forbes2016}{}
-2. \textbf{Forbes JD}, \textbf{Domselaar GV}, \textbf{Bernstein CN}.
-2016. The gut microbiota in immune-mediated inflammatory diseases.
-Frontiers in Microbiology \textbf{7}.
-doi:\href{https://doi.org/10.3389/fmicb.2016.01081}{10.3389/fmicb.2016.01081}.
+2. Forbes JD, Domselaar GV, Bernstein CN. The gut microbiota in
+immune-mediated inflammatory diseases. Frontiers in Microbiology
+{[}Internet{]}. Frontiers Media SA; 2016;7. Available from:
+\url{https://doi.org/10.3389/fmicb.2016.01081}
 
 \hypertarget{ref-Halfvarson2017}{}
-3. \textbf{Halfvarson J}, \textbf{Brislawn CJ}, \textbf{Lamendella R},
-\textbf{Vazquez-Baeza Y}, \textbf{Walters WA}, \textbf{Bramer LM},
-\textbf{DAmato M}, \textbf{Bonfiglio F}, \textbf{McDonald D},
-\textbf{Gonzalez A}, \textbf{McClure EE}, \textbf{Dunklebarger MF},
-\textbf{Knight R}, \textbf{Jansson JK}. 2017. Dynamics of the human gut
-microbiome in inflammatory bowel disease. Nature Microbiology
-\textbf{2}:17004.
-doi:\href{https://doi.org/10.1038/nmicrobiol.2017.4}{10.1038/nmicrobiol.2017.4}.
-
-\hypertarget{ref-Benedix2010}{}
-4. \textbf{Benedix F}, \textbf{Kube R}, \textbf{Meyer F},
-\textbf{Schmidt U}, \textbf{Gastinger I}, \textbf{Lippert H}. 2010.
-Comparison of 17, 641 patients with right- and left-sided colon cancer:
-Differences in epidemiology, perioperative course, histology, and
-survival. Diseases of the Colon \& Rectum \textbf{53}:57--64.
-doi:\href{https://doi.org/10.1007/dcr.0b013e3181c703a4}{10.1007/dcr.0b013e3181c703a4}.
+3. Halfvarson J, Brislawn CJ, Lamendella R, Vazquez-Baeza Y, Walters WA,
+Bramer LM, et al. Dynamics of the human gut microbiome in inflammatory
+bowel disease. Nature Microbiology {[}Internet{]}. Springer Nature;
+2017;2:17004. Available from:
+\url{https://doi.org/10.1038/nmicrobiol.2017.4}
 
 \hypertarget{ref-Albenberg2014}{}
-5. \textbf{Albenberg L}, \textbf{Esipova TV}, \textbf{Judge CP},
-\textbf{Bittinger K}, \textbf{Chen J}, \textbf{Laughlin A},
-\textbf{Grunberg S}, \textbf{Baldassano RN}, \textbf{Lewis JD},
-\textbf{Li H}, \textbf{Thom SR}, \textbf{Bushman FD}, \textbf{Vinogradov
-SA}, \textbf{Wu GD}. 2014. Correlation between intraluminal oxygen
-gradient and radial partitioning of intestinal microbiota.
-Gastroenterology \textbf{147}:1055--1063.e8.
-doi:\href{https://doi.org/10.1053/j.gastro.2014.07.020}{10.1053/j.gastro.2014.07.020}.
+4. Albenberg L, Esipova TV, Judge CP, Bittinger K, Chen J, Laughlin A,
+et al. Correlation between intraluminal oxygen gradient and radial
+partitioning of intestinal microbiota. Gastroenterology {[}Internet{]}.
+Elsevier BV; 2014;147:1055--1063.e8. Available from:
+\url{https://doi.org/10.1053/j.gastro.2014.07.020}
 
 \hypertarget{ref-Donaldson2015}{}
-6. \textbf{Donaldson GP}, \textbf{Lee SM}, \textbf{Mazmanian SK}. 2015.
-Gut biogeography of the bacterial microbiota. Nature Reviews
-Microbiology \textbf{14}:20--32.
-doi:\href{https://doi.org/10.1038/nrmicro3552}{10.1038/nrmicro3552}.
+5. Donaldson GP, Lee SM, Mazmanian SK. Gut biogeography of the bacterial
+microbiota. Nature Reviews Microbiology {[}Internet{]}. Springer Nature;
+2015;14:20--32. Available from:
+\url{https://doi.org/10.1038/nrmicro3552}
+
+\hypertarget{ref-Kostic_2013}{}
+6. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
+et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
+modulates the tumor-immune microenvironment. Cell Host \& Microbe
+{[}Internet{]}. Elsevier BV; 2013;14:207--15. Available from:
+\url{https://doi.org/10.1016\%2Fj.chom.2013.07.007}
 
 \hypertarget{ref-Dejea2014}{}
-7. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
-\textbf{White JR}, \textbf{Welch JLM}, \textbf{Rossetti BJ},
-\textbf{Peterson SN}, \textbf{Snesrud EC}, \textbf{Borisy GG},
-\textbf{Lazarev M}, \textbf{Stein E}, \textbf{Vadivelu J},
-\textbf{Roslani AC}, \textbf{Malik AA}, \textbf{Wanyiri JW}, \textbf{Goh
-KL}, \textbf{Thevambiga I}, \textbf{Fu K}, \textbf{Wan F}, \textbf{Llosa
-N}, \textbf{Housseau F}, \textbf{Romans K}, \textbf{Wu X},
-\textbf{McAllister FM}, \textbf{Wu S}, \textbf{Vogelstein B},
-\textbf{Kinzler KW}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2014.
-Microbiota organization is a distinct feature of proximal colorectal
-cancers. Proceedings of the National Academy of Sciences
-\textbf{111}:18321--18326.
-doi:\href{https://doi.org/10.1073/pnas.1406199111}{10.1073/pnas.1406199111}.
+7. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Welch JLM, Rossetti
+BJ, et al. Microbiota organization is a distinct feature of proximal
+colorectal cancers. Proceedings of the National Academy of Sciences
+{[}Internet{]}. Proceedings of the National Academy of Sciences;
+2014;111:18321--6. Available from:
+\url{https://doi.org/10.1073/pnas.1406199111}
+
+\hypertarget{ref-McCoy_2013}{}
+8. McCoy AN, Araújo-Pérez F, Azcárate-Peril A, Yeh JJ, Sandler RS, Keku
+TO. Fusobacterium is associated with colorectal adenomas. Goel A,
+editor. PLoS ONE {[}Internet{]}. Public Library of Science (PLoS);
+2013;8:e53653. Available from:
+\url{https://doi.org/10.1371\%2Fjournal.pone.0053653}
 
 \hypertarget{ref-Baxter2016}{}
-8. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+9. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine {[}Internet{]}. Springer Nature;
+2016;8. Available from: \url{https://doi.org/10.1186/s13073-016-0290-3}
 
 \hypertarget{ref-Strauss2011}{}
-9. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
-\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
-\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
-gut mucosa-derived fusobacterium nucleatum positively correlates with
-IBD status of the host. Inflammatory Bowel Diseases
-\textbf{17}:1971--1978.
-doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.
+10. Strauss J, Kaplan GG, Beck PL, Rioux K, Panaccione R, DeVinney R, et
+al. Invasive potential of gut mucosa-derived fusobacterium nucleatum
+positively correlates with IBD status of the host. Inflammatory Bowel
+Diseases {[}Internet{]}. Ovid Technologies (Wolters Kluwer Health);
+2011;17:1971--8. Available from: \url{https://doi.org/10.1002/ibd.21606}
+
+\hypertarget{ref-Brennan_2016}{}
+11. Brennan CA, Garrett WS. Gut microbiota, inflammation, and colorectal
+cancer. Annual Review of Microbiology {[}Internet{]}. Annual Reviews;
+2016;70:395--411. Available from:
+\url{https://doi.org/10.1146\%2Fannurev-micro-102215-095513}
 
 \hypertarget{ref-Jalanka2014}{}
-10. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
-\textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
-\textbf{Gowland P}, \textbf{Hoad C}, \textbf{Garsed K}, \textbf{Lam C},
-\textbf{Palva A}, \textbf{Spiller RC}, \textbf{Vos WM de}. 2014. Effects
-of bowel cleansing on the intestinal microbiota. Gut
-\textbf{64}:1562--1568.
-doi:\href{https://doi.org/10.1136/gutjnl-2014-307240}{10.1136/gutjnl-2014-307240}.
+12. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
+et al. Effects of bowel cleansing on the intestinal microbiota. Gut
+{[}Internet{]}. BMJ; 2014;64:1562--8. Available from:
+\url{https://doi.org/10.1136/gutjnl-2014-307240}
 
 \hypertarget{ref-Harrell2012}{}
-11. \textbf{Harrell L}, \textbf{Wang Y}, \textbf{Antonopoulos D},
-\textbf{Young V}, \textbf{Lichtenstein L}, \textbf{Huang Y},
-\textbf{Hanauer S}, \textbf{Chang E}. 2012. Standard colonic lavage
-alters the natural state of mucosal-associated microbiota in the human
-colon. PLoS ONE \textbf{7}:e32545.
-doi:\href{https://doi.org/10.1371/journal.pone.0032545}{10.1371/journal.pone.0032545}.
+13. Harrell L, Wang Y, Antonopoulos D, Young V, Lichtenstein L, Huang Y,
+et al. Standard colonic lavage alters the natural state of
+mucosal-associated microbiota in the human colon. Singh SR, editor. PLoS
+ONE {[}Internet{]}. Public Library of Science (PLoS); 2012;7:e32545.
+Available from: \url{https://doi.org/10.1371/journal.pone.0032545}
 
 \hypertarget{ref-LloydPrice2016}{}
-12. \textbf{Lloyd-Price J}, \textbf{Abu-Ali G}, \textbf{Huttenhower C}.
-2016. The healthy human microbiome. Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0307-y}{10.1186/s13073-016-0307-y}.
+14. Lloyd-Price J, Abu-Ali G, Huttenhower C. The healthy human
+microbiome. Genome Medicine {[}Internet{]}. Springer Nature; 2016;8.
+Available from: \url{https://doi.org/10.1186/s13073-016-0307-y}
 
 \hypertarget{ref-Eckburg2005}{}
-13. \textbf{Eckburg PB}. 2005. Diversity of the human intestinal
-microbial flora. Science \textbf{308}:1635--1638.
-doi:\href{https://doi.org/10.1126/science.1110591}{10.1126/science.1110591}.
+15. Eckburg PB. Diversity of the human intestinal microbial flora.
+Science {[}Internet{]}. American Association for the Advancement of
+Science (AAAS); 2005;308:1635--8. Available from:
+\url{https://doi.org/10.1126/science.1110591}
 
 \hypertarget{ref-deCarcer2010}{}
-14. \textbf{Cárcer DA de}, \textbf{Cuív PÓ}, \textbf{Wang T},
-\textbf{Kang S}, \textbf{Worthley D}, \textbf{Whitehall V},
-\textbf{Gordon I}, \textbf{McSweeney C}, \textbf{Leggett B},
-\textbf{Morrison M}. 2010. Numerical ecology validates a biogeographical
-distribution and gender-based effect on mucosa-associated bacteria along
-the human colon. The ISME Journal \textbf{5}:801--809.
-doi:\href{https://doi.org/10.1038/ismej.2010.177}{10.1038/ismej.2010.177}.
+16. Cárcer DA de, Cuív PÓ, Wang T, Kang S, Worthley D, Whitehall V, et
+al. Numerical ecology validates a biogeographical distribution and
+gender-based effect on mucosa-associated bacteria along the human colon.
+The ISME Journal {[}Internet{]}. Springer Nature; 2010;5:801--9.
+Available from: \url{https://doi.org/10.1038/ismej.2010.177}
 
 \hypertarget{ref-Zhang2013}{}
-15. \textbf{Zhang Z}, \textbf{Geng J}, \textbf{Tang X}, \textbf{Fan H},
-\textbf{Xu J}, \textbf{Wen X}, \textbf{Ma Z (Sam)}, \textbf{Shi P}.
-2013. Spatial heterogeneity and co-occurrence patterns of human
-mucosal-associated intestinal microbiota. The ISME Journal
-\textbf{8}:881--893.
-doi:\href{https://doi.org/10.1038/ismej.2013.185}{10.1038/ismej.2013.185}.
+17. Zhang Z, Geng J, Tang X, Fan H, Xu J, Wen X, et al. Spatial
+heterogeneity and co-occurrence patterns of human mucosal-associated
+intestinal microbiota. The ISME Journal {[}Internet{]}. Springer Nature;
+2013;8:881--93. Available from:
+\url{https://doi.org/10.1038/ismej.2013.185}
 
 \hypertarget{ref-Castellarin2011}{}
-16. \textbf{Castellarin M}, \textbf{Warren RL}, \textbf{Freeman JD},
-\textbf{Dreolini L}, \textbf{Krzywinski M}, \textbf{Strauss J},
-\textbf{Barnes R}, \textbf{Watson P}, \textbf{Allen-Vercoe E},
-\textbf{Moore RA}, \textbf{Holt RA}. 2011. Fusobacterium nucleatum
-infection is prevalent in human colorectal carcinoma. Genome Research
-\textbf{22}:299--306.
-doi:\href{https://doi.org/10.1101/gr.126516.111}{10.1101/gr.126516.111}.
+18. Castellarin M, Warren RL, Freeman JD, Dreolini L, Krzywinski M,
+Strauss J, et al. Fusobacterium nucleatum infection is prevalent in
+human colorectal carcinoma. Genome Research {[}Internet{]}. Cold Spring
+Harbor Laboratory Press; 2011;22:299--306. Available from:
+\url{https://doi.org/10.1101/gr.126516.111}
 
 \hypertarget{ref-Lee2016}{}
-17. \textbf{Lee Y}, \textbf{Eun CS}, \textbf{Lee AR}, \textbf{Park CH},
-\textbf{Han DS}. 2016. FusobacteriumIsolates recovered from colonic
-biopsies of inflammatory bowel disease patients in korea. Annals of
-Laboratory Medicine \textbf{36}:387.
-doi:\href{https://doi.org/10.3343/alm.2016.36.4.387}{10.3343/alm.2016.36.4.387}.
+19. Lee Y, Eun CS, Lee AR, Park CH, Han DS. FusobacteriumIsolates
+recovered from colonic biopsies of inflammatory bowel disease patients
+in korea. Annals of Laboratory Medicine {[}Internet{]}. Korean Society
+for Laboratory Medicine (KAMJE); 2016;36:387. Available from:
+\url{https://doi.org/10.3343/alm.2016.36.4.387}
 
 \hypertarget{ref-Hong2011}{}
-18. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
-\textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
-analysis of the mucosal microbiota in healthy individuals reveals
-ubiquitous bacterial groups and micro-heterogeneity. PLoS ONE
-\textbf{6}:e25042.
-doi:\href{https://doi.org/10.1371/journal.pone.0025042}{10.1371/journal.pone.0025042}.
+20. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
+Pyrosequencing-based analysis of the mucosal microbiota in healthy
+individuals reveals ubiquitous bacterial groups and micro-heterogeneity.
+Ahmed N, editor. PLoS ONE {[}Internet{]}. Public Library of Science
+(PLoS); 2011;6:e25042. Available from:
+\url{https://doi.org/10.1371/journal.pone.0025042}
 
 \hypertarget{ref-Stearns2011}{}
-19. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
-\textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
-\textbf{Croitoru K}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}.
-2011. Bacterial biogeography of the human digestive tract. Scientific
-Reports \textbf{1}.
-doi:\href{https://doi.org/10.1038/srep00170}{10.1038/srep00170}.
+21. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
+Cvitkovitch DG, et al. Bacterial biogeography of the human digestive
+tract. Scientific Reports {[}Internet{]}. Springer Nature; 2011;1.
+Available from: \url{https://doi.org/10.1038/srep00170}
 
 \hypertarget{ref-Sears2014}{}
-20. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
-and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
-doi:\href{https://doi.org/10.1016/j.chom.2014.02.007}{10.1016/j.chom.2014.02.007}.
+22. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
+Host \& Microbe {[}Internet{]}. Elsevier BV; 2014;15:317--28. Available
+from: \url{https://doi.org/10.1016/j.chom.2014.02.007}
 
 \hypertarget{ref-Mima2016}{}
-21. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
-\textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
-\textbf{Silva A da}, \textbf{Gu M}, \textbf{Li W}, \textbf{Hamada T},
-\textbf{Kosumi K}, \textbf{Hanyuda A}, \textbf{Liu L}, \textbf{Kostic
-AD}, \textbf{Giannakis M}, \textbf{Bullman S}, \textbf{Brennan CA},
-\textbf{Milner DA}, \textbf{Baba H}, \textbf{Garraway LA},
-\textbf{Meyerhardt JA}, \textbf{Garrett WS}, \textbf{Huttenhower C},
-\textbf{Meyerson M}, \textbf{Giovannucci EL}, \textbf{Fuchs CS},
-\textbf{Nishihara R}, \textbf{Ogino S}. 2016. Fusobacterium nucleatum in
-colorectal carcinoma tissue according to tumor location. Clinical and
-Translational Gastroenterology \textbf{7}:e200.
-doi:\href{https://doi.org/10.1038/ctg.2016.53}{10.1038/ctg.2016.53}.
+23. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
+Fusobacterium nucleatum in colorectal carcinoma tissue according to
+tumor location. Clinical and Translational Gastroenterology
+{[}Internet{]}. Springer Nature; 2016;7:e200. Available from:
+\url{https://doi.org/10.1038/ctg.2016.53}
+
+\hypertarget{ref-Tahara_2014}{}
+24. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
+al. Fusobacterium in colonic flora and molecular features of colorectal
+carcinoma. Cancer Research {[}Internet{]}. American Association for
+Cancer Research (AACR); 2014;74:1311--8. Available from:
+\url{https://doi.org/10.1158\%2F0008-5472.can-13-1865}
 
 \hypertarget{ref-Whitmore2014}{}
-22. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
-cancer. PLoS Pathogens \textbf{10}:e1003933.
-doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.
+25. Whitmore SE, Lamont RJ. Oral bacteria and cancer. Goldman WE,
+editor. PLoS Pathogens {[}Internet{]}. Public Library of Science (PLoS);
+2014;10:e1003933. Available from:
+\url{https://doi.org/10.1371/journal.ppat.1003933}
 
 \hypertarget{ref-Flynn2016}{}
-23. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
-Metabolic and community synergy of oral bacteria in colorectal cancer.
-mSphere \textbf{1}:e00102--16.
-doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
+26. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
+oral bacteria in colorectal cancer. McMahon K, editor. mSphere
+{[}Internet{]}. American Society for Microbiology; 2016;1:e00102--16.
+Available from: \url{https://doi.org/10.1128/msphere.00102-16}
 
 \hypertarget{ref-Dharmani2011}{}
-24. \textbf{Dharmani P}, \textbf{Strauss J}, \textbf{Ambrose C},
-\textbf{Allen-Vercoe E}, \textbf{Chadee K}. 2011. Fusobacterium
-nucleatum infection of colonic cells stimulates MUC2 mucin and tumor
-necrosis factor alpha. Infection and Immunity \textbf{79}:2597--2607.
-doi:\href{https://doi.org/10.1128/iai.05118-11}{10.1128/iai.05118-11}.
+27. Dharmani P, Strauss J, Ambrose C, Allen-Vercoe E, Chadee K.
+Fusobacterium nucleatum infection of colonic cells stimulates MUC2 mucin
+and tumor necrosis factor alpha. Infection and Immunity {[}Internet{]}.
+American Society for Microbiology; 2011;79:2597--607. Available from:
+\url{https://doi.org/10.1128/iai.05118-11}
 
 \hypertarget{ref-Ohkusa2003}{}
-25. \textbf{Ohkusa T}. 2003. Induction of experimental ulcerative
-colitis by fusobacterium varium isolated from colonic mucosa of patients
-with ulcerative colitis. Gut \textbf{52}:79--83.
-doi:\href{https://doi.org/10.1136/gut.52.1.79}{10.1136/gut.52.1.79}.
+28. Ohkusa T. Induction of experimental ulcerative colitis by
+fusobacterium varium isolated from colonic mucosa of patients with
+ulcerative colitis. Gut {[}Internet{]}. BMJ; 2003;52:79--83. Available
+from: \url{https://doi.org/10.1136/gut.52.1.79}
 
 \hypertarget{ref-Ohkusa2002}{}
-26. \textbf{Ohkusa T}, \textbf{Sato N}, \textbf{Ogihara T},
-\textbf{Morita K}, \textbf{Ogawa M}, \textbf{Okayasu I}. 2002.
+29. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
 Fusobacterium varium localized in the colonic mucosa of patients with
 ulcerative colitis stimulates species-specific antibody. Journal of
-Gastroenterology and Hepatology \textbf{17}:849--853.
-doi:\href{https://doi.org/10.1046/j.1440-1746.2002.02834.x}{10.1046/j.1440-1746.2002.02834.x}.
+Gastroenterology and Hepatology {[}Internet{]}. Wiley-Blackwell;
+2002;17:849--53. Available from:
+\url{https://doi.org/10.1046/j.1440-1746.2002.02834.x}
 
 \hypertarget{ref-Shobar2016}{}
-27. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
-\textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
-\textbf{Losurdo J}, \textbf{Engen PA}, \textbf{Sun Y}, \textbf{Koenig
-L}, \textbf{Mutlu EA}. 2016. The effects of bowel preparation on
-microbiota-related metrics differ in health and in inflammatory bowel
-disease and for the mucosal and luminal microbiota compartments.
-Clinical and Translational Gastroenterology \textbf{7}:e143.
-doi:\href{https://doi.org/10.1038/ctg.2015.54}{10.1038/ctg.2015.54}.
+30. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
+JE, et al. The effects of bowel preparation on microbiota-related
+metrics differ in health and in inflammatory bowel disease and for the
+mucosal and luminal microbiota compartments. Clinical and Translational
+Gastroenterology {[}Internet{]}. Springer Nature; 2016;7:e143. Available
+from: \url{https://doi.org/10.1038/ctg.2015.54}
 
 \hypertarget{ref-Kozich2013}{}
-28. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq illumina sequencing platform.
-Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+31. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+Development of a dual-index sequencing strategy and curation pipeline
+for analyzing amplicon sequence data on the MiSeq illumina sequencing
+platform. Applied and Environmental Microbiology {[}Internet{]}.
+American Society for Microbiology; 2013;79:5112--20. Available from:
+\url{https://doi.org/10.1128/aem.01043-13}
 
 \hypertarget{ref-Schloss2009}{}
-29. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-source, platform-independent,
+32. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+EB, et al. Introducing mothur: Open-source, platform-independent,
 community-supported software for describing and comparing microbial
-communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+communities. Applied and Environmental Microbiology {[}Internet{]}.
+American Society for Microbiology; 2009;75:7537--41. Available from:
+\url{https://doi.org/10.1128/aem.01541-09}
 
 \hypertarget{ref-Wang2007}{}
-30. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
-\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
-of rRNA sequences into the new bacterial taxonomy. Applied and
-Environmental Microbiology \textbf{73}:5261--5267.
-doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
+33. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive bayesian classifier
+for rapid assignment of rRNA sequences into the new bacterial taxonomy.
+Applied and Environmental Microbiology {[}Internet{]}. American Society
+for Microbiology; 2007;73:5261--7. Available from:
+\url{https://doi.org/10.1128/aem.00062-07}
 
-\hypertarget{ref-Liaw2002}{}
-31. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
-regression by randomForest. R News: The Newsletter of the R Project
-\textbf{2}:18--22.
+\hypertarget{ref-Calle_2011}{}
+34. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
+for genomic profiling with random forest. Human Heredity {[}Internet{]}.
+S. Karger AG; 2011;72:121--32. Available from:
+\url{https://doi.org/10.1159\%2F000330778}
+
+\hypertarget{ref-Robin_2011}{}
+35. Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez J-C, et
+al. pROC: An open-source package for r and s to analyze and compare ROC
+curves. BMC Bioinformatics {[}Internet{]}. Springer Nature; 2011;12:77.
+Available from: \url{https://doi.org/10.1186\%2F1471-2105-12-77}
 

### 2017-11-02 submission/manuscript.Rmd
made all text and most figure edits

@@
 
-Figures:
+Fig.s:
 
-Supplementary figures:
+Supplementary Fig.s:
 
@@
 
-**Background and Aims:** The microbiome has been implicated in the development of colorectal cancer (CRC) and inflammatory bowel diseases (IBD).
+The microbiome has been implicated in the development of colorectal cancer (CRC) and inflammatory bowel diseases (IBD).
 The specific traits of these diseases vary along the axis of the digestive tract.
@@
 Here we profiled the microbiota of the healthy proximal and distal mucosa and lumen to better understand how bacterial populations vary along the colon.
-
-**Methods:** We used a two-colonoscope approach to sample proximal and distal mucosal and luminal contents from the colons of 20 healthy subjects that had not undergone any bowel preparation procedure.
+We used a two-colonoscope approach to sample proximal and distal mucosal and luminal contents from the colons of 20 healthy subjects that had not undergone any bowel preparation procedure.
 The biopsies and home-collected stool were subjected to 16S rRNA gene sequencing and Random Forest classification models were built using taxa abundance and location to identify microbiota specific to each site.
-
-**Results:** : The right mucosa and lumen had the most similar community structures of the five sites we considered from each subject.
+The right mucosa and lumen had the most similar community structures of the five sites we considered from each subject.
 The distal mucosa had higher relative abundance of Finegoldia, Murdochiella, Peptoniphilus, Porphyromonas and Anaerococcus.
@@
 Separating proximal and distal luminal samples proved more challenging (AUC=0.580) and specific microbiota that differentiated the two were hard to identify.
-
-**Conclusions:** By sampling the unprepped colon, we identified distinct bacterial populations native to the proximal and distal sides.
+By sampling the unprepped colon, we identified distinct bacterial populations native to the proximal and distal sides.
 Further investigation of these bacteria may elucidate if and how these groups contribute to different disease processes on their respective sides of the colon.
@@
 
-**Keywords**: microbiome; colon cancer; proximal and distal colon
+Keywords: microbiome; colon cancer; proximal and distal colon
 
@@
 
+### Methods
+#### Human subjects
+The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number HUM00082721.
+Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study.
+Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl or Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive.
+20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure.
+There were 13 female and 7 male subjects ranging in age from 25 to 64.
+
+#### Sample collection
+At a baseline visit, subjects gave consent and were given a home collection stool kit (Zymo).
+At least one week prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice.
+Notably, subjects did not undergo any bowel preparation method prior to sampling.
+On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System.
+Subjects were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate.
+A flexible sigmoidoscope was first inserted about 25cm into the colon and jumbo biopsy forceps used to collect the luminal contents.
+Two luminal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen.
+The forceps were withdrawn and new biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
+Three mucosal biopsies were collected and flash-frozen in RNAlater.
+These samples comprised the distal colon samples.
+The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the proximal colon.
+Samples were then collected in the same manner as was done in the distal colon and the colonoscope withdrawn.
+All samples were stored at -80$^\circ$C.
+
+#### Sample processing, sequencing and analysis
+DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories).
+For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples.
+The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013.
+Sequences were curated using the mothur software as described previously @Schloss2009.
+The sequences were assigned a taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
+Sequencing and analysis of a mock community revealed the error rate to be 0.018%.
+Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.
+
+Diversity analysis was performed using the Simpson diversity calculator and $\theta$<sub>YC</sub> calculator metrics in mothur version 1.39.5 @Schloss2009.
+$\theta$<sub>YC</sub> distances were calculated to determine the dissimilarity between two samples.
+Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011.
+For each model the data was split into a 19-subject training set and a 1-subject test set.
+The model was built and cross-validated using AUCRFcv on the training set.
+The model was then tested on the left-out patient.
+This process was repeated iteratively for all subjects and results plotted as Reciever Operator Characteristic curves using the pROC R package @Robin_2011.
+Resultant models were used to identify the OTUs that were most important for classifying each location.
+Species-level information for sequences of interest was obtained by aligning the sequences to the GenBank nucleotide databse using blastn.
+The species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.
+
+#### Statistical analysis
+Differences in community membership at the phyla level were tested using the analysis of molecular variance (AMOVA) metric in mothur.
+Differences in $\theta$<sub>YC</sub> distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
+
+#### Data availability
+16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number XXXX.
+A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017.
+
 ### Results
@@
 #### Microbial membership and diversity of the proximal and distal colon 
-Luminal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1).
+Luminal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Fig.
+1).
 Subjects also collected stool at home one week prior to the procedure.
 To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on DNA extracted from each sample.
-As expected, each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A) @LloydPrice2016.
-Samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B).
+As expected, each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Fig.
+2A) @LloydPrice2016.
+Samples had varying levels of diversity at each site, irrespective of the individual (Fig.
+2B).
 For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others.
@@
 To compare similarity between the proximal and distal sides and within the lumen and mucosa, we compared the community structure of these sites based on the relative abundances of individual Operational Taxonomic Units (OTUs).
-Across all subjects we observed wide variation when comparing sample locations (Figure 3A).
+Across all subjects we observed wide variation when comparing sample locations (Fig.
+3A).
 Those ranges did not follow a clear pattern on an individual basis.
@@
 Next, we compared the luminal and mucosal samples to the fecal sample of each subject.
-Amidst the large inter-subject variation, we did identify significantly less dissimilarity between the distal luminal sample and the feces (Figure 3B, P < 0.05, Wilcoxon, BH adjustment).
+Amidst the large inter-subject variation, we did identify significantly less dissimilarity between the distal luminal sample and the feces (Fig.
+3B, P < 0.05, Wilcoxon, BH adjustment).
 Furthermore, there was an even larger difference in the communities found in the distal mucosa compared to the fecal communities, indicating that the mucosa is as different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment).
@@
 To determine what factors may have driven the differences seen among the samples, we compared the community dissimilarity between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal).
-We found that samples from one individual were far more similar to each other than to matched samples from the other subjects (Figure 3C); this is consistent with previous human microbiome studies that have sampled multiple sites of the human colon [@Eckburg2005; @deCarcer2010; @Zhang2013].
+We found that samples from one individual were far more similar to each other than to matched samples from the other subjects (Fig.
+3C); this is consistent with previous human microbiome studies that have sampled multiple sites of the human colon [@Eckburg2005; @deCarcer2010; @Zhang2013].
 Thus interpersonal variation drove the differences between samples more than whether the sample came from the proximal or distal side of the colon or from the lumen or mucosa.
@@
 To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances.
-We used 10-fold cross validation to build the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Figure S1A).
+We used 10-fold cross validation to build the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Fig.
+4A).
 The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908).
 The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
-For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ were included in the best model (Figure 4A).
-The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ (Figure 4B).
+For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ were included in the best model (Fig.
+5A).
+The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ (Fig.
+5B).
 These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.
@@
 Next, we built a Random Forest model to differentiate the proximal and distal luminal samples using 10-fold cross validation.
-The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = 0.580).
-The OTUs included in the model differentiating the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ (Figure 5A).
-The model that attempted to separate the the proximal and distal lumen included OTUs affiliated with the _Bacteroides_, _Clostridium IV_ and _Oscillibacter_ (Figure 5B).
+The model performed best when distinguishing the proximal versus distal mucosa (Fig.
+4B, AUC = 0.850) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = 0.580).
+The OTUs included in the model differentiating the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ (Fig.
+6A).
+The model that attempted to separate the the proximal and distal lumen included OTUs affiliated with the _Bacteroides_, _Clostridium IV_ and _Oscillibacter_ (Fig.
+6B).
 Interestingly, _Anaerococcus_ and _Finegoldia_ were distinct between the mucosa and lumen and also helped to differentiate between the proximal and distal sides.
@@
 nucleatum subsp.
-animalis_, four of the samples were from the proximal mucosa and three from the distal mucosa (Figure S2A).
+animalis_, four of the samples were from the proximal mucosa and three from the distal mucosa (Supplementary Fig.
+S1A).
 The second most frequent _Fusobacterium_ sequence was OTU472, which aligned with 99% identity to _F.
@@
 Four subjects harbored _F.
-varium_ and the samples were split evenly between the proximal and distal mucosa (Figure S2B).
+varium_ and the samples were split evenly between the proximal and distal mucosa (Supplementary Fig.
+S1B).
 OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (99% over full length), another bacterium commonly detected and isolated from colorectal tumors.
-OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure S2C).
+OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Supplementary Fig.
+S1C).
 Among the 11 distal mucosa samples that were positive for _P.
@@
 _F.
-varium_ was only detected in three subjects and two of those samples were isolated form the proximal mucosa (Figure S2B).
+varium_ was only detected in three subjects and two of those samples were isolated form the proximal mucosa (Supplementary Fig.
+S1B).
 _F.
@@
 
-### Methods
-#### Human subjects
-The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number HUM00082721.
-Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study.
-Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl or Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive.
-20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure.
-There were 13 female and 7 male subjects ranging in age from 25 to 64.
+\newpage
 
-#### Sample collection
-At a baseline visit, subjects gave consent and were given a home collection stool kit (Zymo).
-At least one week prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice.
-Notably, subjects did not undergo any bowel preparation method prior to sampling.
-On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System.
-Subjects were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate.
-A flexible sigmoidoscope was first inserted about 25cm into the colon and jumbo biopsy forceps used to collect the luminal contents.
-Two luminal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen.
-The forceps were withdrawn and new biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
-Three mucosal biopsies were collected and flash-frozen in RNAlater.
-These samples comprised the distal colon samples.
-The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the proximal colon.
-Samples were then collected in the same manner as was done in the distal colon and the colonoscope withdrawn.
-All samples were stored at -80$^\circ$C.
+### References
 
-#### Sample processing, sequencing and analysis
-DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories).
-For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples.
-The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013.
-Sequences were curated using the mothur software as described previously @Schloss2009.
-The sequences were assigned a taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff.
-Sequencing and analysis of a mock community revealed the error rate to be 0.018%.
-Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.
-
-Diversity analysis was performed using the Simpson diversity calculator and $\theta$<sub>YC</sub> calculator metrics in mothur version 1.39.5 @Schloss2009.
-$\theta$<sub>YC</sub> distances were calculated to determine the dissimilarity between two samples.
-Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011.
-For each model the data was split into a 19-subject training set and a 1-subject test set.
-The model was built and cross-validated using AUCRFcv on the training set.
-The model was then tested on the left-out patient.
-This process was repeated iteratively for all subjects and results plotted as Reciever Operator Characteristic curves using the pROC R package @Robin_2011.
-Resultant models were used to identify the OTUs that were most important for classifying each location.
-Species-level information for sequences of interest was obtained by aligning the sequences to the GenBank nucleotide databse using blastn.
-The species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.
-
-#### Statistical analysis
-Differences in community membership at the phyla level were tested using the analysis of molecular variance (AMOVA) metric in mothur.
-Differences in $\theta$<sub>YC</sub> distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
-
-#### Data availability
-16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number XXXX.
-A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017.
+<div id="refs"></div>
 
@@
 ####Figure 1
-
 Sampling strategy.
@@
 
+![Fig 1](../submission/fig1.jpg)
+
 \newpage
@@
 The horizontal lines represent the median values.
+
+![Fig 2](../submission/figure_2.pdf)
 
@@
 
+![Fig 3](../submission/figure_3.pdf)
+
 \newpage
@@
 ####Figure 4
+Random Forest classifies locations in the colon.
+A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal (red) and proximal (blue) sides of the colon.
+(B) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal mucosa vs proximal mucosa (green) and distal lumen versus proximal lumen (purple).
+
+![Fig 4](../submission/figure_4.pdf)
+
+\newpage
+####Figure 5
 Taxa specific to the distal and proximal sides of the colon.
@@
 
+![Fig 5](../submission/figure_5.pdf)
+
 \newpage
 
-####Figure 5 
+####Figure 6 
 Taxa specific to the distal and proximal mucosa and lumen.
@@
 The vertical lines represent the median values for each OTU.
+
+![Fig 6](../submission/figure_6.pdf)
+
 \newpage
@@
 ####Figure S1
-Random Forest classifies locations in the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal (red) and proximal (blue) sides of the colon.
-(B) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal mucosa vs proximal mucosa (green) and distal lumen versus proximal lumen (purple).
-
-\newpage
-
-####Figure S2
 Location and relative abundance of cancer-associated OTUs.
@@
 animalis_ (B) _Fusobacterium varium_ and (C) _Porphyromonas asacharolytica_
+![Fig S1](../submission/figure_S1.pdf)
 
-### References
+\newpage

### 2017-11-02 submission/manuscript.tex
made all text and most figure edits

@@
 
-2. Pennsylvania State University, Hershey, Pennsylvania ??
+2. Department of Family and Community Medicine, College of Medicine,
+Pennsylvania State University, Hershey, Pennsylvania 17033
 
@@
 University of Michigan Medical School, Ann Arbor, Michigan
+
+Grant support: Rose and Lawrence C. Page Foundation
+
+Disclosures: nothing to disclose
+
+Word count:
+
+Fig.s:
+
+Supplementary Fig.s:
 
@@
 
-\textbf{Background and Aims:} The microbiome has been implicated in the
-development of colorectal cancer (CRC) and inflammatory bowel diseases
-(IBD). The specific traits of these diseases vary along the axis of the
-digestive tract. Further, variation in the structure of the gut
-microbiota has been associated with both diseases. Here we profiled the
-microbiota of the healthy proximal and distal mucosa and lumen to better
-understand how bacterial populations vary along the colon.
-
-\textbf{Methods:} We used a two-colonoscope approach to sample proximal
-and distal mucosal and luminal contents from the colons of 20 healthy
-subjects that had not undergone any bowel preparation procedure. The
-biopsies and home-collected stool were subjected to 16S rRNA gene
-sequencing and Random Forest classification models were built using taxa
-abundance and location to identify microbiota specific to each site.
-
-\textbf{Results:} : The right mucosa and lumen had the most similar
-community structures of the five sites we considered from each subject.
-The distal mucosa had higher relative abundance of Finegoldia,
-Murdochiella, Peptoniphilus, Porphyromonas and Anaerococcus. The
-proximal mucosa had more of the genera Enterobacteriaceae, Bacteroides
-and Pseudomonas. The classification model performed well when
-classifying mucosal samples into proximal or distal sides (AUC=0.850).
-Separating proximal and distal luminal samples proved more challenging
-(AUC=0.580) and specific microbiota that differentiated the two were
-hard to identify.
-
-\textbf{Conclusions:} By sampling the unprepped colon, we identified
-distinct bacterial populations native to the proximal and distal sides.
-Further investigation of these bacteria may elucidate if and how these
-groups contribute to different disease processes on their respective
-sides of the colon.
+The microbiome has been implicated in the development of colorectal
+cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits
+of these diseases vary along the axis of the digestive tract. Further,
+variation in the structure of the gut microbiota has been associated
+with both diseases. Here we profiled the microbiota of the healthy
+proximal and distal mucosa and lumen to better understand how bacterial
+populations vary along the colon. We used a two-colonoscope approach to
+sample proximal and distal mucosal and luminal contents from the colons
+of 20 healthy subjects that had not undergone any bowel preparation
+procedure. The biopsies and home-collected stool were subjected to 16S
+rRNA gene sequencing and Random Forest classification models were built
+using taxa abundance and location to identify microbiota specific to
+each site. The right mucosa and lumen had the most similar community
+structures of the five sites we considered from each subject. The distal
+mucosa had higher relative abundance of Finegoldia, Murdochiella,
+Peptoniphilus, Porphyromonas and Anaerococcus. The proximal mucosa had
+more of the genera Enterobacteriaceae, Bacteroides and Pseudomonas. The
+classification model performed well when classifying mucosal samples
+into proximal or distal sides (AUC=0.850). Separating proximal and
+distal luminal samples proved more challenging (AUC=0.580) and specific
+microbiota that differentiated the two were hard to identify. By
+sampling the unprepped colon, we identified distinct bacterial
+populations native to the proximal and distal sides. Further
+investigation of these bacteria may elucidate if and how these groups
+contribute to different disease processes on their respective sides of
+the colon.
 
@@
 
-\textbf{Keywords}: microbiome; colon cancer; proximal and distal colon
+Keywords: microbiome; colon cancer; proximal and distal colon
 
@@
 
-\subsubsection{Results}\label{results}
-
-\paragraph{Microbial membership and diversity of the proximal and distal
-colon}\label{microbial-membership-and-diversity-of-the-proximal-and-distal-colon}
-
-Luminal and mucosal samples were collected from the proximal and distal
-colon of 20 healthy humans that had not undergone bowel preparation
-(Figure 1). Subjects also collected stool at home one week prior to the
-procedure. To characterize the bacterial communities present at these
-sites, 16S rRNA gene sequencing was performed on DNA extracted from each
-sample. As expected, each site was primarily dominated by
-\emph{Firmicutes} and \emph{Bacteriodetes} (Figure 2A) (14). Samples had
-varying levels of diversity at each site, irrespective of the individual
-(Figure 2B). For example, the proximal mucosa was more diverse than the
-distal for some individuals while the opposite was true for others.
-Therefore we could not identify a clear pattern of changes in microbial
-diversity along the gut axis.
-
-To compare similarity between the proximal and distal sides and within
-the lumen and mucosa, we compared the community structure of these sites
-based on the relative abundances of individual Operational Taxonomic
-Units (OTUs). Across all subjects we observed wide variation when
-comparing sample locations (Figure 3A). Those ranges did not follow a
-clear pattern on an individual basis. However, when comparing median
-dissimilarity between the communities found in the proximal lumen and
-mucosa, the proximal and distal lumen, the proximal and distal mucosa,
-and the distal lumen and mucosa, we found that the proximal lumen and
-mucosa were most similar to each other than to the other samples (P
-\textless{} 0.005, Wilcoxon, BH adjustment).
-
-\paragraph{Fecal samples resemble luminal samples from the distal
-colon}\label{fecal-samples-resemble-luminal-samples-from-the-distal-colon}
-
-Next, we compared the luminal and mucosal samples to the fecal sample of
-each subject. Amidst the large inter-subject variation, we did identify
-significantly less dissimilarity between the distal luminal sample and
-the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH adjustment).
-Furthermore, there was an even larger difference in the communities
-found in the distal mucosa compared to the fecal communities, indicating
-that the mucosa is as different from the stool as compared to lumen (P
-\textless{} 0.0005, Wilcoxon, BH adjustment). These results suggest that
-the contents of the distal lumen were most representative of the
-subjects' feces, and the mucosal microbiota are distinct from the fecal
-and luminal communities.
-
-\paragraph{Interpersonal community variation is greater than the
-variation between
-sites}\label{interpersonal-community-variation-is-greater-than-the-variation-between-sites}
-
-To determine what factors may have driven the differences seen among the
-samples, we compared the community dissimilarity between samples from
-all subjects (interpersonal) versus samples from within one subject
-(intrapersonal). We found that samples from one individual were far more
-similar to each other than to matched samples from the other subjects
-(Figure 3C); this is consistent with previous human microbiome studies
-that have sampled multiple sites of the human colon (15--17). Thus
-interpersonal variation drove the differences between samples more than
-whether the sample came from the proximal or distal side of the colon or
-from the lumen or mucosa.
-
-\paragraph{Random Forest classification models identify important
-Operational Taxonomic Units (OTUs) on each
-side}\label{random-forest-classification-models-identify-important-operational-taxonomic-units-otus-on-each-side}
-
-To identify OTUs that were distinct at each site, we constructed several
-Random Forest models trained using OTU relative abundances. We used
-10-fold cross validation to build the first model to classify the
-luminal versus mucosal samples for the proximal and distal sides,
-independently (Figure S1A). The models performed well when classifying
-these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
-were most predictive of each site were identified by their greatest mean
-decrease in accuracy when removed from the model. For distinguishing the
-proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
-\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
-included in the best model (Figure 4A). The model to differentiate
-between the distal lumen and mucosa included OTUs affiliated with the
-\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
-\emph{Anaerococcus} (Figure 4B). These results indicated that there were
-fine differences between the different sites of the colon, and that
-these could be traced to specific OTUs on each side.
-
-Next, we built a Random Forest model to differentiate the proximal and
-distal luminal samples using 10-fold cross validation. The model
-performed best when distinguishing the proximal versus distal mucosa
-(Figure S1B, AUC = 0.850) whereas the model to differentiate between the
-proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
-included in the model differentiating the distal and proximal mucosa
-included members of the \emph{Porphyromonas}, \emph{Murdochiella},
-\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Figure
-5A). The model that attempted to separate the the proximal and distal
-lumen included OTUs affiliated with the \emph{Bacteroides},
-\emph{Clostridium IV} and \emph{Oscillibacter} (Figure 5B).
-Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were distinct
-between the mucosa and lumen and also helped to differentiate between
-the proximal and distal sides.
-
-\paragraph{Bacterial OTUs associated with CRC and IBD are found in
-healthy
-individuals}\label{bacterial-otus-associated-with-crc-and-ibd-are-found-in-healthy-individuals}
-
-Given that specific bacterial species have been associated with
-colorectal cancer and IBD, we probed our sample set for these OTUs.
-Among our 100 samples, the most frequent sequence associated with the
-\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
-\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
-This is the only species of \emph{Fusobacterium} known to have oncogenic
-properties and be found on the surfaces of colorectal cancer tumors
-(18). There were 14 samples from 8 subjects with the \emph{F. nucleatum
-subsp. animalis} sequences. Of the samples with the highest relative
-abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
-were from the proximal mucosa and three from the distal mucosa (Figure
-S2A). The second most frequent \emph{Fusobacterium} sequence was OTU472,
-which aligned with 99\% identity to \emph{F. varium}. In addition to
-\emph{F. nucleatum}, \emph{F. varium} has been associated with IBD (19).
-Four subjects harbored \emph{F. varium} and the samples were split
-evenly between the proximal and distal mucosa (Figure S2B). OTU152 was
-similar to the members of the \emph{Porphyromonas} genus and the most
-frequent sequence in that OTU aligned to \emph{Porphyromonas
-asacharolytica} (99\% over full length), another bacterium commonly
-detected and isolated from colorectal tumors. OTU152 was only detected
-on the distal mucosa, and in fact was one of the OTUs the classification
-model identified as separating distal and proximal sides (Figure S2C).
-Among the 11 distal mucosa samples that were positive for \emph{P.
-asacharolytica}, the relative abundances for this OTU ranged from 0.01\%
-to 16\%. Thus, disease-associated OTUs could be found in our sample set
-of 20 healthy individuals.
-
-\subsubsection{Discussion}\label{discussion}
-
-Here we identified bacterial taxa that were specific to the lumen and
-mucosa of the proximal and distal sides of the human colon using samples
-collected during an unprepared colonoscopy of healthy subjects. We found
-that all locations contained a range of phyla relative abundances and a
-range of diversity, but that there was a wide variability between
-subjects. Pairwise comparisons of each of the sites revealed that the
-proximal mucosa and lumen were most similar to each other. Further,
-comparison of colonoscopy-collected samples with fecal samples
-demonstrated that the distal lumen was most similar to feces. Random
-Forest models built using OTU relative abundances from each sample
-identified microbiota that were particular to each location of the
-colon. Finally, we were able to detect some bacterial OTUs associated
-with colonic disease in our healthy cohort. Using unprepped
-colonoscopies and machine learning, we have identified bacterial taxa
-specific to the healthy proximal and distal human colon.
-
-When examining the relative abundance of the dominant phyla at each site
-(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
-of variation. This likely reflects not only the variability between
-human subjects, casued by differences in age, sex, diet, but also
-spatial ``patchiness'' in the gut microbiome. One study noted that the
-bacteria recoverable from the same mucosal sample location can be vastly
-different when the samples are taken just 1 cm away from each other
-(20). Similar patchiness was also observed in luminal contents and fecal
-samples themselves; there was separation of different interacting
-microbes along the length of a stool sample, for instance (21). That
-said, across our samples, the mucosal samples harbor more
-\emph{Proteobacteria}, consistent with previous studies comparing
-mucosal swabs to luminal content in humans (4). Hence, the conclusions
-we were able to draw from phyla analysis may have been impacted by
-inter-subject patchiness.
-
-To get around the noisiness from a diverse set of samples, we built
-Random Forest classification models to identify the microbiota that were
-specific to each side and in the lumen and mucosa. For each comparison
-we identified the top five OTUs that were strongly predictive of one
-site or another. Generally, OTUs identified in each location were
-consistent with known physiological gradients along the gut axis (5).
-For instance, the proximal mucosa contains the highest oxygen
-concentrations of the colon and harbored mucosa-associated facultative
-anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
-aerobic \emph{Psuedomonas}. The distal mucosa was far more likely to
-host strictly anaerobic species such as \emph{Porphyromonas},
-\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. Thus
-the gut microenvironment of each location likely enriches for these
-specific microbiota.
-
-In addition to identifying features that are specific to each side of
-the gut, the ability of the Random Forest to classify samples can serve
-as a proxy for similarity. That is, a higher AUC value indicates the
-samples are more efficiently classfied (and thus more different) than a
-model with a lower AUC value. For instance, the model separating the
-proximal and distal mucosa had an AUC of 0.850 whereas the model for
-classifying the proximal and distal lumen had a much lower AUC of 0.580.
-Further, the latter model required 44 OTUs to best separate the samples
-whereas the models separating the mucosa only needed 10 OTUs. The much
-lower AUC and need for a high number of features compared to other
-models suggest these locations are the most similar of the comparisons
-tested. We speculate that the model was less effective at classifying
-the proximal and distal luminal contents because the mucosal
-microenvironments have variable selective pressure along the colon than
-the luminal microenvironments.
-
-We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
-of our subjects, respectively. These bacteria have been shown to be
-predictive of colorectal cancer in humans (9) and have oncogenic
-properties in cell culture and in mice (22). Interestingly, while
-\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
-asacharolytica} was only detected in the distal mucosa. Not much is
-known about the distribution of \emph{P. asacharolytica} along the
-colon, but given its anaerobic lifestyle and asacharolytic metabolism,
-it is not surprising that it resides in the less-oygen-rich and
-protein-rich distal mucosa (4). In studies examining bacteria on
-colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
-on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreased along the colon to rectum (23). In another study,
-\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
-mutations, molecular features of proximal CRC (24). Of the 8 (40\%)
-individuals positive for \emph{F. nucleatum} in the present study, the
-bacterium was spread across the proximal mucosa, distal lumen and distal
-mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
-suggests that \emph{Fusobacteria} species are more commonly found on
-proximal tumors and in biofilms, indicating that it is not only the
-presence of the bacterium but the structure of the tumor community that
-contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
-Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
-only co-occur on CRC tumors but also to synergize to promote
-tumorigenesis in an oral cancer model (25) (26). Further analysis of the
-distribution and activities of these pathogens along the colon is needed
-to elucidate a mechanism for development of CRC or IBD subtypes in the
-proximal or distal colon.
-
-The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
-been commonly isolated from mucosal biopsies of patients with IBD (19).
-Laboratory experiments with these isolates have shown that
-disease-isolated \emph{F. nucleatum} are more invasive and stimulate
-more TNF-\(\alpha\) production than strains from healthy individuals
-(10), suggesting the bacteria may increase inflammation in the gut as
-well (27). \emph{F. varium} isolated from UC patients caused colonic
-ulcers in an experimental mouse model (28). \emph{F. varium} was only
-detected in three subjects and two of those samples were isolated form
-the proximal mucosa (Figure S2B). \emph{F. varium} is most commonly
-isolated from UC patient biopsies from the ileum or cecum (29),
-suggesting this species may exhibit preference for the different
-environmental conditions of these gastrointestinal sites. Further work
-will assess how gut environment may select for species which may then
-cause localized disease.
-
-Specific comparisons of our findings to previously published studies of
-spatial variation are confounded by the use of bowel preparation
-methods. A rare report of a matched-colonoscopy study sampled 18
-patient's colonic mucosa and luminal contents prior to and after bowel
-cleansing (30). This study found that mucosal and luminal samples were
-distinguishable prior to bowel cleansing, but that bowel preparation
-resulted in an increase in shared OTUs between each site (30). After
-seven days, bowel cleansing not only made the samples more difficult to
-distinguish, but it also decreased the diversity observed across sites.
-Bowel preparation clearly biases the representation of microbiota
-recovered from sampling the lumen or mucosa.
-
-By revealing specific differences in microbial populations at each
-location in the gut via sampling an unprepared bowel, we can begin to
-form hypotheses about how specific host-microbe interactions can affect
-disease progression of proximal and distal CRC and IBD subtypes. A
-better understanding of microbial activities in the gut can enhance
-microbiome-based screening and treatment modalities for these colon
-diseases.
-
-\subsubsection{Acknowledgments}\label{acknowledgments}
-
-We thank all the individuals who volunteered for the study. This work
-was supported by the Rose and Lawrence C. Page Foundation (DKT). We
-would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman
-for their roles in study coordination, subject recruitment, sample
-collection and sample processing.
-
 \subsubsection{Methods}\label{methods}
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described (31). Sequences were curated
-using the mothur software as described previously (32). The sequences
+Illumina MiSeq as previously described (14). Sequences were curated
+using the mothur software as described previously (15). The sequences
 were assigned a taxonomic classification using a naive Bayesian
 classifier trained using a 16S rRNA gene training set from the Ribosomal
-Database Project (RDP) (33) and clustered into operational taxonomic
+Database Project (RDP) (16) and clustered into operational taxonomic
 units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
@@
 Diversity analysis was performed using the Simpson diversity calculator
-and \(\theta\)YC calculator metrics in mothur version 1.39.5 (32).
+and \(\theta\)YC calculator metrics in mothur version 1.39.5 (15).
 \(\theta\)YC distances were calculated to determine the dissimilarity
 between two samples. Random Forest classification models were built
-using the AUCRF R package using a leave-one-subject out approach (34).
+using the AUCRF R package using a leave-one-subject out approach (17).
 For each model the data was split into a 19-subject training set and a
@@
 results plotted as Reciever Operator Characteristic curves using the
-pROC R package (35). Resultant models were used to identify the OTUs
+pROC R package (18). Resultant models were used to identify the OTUs
 that were most important for classifying each location. Species-level
@@
 
-\newpage
+\subsubsection{Results}\label{results}
 
-\subsubsection{Figures}\label{figures}
+\paragraph{Microbial membership and diversity of the proximal and distal
+colon}\label{microbial-membership-and-diversity-of-the-proximal-and-distal-colon}
 
-\paragraph{Figure 1}\label{figure-1}
+Luminal and mucosal samples were collected from the proximal and distal
+colon of 20 healthy humans that had not undergone bowel preparation
+(Fig. 1). Subjects also collected stool at home one week prior to the
+procedure. To characterize the bacterial communities present at these
+sites, 16S rRNA gene sequencing was performed on DNA extracted from each
+sample. As expected, each site was primarily dominated by
+\emph{Firmicutes} and \emph{Bacteriodetes} (Fig. 2A) (19). Samples had
+varying levels of diversity at each site, irrespective of the individual
+(Fig. 2B). For example, the proximal mucosa was more diverse than the
+distal for some individuals while the opposite was true for others.
+Therefore we could not identify a clear pattern of changes in microbial
+diversity along the gut axis.
 
-Sampling strategy. A flexible sigmoidoscope was used to sample the
-distal colonic luminal contents and mucosa. The scope was inserted
-\textasciitilde{} 25cm into the subject and biopsy forceps were used to
-sample the luminal contents (D, inset). A separate set of biopsy forceps
-was used to sample the distal mucosa (D, inset). The sigmoidoscope was
-removed. A pediatric colonoscope was inserted and used to access the
-proximal colon (P, inset). Biopsies were taken of the proximal luminal
-contents and mucosa as described. One week prior to the procedure stool
-was collected at home and sent into the laboratory. Representative
-images from one individual are shown.
+To compare similarity between the proximal and distal sides and within
+the lumen and mucosa, we compared the community structure of these sites
+based on the relative abundances of individual Operational Taxonomic
+Units (OTUs). Across all subjects we observed wide variation when
+comparing sample locations (Fig. 3A). Those ranges did not follow a
+clear pattern on an individual basis. However, when comparing median
+dissimilarity between the communities found in the proximal lumen and
+mucosa, the proximal and distal lumen, the proximal and distal mucosa,
+and the distal lumen and mucosa, we found that the proximal lumen and
+mucosa were most similar to each other than to the other samples (P
+\textless{} 0.005, Wilcoxon, BH adjustment).
+
+\paragraph{Fecal samples resemble luminal samples from the distal
+colon}\label{fecal-samples-resemble-luminal-samples-from-the-distal-colon}
+
+Next, we compared the luminal and mucosal samples to the fecal sample of
+each subject. Amidst the large inter-subject variation, we did identify
+significantly less dissimilarity between the distal luminal sample and
+the feces (Fig. 3B, P \textless{} 0.05, Wilcoxon, BH adjustment).
+Furthermore, there was an even larger difference in the communities
+found in the distal mucosa compared to the fecal communities, indicating
+that the mucosa is as different from the stool as compared to lumen (P
+\textless{} 0.0005, Wilcoxon, BH adjustment). These results suggest that
+the contents of the distal lumen were most representative of the
+subjects' feces, and the mucosal microbiota are distinct from the fecal
+and luminal communities.
+
+\paragraph{Interpersonal community variation is greater than the
+variation between
+sites}\label{interpersonal-community-variation-is-greater-than-the-variation-between-sites}
+
+To determine what factors may have driven the differences seen among the
+samples, we compared the community dissimilarity between samples from
+all subjects (interpersonal) versus samples from within one subject
+(intrapersonal). We found that samples from one individual were far more
+similar to each other than to matched samples from the other subjects
+(Fig. 3C); this is consistent with previous human microbiome studies
+that have sampled multiple sites of the human colon (20--22). Thus
+interpersonal variation drove the differences between samples more than
+whether the sample came from the proximal or distal side of the colon or
+from the lumen or mucosa.
+
+\paragraph{Random Forest classification models identify important
+Operational Taxonomic Units (OTUs) on each
+side}\label{random-forest-classification-models-identify-important-operational-taxonomic-units-otus-on-each-side}
+
+To identify OTUs that were distinct at each site, we constructed several
+Random Forest models trained using OTU relative abundances. We used
+10-fold cross validation to build the first model to classify the
+luminal versus mucosal samples for the proximal and distal sides,
+independently (Fig. 4A). The models performed well when classifying
+these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
+were most predictive of each site were identified by their greatest mean
+decrease in accuracy when removed from the model. For distinguishing the
+proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
+\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
+included in the best model (Fig. 5A). The model to differentiate between
+the distal lumen and mucosa included OTUs affiliated with the
+\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
+\emph{Anaerococcus} (Fig. 5B). These results indicated that there were
+fine differences between the different sites of the colon, and that
+these could be traced to specific OTUs on each side.
+
+Next, we built a Random Forest model to differentiate the proximal and
+distal luminal samples using 10-fold cross validation. The model
+performed best when distinguishing the proximal versus distal mucosa
+(Fig. 4B, AUC = 0.850) whereas the model to differentiate between the
+proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
+included in the model differentiating the distal and proximal mucosa
+included members of the \emph{Porphyromonas}, \emph{Murdochiella},
+\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Fig.
+6A). The model that attempted to separate the the proximal and distal
+lumen included OTUs affiliated with the \emph{Bacteroides},
+\emph{Clostridium IV} and \emph{Oscillibacter} (Fig. 6B). Interestingly,
+\emph{Anaerococcus} and \emph{Finegoldia} were distinct between the
+mucosa and lumen and also helped to differentiate between the proximal
+and distal sides.
+
+\paragraph{Bacterial OTUs associated with CRC and IBD are found in
+healthy
+individuals}\label{bacterial-otus-associated-with-crc-and-ibd-are-found-in-healthy-individuals}
+
+Given that specific bacterial species have been associated with
+colorectal cancer and IBD, we probed our sample set for these OTUs.
+Among our 100 samples, the most frequent sequence associated with the
+\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
+\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
+This is the only species of \emph{Fusobacterium} known to have oncogenic
+properties and be found on the surfaces of colorectal cancer tumors
+(23). There were 14 samples from 8 subjects with the \emph{F. nucleatum
+subsp. animalis} sequences. Of the samples with the highest relative
+abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
+were from the proximal mucosa and three from the distal mucosa
+(Supplementary Fig. S1A). The second most frequent \emph{Fusobacterium}
+sequence was OTU472, which aligned with 99\% identity to \emph{F.
+varium}. In addition to \emph{F. nucleatum}, \emph{F. varium} has been
+associated with IBD (24). Four subjects harbored \emph{F. varium} and
+the samples were split evenly between the proximal and distal mucosa
+(Supplementary Fig. S1B). OTU152 was similar to the members of the
+\emph{Porphyromonas} genus and the most frequent sequence in that OTU
+aligned to \emph{Porphyromonas asacharolytica} (99\% over full length),
+another bacterium commonly detected and isolated from colorectal tumors.
+OTU152 was only detected on the distal mucosa, and in fact was one of
+the OTUs the classification model identified as separating distal and
+proximal sides (Supplementary Fig. S1C). Among the 11 distal mucosa
+samples that were positive for \emph{P. asacharolytica}, the relative
+abundances for this OTU ranged from 0.01\% to 16\%. Thus,
+disease-associated OTUs could be found in our sample set of 20 healthy
+individuals.
+
+\subsubsection{Discussion}\label{discussion}
+
+Here we identified bacterial taxa that were specific to the lumen and
+mucosa of the proximal and distal sides of the human colon using samples
+collected during an unprepared colonoscopy of healthy subjects. We found
+that all locations contained a range of phyla relative abundances and a
+range of diversity, but that there was a wide variability between
+subjects. Pairwise comparisons of each of the sites revealed that the
+proximal mucosa and lumen were most similar to each other. Further,
+comparison of colonoscopy-collected samples with fecal samples
+demonstrated that the distal lumen was most similar to feces. Random
+Forest models built using OTU relative abundances from each sample
+identified microbiota that were particular to each location of the
+colon. Finally, we were able to detect some bacterial OTUs associated
+with colonic disease in our healthy cohort. Using unprepped
+colonoscopies and machine learning, we have identified bacterial taxa
+specific to the healthy proximal and distal human colon.
+
+When examining the relative abundance of the dominant phyla at each site
+(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
+of variation. This likely reflects not only the variability between
+human subjects, casued by differences in age, sex, diet, but also
+spatial ``patchiness'' in the gut microbiome. One study noted that the
+bacteria recoverable from the same mucosal sample location can be vastly
+different when the samples are taken just 1 cm away from each other
+(25). Similar patchiness was also observed in luminal contents and fecal
+samples themselves; there was separation of different interacting
+microbes along the length of a stool sample, for instance (26). That
+said, across our samples, the mucosal samples harbor more
+\emph{Proteobacteria}, consistent with previous studies comparing
+mucosal swabs to luminal content in humans (4). Hence, the conclusions
+we were able to draw from phyla analysis may have been impacted by
+inter-subject patchiness.
+
+To get around the noisiness from a diverse set of samples, we built
+Random Forest classification models to identify the microbiota that were
+specific to each side and in the lumen and mucosa. For each comparison
+we identified the top five OTUs that were strongly predictive of one
+site or another. Generally, OTUs identified in each location were
+consistent with known physiological gradients along the gut axis (5).
+For instance, the proximal mucosa contains the highest oxygen
+concentrations of the colon and harbored mucosa-associated facultative
+anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
+aerobic \emph{Psuedomonas}. The distal mucosa was far more likely to
+host strictly anaerobic species such as \emph{Porphyromonas},
+\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. Thus
+the gut microenvironment of each location likely enriches for these
+specific microbiota.
+
+In addition to identifying features that are specific to each side of
+the gut, the ability of the Random Forest to classify samples can serve
+as a proxy for similarity. That is, a higher AUC value indicates the
+samples are more efficiently classfied (and thus more different) than a
+model with a lower AUC value. For instance, the model separating the
+proximal and distal mucosa had an AUC of 0.850 whereas the model for
+classifying the proximal and distal lumen had a much lower AUC of 0.580.
+Further, the latter model required 44 OTUs to best separate the samples
+whereas the models separating the mucosa only needed 10 OTUs. The much
+lower AUC and need for a high number of features compared to other
+models suggest these locations are the most similar of the comparisons
+tested. We speculate that the model was less effective at classifying
+the proximal and distal luminal contents because the mucosal
+microenvironments have variable selective pressure along the colon than
+the luminal microenvironments.
+
+We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
+of our subjects, respectively. These bacteria have been shown to be
+predictive of colorectal cancer in humans (9) and have oncogenic
+properties in cell culture and in mice (27). Interestingly, while
+\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
+asacharolytica} was only detected in the distal mucosa. Not much is
+known about the distribution of \emph{P. asacharolytica} along the
+colon, but given its anaerobic lifestyle and asacharolytic metabolism,
+it is not surprising that it resides in the less-oygen-rich and
+protein-rich distal mucosa (4). In studies examining bacteria on
+colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
+on proximal-sided tumors, and distribution of \emph{F. nucleatum}
+decreased along the colon to rectum (28). In another study,
+\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
+mutations, molecular features of proximal CRC (29). Of the 8 (40\%)
+individuals positive for \emph{F. nucleatum} in the present study, the
+bacterium was spread across the proximal mucosa, distal lumen and distal
+mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
+suggests that \emph{Fusobacteria} species are more commonly found on
+proximal tumors and in biofilms, indicating that it is not only the
+presence of the bacterium but the structure of the tumor community that
+contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
+Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
+only co-occur on CRC tumors but also to synergize to promote
+tumorigenesis in an oral cancer model (30) (31). Further analysis of the
+distribution and activities of these pathogens along the colon is needed
+to elucidate a mechanism for development of CRC or IBD subtypes in the
+proximal or distal colon.
+
+The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
+been commonly isolated from mucosal biopsies of patients with IBD (24).
+Laboratory experiments with these isolates have shown that
+disease-isolated \emph{F. nucleatum} are more invasive and stimulate
+more TNF-\(\alpha\) production than strains from healthy individuals
+(10), suggesting the bacteria may increase inflammation in the gut as
+well (32). \emph{F. varium} isolated from UC patients caused colonic
+ulcers in an experimental mouse model (33). \emph{F. varium} was only
+detected in three subjects and two of those samples were isolated form
+the proximal mucosa (Supplementary Fig. S1B). \emph{F. varium} is most
+commonly isolated from UC patient biopsies from the ileum or cecum (34),
+suggesting this species may exhibit preference for the different
+environmental conditions of these gastrointestinal sites. Further work
+will assess how gut environment may select for species which may then
+cause localized disease.
+
+Specific comparisons of our findings to previously published studies of
+spatial variation are confounded by the use of bowel preparation
+methods. A rare report of a matched-colonoscopy study sampled 18
+patient's colonic mucosa and luminal contents prior to and after bowel
+cleansing (35). This study found that mucosal and luminal samples were
+distinguishable prior to bowel cleansing, but that bowel preparation
+resulted in an increase in shared OTUs between each site (35). After
+seven days, bowel cleansing not only made the samples more difficult to
+distinguish, but it also decreased the diversity observed across sites.
+Bowel preparation clearly biases the representation of microbiota
+recovered from sampling the lumen or mucosa.
+
+By revealing specific differences in microbial populations at each
+location in the gut via sampling an unprepared bowel, we can begin to
+form hypotheses about how specific host-microbe interactions can affect
+disease progression of proximal and distal CRC and IBD subtypes. A
+better understanding of microbial activities in the gut can enhance
+microbiome-based screening and treatment modalities for these colon
+diseases.
+
+\subsubsection{Acknowledgments}\label{acknowledgments}
+
+We thank all the individuals who volunteered for the study. This work
+was supported by the Rose and Lawrence C. Page Foundation (DKT). We
+would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman
+for their roles in study coordination, subject recruitment, sample
+collection and sample processing.
 
@@
 
-\paragraph{Figure 2}\label{figure-2}
-
-Phylum-level relative abundance and diversity in the proximal and distal
-human colon. A) Relative abundance of the top five bacterial phyla in
-each sampling site. Each box represents the median and interquartile
-range. B) Simpson diversity of the microbial communities at each
-location. The horizontal lines represent the median values.
-
-\newpage
-
-\paragraph{Figure 3}\label{figure-3}
-
-Comparison of microbial community structure between sites of the gut.
-\(\theta\)YC distances are shown to indicate the interpersonal
-dissimilarities between two sites -- each point represents one
-individual. In (A), comparisons of the proximal and distal mucosal and
-lumen are shown. In (B), comparisons of each site to the exit stool are
-shown. In (C), comparisons of samples from all subjects to each other
-(interpersonal) or within one subject (intrapersonal) are shown.
-
-\newpage
-
-\paragraph{Figure 4}\label{figure-4}
-
-Taxa specific to the distal and proximal sides of the colon. Top five
-OTUs that are most important for the classification model for the distal
-mucosa and lumen (A) and the proximal mucosa and lumen (B). The vertical
-lines represent the median values for each OTU.
-
-\newpage
-
-\paragraph{Figure 5}\label{figure-5}
-
-Taxa specific to the distal and proximal mucosa and lumen. The five OTUs
-that were most important differentiating the distal and proximal mucosa
-(A) and the distal and proximal lumen (B). The vertical lines represent
-the median values for each OTU. \newpage
-
-\paragraph{Figure S1}\label{figure-s1}
-
-Random Forest classifies locations in the colon. A) Receiver Operator
-Characteristic curves are shown for the 10-fold cross validation of the
-Random Forest model classifying lumen and mucosal samples for the distal
-(red) and proximal (blue) sides of the colon. (B) Receiver Operator
-Characteristic curves are shown for the 10-fold cross validation of the
-Random Forest model classifying distal mucosa vs proximal mucosa (green)
-and distal lumen versus proximal lumen (purple).
-
-\newpage
-
-\paragraph{Figure S2}\label{figure-s2}
-
-Location and relative abundance of cancer-associated OTUs. Relative
-abundance was calculated and plotted by sample site for each OTU of
-interest: (A) \emph{Fusobacterium nucleatum subsp. animalis} (B)
-\emph{Fusobacterium varium} and (C) \emph{Porphyromonas asacharolytica}
-
-\subsubsection*{References}\label{references}
-\addcontentsline{toc}{subsubsection}{References}
+\subsubsection{References}\label{references}
 
@@
 
+\hypertarget{ref-Kozich2013}{}
+14. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+Development of a dual-index sequencing strategy and curation pipeline
+for analyzing amplicon sequence data on the MiSeq illumina sequencing
+platform. Applied and Environmental Microbiology {[}Internet{]}.
+American Society for Microbiology; 2013;79:5112--20. Available from:
+\url{https://doi.org/10.1128/aem.01043-13}
+
+\hypertarget{ref-Schloss2009}{}
+15. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+EB, et al. Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology {[}Internet{]}.
+American Society for Microbiology; 2009;75:7537--41. Available from:
+\url{https://doi.org/10.1128/aem.01541-09}
+
+\hypertarget{ref-Wang2007}{}
+16. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive bayesian classifier
+for rapid assignment of rRNA sequences into the new bacterial taxonomy.
+Applied and Environmental Microbiology {[}Internet{]}. American Society
+for Microbiology; 2007;73:5261--7. Available from:
+\url{https://doi.org/10.1128/aem.00062-07}
+
+\hypertarget{ref-Calle_2011}{}
+17. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
+for genomic profiling with random forest. Human Heredity {[}Internet{]}.
+S. Karger AG; 2011;72:121--32. Available from:
+\url{https://doi.org/10.1159\%2F000330778}
+
+\hypertarget{ref-Robin_2011}{}
+18. Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez J-C, et
+al. pROC: An open-source package for r and s to analyze and compare ROC
+curves. BMC Bioinformatics {[}Internet{]}. Springer Nature; 2011;12:77.
+Available from: \url{https://doi.org/10.1186\%2F1471-2105-12-77}
+
 \hypertarget{ref-LloydPrice2016}{}
-14. Lloyd-Price J, Abu-Ali G, Huttenhower C. The healthy human
+19. Lloyd-Price J, Abu-Ali G, Huttenhower C. The healthy human
 microbiome. Genome Medicine {[}Internet{]}. Springer Nature; 2016;8.
@@
 \hypertarget{ref-Eckburg2005}{}
-15. Eckburg PB. Diversity of the human intestinal microbial flora.
+20. Eckburg PB. Diversity of the human intestinal microbial flora.
 Science {[}Internet{]}. American Association for the Advancement of
@@
 \hypertarget{ref-deCarcer2010}{}
-16. Cárcer DA de, Cuív PÓ, Wang T, Kang S, Worthley D, Whitehall V, et
+21. Cárcer DA de, Cuív PÓ, Wang T, Kang S, Worthley D, Whitehall V, et
 al. Numerical ecology validates a biogeographical distribution and
@@
 \hypertarget{ref-Zhang2013}{}
-17. Zhang Z, Geng J, Tang X, Fan H, Xu J, Wen X, et al. Spatial
+22. Zhang Z, Geng J, Tang X, Fan H, Xu J, Wen X, et al. Spatial
 heterogeneity and co-occurrence patterns of human mucosal-associated
@@
 \hypertarget{ref-Castellarin2011}{}
-18. Castellarin M, Warren RL, Freeman JD, Dreolini L, Krzywinski M,
+23. Castellarin M, Warren RL, Freeman JD, Dreolini L, Krzywinski M,
 Strauss J, et al. Fusobacterium nucleatum infection is prevalent in
@@
 \hypertarget{ref-Lee2016}{}
-19. Lee Y, Eun CS, Lee AR, Park CH, Han DS. FusobacteriumIsolates
+24. Lee Y, Eun CS, Lee AR, Park CH, Han DS. FusobacteriumIsolates
 recovered from colonic biopsies of inflammatory bowel disease patients
@@
 \hypertarget{ref-Hong2011}{}
-20. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
+25. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
 Pyrosequencing-based analysis of the mucosal microbiota in healthy
@@
 \hypertarget{ref-Stearns2011}{}
-21. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
+26. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
 Cvitkovitch DG, et al. Bacterial biogeography of the human digestive
@@
 \hypertarget{ref-Sears2014}{}
-22. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
+27. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
 Host \& Microbe {[}Internet{]}. Elsevier BV; 2014;15:317--28. Available
@@
 \hypertarget{ref-Mima2016}{}
-23. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
+28. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
 Fusobacterium nucleatum in colorectal carcinoma tissue according to
@@
 \hypertarget{ref-Tahara_2014}{}
-24. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
+29. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
 al. Fusobacterium in colonic flora and molecular features of colorectal
@@
 \hypertarget{ref-Whitmore2014}{}
-25. Whitmore SE, Lamont RJ. Oral bacteria and cancer. Goldman WE,
+30. Whitmore SE, Lamont RJ. Oral bacteria and cancer. Goldman WE,
 editor. PLoS Pathogens {[}Internet{]}. Public Library of Science (PLoS);
@@
 \hypertarget{ref-Flynn2016}{}
-26. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
+31. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
 oral bacteria in colorectal cancer. McMahon K, editor. mSphere
@@
 \hypertarget{ref-Dharmani2011}{}
-27. Dharmani P, Strauss J, Ambrose C, Allen-Vercoe E, Chadee K.
+32. Dharmani P, Strauss J, Ambrose C, Allen-Vercoe E, Chadee K.
 Fusobacterium nucleatum infection of colonic cells stimulates MUC2 mucin
@@
 \hypertarget{ref-Ohkusa2003}{}
-28. Ohkusa T. Induction of experimental ulcerative colitis by
+33. Ohkusa T. Induction of experimental ulcerative colitis by
 fusobacterium varium isolated from colonic mucosa of patients with
@@
 \hypertarget{ref-Ohkusa2002}{}
-29. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
+34. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
 Fusobacterium varium localized in the colonic mucosa of patients with
@@
 \hypertarget{ref-Shobar2016}{}
-30. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
+35. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
 JE, et al. The effects of bowel preparation on microbiota-related
@@
 
-\hypertarget{ref-Kozich2013}{}
-31. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
-Development of a dual-index sequencing strategy and curation pipeline
-for analyzing amplicon sequence data on the MiSeq illumina sequencing
-platform. Applied and Environmental Microbiology {[}Internet{]}.
-American Society for Microbiology; 2013;79:5112--20. Available from:
-\url{https://doi.org/10.1128/aem.01043-13}
+\newpage
 
-\hypertarget{ref-Schloss2009}{}
-32. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
-EB, et al. Introducing mothur: Open-source, platform-independent,
-community-supported software for describing and comparing microbial
-communities. Applied and Environmental Microbiology {[}Internet{]}.
-American Society for Microbiology; 2009;75:7537--41. Available from:
-\url{https://doi.org/10.1128/aem.01541-09}
+\subsubsection{Figures}\label{figures}
 
-\hypertarget{ref-Wang2007}{}
-33. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive bayesian classifier
-for rapid assignment of rRNA sequences into the new bacterial taxonomy.
-Applied and Environmental Microbiology {[}Internet{]}. American Society
-for Microbiology; 2007;73:5261--7. Available from:
-\url{https://doi.org/10.1128/aem.00062-07}
+\paragraph{Figure 1}\label{figure-1}
 
-\hypertarget{ref-Calle_2011}{}
-34. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
-for genomic profiling with random forest. Human Heredity {[}Internet{]}.
-S. Karger AG; 2011;72:121--32. Available from:
-\url{https://doi.org/10.1159\%2F000330778}
+Sampling strategy. A flexible sigmoidoscope was used to sample the
+distal colonic luminal contents and mucosa. The scope was inserted
+\textasciitilde{} 25cm into the subject and biopsy forceps were used to
+sample the luminal contents (D, inset). A separate set of biopsy forceps
+was used to sample the distal mucosa (D, inset). The sigmoidoscope was
+removed. A pediatric colonoscope was inserted and used to access the
+proximal colon (P, inset). Biopsies were taken of the proximal luminal
+contents and mucosa as described. One week prior to the procedure stool
+was collected at home and sent into the laboratory. Representative
+images from one individual are shown.
 
-\hypertarget{ref-Robin_2011}{}
-35. Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez J-C, et
-al. pROC: An open-source package for r and s to analyze and compare ROC
-curves. BMC Bioinformatics {[}Internet{]}. Springer Nature; 2011;12:77.
-Available from: \url{https://doi.org/10.1186\%2F1471-2105-12-77}
+\begin{figure}[htbp]
+\centering
+\includegraphics{../submission/fig1.jpg}
+\caption{Fig 1}
+\end{figure}
+
+\newpage
+
+\paragraph{Figure 2}\label{figure-2}
+
+Phylum-level relative abundance and diversity in the proximal and distal
+human colon. A) Relative abundance of the top five bacterial phyla in
+each sampling site. Each box represents the median and interquartile
+range. B) Simpson diversity of the microbial communities at each
+location. The horizontal lines represent the median values.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{../submission/figure_2.pdf}
+\caption{Fig 2}
+\end{figure}
+
+\newpage
+
+\paragraph{Figure 3}\label{figure-3}
+
+Comparison of microbial community structure between sites of the gut.
+\(\theta\)YC distances are shown to indicate the interpersonal
+dissimilarities between two sites -- each point represents one
+individual. In (A), comparisons of the proximal and distal mucosal and
+lumen are shown. In (B), comparisons of each site to the exit stool are
+shown. In (C), comparisons of samples from all subjects to each other
+(interpersonal) or within one subject (intrapersonal) are shown.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{../submission/figure_3.pdf}
+\caption{Fig 3}
+\end{figure}
+
+\newpage
+
+\paragraph{Figure 4}\label{figure-4}
+
+Random Forest classifies locations in the colon. A) Receiver Operator
+Characteristic curves are shown for the 10-fold cross validation of the
+Random Forest model classifying lumen and mucosal samples for the distal
+(red) and proximal (blue) sides of the colon. (B) Receiver Operator
+Characteristic curves are shown for the 10-fold cross validation of the
+Random Forest model classifying distal mucosa vs proximal mucosa (green)
+and distal lumen versus proximal lumen (purple).
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{../submission/figure_4.pdf}
+\caption{Fig 4}
+\end{figure}
+
+\newpage
+
+\paragraph{Figure 5}\label{figure-5}
+
+Taxa specific to the distal and proximal sides of the colon. Top five
+OTUs that are most important for the classification model for the distal
+mucosa and lumen (A) and the proximal mucosa and lumen (B). The vertical
+lines represent the median values for each OTU.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{../submission/figure_5.pdf}
+\caption{Fig 5}
+\end{figure}
+
+\newpage
+
+\paragraph{Figure 6}\label{figure-6}
+
+Taxa specific to the distal and proximal mucosa and lumen. The five OTUs
+that were most important differentiating the distal and proximal mucosa
+(A) and the distal and proximal lumen (B). The vertical lines represent
+the median values for each OTU.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{../submission/figure_6.pdf}
+\caption{Fig 6}
+\end{figure}
+
+\newpage
+
+\paragraph{Figure S1}\label{figure-s1}
+
+Location and relative abundance of cancer-associated OTUs. Relative
+abundance was calculated and plotted by sample site for each OTU of
+interest: (A) \emph{Fusobacterium nucleatum subsp. animalis} (B)
+\emph{Fusobacterium varium} and (C) \emph{Porphyromonas asacharolytica}
+\includegraphics{../submission/figure_S1.pdf}
+
+\newpage
 

### 2017-11-02 code/build_figure2.R

@@ -78,5 +78,5 @@ dev.off()
 
 ##### Cowplot way to save the plots! 
 fig2 <- plot_grid(phy_plot, simp_plot, labels = c("A", "B"), label_size=16, ncol = 1, align = "v")  
-save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_2.pdf', fig2, ncol=1, nrow=2, base_width=7, base_height = 4)
+save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_2.pdf', fig2, ncol=1, nrow=2, base_width=6, base_height = 3)
 

### 2017-11-02 code/build_figure3.R

@@ -36,19 +36,19 @@ exittyc <- subset(tyc, samp2 == 'SS')
 #plots
 tycpositions <- c("RB_RS", "LS_RS", "LB_RB", "LB_LS")
 lvr_plot <- ggplot(lvsr, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() + 
-  theme(legend.position="none", axis.title.x=element_blank(), axis.line = element_line(colour = "black"), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
+  theme(legend.position="none", axis.title.x=element_blank(), axis.line = element_line(colour = "black"), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
   scale_x_discrete(limits = tycpositions, breaks = tycpositions,
-                   labels=c("P Mucosa vs P Lumen", "D Lumen vs P Lumen", "D Mucosa vs P Mucosa", "D Mucosa vs D Lumen")) +
-  ylab("ThetaYC distance") 
+                   labels=c("P Muc vs P Lum", "D Lum vs P Lum", "D Muc vs P Muc", "D Muc vs D Lum")) +
+  ylab(expression(theta["YC"]* " dissimilarity"))
 
 
 
 exitpositions <- c("RB_SS", "RS_SS", "LB_SS", "LS_SS")
 exit_plot <- ggplot(exittyc, aes(x=match, y=thetayc)) + geom_boxplot(width=0.8) +theme_bw() +
-  theme(legend.position="none", axis.title.x=element_blank(), axis.line = element_line(colour = "black"), axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
+  theme(legend.position="none", axis.title.x=element_blank(), axis.line = element_line(colour = "black"), panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
   scale_x_discrete(limits=exitpositions, breaks=exitpositions,
-                   labels=c("P Mucosa vs Stool", "P Lumen vs Stool", "D Mucosa vs Stool", "D Lumen vs Stool")) +
-  theme(axis.title.x=element_blank()) +ylab("ThetaYC distance")
[…]

### 2017-11-02 code/oncogenic_otus.R

@@ -107,7 +107,7 @@ dev.off()
 
 ####Cowplot way!
 
-figS2 <- plot_grid(fuso_plot, fusoviv_plot, porphy_plot, labels = c("A", "B", "C"), ncol = 1, align = "v")  
-save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_S2.pdf', figS2, ncol=1, nrow=3, base_width = 6, base_height=3)
+figS1 <- plot_grid(fuso_plot, fusoviv_plot, porphy_plot, labels = c("A", "B", "C"), ncol = 1, align = "v")  
+save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_S1.pdf', figS1, ncol=1, nrow=3, base_width = 6, base_height=3)
 
 

### 2017-11-02 code/random_forest.R

@@ -591,7 +591,7 @@ legend('topright', legend=c("Left lumen", "Right lumen"), pch=c(21, 21), pt.bg=c
 ######################################################## build and export figure 
 #export as PDF
 
-plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
+plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_5.pdf'
 pdf(file=plot_file, width=7, height=9)
 layout(matrix(c(1,
                 2), 
@@ -646,10 +646,10 @@ mtext('B', side=2, line=7.5, las=1, adj=2, padj=-6.5, cex=2, font=2)
 
 dev.off()
 
-####################################### Figure 5 - updated 
+####################################### Figure 6 - updated 
 #export as PDF
 
-plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_5.pdf'
+plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_6.pdf'
 pdf(file=plot_file, width=7, height=9)
 layout(matrix(c(1,
                 2), 

### 2017-11-02 code/rf_optimized.R

@@ -148,7 +148,7 @@ legend('bottom', legend=c(sprintf('P lumen vs D lumen, hold-one-out, AUC = 0.580
 
 ###########saving figures as proper size
 
-plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_S1.pdf'
+plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
 pdf(file=plot_file, width=7, height=9)
 layout(matrix(c(1,
                 2), 

### 2017-11-02 code/build_figure3.R
now with figs 1-4 updated

@@ -167,6 +167,6 @@ dev.off()
 
 
 fig3 <- plot_grid(lvr_plot, exit_plot, inter_plot, labels = c("A", "B", "C"), label_size= 16, ncol = 1, align = "v")  
-save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_3.pdf', fig3, ncol=1, nrow=3, base_width=5, base_height = 3)
+save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_3.pdf', fig3, ncol=1, nrow=3, base_width=5, base_height = 2)
 
 

### 2017-11-02 code/rf_optimized.R

@@ -149,7 +149,7 @@ legend('bottom', legend=c(sprintf('P lumen vs D lumen, hold-one-out, AUC = 0.580
 ###########saving figures as proper size
 
 plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_4.pdf'
-pdf(file=plot_file, width=7, height=9)
+pdf(file=plot_file, width=4, height=6)
 layout(matrix(c(1,
                 2), 
               nrow=2, byrow = TRUE))
@@ -158,30 +158,30 @@ layout(matrix(c(1,
 #Left and right new method plot
 par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(left_roc, col='red', lwd=3, add=T, lty=1)
-plot(right_roc, col = 'blue', lwd=3, add=T, lty=1)
+plot(left_roc, col='red', add=T, lty=1)
+plot(right_roc, col = 'blue', add=T, lty=1)
 mtext(side=2, text="Sensitivity", line=2.5)
 mtext(side=1, text="Specificity", line=2.5)
-legend('bottom', legend=c(
-  sprintf('D Lumen vs D Mucosa, AUC = 0.9079'),
-  sprintf('P Lumen vs P Mucosa, AUC = 0.7645')
-),lty=c(1, 1, 1), lwd=3, col=c('red', 'blue'), bty='n')
+legend('bottomright', legend=c(
+  sprintf('D Lum vs D Muc, AUC = 0.908'),
+  sprintf('P Lum vs P Muc, AUC = 0.764')
+),lty=1, cex=0.8, col=c('red', 'blue'), bty='n')
 
-mtext('A', side=2, line=1.5, las=1, adj=1.7, padj=-7, cex=2, font=2)
+mtext('A', side=2, line=2, las=1, adj=1.5, padj=-5, cex=1.5, font=2)
 
 #stool and mucosa new method plot 
 par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(muc_roc, col='darkgreen', lwd=3, add=T, lty=1)
[…]

### 2017-11-02 code/random_forest.R
almost all figures finished

@@ -592,7 +592,7 @@ legend('topright', legend=c("Left lumen", "Right lumen"), pch=c(21, 21), pt.bg=c
 #export as PDF
 
 plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_5.pdf'
-pdf(file=plot_file, width=7, height=9)
+pdf(file=plot_file, width=6, height=7)
 layout(matrix(c(1,
                 2), 
               nrow=2, byrow = TRUE))
@@ -601,6 +601,8 @@ layout(matrix(c(1,
 
 right_otu_feat <- colnames(aucrf_data_right_bs[2:6])
 otu_taxa_right <- get_tax(1, right_otu_feat, tax_file)
+otu_taxa_right <- separate(otu_taxa_right, tax_label, into = c("OTU", "otu_num"), sep = "\\(")
+formatted4 <- lapply(1:nrow(otu_taxa_right), function(i) bquote(paste(italic(.(otu_taxa_right$OTU[i])), "(", .(otu_taxa_right$otu_num[i]), sep=" ")))
 #Abundance stripchart or most predictive otus
 rs_abunds <- shared_meta[shared_meta$location=='RS', right_otu_feat]/10000 + 1e-4
 rb_abunds <- shared_meta[shared_meta$location=='RB', right_otu_feat]/10000 + 1e-4
@@ -615,7 +617,7 @@ for(i in right_otu_feat){
   segments(median(rb_abunds[,i]),index+0.7,median(rb_abunds[,i]),index, lwd=3)
   index <- index + 2
 }
-axis(2, at=seq(1,index-2,2), labels=otu_taxa_right$tax_label, las=1, line=-0.5, tick=F, cex.axis=0.9)
+axis(2, at=seq(1,index-2,2), labels=do.call(expression,formatted4), las=1, line=-0.5, tick=F, cex.axis=0.9)
 axis(1, at=c(1e-4, 1e-3, 1e-2, 1e-1, 1), label=c("0", "0.1", "1", "10", "100"))
 legend('topright', legend=c("P Mucosa", "P Lumen"), pch=c(21, 21), pt.bg=c("black","white"), cex=0.8)
 
@@ -624,6 +626,8 @@ mtext('A', side=2, line=7.5, las=1, adj=2, padj=-6.5, cex=2, font=2)
 #just LB vs LS 
[…]

### 2017-11-03 code/oncogenic_otus.R
manuscript edits nearly complete

@@ -34,12 +34,14 @@ fuso179 <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient"
 fuso179[,5] <- (fuso179[,4]/4321)*100
 names(fuso179)[5] <- "Otu00179_relAbund"
 
+fuso_y_title <- expression(paste(italic("F. nucleatum"), " Rel. Abund."))
 positions <- c("RB", "RS", "LB", "LS", "SS")
 fuso_plot <- ggplot(fuso179, aes(x=location, y=Otu00179_relAbund)) +geom_jitter(width=0.3) +theme_bw() +
-  ylab("F. nucleatum Relative Abundance") +
+  ylab(fuso_y_title) +
   scale_x_discrete(limits = positions, breaks=positions, 
-                   labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
-  theme(legend.position='none', axis.title.x=element_blank(), axis.text = element_text(size= 10), axis.title= element_text(size=10), panel.grid.major = element_blank(),panel.grid.minor = element_blank())
+                   labels=c("P Muc", "P Lum", "D Muc", "D Lum", "Feces")) +
+  theme(legend.position='none', axis.title.x=element_blank(), axis.title.y=element_text(size=8),panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
+  scale_y_log10()
 
 #IBD fuso OTU - F. varium 
 
@@ -47,12 +49,15 @@ fusoviv <- subsampled_meta[, colnames(subsampled_meta) %in% c("group", "patient"
 fusoviv[,5] <- (fusoviv[,4]/4321)*100
 names(fusoviv)[5] <- "Otu00472_relAbund"
 
+fusoviv_y_title <- expression(paste(italic("F. varium"), " Rel. Abund."))
+
 positions <- c("RB", "RS", "LB", "LS", "SS")
 fusoviv_plot <- ggplot(fusoviv, aes(x=location, y=Otu00472_relAbund)) +geom_jitter(width=0.3) +theme_bw() +
-  ylab("F. varium Relative Abundance") +
+  ylab(fusoviv_y_title) +
   scale_x_discrete(limits = positions, breaks=positions, 
-                   labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
[…]

### 2017-11-03 code/random_forest.R

@@ -619,9 +619,9 @@ for(i in right_otu_feat){
 }
 axis(2, at=seq(1,index-2,2), labels=do.call(expression,formatted4), las=1, line=-0.5, tick=F, cex.axis=0.9)
 axis(1, at=c(1e-4, 1e-3, 1e-2, 1e-1, 1), label=c("0", "0.1", "1", "10", "100"))
-legend('topright', legend=c("P Mucosa", "P Lumen"), pch=c(21, 21), pt.bg=c("black","white"), cex=0.8)
+legend('topright', legend=c("P Muc", "P Lum"), pch=c(21, 21), pt.bg=c("black","white"), cex=0.8)
 
-mtext('A', side=2, line=7.5, las=1, adj=2, padj=-6.5, cex=2, font=2)
+mtext('A', side=2, line=7.5, las=1, adj=2, padj=-4.5, cex=2, font=2)
 
 #just LB vs LS 
 left_otu_feat <- colnames(aucrf_data_left_bs[2:6])
@@ -644,9 +644,9 @@ for(i in left_otu_feat){
 }
 axis(2, at=seq(1,index-2,2), labels=do.call(expression, formatted3), las=1, line=-0.5, tick=F, cex.axis=0.9)
 axis(1, at=c(1e-4, 1e-3, 1e-2, 1e-1, 1), label=c("0", "0.1", "1", "10", "100"))
-legend('topright', legend=c("D Mucosa", "D Lumen"), pch=c(21, 21), pt.bg=c("black","white"), cex=0.8)
+legend('topright', legend=c("D Muc", "D Lum"), pch=c(21, 21), pt.bg=c("black","white"), cex=0.8)
 
-mtext('B', side=2, line=7.5, las=1, adj=2, padj=-6.5, cex=2, font=2)
+mtext('B', side=2, line=7.5, las=1, adj=2, padj=-4.5, cex=2, font=2)
 
 dev.off()
 
@@ -654,7 +654,7 @@ dev.off()
 #export as PDF
 
 plot_file <- '~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_6.pdf'
-pdf(file=plot_file, width=7, height=9)
+pdf(file=plot_file, width=6, height=7)
 layout(matrix(c(1,
                 2), 
               nrow=2, byrow = TRUE))
@@ -680,9 +680,9 @@ for(i in LRbowel_otu_feat){
 }
[…]

### 2017-11-09 submission/manuscript.Rmd
now with figs 1-4 updated
almost done with last pat edits

@@
 ---
-\vspace{15mm}
+\vspace{8mm}
 
@@
 
-\vspace{15mm}
+\vspace{8mm}
 
@@
 
-Word count:
+Word count: 3,965/5,000
 
-Fig.s:
-
-Supplementary Fig.s:
+Figures: 6
 
@@
 The right mucosa and lumen had the most similar community structures of the five sites we considered from each subject.
-The distal mucosa had higher relative abundance of Finegoldia, Murdochiella, Peptoniphilus, Porphyromonas and Anaerococcus.
-The proximal mucosa had more of the genera Enterobacteriaceae, Bacteroides and Pseudomonas.
+The distal mucosa had higher relative abundance of _Finegoldia, Murdochiella, Peptoniphilus, Porphyromonas_ and _Anaerococcus_.
+The proximal mucosa had more of the genera _Enterobacteriaceae, Bacteroides_ and _Pseudomonas_.
 The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC=0.850).
@@
 Microenvironments differ not only longitudinally along the colon, but also radially from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
-The identity of these specific microbiota and communities are important for understanding the etiology of complex colon diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD).
+The identity of these specific microbiota and communities are important for understanding the etiology of complex diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD).
 CRC and IBD can be preceded or accelerated by perturbations the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017].
@@
 These studies often examined only shed human stool or the small intestine, preventing fine-resolution analysis of paired samples from the proximal and distal sides of the colon.
-Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same, disease-baring patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.
+Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same, disease-baring patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site @Flemer633.
 Due to these limitations, the contribution of the gut microbiota to CRC and IBD disease location in the colon is largely undefined.
@@
 The few existing profiles of the microbial spatial variation of the colon have been limited by sample collection methods.
-The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy or surgery.
+The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy or surgery @Donaldson2015.
 While the latter method allows investigators to acquire samples from inside the human colon, typically these procedures are preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
@@
 A flexible sigmoidoscope was first inserted about 25cm into the colon and jumbo biopsy forceps used to collect the luminal contents.
-Two luminal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen.
+Two luminal samples were collected and the contents immediately deposited into RNAlater (Fisher) and flash-frozen in liquid nitrogen.
 The forceps were withdrawn and new biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter.
@@
 Sequencing and analysis of a mock community revealed the error rate to be 0.018%.
-Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.
+Samples were rarefied to 4231 sequences per sample in order to reduce the effects of uneven sampling bias.
 
-Diversity analysis was performed using the Simpson diversity calculator and $\theta$<sub>YC</sub> calculator metrics in mothur version 1.39.5 @Schloss2009.
-$\theta$<sub>YC</sub> distances were calculated to determine the dissimilarity between two samples.
+Diversity analysis was performed using the Simpson diversity calculator and $\theta$~YC~ calculator metrics in mothur version 1.39.5 @Schloss2009.
+$\theta$~YC~ distances were calculated to determine the dissimilarity between two samples.
 Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011.
-For each model the data was split into a 19-subject training set and a 1-subject test set.
+For each model the data were split into a 19-subject training set and a 1-subject test set.
 The model was built and cross-validated using AUCRFcv on the training set.
@@
 Differences in community membership at the phyla level were tested using the analysis of molecular variance (AMOVA) metric in mothur.
-Differences in $\theta$<sub>YC</sub> distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
+Differences in $\theta$~YC~ distances by location were tested using the Wilcoxon rank-sum test adjusted for multiple comparisons using the Benjamini-Hochberg procedure.
 
@@
 16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number XXXX.
-A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017.
+A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017.
 
@@
 #### Microbial membership and diversity of the proximal and distal colon 
-Luminal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Fig.
+Luminal and mucosal samples were collected from the proximal and distal colon of 20 healthy individuals who had not undergone bowel preparation (Fig.
 1).
@@
 
-#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
+#### Random Forest classification models identify important OTUs on each side
 To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances.
-We used 10-fold cross validation to build the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Fig.
+We built the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Fig.
 4A).
@@
 
-Next, we built a Random Forest model to differentiate the proximal and distal luminal samples using 10-fold cross validation.
+Next, we built a Random Forest model to differentiate the proximal and distal luminal samples.
 The model performed best when distinguishing the proximal versus distal mucosa (Fig.
@@
 ### Discussion
-Here we identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon using samples collected during an unprepared colonoscopy of healthy subjects.
-We found that all locations contained a range of phyla relative abundances and a range of diversity, but that there was a wide variability between subjects.
+We identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon using samples collected during an unprepared colonoscopy of healthy subjects.
+We found that all locations contained a range of phylum relative abundances and a range of diversity, but that there was a wide variability between subjects.
 Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other.
@@
 Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacterium but the structure of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014.
-Finally, _Fusobacterium_ and _Porphyromonas_ populations not only co-occur on CRC tumors but also to synergize to promote tumorigenesis in an oral cancer model @Whitmore2014 @Flynn2016.
+Finally, _Fusobacterium_ and _Porphyromonas_ populations not only co-occur on CRC tumors but also to synergize to promote tumorigenesis in an oral cancer model [@Whitmore2014; @Flynn2016].
 Further analysis of the distribution and activities of these pathogens along the colon is needed to elucidate a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.
@@
 We would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman for their roles in study coordination, subject recruitment, sample collection and sample processing.
-<need grant # here>
 
@@
 Comparison of microbial community structure between sites of the gut.
-$\theta$<sub>YC</sub> distances are shown to indicate the interpersonal dissimilarities between two sites – each point represents one individual.
+$\theta$~YC~ distances are shown to indicate the interpersonal dissimilarities between two sites – each point represents one individual.
 In (A), comparisons of the proximal and distal mucosal and lumen are shown.
@@
 Random Forest classifies locations in the colon.
-A) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying lumen and mucosal samples for the distal (red) and proximal (blue) sides of the colon.
+A) Receiver Operator Characteristic curves are shown for the Random Forest model classifying lumen and mucosal samples for the distal (red) and proximal (blue) sides of the colon.
 (B) Receiver Operator Characteristic curves are shown for the 10-fold cross validation of the Random Forest model classifying distal mucosa vs proximal mucosa (green) and distal lumen versus proximal lumen (purple).

### 2017-11-09 submission/manuscript.tex
now with figs 1-4 updated
almost done with last pat edits

@@
 structures of the five sites we considered from each subject. The distal
-mucosa had higher relative abundance of Finegoldia, Murdochiella,
-Peptoniphilus, Porphyromonas and Anaerococcus. The proximal mucosa had
-more of the genera Enterobacteriaceae, Bacteroides and Pseudomonas. The
-classification model performed well when classifying mucosal samples
-into proximal or distal sides (AUC=0.850). Separating proximal and
-distal luminal samples proved more challenging (AUC=0.580) and specific
-microbiota that differentiated the two were hard to identify. By
-sampling the unprepped colon, we identified distinct bacterial
-populations native to the proximal and distal sides. Further
-investigation of these bacteria may elucidate if and how these groups
-contribute to different disease processes on their respective sides of
-the colon.
+mucosa had higher relative abundance of \emph{Finegoldia, Murdochiella,
+Peptoniphilus, Porphyromonas} and \emph{Anaerococcus}. The proximal
+mucosa had more of the genera \emph{Enterobacteriaceae, Bacteroides} and
+\emph{Pseudomonas}. The classification model performed well when
+classifying mucosal samples into proximal or distal sides (AUC=0.850).
+Separating proximal and distal luminal samples proved more challenging
+(AUC=0.580) and specific microbiota that differentiated the two were
+hard to identify. By sampling the unprepped colon, we identified
+distinct bacterial populations native to the proximal and distal sides.
+Further investigation of these bacteria may elucidate if and how these
+groups contribute to different disease processes on their respective
+sides of the colon.
 
@@
 Diversity analysis was performed using the Simpson diversity calculator
-and \(\theta\)YC calculator metrics in mothur version 1.39.5 (15).
-\(\theta\)YC distances were calculated to determine the dissimilarity
-between two samples. Random Forest classification models were built
-using the AUCRF R package using a leave-one-subject out approach (17).
-For each model the data was split into a 19-subject training set and a
-1-subject test set. The model was built and cross-validated using
-AUCRFcv on the training set. The model was then tested on the left-out
-patient. This process was repeated iteratively for all subjects and
-results plotted as Reciever Operator Characteristic curves using the
-pROC R package (18). Resultant models were used to identify the OTUs
-that were most important for classifying each location. Species-level
-information for sequences of interest was obtained by aligning the
-sequences to the GenBank nucleotide databse using blastn. The species
-name was only used if the identity score was \(\ge\) 99\% over the
-full-length of the contig and matched a single reference.
+and \(\theta\)\textsubscript{YC} calculator metrics in mothur version
+1.39.5 (15). \(\theta\)\textsubscript{YC} distances were calculated to
+determine the dissimilarity between two samples. Random Forest
+classification models were built using the AUCRF R package using a
+leave-one-subject out approach (17). For each model the data was split
+into a 19-subject training set and a 1-subject test set. The model was
+built and cross-validated using AUCRFcv on the training set. The model
+was then tested on the left-out patient. This process was repeated
+iteratively for all subjects and results plotted as Reciever Operator
+Characteristic curves using the pROC R package (18). Resultant models
+were used to identify the OTUs that were most important for classifying
+each location. Species-level information for sequences of interest was
+obtained by aligning the sequences to the GenBank nucleotide databse
+using blastn. The species name was only used if the identity score was
+\(\ge\) 99\% over the full-length of the contig and matched a single
+reference.
 
@@
 the analysis of molecular variance (AMOVA) metric in mothur. Differences
-in \(\theta\)YC distances by location were tested using the Wilcoxon
-rank-sum test adjusted for multiple comparisons using the
+in \(\theta\)\textsubscript{YC} distances by location were tested using
+the Wilcoxon rank-sum test adjusted for multiple comparisons using the
 Benjamini-Hochberg procedure.

### 2017-11-09 code/build_figure2.R

@@ -41,7 +41,7 @@ positions <- c("RB", "RS", "LB", "LS", "SS")
 phy_plot <- ggplot(phyla_RAmelt, aes(x=location, y=value)) + geom_boxplot(aes(color=variable)) + 
   scale_color_discrete(guide=FALSE)+
   geom_boxplot(aes(fill=variable), outlier.shape=21) + theme_bw() + 
-  theme(axis.text = element_text(size= 10), axis.title= element_text(size=12)) +
+  theme(axis.text = element_text(size= 10), axis.title= element_text(size=12), panel.border = element_blank()) + 
   scale_x_discrete(limits = positions, breaks=positions, 
                    labels=c("P Mucosa", "P Lumen", "D Mucosa", "D Lumen", "Stool")) +
   theme(axis.title.x=element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + scale_fill_brewer(palette="Dark2", name="Phylum") +
@@ -77,6 +77,24 @@ simp_plot
 dev.off()
 
 ##### Cowplot way to save the plots! 
-fig2 <- plot_grid(phy_plot, simp_plot, labels = c("A", "B"), label_size=16, ncol = 1, align = "v")  
+#fig2 <- plot_grid(phy_plot, simp_plot, labels = c("A", "B"), label_size=16, ncol = 1, align = "v") 
+
+theme_set(theme_minimal())
+
+fig2 <- plot_grid(
+  plot_grid(
+    phy_plot + theme(legend.position = "none")
+    , simp_plot
+    , ncol = 1
+    , align = "hv"
+    , labels=c("A", "B")
+    , label_size = 16)
+  , plot_grid(
+    get_legend(phy_plot)
+    , ggplot()
+    , ncol =1)
+  , rel_widths = c(7,3)
+)
+
 save_plot('~/Documents/Flynn_LRColon_XXXX_2017/submission/figure_2.pdf', fig2, ncol=1, nrow=2, base_width=6, base_height = 3)
 

### 2017-11-09 code/rf_optimized.R

@@ -165,7 +165,7 @@ mtext(side=1, text="Specificity", line=2.5)
 legend('bottomright', legend=c(
   sprintf('D Lum vs D Muc, AUC = 0.908'),
   sprintf('P Lum vs P Muc, AUC = 0.764')
-),lty=1, cex=0.8, col=c('red', 'blue'), bty='n')
+),lty=1, lwd = 2, cex=0.7, col=c('red', 'blue'), bty='n')
 
 mtext('A', side=2, line=2, las=1, adj=1.5, padj=-5, cex=1.5, font=2)
 
@@ -179,7 +179,7 @@ mtext(side=1, text="Specificity", line=2.5)
 legend('bottomright', legend=c(
   sprintf('D Muc vs P Muc, AUC = 0.850'),
   sprintf('D Lum vs P Lum, AUC = 0.580')
-), lty=1, cex=0.8, col=c('darkgreen', 'purple'), bty='n')
+), lty=1, lwd=2, cex=0.7, col=c('darkgreen', 'purple'), bty='n')
 
 mtext('B', side=2, line=2, las=1, adj=1.5, padj=-5, cex=1.5, font=2)
 

### 2017-11-09 submission/manuscript.tex
version sent to collabs

@@
 
-Word count: 3,965/5,000
+Word count: 3,953/5,000
 
@@
 The human colon is an ecosystem comprised of numerous microenvironments
-that select for different microbiota. Concentrations of oxygen, water
+that select for different microbiota. Concentrations of oxygen, water,
 and anti-microbial peptides change along the gut axis and influence
@@
 specific microbiota and communities are important for understanding the
-etiology of complex colon diseases such as Colorectal Cancer (CRC) and
+etiology of complex diseases such as Colorectal Cancer (CRC) and
 Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
@@
 what conclusions can be drawn about the colonic microbiome overall, let
-alone at that specific site. Due to these limitations, the contribution
-of the gut microbiota to CRC and IBD disease location in the colon is
-largely undefined. Characterizing these communities in healthy
-individuals could provide needed insight into disease etiology,
+alone at that specific site (12). Due to these limitations, the
+contribution of the gut microbiota to CRC and IBD disease location in
+the colon is largely undefined. Characterizing these communities in
+healthy individuals could provide needed insight into disease etiology,
 including how the disruption of the healthy community could promote the
@@
 human gut microbiome studies have been performed on whole shed feces or
-on samples collected during colonoscopy or surgery. While the latter
+on samples collected during colonoscopy or surgery (5). While the latter
 method allows investigators to acquire samples from inside the human
@@
 as the chemicals strip the bowel of contents and disrupt the mucosal
-layer (12,13). As such, what little information we do have about the
+layer (13,14). As such, what little information we do have about the
 spatial distribution of the microbiota in the proximal and distal colon
@@
 were collected and the contents immediately deposited into RNAlater
-(Fischer) and flash-frozen in liquid nitrogen. The forceps were
-withdrawn and new biopsy forceps were used to collect mucosal biopsies
-on sections of the colon that were pink and free of stool matter. Three
-mucosal biopsies were collected and flash-frozen in RNAlater. These
-samples comprised the distal colon samples. The sigmoidoscope was then
-withdrawn and a pediatric colonoscope was inserted to reach the proximal
-colon. Samples were then collected in the same manner as was done in the
-distal colon and the colonoscope withdrawn. All samples were stored at
+(Fisher) and flash-frozen in liquid nitrogen. The forceps were withdrawn
+and new biopsy forceps were used to collect mucosal biopsies on sections
+of the colon that were pink and free of stool matter. Three mucosal
+biopsies were collected and flash-frozen in RNAlater. These samples
+comprised the distal colon samples. The sigmoidoscope was then withdrawn
+and a pediatric colonoscope was inserted to reach the proximal colon.
+Samples were then collected in the same manner as was done in the distal
+colon and the colonoscope withdrawn. All samples were stored at
 -80\(^\circ\)C.
@@
 V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described (14). Sequences were curated
-using the mothur software as described previously (15). The sequences
+Illumina MiSeq as previously described (15). Sequences were curated
+using the mothur software as described previously (16). The sequences
 were assigned a taxonomic classification using a naive Bayesian
 classifier trained using a 16S rRNA gene training set from the Ribosomal
-Database Project (RDP) (16) and clustered into operational taxonomic
+Database Project (RDP) (17) and clustered into operational taxonomic
 units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
 of a mock community revealed the error rate to be 0.018\%. Samples were
-rarefied to 4231 sequences per sample in order to reduce uneven sampling
-bias.
+rarefied to 4231 sequences per sample in order to reduce the effects of
+uneven sampling bias.
 
@@
 and \(\theta\)\textsubscript{YC} calculator metrics in mothur version
-1.39.5 (15). \(\theta\)\textsubscript{YC} distances were calculated to
+1.39.5 (16). \(\theta\)\textsubscript{YC} distances were calculated to
 determine the dissimilarity between two samples. Random Forest
 classification models were built using the AUCRF R package using a
-leave-one-subject out approach (17). For each model the data was split
+leave-one-subject out approach (18). For each model the data were split
 into a 19-subject training set and a 1-subject test set. The model was
@@
 iteratively for all subjects and results plotted as Reciever Operator
-Characteristic curves using the pROC R package (18). Resultant models
+Characteristic curves using the pROC R package (19). Resultant models
 were used to identify the OTUs that were most important for classifying
@@
 reproducible data analysis pipeline can be found at
-\url{https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017}.
+\url{https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017}.
 
@@
 Luminal and mucosal samples were collected from the proximal and distal
-colon of 20 healthy humans that had not undergone bowel preparation
+colon of 20 healthy individuals who had not undergone bowel preparation
 (Fig. 1). Subjects also collected stool at home one week prior to the
@@
 sample. As expected, each site was primarily dominated by
-\emph{Firmicutes} and \emph{Bacteriodetes} (Fig. 2A) (19). Samples had
+\emph{Firmicutes} and \emph{Bacteriodetes} (Fig. 2A) (20). Samples had
 varying levels of diversity at each site, irrespective of the individual
@@
 (Fig. 3C); this is consistent with previous human microbiome studies
-that have sampled multiple sites of the human colon (20--22). Thus
+that have sampled multiple sites of the human colon (21--23). Thus
 interpersonal variation drove the differences between samples more than
@@
 
-\paragraph{Random Forest classification models identify important
-Operational Taxonomic Units (OTUs) on each
-side}\label{random-forest-classification-models-identify-important-operational-taxonomic-units-otus-on-each-side}
+\paragraph{Random Forest classification models identify important OTUs
+on each
+side}\label{random-forest-classification-models-identify-important-otus-on-each-side}
 
 To identify OTUs that were distinct at each site, we constructed several
-Random Forest models trained using OTU relative abundances. We used
-10-fold cross validation to build the first model to classify the
-luminal versus mucosal samples for the proximal and distal sides,
-independently (Fig. 4A). The models performed well when classifying
-these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
-were most predictive of each site were identified by their greatest mean
-decrease in accuracy when removed from the model. For distinguishing the
-proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
-\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
-included in the best model (Fig. 5A). The model to differentiate between
-the distal lumen and mucosa included OTUs affiliated with the
-\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
-\emph{Anaerococcus} (Fig. 5B). These results indicated that there were
-fine differences between the different sites of the colon, and that
-these could be traced to specific OTUs on each side.
+Random Forest models trained using OTU relative abundances. We built the
+first model to classify the luminal versus mucosal samples for the
+proximal and distal sides, independently (Fig. 4A). The models performed
+well when classifying these samples (proximal AUC = 0.764, distal AUC =
+0.908). The OTUs that were most predictive of each site were identified
+by their greatest mean decrease in accuracy when removed from the model.
+For distinguishing the proximal lumen and mucosa, OTUs affiliated with
+the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
+\emph{Enterobacteraceae} were included in the best model (Fig. 5A). The
+model to differentiate between the distal lumen and mucosa included OTUs
+affiliated with the \emph{Turicibacter}, \emph{Finegoldia},
+\emph{Peptoniphilus} and \emph{Anaerococcus} (Fig. 5B). These results
+indicated that there were fine differences between the different sites
+of the colon, and that these could be traced to specific OTUs on each
+side.
 
 Next, we built a Random Forest model to differentiate the proximal and
-distal luminal samples using 10-fold cross validation. The model
-performed best when distinguishing the proximal versus distal mucosa
-(Fig. 4B, AUC = 0.850) whereas the model to differentiate between the
-proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
-included in the model differentiating the distal and proximal mucosa
-included members of the \emph{Porphyromonas}, \emph{Murdochiella},
-\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Fig.
-6A). The model that attempted to separate the the proximal and distal
-lumen included OTUs affiliated with the \emph{Bacteroides},
-\emph{Clostridium IV} and \emph{Oscillibacter} (Fig. 6B). Interestingly,
-\emph{Anaerococcus} and \emph{Finegoldia} were distinct between the
-mucosa and lumen and also helped to differentiate between the proximal
-and distal sides.
+distal luminal samples. The model performed best when distinguishing the
+proximal versus distal mucosa (Fig. 4B, AUC = 0.850) whereas the model
+to differentiate between the proximal versus distal lumen performed
+poorly (AUC = 0.580). The OTUs included in the model differentiating the
+distal and proximal mucosa included members of the \emph{Porphyromonas},
+\emph{Murdochiella}, \emph{Finegoldia}, \emph{Anaerococcus} and
+\emph{Peptoniphilus} (Fig. 6A). The model that attempted to separate the
+the proximal and distal lumen included OTUs affiliated with the
+\emph{Bacteroides}, \emph{Clostridium IV} and \emph{Oscillibacter} (Fig.
+6B). Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were
+distinct between the mucosa and lumen and also helped to differentiate
+between the proximal and distal sides.
 
@@
 properties and be found on the surfaces of colorectal cancer tumors
-(23). There were 14 samples from 8 subjects with the \emph{F. nucleatum
+(24). There were 14 samples from 8 subjects with the \emph{F. nucleatum
 subsp. animalis} sequences. Of the samples with the highest relative
@@
 varium}. In addition to \emph{F. nucleatum}, \emph{F. varium} has been
-associated with IBD (24). Four subjects harbored \emph{F. varium} and
+associated with IBD (25). Four subjects harbored \emph{F. varium} and
 the samples were split evenly between the proximal and distal mucosa
@@
 
-Here we identified bacterial taxa that were specific to the lumen and
-mucosa of the proximal and distal sides of the human colon using samples
+We identified bacterial taxa that were specific to the lumen and mucosa
+of the proximal and distal sides of the human colon using samples
 collected during an unprepared colonoscopy of healthy subjects. We found
-that all locations contained a range of phyla relative abundances and a
+that all locations contained a range of phylum relative abundances and a
 range of diversity, but that there was a wide variability between
@@
 different when the samples are taken just 1 cm away from each other
-(25). Similar patchiness was also observed in luminal contents and fecal
+(26). Similar patchiness was also observed in luminal contents and fecal
 samples themselves; there was separation of different interacting
-microbes along the length of a stool sample, for instance (26). That
+microbes along the length of a stool sample, for instance (27). That
 said, across our samples, the mucosal samples harbor more
@@
 predictive of colorectal cancer in humans (9) and have oncogenic
-properties in cell culture and in mice (27). Interestingly, while
+properties in cell culture and in mice (28). Interestingly, while
 \emph{F. nucleatum} was found on both sides of the colon, \emph{P.
@@
 on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreased along the colon to rectum (28). In another study,
+decreased along the colon to rectum (29). In another study,
 \emph{Fusobacterium} was associated with MSI with BRAF and KRAS
-mutations, molecular features of proximal CRC (29). Of the 8 (40\%)
+mutations, molecular features of proximal CRC (30). Of the 8 (40\%)
 individuals positive for \emph{F. nucleatum} in the present study, the
@@
 only co-occur on CRC tumors but also to synergize to promote
-tumorigenesis in an oral cancer model (30) (31). Further analysis of the
+tumorigenesis in an oral cancer model (31,32). Further analysis of the
 distribution and activities of these pathogens along the colon is needed
@@
 The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
-been commonly isolated from mucosal biopsies of patients with IBD (24).
+been commonly isolated from mucosal biopsies of patients with IBD (25).
 Laboratory experiments with these isolates have shown that
@@
 (10), suggesting the bacteria may increase inflammation in the gut as
-well (32). \emph{F. varium} isolated from UC patients caused colonic
-ulcers in an experimental mouse model (33). \emph{F. varium} was only
+well (33). \emph{F. varium} isolated from UC patients caused colonic
+ulcers in an experimental mouse model (34). \emph{F. varium} was only
 detected in three subjects and two of those samples were isolated form
 the proximal mucosa (Supplementary Fig. S1B). \emph{F. varium} is most
-commonly isolated from UC patient biopsies from the ileum or cecum (34),
+commonly isolated from UC patient biopsies from the ileum or cecum (35),
 suggesting this species may exhibit preference for the different
@@
 patient's colonic mucosa and luminal contents prior to and after bowel
-cleansing (35). This study found that mucosal and luminal samples were
+cleansing (36). This study found that mucosal and luminal samples were
 distinguishable prior to bowel cleansing, but that bowel preparation
-resulted in an increase in shared OTUs between each site (35). After
+resulted in an increase in shared OTUs between each site (36). After
 seven days, bowel cleansing not only made the samples more difficult to
@@
 
+\hypertarget{ref-Flemer633}{}
+12. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
+al. Tumour-associated and non-tumour-associated microbiota in colorectal
+cancer. Gut {[}Internet{]}. BMJ Publishing Group; 2017;66:633--43.
+Available from: \url{http://gut.bmj.com/content/66/4/633}
+
 \hypertarget{ref-Jalanka2014}{}
-12. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
+13. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
 et al. Effects of bowel cleansing on the intestinal microbiota. Gut
@@
 \hypertarget{ref-Harrell2012}{}
-13. Harrell L, Wang Y, Antonopoulos D, Young V, Lichtenstein L, Huang Y,
+14. Harrell L, Wang Y, Antonopoulos D, Young V, Lichtenstein L, Huang Y,
 et al. Standard colonic lavage alters the natural state of
@@
 \hypertarget{ref-Kozich2013}{}
-14. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+15. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-Schloss2009}{}
-15. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+16. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, et al. Introducing mothur: Open-source, platform-independent,
@@
 \hypertarget{ref-Wang2007}{}
-16. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive bayesian classifier
+17. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive bayesian classifier
 for rapid assignment of rRNA sequences into the new bacterial taxonomy.
@@
 \hypertarget{ref-Calle_2011}{}
-17. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
+18. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
 for genomic profiling with random forest. Human Heredity {[}Internet{]}.
@@
 \hypertarget{ref-Robin_2011}{}
-18. Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez J-C, et
+19. Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez J-C, et
 al. pROC: An open-source package for r and s to analyze and compare ROC
@@
 \hypertarget{ref-LloydPrice2016}{}
-19. Lloyd-Price J, Abu-Ali G, Huttenhower C. The healthy human
+20. Lloyd-Price J, Abu-Ali G, Huttenhower C. The healthy human
 microbiome. Genome Medicine {[}Internet{]}. Springer Nature; 2016;8.
@@
 \hypertarget{ref-Eckburg2005}{}
-20. Eckburg PB. Diversity of the human intestinal microbial flora.
+21. Eckburg PB. Diversity of the human intestinal microbial flora.
 Science {[}Internet{]}. American Association for the Advancement of
@@
 \hypertarget{ref-deCarcer2010}{}
-21. Cárcer DA de, Cuív PÓ, Wang T, Kang S, Worthley D, Whitehall V, et
+22. Cárcer DA de, Cuív PÓ, Wang T, Kang S, Worthley D, Whitehall V, et
 al. Numerical ecology validates a biogeographical distribution and
@@
 \hypertarget{ref-Zhang2013}{}
-22. Zhang Z, Geng J, Tang X, Fan H, Xu J, Wen X, et al. Spatial
+23. Zhang Z, Geng J, Tang X, Fan H, Xu J, Wen X, et al. Spatial
 heterogeneity and co-occurrence patterns of human mucosal-associated
@@
 \hypertarget{ref-Castellarin2011}{}
-23. Castellarin M, Warren RL, Freeman JD, Dreolini L, Krzywinski M,
+24. Castellarin M, Warren RL, Freeman JD, Dreolini L, Krzywinski M,
 Strauss J, et al. Fusobacterium nucleatum infection is prevalent in
@@
 \hypertarget{ref-Lee2016}{}
-24. Lee Y, Eun CS, Lee AR, Park CH, Han DS. FusobacteriumIsolates
+25. Lee Y, Eun CS, Lee AR, Park CH, Han DS. FusobacteriumIsolates
 recovered from colonic biopsies of inflammatory bowel disease patients
@@
 \hypertarget{ref-Hong2011}{}
-25. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
+26. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
 Pyrosequencing-based analysis of the mucosal microbiota in healthy
@@
 \hypertarget{ref-Stearns2011}{}
-26. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
+27. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
 Cvitkovitch DG, et al. Bacterial biogeography of the human digestive
@@
 \hypertarget{ref-Sears2014}{}
-27. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
+28. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
 Host \& Microbe {[}Internet{]}. Elsevier BV; 2014;15:317--28. Available
@@
 \hypertarget{ref-Mima2016}{}
-28. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
+29. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
 Fusobacterium nucleatum in colorectal carcinoma tissue according to
@@
 \hypertarget{ref-Tahara_2014}{}
-29. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
+30. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
 al. Fusobacterium in colonic flora and molecular features of colorectal
@@
 \hypertarget{ref-Whitmore2014}{}
-30. Whitmore SE, Lamont RJ. Oral bacteria and cancer. Goldman WE,
+31. Whitmore SE, Lamont RJ. Oral bacteria and cancer. Goldman WE,
 editor. PLoS Pathogens {[}Internet{]}. Public Library of Science (PLoS);
@@
 \hypertarget{ref-Flynn2016}{}
-31. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
+32. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
 oral bacteria in colorectal cancer. McMahon K, editor. mSphere
@@
 \hypertarget{ref-Dharmani2011}{}
-32. Dharmani P, Strauss J, Ambrose C, Allen-Vercoe E, Chadee K.
+33. Dharmani P, Strauss J, Ambrose C, Allen-Vercoe E, Chadee K.
 Fusobacterium nucleatum infection of colonic cells stimulates MUC2 mucin
@@
 \hypertarget{ref-Ohkusa2003}{}
-33. Ohkusa T. Induction of experimental ulcerative colitis by
+34. Ohkusa T. Induction of experimental ulcerative colitis by
 fusobacterium varium isolated from colonic mucosa of patients with
@@
 \hypertarget{ref-Ohkusa2002}{}
-34. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
+35. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
 Fusobacterium varium localized in the colonic mucosa of patients with
@@
 \hypertarget{ref-Shobar2016}{}
-35. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
+36. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
 JE, et al. The effects of bowel preparation on microbiota-related
@@
 Comparison of microbial community structure between sites of the gut.
-\(\theta\)YC distances are shown to indicate the interpersonal
-dissimilarities between two sites -- each point represents one
-individual. In (A), comparisons of the proximal and distal mucosal and
-lumen are shown. In (B), comparisons of each site to the exit stool are
-shown. In (C), comparisons of samples from all subjects to each other
-(interpersonal) or within one subject (intrapersonal) are shown.
+\(\theta\)\textsubscript{YC} distances are shown to indicate the
+interpersonal dissimilarities between two sites -- each point represents
+one individual. In (A), comparisons of the proximal and distal mucosal
+and lumen are shown. In (B), comparisons of each site to the exit stool
+are shown. In (C), comparisons of samples from all subjects to each
+other (interpersonal) or within one subject (intrapersonal) are shown.
 
@@
 Random Forest classifies locations in the colon. A) Receiver Operator
-Characteristic curves are shown for the 10-fold cross validation of the
-Random Forest model classifying lumen and mucosal samples for the distal
-(red) and proximal (blue) sides of the colon. (B) Receiver Operator
-Characteristic curves are shown for the 10-fold cross validation of the
-Random Forest model classifying distal mucosa vs proximal mucosa (green)
-and distal lumen versus proximal lumen (purple).
+Characteristic curves are shown for the Random Forest model classifying
+lumen and mucosal samples for the distal (red) and proximal (blue) sides
+of the colon. (B) Receiver Operator Characteristic curves are shown for
+the 10-fold cross validation of the Random Forest model classifying
+distal mucosa vs proximal mucosa (green) and distal lumen versus
+proximal lumen (purple).
 

### 2017-11-09 code/random_forest.R

@@ -612,14 +612,14 @@ plot(1, type="n", ylim=c(0,length(right_otu_feat)*2), xlim=c(1e-4,3), log="x", y
 index <- 1
 for(i in right_otu_feat){
   stripchart(at=index-0.35, jitter(rs_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, lwd=0.5)
-  stripchart(at=index+0.35, jitter(rb_abunds[,i], amount=1e-5), pch=21, bg="black", method="jitter", jitter=0.2, add=T, lwd=0.5)
-  segments(median(rs_abunds[,i]),index-0.7,median(rs_abunds[,i]),index, lwd=3)
-  segments(median(rb_abunds[,i]),index+0.7,median(rb_abunds[,i]),index, lwd=3)
+  stripchart(at=index+0.35, jitter(rb_abunds[,i], amount=1e-5), pch=21, bg="gray29", method="jitter", jitter=0.2, add=T, lwd=0.5)
+  segments(median(rs_abunds[,i]),index-0.8,median(rs_abunds[,i]),index, lwd=3)
+  segments(median(rb_abunds[,i]),index+0.8,median(rb_abunds[,i]),index, lwd=3)
   index <- index + 2
 }
 axis(2, at=seq(1,index-2,2), labels=do.call(expression,formatted4), las=1, line=-0.5, tick=F, cex.axis=0.9)
 axis(1, at=c(1e-4, 1e-3, 1e-2, 1e-1, 1), label=c("0", "0.1", "1", "10", "100"))
-legend('topright', legend=c("P Muc", "P Lum"), pch=c(21, 21), pt.bg=c("black","white"), cex=0.8)
+legend('topright', legend=c("P Muc", "P Lum"), pch=c(21, 21), pt.bg=c("gray29","white"), cex=0.7)
 
 mtext('A', side=2, line=7.5, las=1, adj=2, padj=-4.5, cex=2, font=2)
 
@@ -637,14 +637,14 @@ plot(1, type="n", ylim=c(0,length(left_otu_feat)*2), xlim=c(1e-4,3), log="x", yl
 index <- 1
 for(i in left_otu_feat){
   stripchart(at=index-0.35, jitter(ls_abunds[,i], amount=1e-5), pch=21, bg="white", method="jitter", jitter=0.2, add=T, lwd=0.5)
[…]

### 2017-11-14 submission/manuscript.Rmd
version sent to collabs
first submission draft complete

@@
   word_document: default
+sansfont: Arial
 header-includes:
@@
 
-Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, D. Kim Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^
+Kaitlin J. Flynn^1^, Mack T. Ruffin IV^2^, D. Kim Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^
 
@@
 The human colon is an ecosystem comprised of numerous microenvironments that select for different microbiota.
-Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence which microbiota reside in each location.
+Concentrations of oxygen, water, and anti-microbial peptides change along the gut axis and influence which microbiota reside in each location.
 Microenvironments differ not only longitudinally along the colon, but also radially from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.
@@
 Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study.
-Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl or Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive.
+Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl, Versed and Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive.
 20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure.
@@
 #### Data availability
-16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number XXXX.
+16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number SRP124947 and PRJNA418115.
 A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017.
@@
 We would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman for their roles in study coordination, subject recruitment, sample collection and sample processing.
+Thanks also to GI Fellows Drs Amy Hosmer, Alexander Larson and Anna Tavakkoli who assisted Dr.
+Turgeon with the colonoscopies.
 

### 2017-11-14 submission/manuscript.tex
first submission draft complete

@@
   \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+    \setsansfont[]{Arial}
 \fi
@@
 
-Kaitlin J. Flynn\textsuperscript{1}, Charles C.
-Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2}, D.
-Kim Turgeon\textsuperscript{3\(\dagger\)}, and Patrick D.
-Schloss\textsuperscript{1\(\dagger\)}
+Kaitlin J. Flynn\textsuperscript{1}, Mack T. Ruffin
+IV\textsuperscript{2}, D. Kim Turgeon\textsuperscript{3\(\dagger\)}, and
+Patrick D. Schloss\textsuperscript{1\(\dagger\)}
 
@@
 days, use of antibiotics within 3 months, current use of anticoagulants,
-known allergies to Fentanyl or Benadryl, prior history of colon disease,
-diabetes, abdominal surgery, respiratory, liver, kidney or brain
-impairments, undergoing current chemotherapy or radiation treatment and
-subjects that were pregnant or trying to conceive. 20 subjects that met
-the criteria were selected and provided signed informed consent prior to
-the procedure. There were 13 female and 7 male subjects ranging in age
-from 25 to 64.
+known allergies to Fentanyl, Versed and Benadryl, prior history of colon
+disease, diabetes, abdominal surgery, respiratory, liver, kidney or
+brain impairments, undergoing current chemotherapy or radiation
+treatment and subjects that were pregnant or trying to conceive. 20
+subjects that met the criteria were selected and provided signed
+informed consent prior to the procedure. There were 13 female and 7 male
+subjects ranging in age from 25 to 64.
 
@@
 16S rRNA gene sequence reads and experiment metadata are available on
-the NCBI Sequence Read Archive (SRA) with accession number XXXX. A
-reproducible data analysis pipeline can be found at
+the NCBI Sequence Read Archive (SRA) with accession number SRP124947 and
+PRJNA418115. A reproducible data analysis pipeline can be found at
 \url{https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017}.
@@
 for their roles in study coordination, subject recruitment, sample
-collection and sample processing.
+collection and sample processing. Thanks also to GI Fellows Drs Amy
+Hosmer, Alexander Larson and Anna Tavakkoli who assisted Dr.~Turgeon
+with the colonoscopies.
 

### 2017-11-14 HIB.R
fixed readme

@@ -1,6 +1,0 @@
-#stuff for jay
-
-
-#trying to get file imported in a workable format
-myfile2 <- "sequence.gb"
-test2 <- read.dna(myfile2, format="fasta")

### 2018-01-27 submission/manuscript_with_revisions.Rmd
updated manuscript

--- 2017-11-14 submission/manuscript.Rmd
@@
 
-### Figures
+### Figure legends
 
@@
 
-![Fig 1](../submission/fig1.jpg)
-
-\newpage
-
 ####Figure 2
@@
 The horizontal lines represent the median values.
-
-![Fig 2](../submission/figure_2.pdf)
-
-\newpage
 
@@
 
-![Fig 3](../submission/figure_3.pdf)
-
-\newpage
-
 ####Figure 4
@@
 
-![Fig 4](../submission/figure_4.pdf)
-
-\newpage
 ####Figure 5
@@
 The vertical lines represent the median values for each OTU.
-
-![Fig 5](../submission/figure_5.pdf)
-
-\newpage
 
@@
 
-![Fig 6](../submission/figure_6.pdf)
-
-\newpage
-
 ####Figure S1
@@
 animalis_ (B) _Fusobacterium varium_ and (C) _Porphyromonas asacharolytica_
-![Fig S1](../submission/figure_S1.pdf)
-
-\newpage

### 2018-01-29 submission/manuscript.Rmd
manuscript edits

@@
 nucleatum_ decreased along the colon to rectum @Mima2016.
-In another study, _Fusobacterium_ was associated with MSI with BRAF and KRAS mutations, molecular features of proximal CRC @Tahara_2014.
 Of the 8 (40%) individuals positive for _F.
@@
 By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypotheses about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes.
-A better understanding of microbial activities in the gut can enhance microbiome-based screening and treatment modalities for these colon diseases.
+Further investigation of these samples using metagenomics and metatranscriptomics would illuminate the microbial activities in these gut microenvironments and could enhance microbiome-based screening and treatment modalities for colon disease.
 

### 2018-01-30 submission/manuscript.Rmd
paper discussion revisions

@@
 These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014.
-Interestingly, while _F.
-nucleatum_ was found on both sides of the colon, _P.
+Though the bacteria are known to co-localize on CRC tumors, in our study _F.
+nucleatum_ was found on both sides of the colon while _P.
 asacharolytica_ was only detected in the distal mucosa.
 Not much is known about the distribution of _P.
-asacharolytica_ along the colon, but given its anaerobic lifestyle and asacharolytic metabolism, it is not surprising that it resides in the less-oygen-rich and protein-rich distal mucosa @Albenberg2014.
+asacharolytica_ along the healthy colon, but given its anaerobic lifestyle and asacharolytic metabolism, it is perhaps not surprising that our study detected the bacteria primarily in the less-oygen-rich and protein-rich distal mucosa @Albenberg2014.
 In studies examining bacteria on colorectal cancer tumors, _F.
@@
 Of the 8 (40%) individuals positive for _F.
-nucleatum_ in the present study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa.
-Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacterium but the structure of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014.
-Finally, _Fusobacterium_ and _Porphyromonas_ populations not only co-occur on CRC tumors but also to synergize to promote tumorigenesis in an oral cancer model [@Whitmore2014; @Flynn2016].
-Further analysis of the distribution and activities of these pathogens along the colon is needed to elucidate a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.
-
-The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD @Lee2016.
-Laboratory experiments with these isolates have shown that disease-isolated _F.
-nucleatum_ are more invasive and stimulate more TNF-$\alpha$ production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011.
-_F.
-varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model @Ohkusa2003.
-_F.
+nucleatum_ in our study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa.
+The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD and UC [@Lee2016; @Ohkusa2003].
+In our study, _F.
 varium_ was only detected in three subjects and two of those samples were isolated form the proximal mucosa (Supplementary Fig.
@@
 _F.
-varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum @Ohkusa2002, suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites.
-Further work will assess how gut environment may select for species which may then cause localized disease.
+varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum (adjacent to the proximal colon) @Ohkusa2002, suggesting this species may exhibit preference for the higher oxygen content of these gastrointestinal sites.
+
+The distribution of these disease-associated microbes in healthy patients is of interest their presence can be predictive of disease prior to the onset of symptoms @Baxter2016.
+Spatial organization of _Fusobacteria_ and other bacterial species into polymicrobial biofilms that can invade the gut mucosa have been linked CRC (cite new paper).
+The organization into biofilms promotes tumorigenesis by allowing bacteria to grow near the epithelium, promoting inflammation, genotoxicity and metabolic changes that favor tumor cell growth (cite new paper and kaitlin).
+In one study of CRC biofilms and tumors, all of the proximal tumors examined contained a polymicrobial biofilm on the tumor mucosa @Dejea2014.
+Further examination of these tumors identified _Fusobacteria_ species as members of the biofilm.
+These results indicate that it is not only the presence of the bacterium but the tumor community as a whole that contributes to tumorigenesis @Dejea2014.
+Driver passeneger model (cite new paper and KF).
+Need to study healthy people/biofilms more to understand how can affect disease.
+Further analysis of the distribution and activities of these pathogens along the colon is needed to elucidate a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.
+
+paragraph on bacterial organization/spatial variation along colon vs at tumor site/biofilms
 

### 2018-01-30 submission/manuscript.Rmd
discussion updated revision

@@
 
-The distribution of these disease-associated microbes in healthy patients is of interest their presence can be predictive of disease prior to the onset of symptoms @Baxter2016.
 Spatial organization of _Fusobacteria_ and other bacterial species into polymicrobial biofilms that can invade the gut mucosa have been linked CRC (cite new paper).
@@
 These results indicate that it is not only the presence of the bacterium but the tumor community as a whole that contributes to tumorigenesis @Dejea2014.
-Driver passeneger model (cite new paper and KF).
-Need to study healthy people/biofilms more to understand how can affect disease.
-Further analysis of the distribution and activities of these pathogens along the colon is needed to elucidate a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.
-
-paragraph on bacterial organization/spatial variation along colon vs at tumor site/biofilms
+A 'driver-passenger' model has been proposed as a mechanism for biofilm assembly in the gut (Tjalsma 2012, flynn).
+In this model, 'driver' species such as _Fusobacterium spp_ and _Porphyromonas spp_ exert tumorigenic effects locally and create a niche for colonization or adherence of 'passenger' species that comprise the rest of the biofilm (Tjalsma and Flynn).
+Thus the the distribution of these disease-associated microbes in healthy patients is of interest as their presence can be predictive of disease prior to the onset of symptoms @Baxter2016.
+A better understanding of the early microbial changes in the gut microboiome is essential for elucidating a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.
 

### 2018-01-30 submission/manuscript.Rmd
reviewer revisions completed

@@
 The identity of these specific microbiota and communities are important for understanding the etiology of complex diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD).
-CRC and IBD can be preceded or accelerated by perturbations the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017].
+CRC and IBD can be preceded or accelerated by perturbations of the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017].
 The manifestations of these diseases are known to vary based upon the location in which they occur.
 For instance, CRC that arises in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012.
-In contrast, CRC arising in the proximal (right) colon tumors are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012.
+In contrast, CRC arising in the proximal (right) colon are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012.
 In addition to the environmental gradients within the colon, the distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively [@Yamauchi2012; @Albenberg2014; @Donaldson2015].
@@
 The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy or surgery @Donaldson2015.
-While the latter method allows investigators to acquire samples from inside the human colon, typically these procedures are preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
+While invasive methods allow investigators to acquire samples from inside the human colon, typically these procedures are preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel.
 Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer [@Jalanka2014; @Harrell2012].
@@
 There were 13 female and 7 male subjects ranging in age from 25 to 64.
+18 of the 20 subjects had not used antibiotics within a year prior to the collection date and 2 had not used antibiotics within 6 months.
+None of the subjects had medical conditions requiring frequent or extended antibiotic use.
 
@@
 At a baseline visit, subjects gave consent and were given a home collection stool kit (Zymo).
-At least one week prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice.
+One to seven days prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice.
 Notably, subjects did not undergo any bowel preparation method prior to sampling.
@@
 The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the proximal colon.
+The proximal samples were taken from the ascending colon proximal to the hepatic flexure at 75-120cm depending on the subject.
 Samples were then collected in the same manner as was done in the distal colon and the colonoscope withdrawn.
@@
 For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples.
-The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013.
+The resulting DNA was normalized to equal concentrations across all samples and used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013.
 Sequences were curated using the mothur software as described previously @Schloss2009.
@@
 Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011.
+The Random Forest models were built using the full, non-rarefied, dataset as input.
 For each model the data were split into a 19-subject training set and a 1-subject test set.
-The model was built and cross-validated using AUCRFcv on the training set.
-The model was then tested on the left-out patient.
+The model was built and cross-validated using 10-fold k cross-validation (AUCRFcv) on the training set to estimate the prediction error of the model.
+The resultant model was then used to predict the outcome the left-out subject.
 This process was repeated iteratively for all subjects and results plotted as Reciever Operator Characteristic curves using the pROC R package @Robin_2011.
@@
 _Bacteriodes_ and _Firmicutes_), there was a wide amount of variation.
-This likely reflects not only the variability between human subjects, casued by differences in age, sex, diet, but also spatial "patchiness" in the gut microbiome.
+This likely reflects not only the variability between human subjects, caused by differences in age, sex, and diet, but may also reflect spatial patchiness in the gut microbiome within a subject.
+Patchiness refers to inconsistent distribution of microbial populations due to fluctuations in local resources @Rougharden1977.
 One study noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011.
 Similar patchiness was also observed in luminal contents and fecal samples themselves; there was separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011.
-That said, across our samples, the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to luminal content in humans @Albenberg2014.
-Hence, the conclusions we were able to draw from phyla analysis may have been impacted by inter-subject patchiness.
+A third study that sampled six mucosal sites along the colon observed such patchiness in two of the three study subjects @Eckburg2005.
+While our subjects were not sampled frequently enough to draw specific conclusions about patchiness along the unprepped colon, we did observe some specific differences in mucosal versus luminal samples at the phylum level.
+The mucosal samples harbored more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to luminal content in humans @Albenberg2014.
+However, we must still consider that the results from phyla analysis may have been impacted by inter-subject patchiness.
 
@@
 The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested.
-We speculate that the model was less effective at classifying the proximal and distal luminal contents because the mucosal microenvironments have variable selective pressure along the colon than the luminal microenvironments.
+We speculate that the model was less effective at classifying the proximal and distal luminal contents because the mucosal microenvironments have more variable selective pressure along the colon than the luminal microenvironments.
 
@@
 
-Spatial organization of _Fusobacteria_ and other bacterial species into polymicrobial biofilms that can invade the gut mucosa have been linked CRC (cite new paper).
-The organization into biofilms promotes tumorigenesis by allowing bacteria to grow near the epithelium, promoting inflammation, genotoxicity and metabolic changes that favor tumor cell growth (cite new paper and kaitlin).
+Spatial organization of _Fusobacteria_ and other bacterial species into polymicrobial biofilms that can invade the gut mucosa have been linked to CRC @Li2017.
+The biofilms promote tumorigenesis by allowing bacteria to grow near the epithelium, inducing inflammation, genotoxicity and metabolic changes that favor tumor cell growth @Li2017.
 In one study of CRC biofilms and tumors, all of the proximal tumors examined contained a polymicrobial biofilm on the tumor mucosa @Dejea2014.
@@
 These results indicate that it is not only the presence of the bacterium but the tumor community as a whole that contributes to tumorigenesis @Dejea2014.
-A 'driver-passenger' model has been proposed as a mechanism for biofilm assembly in the gut (Tjalsma 2012, flynn).
-In this model, 'driver' species such as _Fusobacterium spp_ and _Porphyromonas spp_ exert tumorigenic effects locally and create a niche for colonization or adherence of 'passenger' species that comprise the rest of the biofilm (Tjalsma and Flynn).
+A 'driver-passenger' model has been proposed as a mechanism for biofilm assembly in the gut [@Tjalsma2012; @Flynn2016].
+In this model, 'driver' species such as _Fusobacterium spp_ and _Porphyromonas spp_ exert tumorigenic effects locally and create a niche for adherence of 'passenger' species that comprise the rest of the biofilm [@Tjalsma2012; @Flynn2016].
 Thus the the distribution of these disease-associated microbes in healthy patients is of interest as their presence can be predictive of disease prior to the onset of symptoms @Baxter2016.
@@
 By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypotheses about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes.
-Further investigation of these samples using metagenomics and metatranscriptomics would illuminate the microbial activities in these gut microenvironments and could enhance microbiome-based screening and treatment modalities for colon disease.
+Future investigation of these samples using metagenomics and metatranscriptomics would illuminate the microbial activities in these gut microenvironments.
+Further, combining this approach with a more comprehensive sampling strategy along the unprepped colon could enhance microbiome-based screening and treatment modalities for colon disease.
 

### 2018-01-30 submission/manuscript.tex
took figures out of manuscript.Rmd'
reviewer revisions completed

@@
 Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
-accelerated by perturbations the structure of the gut microbiota (1--3).
-The manifestations of these diseases are known to vary based upon the
-location in which they occur. For instance, CRC that arises in the
-distal (left) colon are of hindgut origin and tend to have large
+accelerated by perturbations of the structure of the gut microbiota
+(1--3). The manifestations of these diseases are known to vary based
+upon the location in which they occur. For instance, CRC that arises in
+the distal (left) colon are of hindgut origin and tend to have large
 chromosomal alterations indicative of chromosomal instability (1). In
-contrast, CRC arising in the proximal (right) colon tumors are of midgut
-origin and tend to be sessile and microsatellite instable (MSI with BRAF
-and KRAS mutations) (1). In addition to the environmental gradients
-within the colon, the distal and proximal sides of the colon differ in
-the amount of inflammation present and the genomic instability of
+contrast, CRC arising in the proximal (right) colon are of midgut origin
+and tend to be sessile and microsatellite instable (MSI with BRAF and
+KRAS mutations) (1). In addition to the environmental gradients within
+the colon, the distal and proximal sides of the colon differ in the
+amount of inflammation present and the genomic instability of
 precancerous cells, respectively (1,4,5). In IBD patients, disease
@@
 human gut microbiome studies have been performed on whole shed feces or
-on samples collected during colonoscopy or surgery (5). While the latter
-method allows investigators to acquire samples from inside the human
+on samples collected during colonoscopy or surgery (5). While invasive
+methods allow investigators to acquire samples from inside the human
 colon, typically these procedures are preceded by the use of bowel
@@
 informed consent prior to the procedure. There were 13 female and 7 male
-subjects ranging in age from 25 to 64.
+subjects ranging in age from 25 to 64. 18 of the 20 subjects had not
+used antibiotics within a year prior to the collection date and 2 had
+not used antibiotics within 6 months. None of the subjects had medical
+conditions requiring frequent or extended antibiotic use.
 
@@
 At a baseline visit, subjects gave consent and were given a home
-collection stool kit (Zymo). At least one week prior to the scheduled
+collection stool kit (Zymo). One to seven days prior to the scheduled
 colonoscopy, subjects collected whole stool at home and shipped the
@@
 and a pediatric colonoscope was inserted to reach the proximal colon.
-Samples were then collected in the same manner as was done in the distal
-colon and the colonoscope withdrawn. All samples were stored at
--80\(^\circ\)C.
+The proximal samples were taken from the ascending colon proximal to the
+hepatic flexure at 75-120cm depending on the subject. Samples were then
+collected in the same manner as was done in the distal colon and the
+colonoscope withdrawn. All samples were stored at -80\(^\circ\)C.
 
@@
 added to the bead beating step to enhance DNA recovery from mucosal
-samples. The resulting DNA was used as template for amplification of the
-V4 region of the 16S rRNA gene and fragments were sequenced on an
-Illumina MiSeq as previously described (15). Sequences were curated
-using the mothur software as described previously (16). The sequences
-were assigned a taxonomic classification using a naive Bayesian
-classifier trained using a 16S rRNA gene training set from the Ribosomal
-Database Project (RDP) (17) and clustered into operational taxonomic
-units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
-of a mock community revealed the error rate to be 0.018\%. Samples were
-rarefied to 4231 sequences per sample in order to reduce the effects of
-uneven sampling bias.
+samples. The resulting DNA was normalized to equal concentrations across
+all samples and used as template for amplification of the V4 region of
+the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as
+previously described (15). Sequences were curated using the mothur
+software as described previously (16). The sequences were assigned a
+taxonomic classification using a naive Bayesian classifier trained using
+a 16S rRNA gene training set from the Ribosomal Database Project (RDP)
+(17) and clustered into operational taxonomic units (OTUs) based on a
+97\% similarity cutoff. Sequencing and analysis of a mock community
+revealed the error rate to be 0.018\%. Samples were rarefied to 4231
+sequences per sample in order to reduce the effects of uneven sampling
+bias.
 
@@
 classification models were built using the AUCRF R package using a
-leave-one-subject out approach (18). For each model the data were split
-into a 19-subject training set and a 1-subject test set. The model was
-built and cross-validated using AUCRFcv on the training set. The model
-was then tested on the left-out patient. This process was repeated
-iteratively for all subjects and results plotted as Reciever Operator
-Characteristic curves using the pROC R package (19). Resultant models
-were used to identify the OTUs that were most important for classifying
-each location. Species-level information for sequences of interest was
-obtained by aligning the sequences to the GenBank nucleotide databse
-using blastn. The species name was only used if the identity score was
-\(\ge\) 99\% over the full-length of the contig and matched a single
-reference.
+leave-one-subject out approach (18). The Random Forest models were built
+using the full, non-rarefied, dataset as input. For each model the data
+were split into a 19-subject training set and a 1-subject test set. The
+model was built and cross-validated using 10-fold k cross-validation
+(AUCRFcv) on the training set to estimate the prediction error of the
+model. The resultant model was then used to predict the outcome the
+left-out subject. This process was repeated iteratively for all subjects
+and results plotted as Reciever Operator Characteristic curves using the
+pROC R package (19). Resultant models were used to identify the OTUs
+that were most important for classifying each location. Species-level
+information for sequences of interest was obtained by aligning the
+sequences to the GenBank nucleotide databse using blastn. The species
+name was only used if the identity score was \(\ge\) 99\% over the
+full-length of the contig and matched a single reference.
 
@@
 of variation. This likely reflects not only the variability between
-human subjects, casued by differences in age, sex, diet, but also
-spatial ``patchiness'' in the gut microbiome. One study noted that the
+human subjects, caused by differences in age, sex, and diet, but may
+also reflect spatial patchiness in the gut microbiome within a subject.
+Patchiness refers to inconsistent distribution of microbial populations
+due to fluctuations in local resources (26). One study noted that the
 bacteria recoverable from the same mucosal sample location can be vastly
 different when the samples are taken just 1 cm away from each other
-(26). Similar patchiness was also observed in luminal contents and fecal
+(27). Similar patchiness was also observed in luminal contents and fecal
 samples themselves; there was separation of different interacting
-microbes along the length of a stool sample, for instance (27). That
-said, across our samples, the mucosal samples harbor more
-\emph{Proteobacteria}, consistent with previous studies comparing
-mucosal swabs to luminal content in humans (4). Hence, the conclusions
-we were able to draw from phyla analysis may have been impacted by
-inter-subject patchiness.
+microbes along the length of a stool sample, for instance (28). A third
+study that sampled six mucosal sites along the colon observed such
+patchiness in two of the three study subjects (21). While our subjects
+were not sampled frequently enough to draw specific conclusions about
+patchiness along the unprepped colon, we did observe some specific
+differences in mucosal versus luminal samples at the phylum level. The
+mucosal samples harbored more \emph{Proteobacteria}, consistent with
+previous studies comparing mucosal swabs to luminal content in humans
+(4). However, we must still consider that the results from phyla
+analysis may have been impacted by inter-subject patchiness.
 
@@
 the proximal and distal luminal contents because the mucosal
-microenvironments have variable selective pressure along the colon than
-the luminal microenvironments.
+microenvironments have more variable selective pressure along the colon
+than the luminal microenvironments.
 
@@
 predictive of colorectal cancer in humans (9) and have oncogenic
-properties in cell culture and in mice (28). Interestingly, while
-\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
-asacharolytica} was only detected in the distal mucosa. Not much is
-known about the distribution of \emph{P. asacharolytica} along the
-colon, but given its anaerobic lifestyle and asacharolytic metabolism,
-it is not surprising that it resides in the less-oygen-rich and
-protein-rich distal mucosa (4). In studies examining bacteria on
-colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
-on proximal-sided tumors, and distribution of \emph{F. nucleatum}
-decreased along the colon to rectum (29). In another study,
-\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
-mutations, molecular features of proximal CRC (30). Of the 8 (40\%)
-individuals positive for \emph{F. nucleatum} in the present study, the
-bacterium was spread across the proximal mucosa, distal lumen and distal
-mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
-suggests that \emph{Fusobacteria} species are more commonly found on
-proximal tumors and in biofilms, indicating that it is not only the
-presence of the bacterium but the structure of the tumor community that
-contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
-Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
-only co-occur on CRC tumors but also to synergize to promote
-tumorigenesis in an oral cancer model (31,32). Further analysis of the
-distribution and activities of these pathogens along the colon is needed
-to elucidate a mechanism for development of CRC or IBD subtypes in the
-proximal or distal colon.
+properties in cell culture and in mice (29). Though the bacteria are
+known to co-localize on CRC tumors, in our study \emph{F. nucleatum} was
+found on both sides of the colon while \emph{P. asacharolytica} was only
+detected in the distal mucosa. Not much is known about the distribution
+of \emph{P. asacharolytica} along the healthy colon, but given its
+anaerobic lifestyle and asacharolytic metabolism, it is perhaps not
+surprising that our study detected the bacteria primarily in the
+less-oygen-rich and protein-rich distal mucosa (4). In studies examining
+bacteria on colorectal cancer tumors, \emph{F. nucleatum} was more
+commonly detected on proximal-sided tumors, and distribution of \emph{F.
+nucleatum} decreased along the colon to rectum (30). Of the 8 (40\%)
+individuals positive for \emph{F. nucleatum} in our study, the bacterium
+was spread across the proximal mucosa, distal lumen and distal mucosa.
+The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
+been commonly isolated from mucosal biopsies of patients with IBD and UC
+(25,31). In our study, \emph{F. varium} was only detected in three
+subjects and two of those samples were isolated form the proximal mucosa
+(Supplementary Fig. S1B). \emph{F. varium} is most commonly isolated
+from UC patient biopsies from the ileum or cecum (adjacent to the
+proximal colon) (32), suggesting this species may exhibit preference for
+the higher oxygen content of these gastrointestinal sites.
 
-The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
-been commonly isolated from mucosal biopsies of patients with IBD (25).
-Laboratory experiments with these isolates have shown that
-disease-isolated \emph{F. nucleatum} are more invasive and stimulate
-more TNF-\(\alpha\) production than strains from healthy individuals
-(10), suggesting the bacteria may increase inflammation in the gut as
-well (33). \emph{F. varium} isolated from UC patients caused colonic
-ulcers in an experimental mouse model (34). \emph{F. varium} was only
-detected in three subjects and two of those samples were isolated form
-the proximal mucosa (Supplementary Fig. S1B). \emph{F. varium} is most
-commonly isolated from UC patient biopsies from the ileum or cecum (35),
-suggesting this species may exhibit preference for the different
-environmental conditions of these gastrointestinal sites. Further work
-will assess how gut environment may select for species which may then
-cause localized disease.
+Spatial organization of \emph{Fusobacteria} and other bacterial species
+into polymicrobial biofilms that can invade the gut mucosa have been
+linked to CRC (33). The biofilms promote tumorigenesis by allowing
+bacteria to grow near the epithelium, inducing inflammation,
+genotoxicity and metabolic changes that favor tumor cell growth (33). In
+one study of CRC biofilms and tumors, all of the proximal tumors
+examined contained a polymicrobial biofilm on the tumor mucosa (7).
+Further examination of these tumors identified \emph{Fusobacteria}
+species as members of the biofilm. These results indicate that it is not
+only the presence of the bacterium but the tumor community as a whole
+that contributes to tumorigenesis (7). A `driver-passenger' model has
+been proposed as a mechanism for biofilm assembly in the gut (34,35). In
+this model, `driver' species such as \emph{Fusobacterium spp} and
+\emph{Porphyromonas spp} exert tumorigenic effects locally and create a
+niche for adherence of `passenger' species that comprise the rest of the
+biofilm (34,35). Thus the the distribution of these disease-associated
+microbes in healthy patients is of interest as their presence can be
+predictive of disease prior to the onset of symptoms (9). A better
+understanding of the early microbial changes in the gut microboiome is
+essential for elucidating a mechanism for development of CRC or IBD
+subtypes in the proximal or distal colon.
 
@@
 form hypotheses about how specific host-microbe interactions can affect
-disease progression of proximal and distal CRC and IBD subtypes. A
-better understanding of microbial activities in the gut can enhance
-microbiome-based screening and treatment modalities for these colon
-diseases.
+disease progression of proximal and distal CRC and IBD subtypes. Future
+investigation of these samples using metagenomics and
+metatranscriptomics would illuminate the microbial activities in these
+gut microenvironments. Further, combining this approach with a more
+comprehensive sampling strategy along the unprepped colon could enhance
+microbiome-based screening and treatment modalities for colon disease.
 
@@
 
+\hypertarget{ref-Rougharden1977}{}
+26. Rougharden JD. Patchiness in the spatial distribution of a
+population caused by stochastic fluctuations in resources. Oikos
+{[}Internet{]}. JSTOR; 1977;29:52. Available from:
+\url{https://doi.org/10.2307/3543292}
+
 \hypertarget{ref-Hong2011}{}
-26. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
+27. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
 Pyrosequencing-based analysis of the mucosal microbiota in healthy
@@
 \hypertarget{ref-Stearns2011}{}
-27. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
+28. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
 Cvitkovitch DG, et al. Bacterial biogeography of the human digestive
@@
 \hypertarget{ref-Sears2014}{}
-28. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
+29. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
 Host \& Microbe {[}Internet{]}. Elsevier BV; 2014;15:317--28. Available
@@
 \hypertarget{ref-Mima2016}{}
-29. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
+30. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
 Fusobacterium nucleatum in colorectal carcinoma tissue according to
@@
 
-\hypertarget{ref-Tahara_2014}{}
-30. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
-al. Fusobacterium in colonic flora and molecular features of colorectal
-carcinoma. Cancer Research {[}Internet{]}. American Association for
-Cancer Research (AACR); 2014;74:1311--8. Available from:
-\url{https://doi.org/10.1158\%2F0008-5472.can-13-1865}
-
-\hypertarget{ref-Whitmore2014}{}
-31. Whitmore SE, Lamont RJ. Oral bacteria and cancer. Goldman WE,
-editor. PLoS Pathogens {[}Internet{]}. Public Library of Science (PLoS);
-2014;10:e1003933. Available from:
-\url{https://doi.org/10.1371/journal.ppat.1003933}
-
-\hypertarget{ref-Flynn2016}{}
-32. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
-oral bacteria in colorectal cancer. McMahon K, editor. mSphere
-{[}Internet{]}. American Society for Microbiology; 2016;1:e00102--16.
-Available from: \url{https://doi.org/10.1128/msphere.00102-16}
-
-\hypertarget{ref-Dharmani2011}{}
-33. Dharmani P, Strauss J, Ambrose C, Allen-Vercoe E, Chadee K.
-Fusobacterium nucleatum infection of colonic cells stimulates MUC2 mucin
-and tumor necrosis factor alpha. Infection and Immunity {[}Internet{]}.
-American Society for Microbiology; 2011;79:2597--607. Available from:
-\url{https://doi.org/10.1128/iai.05118-11}
-
 \hypertarget{ref-Ohkusa2003}{}
-34. Ohkusa T. Induction of experimental ulcerative colitis by
+31. Ohkusa T. Induction of experimental ulcerative colitis by
 fusobacterium varium isolated from colonic mucosa of patients with
@@
 \hypertarget{ref-Ohkusa2002}{}
-35. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
+32. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
 Fusobacterium varium localized in the colonic mucosa of patients with
@@
 \url{https://doi.org/10.1046/j.1440-1746.2002.02834.x}
+
+\hypertarget{ref-Li2017}{}
+33. Li S, Konstantinov SR, Smits R, Peppelenbosch MP. Bacterial biofilms
+in colorectal cancer initiation and progression. Trends in Molecular
+Medicine {[}Internet{]}. Elsevier BV; 2017;23:18--30. Available from:
+\url{https://doi.org/10.1016/j.molmed.2016.11.004}
+
+\hypertarget{ref-Tjalsma2012}{}
+34. Tjalsma H, Boleij A, Marchesi JR, Dutilh BE. A bacterial
+driverpassenger model for colorectal cancer: Beyond the usual suspects.
+Nature Reviews Microbiology {[}Internet{]}. Springer Nature;
+2012;10:575--82. Available from:
+\url{https://doi.org/10.1038/nrmicro2819}
+
+\hypertarget{ref-Flynn2016}{}
+35. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
+oral bacteria in colorectal cancer. McMahon K, editor. mSphere
+{[}Internet{]}. American Society for Microbiology; 2016;1:e00102--16.
+Available from: \url{https://doi.org/10.1128/msphere.00102-16}
 
@@
 
-\subsubsection{Figures}\label{figures}
+\subsubsection{Figure legends}\label{figure-legends}
 
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{../submission/fig1.jpg}
-\caption{Fig 1}
-\end{figure}
-
-\newpage
-
 \paragraph{Figure 2}\label{figure-2}
@@
 location. The horizontal lines represent the median values.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{../submission/figure_2.pdf}
-\caption{Fig 2}
-\end{figure}
-
-\newpage
 
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{../submission/figure_3.pdf}
-\caption{Fig 3}
-\end{figure}
-
-\newpage
-
 \paragraph{Figure 4}\label{figure-4}
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{../submission/figure_4.pdf}
-\caption{Fig 4}
-\end{figure}
-
-\newpage
-
 \paragraph{Figure 5}\label{figure-5}
@@
 lines represent the median values for each OTU.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{../submission/figure_5.pdf}
-\caption{Fig 5}
-\end{figure}
-
-\newpage
 
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{../submission/figure_6.pdf}
-\caption{Fig 6}
-\end{figure}
-
-\newpage
-
 \paragraph{Figure S1}\label{figure-s1}
@@
 \emph{Fusobacterium varium} and (C) \emph{Porphyromonas asacharolytica}
-\includegraphics{../submission/figure_S1.pdf}
-
-\newpage
 

### 2018-02-08 submission/manuscript.Rmd
model updates

@@
 tyc <- read.table(file ="../data/mothur/kws_final.an.summary", sep = '\t', header = T, row.names=NULL)
-
-```
-
-```{r otu_barplot, echo=FALSE}
-source('../code/Sum_OTU_by_Tax.R')
-source('../code/sum_shared.R')
-
-#use this code to assign phyla to each OTU in the shared file 
-sub <- '../data/mothur/kws_final.an.0.03.subsample.shared'
-shared_phyla <- get_tax_level_shared(sub, tax, 2)
-phyla_met <- merge(meta_file, shared_phyla, by.x='group', by.y='row.names')
-
-#try to get the df organized to work as a boxplot - no median calculation 
-phyla_test <- phyla_met[, c("location","Firmicutes","Bacteroidetes","Proteobacteria","Verrucomicrobia","Actinobacteria","Fusobacteria")]
-subsampled_to <- 4231
-RA <- function(x) 100*x/subsampled_to
-
-phyla_RA <- data.frame(phyla_test[1], apply(phyla_test[2:ncol(phyla_test)],2, RA))
-phylaRAnames <- colnames(phyla_RA[,1:7])
-phyla_RAmelt <- melt(phyla_RA[, phylaRAnames], id.vars=1)
 
@@
 #update this to just be importing end files off of optimization scripts to pull AUCs from
+
 ```

### 2018-02-08 code/bowel_optimized.R

@@ -18,19 +18,34 @@ meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
 shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
 tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
 
+subs_file <- read.table(file='data/mothur/kws_final.an.0.03.subsample.shared', sep = '\t', header = T, row.names=2)
+
 #make OTU abundance file
 #Create df with relative abundances
 shared_file <- subset(shared_file, select = -c(numOtus, label))
 shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
-
 rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+
+#do rel abund calcs for subsampled
+subs_file <- subset(subs_file, select = -c(numOtus, label))
+subs_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+subs_abund <- 100*subs_file/unique(apply(subs_file, 1, sum))
+
 
 #Create vector of OTUs with median abundances >1%
 OTUs_1 <- apply(rel_abund, 2, max) > 1
 OTU_list <- colnames(rel_abund)[OTUs_1]
+
+OTUs_sub <- apply(subs_abund, 2, max) > 1
+OTU_list_sub <- colnames(subs_abund)[OTUs_sub]
+
 #get df of just top OTUs
 rel_abund_top <- rel_abund[, OTUs_1]
 rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+subs_abund_top <- subs_abund[, OTUs_sub]
+subs_meta <- merge(meta_file, subs_abund_top, by.x='group', by.y="row.names")
+
 
 print("loaded and organized data")
 
@@ -40,7 +55,7 @@ n_trees <- 2001
 source('code/random_functions.R')
 source('code/tax_level.R')
 
-testsub <- subset(rel_meta, location %in% c("LB", "RB"))
+testsub <- subset(subs_meta, location %in% c("LB", "RB"))
 testsub$location <- factor(testsub$location)
 levels(testsub$location) <- c(1:length(levels(testsub$location))-1)
 

### 2018-02-08 code/build_figure4.R

@@ -79,8 +79,24 @@ right_optimized_results <- read.table(file = 'data/process/right_optimized.tsv',
 bowel_optimized_results <- read.table(file = 'data/process/bowel_optimized.tsv', sep = '\t')
 stool_optimized_results <- read.table(file = 'data/process/stool_optimized.tsv', sep = '\t')
 
+##############
+# make test datasets for each
+
+testsubS <- subset(subs_meta, location %in% c("RS", "LS"))
+testsubS$location <- factor(testsubS$location)
+levels(testsubS$location) <- c(1:length(levels(testsubS$location))-1)
+
+testsubM <- subset(subs_meta, location %in% c("RB", "LB"))
+testsubM$location <- factor(testsubM$location)
+levels(testsubM$location) <- c(1:length(levels(testsubM$location))-1)
+
+
+
+
 left_roc <- roc(testsub$location ~ left_optimized_results$one)
 right_roc <- roc(testsubR$location ~ right_optimized_results$one)
+
+bowel_optimized_results <- bowel_optimized_results[-13,]
 
 muc_roc <- roc(testsubM$location ~ bowel_optimized_results$one)
 stool_roc <- roc(testsubS$location ~ stool_optimized_results$one)

### 2018-02-08 code/build_figure5.R

@@ -2,7 +2,7 @@
 # Build figure 5 - abundance charts of feature selected OTUs from RF models
 # Kaitlin Flynn, Schloss lab, updated 12-20-17
 
-PUT TAX FILE AND SHARED META UP IN HURR
+# PUT TAX FILE AND SHARED META UP IN HURR
 source('code/random_functions.R')
 source('code/tax_level.R')
 #from optimized model code that is run on the cluster, import file with OTUs and abundances 

### 2018-02-08 code/left_optimized.R

@@ -64,7 +64,7 @@ source('code/tax_level.R')
 #6) aggregate all of the models from the left-out-list to get the AUC table to plot a curve and get an AUC value for the whole model
 # do that for all of the models we have
 
-testsub <- subset(subs_meta, location %in% c("RB", "RS"))
+testsub <- subset(subs_meta, location %in% c("LB", "LS"))
 testsub$location <- factor(testsub$location)
 levels(testsub$location) <- c(1:length(levels(testsub$location))-1)
 
@@ -85,8 +85,8 @@ for(p in unique(testsub$patient)){
   test_results <- rbind(test_results, held_out_results)
 }
 
-right_xopt <- aucrf_test$Xopt
-right_imp_otus <- testsub[,c('location', right_xopt)]
+left_xopt <- aucrf_test$Xopt
+left_imp_otus <- testsub[,c('location', left_xopt)]
 
-r_imp_otus <- write.table(right_imp_otus, file = 'data/process/r_imp_otus.tsv', sep = '\t')
-right_optimized <- write.table(test_results, file = 'data/process/right_optimized.tsv', sep = '\t')
+l_imp_otus <- write.table(left_imp_otus, file = 'data/process/l_imp_otus.tsv', sep = '\t')
+left_optimized <- write.table(test_results, file = 'data/process/left_optimized.tsv', sep = '\t')

### 2018-02-08 code/stool_optimized.R

@@ -18,19 +18,34 @@ meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
 shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
 tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
 
+subs_file <- read.table(file='data/mothur/kws_final.an.0.03.subsample.shared', sep = '\t', header = T, row.names=2)
+
 #make OTU abundance file
 #Create df with relative abundances
 shared_file <- subset(shared_file, select = -c(numOtus, label))
 shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
-
 rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+
+#do rel abund calcs for subsampled
+subs_file <- subset(subs_file, select = -c(numOtus, label))
+subs_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+subs_abund <- 100*subs_file/unique(apply(subs_file, 1, sum))
+
 
 #Create vector of OTUs with median abundances >1%
 OTUs_1 <- apply(rel_abund, 2, max) > 1
 OTU_list <- colnames(rel_abund)[OTUs_1]
+
+OTUs_sub <- apply(subs_abund, 2, max) > 1
+OTU_list_sub <- colnames(subs_abund)[OTUs_sub]
+
 #get df of just top OTUs
 rel_abund_top <- rel_abund[, OTUs_1]
 rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+subs_abund_top <- subs_abund[, OTUs_sub]
+subs_meta <- merge(meta_file, subs_abund_top, by.x='group', by.y="row.names")
+
 
 print("loaded and organized data")
 
@@ -49,7 +64,7 @@ source('code/tax_level.R')
 #6) aggregate all of the models from the left-out-list to get the AUC table to plot a curve and get an AUC value for the whole model
 # do that for all of the models we have
 
-testsub <- subset(rel_meta, location %in% c("LS", "RS"))
+testsub <- subset(subs_meta, location %in% c("LS", "RS"))
 testsub$location <- factor(testsub$location)
 levels(testsub$location) <- c(1:length(levels(testsub$location))-1)
 

### 2018-02-09 code/build_figure4.R
rf model figures fixed

@@ -90,16 +90,29 @@ testsubM <- subset(subs_meta, location %in% c("RB", "LB"))
 testsubM$location <- factor(testsubM$location)
 levels(testsubM$location) <- c(1:length(levels(testsubM$location))-1)
 
+testsubR <- subset(subs_meta, location %in% c("RB", "RS"))
+testsubR$location <- factor(testsubR$location)
+levels(testsubR$location) <- c(1:length(levels(testsubR$location))-1)
+
+testsubL <- subset(subs_meta, location %in% c("LB", "LS"))
+testsubL$location <- factor(testsubL$location)
+levels(testsubL$location) <- c(1:length(levels(testsubL$location))-1)
 
 
-
-left_roc <- roc(testsub$location ~ left_optimized_results$one)
+left_roc <- roc(testsubL$location ~ left_optimized_results$one)
 right_roc <- roc(testsubR$location ~ right_optimized_results$one)
 
 bowel_optimized_results <- bowel_optimized_results[-13,]
 
 muc_roc <- roc(testsubM$location ~ bowel_optimized_results$one)
 stool_roc <- roc(testsubS$location ~ stool_optimized_results$one)
+
+#print AUCs
+auc_muc <- auc(muc_roc)
+auc_stool <- auc(stool_roc)
+
+auc_right <- auc(right_roc)
+auc_left <- auc(left_roc)
 
 #########################################################################
 # Save and export Figure 4
@@ -120,7 +133,7 @@ mtext(side=2, text="Sensitivity", line=2.5)
 mtext(side=1, text="Specificity", line=2.5)
 legend('bottomright', legend=c(
   sprintf('D Lum vs D Muc, AUC = 0.908'),
-  sprintf('P Lum vs P Muc, AUC = 0.764')
+  sprintf('P Lum vs P Muc, AUC = 0.716')
 ),lty=1, lwd = 2, cex=0.7, col=c('red', 'blue'), bty='n')
 
 mtext('A', side=2, line=2, las=1, adj=1.5, padj=-5, cex=1.5, font=2)

### 2018-02-09 code/build_figure5.R

@@ -1,35 +1,180 @@
 #########################################################################
-# Build figure 5 - abundance charts of feature selected OTUs from RF models
+# Build figure 5 and 6 - abundance charts of feature selected OTUs from RF models
 # Kaitlin Flynn, Schloss lab, updated 12-20-17
 
 # PUT TAX FILE AND SHARED META UP IN HURR
 source('code/random_functions.R')
 source('code/tax_level.R')
-#from optimized model code that is run on the cluster, import file with OTUs and abundances 
 
-r_top_otus <- read.table(file='data/process/r_imp_otus.tsv')
+#load in all of the files, get rel abund of >1% 
 
-#####Relative abundance plots#####
-#get top 5 OTUs and plot relative abundance
+meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
+shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
+tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
 
-r_top_names <- colnames(r_top_otus[2:6])
-r_top_labels <- get_tax(1, r_top_names, tax_file)
+subs_file <- read.table(file='data/mothur/kws_final.an.0.03.subsample.shared', sep = '\t', header = T, row.names=2)
 
-rs_abunds <- shared_meta[shared_meta$location=='RS', r_top_names]/10000 + 1e-4
-rb_abunds <- shared_meta[shared_meta$location=='RB', r_top_names]/10000 + 1e-4
+#make OTU abundance file
+#Create df with relative abundances
+shared_file <- subset(shared_file, select = -c(numOtus, label))
+shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
 
[…]

### 2018-02-12 submission/manuscript.Rmd
reviewer edits completed

@@
   html_document: default
-  word_document: default
-sansfont: Arial
 header-includes:
@@
 geometry: margin=1.0in
+sansfont: Arial
 bibliography: references.bib
@@
 
-Word count: 3,953/5,000
+Word count: 4,205/5,000
 
@@
 ```{r datasets, echo=FALSE}
-meta <- '../data/raw/kws_metadata.tsv'
-shared <- '../data/mothur/kws_final.an.shared'
-tax <- '../data/mothur/kws_final.an.cons.taxonomy'
-subsample <- read.table(file='../data/mothur/kws_final.an.0.03.subsample.shared', header=T)
 
@@
 shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+subs_file <- read.table(file='../data/mothur/kws_final.an.0.03.subsample.shared', sep = '\t', header = T, row.names=2)
 
@@
 ```{r randomforest, echo=FALSE}
+#make OTU abundance file
+#Create df with relative abundances
+shared_file <- subset(shared_file, select = -c(numOtus, label))
+shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
 
-#update this to just be importing end files off of optimization scripts to pull AUCs from
+#do rel abund calcs for subsampled
+subs_file <- subset(subs_file, select = -c(numOtus, label))
+subs_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+subs_abund <- 100*subs_file/unique(apply(subs_file, 1, sum))
+
+#Create vector of OTUs with median abundances >1%
+OTUs_1 <- apply(rel_abund, 2, max) > 1
+OTU_list <- colnames(rel_abund)[OTUs_1]
+
+OTUs_sub <- apply(subs_abund, 2, max) > 1
+OTU_list_sub <- colnames(subs_abund)[OTUs_sub]
+
+#get df of just top OTUs
+rel_abund_top <- rel_abund[, OTUs_1]
+rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+subs_abund_top <- subs_abund[, OTUs_sub]
+subs_meta <- merge(meta_file, subs_abund_top, by.x='group', by.y="row.names")
+
+seed <- 1
+n_trees <- 2001
+
+left_optimized_results <- read.table(file = '../data/process/left_optimized.tsv', sep = '\t')
+right_optimized_results <- read.table(file = '../data/process/right_optimized.tsv', sep = '\t')
+
+bowel_optimized_results <- read.table(file = '../data/process/bowel_optimized.tsv', sep = '\t')
+stool_optimized_results <- read.table(file = '../data/process/stool_optimized.tsv', sep = '\t')
+
+##############
+# make test datasets for each
+
+testsubS <- subset(subs_meta, location %in% c("RS", "LS"))
+testsubS$location <- factor(testsubS$location)
+levels(testsubS$location) <- c(1:length(levels(testsubS$location))-1)
+
+testsubM <- subset(subs_meta, location %in% c("RB", "LB"))
+testsubM$location <- factor(testsubM$location)
+levels(testsubM$location) <- c(1:length(levels(testsubM$location))-1)
+
+testsubR <- subset(subs_meta, location %in% c("RB", "RS"))
+testsubR$location <- factor(testsubR$location)
+levels(testsubR$location) <- c(1:length(levels(testsubR$location))-1)
+
+testsubL <- subset(subs_meta, location %in% c("LB", "LS"))
+testsubL$location <- factor(testsubL$location)
+levels(testsubL$location) <- c(1:length(levels(testsubL$location))-1)
+
+left_roc <- roc(testsubL$location ~ left_optimized_results$one)
+right_roc <- roc(testsubR$location ~ right_optimized_results$one)
+
+bowel_optimized_results <- bowel_optimized_results[-13,]
+
+muc_roc <- roc(testsubM$location ~ bowel_optimized_results$one)
+stool_roc <- roc(testsubS$location ~ stool_optimized_results$one)
+
+#print AUCs
+auc_muc <- auc(muc_roc)
+auc_stool <- auc(stool_roc)
+
+auc_right <- auc(right_roc)
+auc_left <- auc(left_roc)
 
@@
 The proximal mucosa had more of the genera _Enterobacteriaceae, Bacteroides_ and _Pseudomonas_.
-The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC=0.850).
-Separating proximal and distal luminal samples proved more challenging (AUC=0.580) and specific microbiota that differentiated the two were hard to identify.
+The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC = `r round(auc_muc, digits=3)`).
+Separating proximal and distal luminal samples proved more challenging (AUC = `r round(auc_stool, digits=3)`) and specific microbiota that differentiated the two were hard to identify.
 By sampling the unprepped colon, we identified distinct bacterial populations native to the proximal and distal sides.
@@
 
-Words: 259/260
+Words: 233/250
 
@@
 4A).
-The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908).
+The models performed well when classifying these samples (proximal AUC = `r round(auc_right, digits=3)`, distal AUC = `r round(auc_left, digits=3)`).
 The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model.
@@
 The model performed best when distinguishing the proximal versus distal mucosa (Fig.
-4B, AUC = 0.850) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = 0.580).
+4B, AUC = `r round(auc_muc, digits=3)`) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = `r round(auc_stool, digits=3)`).
 The OTUs included in the model differentiating the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ (Fig.

### 2018-02-12 submission/manuscript.tex
reviewer edits completed

@@
 \emph{Pseudomonas}. The classification model performed well when
-classifying mucosal samples into proximal or distal sides (AUC=0.850).
+classifying mucosal samples into proximal or distal sides (AUC = 0.808).
 Separating proximal and distal luminal samples proved more challenging
-(AUC=0.580) and specific microbiota that differentiated the two were
+(AUC = 0.599) and specific microbiota that differentiated the two were
 hard to identify. By sampling the unprepped colon, we identified
@@
 proximal and distal sides, independently (Fig. 4A). The models performed
-well when classifying these samples (proximal AUC = 0.764, distal AUC =
-0.908). The OTUs that were most predictive of each site were identified
+well when classifying these samples (proximal AUC = 0.716, distal AUC =
+0.862). The OTUs that were most predictive of each site were identified
 by their greatest mean decrease in accuracy when removed from the model.
@@
 distal luminal samples. The model performed best when distinguishing the
-proximal versus distal mucosa (Fig. 4B, AUC = 0.850) whereas the model
+proximal versus distal mucosa (Fig. 4B, AUC = 0.808) whereas the model
 to differentiate between the proximal versus distal lumen performed
-poorly (AUC = 0.580). The OTUs included in the model differentiating the
+poorly (AUC = 0.599). The OTUs included in the model differentiating the
 distal and proximal mucosa included members of the \emph{Porphyromonas},

### 2018-02-12 code/build_figure4.R

@@ -11,19 +11,34 @@ meta_file <- read.table(file='data/raw/kws_metadata.tsv', header = T)
 shared_file <- read.table(file='data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
 tax_file <- read.table(file='data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
 
+subs_file <- read.table(file='data/mothur/kws_final.an.0.03.subsample.shared', sep = '\t', header = T, row.names=2)
+
 #make OTU abundance file
 #Create df with relative abundances
 shared_file <- subset(shared_file, select = -c(numOtus, label))
 shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
-
 rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))
+
+#do rel abund calcs for subsampled
+subs_file <- subset(subs_file, select = -c(numOtus, label))
+subs_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
+subs_abund <- 100*subs_file/unique(apply(subs_file, 1, sum))
+
 
 #Create vector of OTUs with median abundances >1%
 OTUs_1 <- apply(rel_abund, 2, max) > 1
 OTU_list <- colnames(rel_abund)[OTUs_1]
+
+OTUs_sub <- apply(subs_abund, 2, max) > 1
+OTU_list_sub <- colnames(subs_abund)[OTUs_sub]
+
 #get df of just top OTUs
 rel_abund_top <- rel_abund[, OTUs_1]
 rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")
+
+subs_abund_top <- subs_abund[, OTUs_sub]
+subs_meta <- merge(meta_file, subs_abund_top, by.x='group', by.y="row.names")
+
 
 print("loaded and organized data")
 
@@ -47,7 +62,7 @@ source('code/tax_level.R')
 # code needs to be run on a high performance computer cluster - takes too much memory and time
 
 ################ Run the following on HPCC / flux
-testsub <- subset(rel_meta, location %in% c("LB", "LS"))
+testsub <- subset(subs_meta, location %in% c("LB", "LS"))
[…]

### 2018-02-13 submission/manuscript.Rmd
cleaned up repo

@@
 output:
-  word_document: default
   pdf_document:

### 2018-03-26 submission/manuscript.Rmd
final tweaks

@@
 That is, a higher AUC value indicates the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value.
-For instance, the model separating the proximal and distal mucosa had an AUC of 0.850 whereas the model for classifying the proximal and distal lumen had a much lower AUC of 0.580.
+For instance, the model separating the proximal and distal mucosa had an AUC of `r round(auc_muc, digits=3)` whereas the model for classifying the proximal and distal lumen had a much lower AUC of `r round(auc_stool, digits=3)`.
 Further, the latter model required 44 OTUs to best separate the samples whereas the models separating the mucosa only needed 10 OTUs.

### 2018-07-17 submission/manuscript.tex
cleaned up repo

@@
 model with a lower AUC value. For instance, the model separating the
-proximal and distal mucosa had an AUC of 0.850 whereas the model for
-classifying the proximal and distal lumen had a much lower AUC of 0.580.
+proximal and distal mucosa had an AUC of 0.808 whereas the model for
+classifying the proximal and distal lumen had a much lower AUC of 0.599.
 Further, the latter model required 44 OTUs to best separate the samples
